### 2017-05-24 RESP1.rmd
yes

---
title: "Resp I "
Author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# CBL

## Main Case

18F  wakes up at 6am can't breathe goes into ED. She is *too short of breath to speak*.

Key processes in respiratory system

-  *Respiration*: whole process = {breathing in, breathing out, cellullar respiration,...}
-  *Ventilation (V)*: air going into alveoli
-  *Perfusion (Q)*: air diffusing into blood
-  *V/Q mismatch*: areas of lung are either not getting enough air or blood. E.g. APO (alveoli filled with fluid so no ventilation), PE (no blood flowing (no Q) whilst lung parenchyma are fine), pulmonary arteriole stenosis.
    -  V/Q > 1: physiological dead space
    -  V/Q < 1: physiological shunting

Pathological process of noisy breathing: turbulent air flow against a smaller than usual airway diameter OR obstructed. What can cause this:

-  Inside the lumen: excess mucus secretion, fluid = {blood, pus,...}
-  Wall of the bronchiole: could get fattened due to acute inflammation of the walls $\rightarrow$ change in permeability of the walls due to cytokines $\rightarrow$ oedema $\rightarrow$ narrow.
-  Outside the wall: smooth muscle contraction $\rightarrow$ bronchospasms.

What causes central cyanosis: 30mmHg (increased) deoxyhaemoglobin

DDx: asthma, COPD, anaphylaxis (typically causes obstruction in the large airways), PE, bronchitis, *foreign body obstruction* (common in children where they swallow a jelly bean, small car, or it could be a carcinoma)

Can't breathe: have to immediately stabilise breathing: salbutamol ($\beta_{2}$-agonist) $\pm$ $O_{2}$

### Trigger 2

Explanations for vital signs:

-  Tachypnoeic: hypoxaemic $\rightarrow$ $\Uparrow$respiratory rate.
-  Tachycardic: hypoxaemia (low [O~2~] in the *blood*) $\rightarrow$ sympathetic stimulation $\rightarrow$ catecholamine release $\rightarrow$ postive chronotropy + inotropy.
-  Hypertensive: trying to increase pressure and perfusion to the lungs
-  *Pulsus paradoxus*: normally on inspiration there is increased venous return into RA $\rightarrow$ $\Uparrow$pressure. Paradoxus:  there is a decrease in pressure on inspiration which is weird. When there is an obstruction *inspiration is prolonged*: initial increase in pressure due to inspiration $\rightarrow$ drops as inspiration is prolonged.
-  Wheeze: turbulent air flow
-  Crackles: no oedema = {pus, blood,...}
-  Reduced breath sounds: decreased alveolar ventilation (not moving air) and forced vital capacity (FVC; maximal inspiration) is decreased.
    -  **Early dynamic small airway closure**: when pressure in pleural space > pressure in airways the small airways snap shut $\rightarrow$ can't exhale as much as inspired.
-  Percussion hyper-resonant: lots of air in the lungs (like a drum). If it is dull to percussion then it is filled with fluid.
-  Pulse Oximetry 75%: Widespread bronchodilation $\rightarrow$ decreased FVC and tidal volume $\rightarrow$ decrease oxygenation to pulmonary arterioles $\rightarrow$ decrease overall oxygenation of the blood.
-  Temperature normal: no infn. In asthma increased mucus secretion $\rightarrow$ leads to infn. Infn can also exacerbate asthma.
-  Wheezing higher on expiration: air is moving in but not out
-  ABG analysis: used sometimes in COPD as O2 might kill them.

Suggestive of asthma: wheeze, history

Noisy breathing during inspiration (upper/large airway obstruction): *stridor* . Noisy breathing during exhalation (lower/small airway obstruction): wheeze.

How to look after her:

-  salbutamol inhaled through a spacer or through nebuliser (5mL of liquid salbutamol which "humidifies" the gas with it). Nebuliser $\neg$used when there is no airway obstruction, used when significant airway obstruction.
-  $O_{2}$
-  Oral prednisolone (steroid) or IV hydrocortisone

## Trigger 3

Presenting case to a consultant: **ISBAR** = {introduction, situation, background, assessment, recommendation}

ABG: explanations on the pdf.

Hypoxic: hypoxic-respiratory drive will activate (sensed by chemoceptors?)

Drugs:

-  Sympathetic: redirects blood *away* from GIT, skin
-  $\beta_{2}$-agonist: bronchodilation, vasodilation to skeletal muscle
-  Corticosteroids are *strong anti-inflammatories* by releasing annexin-1 that prevents TXN of phospholipase A
-  Ipratropium bromide = muscarinic antagonist: inhibits PS-NS

Ask her:

-  What were you taking
-  When did you stop taking meds
-  How long she has been on for
-  Assess asthma control: frequency of asthma vs asthma episodes (have you been hospitalised for asthma before [been in ICU; upregulates treatment])

Symptoms of asthma: coughing (night), SOB, chest tightness

Trigger asthma: irritants, infn, exercise, drugs (NSAIDs, $\beta$-blockers), anxiety attack

Indications for CXR: assymetrical chest expansion (side could be full of fluid, pneumothorax), crackles, unequal air entry, febrile + symptoms suggesting pneumonia, haemoptysis.

## Trigger 4

Spirometry can measure:

-  FEV~1~: how much air out in one second

Textbook defn asthma: reversible airway obstruction ($\Delta$FEV~1~ > 200mL | 12%). Role of spirometry in acute: way more sensitive.

Combination of steroid (fluticasone) and $\beta$ agonist (salbutamol, turbutaline). Sometimes have to add a LABA (long acting $\beta$ agonist; salmetarlol, efametolol). Steroids can cause: immunosuppression, cushings, CNS, diabetes, $\therefore$ ICS (inhaled) is a method to get it to the lungs and prevent these systemic effects.

Asthma medication has to be taken everyday as asthma is a *chronic inflammatory* process. If it is not taken then chronic inflammation is still occuring underneath. Corticosteroids prevent scarring,

Explore reasons *why* people are not compliant: socio-economic barriers to health.

Autonomy: four things required for providing consent (VICS)

-  \textbf{V}oluntary
-  \textbf{I}nformed
-  (not) \textbf{C}oerced
-  \textbf{S}pecific: has to be for a specific process rather than just "yeah you can treat my cancer"

## Trigger 5

Asthma NOT resolved. Cellular response occurs 8-12 hours after initial flare up.

## Trigger 6

Silent chest: no air moving at all.

To improve patient adherence: education (written and visual explanation), asthma action plan, multidisciplinary team (abo health worker + pharmacist).

### 2017-05-24 RESP2.rmd
yes

---
title: "Resp II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# Quiz

-  Restrictive: impairment of lung expansion
-  Obstructive: increased resistance to airflow
-  Tiotropium bromide: muscarinic antag
-  Defns
    -  Asthma: hyperreactivity resulting in episodic and reversible bronchoconstriction
    -  Chronic bronchitis: chronic productive cough for at least 3 months over 2 consecutive years
    -  Emphysema: permanent enlargement due to destruction of alveolar walls, without fibrosis
    -  Interstitial lung disease: interstitial fibrosis with a restrictive pattern of spirometry
-  Mech of coughing
    1.  Close glottis in the larynx
    2.  Contract expiratory muscles esp the abdominal
    3.  Increase intrathoracic pressure
    4.  Open glottis
    5.  Air forced out through bronchi and larynx

# Pulmonary mechanics

## Surface tension

Is a mixture of phospholipids and specific surfactant proteins. Has 3 functions:

1.  $\Downarrow$elastic recoil of the lungs $\rightarrow$ reducing pressure needed to inflate them.
2.  Allows surface forces to vary with alveolar SA $\rightarrow$ protecting against atelectasis.
3.  Limits reduction of hydrostatic pressure caused by surface tension $\rightarrow$ reduces forces promoting transudation of fluid and intersitital oedema.

## Forces to overcome during inflation

3 forces:

1.  Elastic recoil
2.  Inertia: negligible
3.  Resistance to flow

Can be divided into work to overcome {elastic forces, flow resistance}.

## Lung failure

-  Type I: more emphysema
-  Type II:

CO2 retainers: chronic increased PaCO2 => decreased sensitivity in the central chemoceptors => hypoxia becomes main breathing drive => DON'T GIVE O2

# Spirometry physiology prac

## Instructions given to pt

-  Upright position (standing or sitting but upright)
-  Shoulder roll to relax
-  Deep breath into until cant breathe anymore
-  Make good seal around the mouthpiece
-  Forcefully exhale as fast as possible for 6s + give encouragement
    -  Uses abdominal and ext intercostal muscles to flatten the ribcage
-  Repeat
-  Compare results to predicted "healthy" values from some random table

## Things affecting forced respiration

-  Recoil of the lung: mainly from alveolar surface tension but also stretched connective tissue when full
-  External muscles
-  Airways

## Results

If FEV1:FVC < 0.7 then can only say *lung fn is impaired*.

## Flow volume loop

Measures Y-axis = flow (L/s); X axis = Volunme (L). The max(Y) is the PEF = peak expiratory flow which should be achieved before 15% of vital capacity has been exhaled due to elastic recoil. PIF = peak inspiratory flow which is not an indicator for anything as it is driven by muscles. In healthy: PEF > PIF due to the recoil of the lungs that promotes expiration and inhibits inspiration.

Males have stronger internal intercostal muscles.

## Pathologies

### Large airway obstruction

Decreases FEV1, normal FVC as it doesn't affect dynamic small airway closure.

### Asthma

Devcreases FEV1 and FVC as there is an increased airway resistance due to obstruction so when exhaling it loses pressure faster $\rightarrow$ early dynamic airway closure $\rightarrow$ air trapping.

PIF is normal as our diaphragm is lowered and rib cage is raised. This decreases the intrapleural pressure which dramatically decreases pressure in the airways making them open even if they were obstructed. Therefore PIF is normal (flow is normal, but volume is not).

### Emphysema

Loss of lung tissue => decreases lung recoil => increased intrapleural pressure => early airway closure, decreased FVC, FEV1. It INCREASES PIF due to decreased recoil.

### Restrictive disease

Characterised by increase in FEV1/FVC.

HOWEVER: Young asian females tend to be skinny, good muscles, small lungs, FEV1 could be 100% but don't have restrictive disease.

# Pathology: POW = pneuomoconiosis

## DDx for dyspnoea

-  CVS: MI, cardiac failure, arrhythmia
-  MSK: broken ribs, pectus excavatum, scoliosis
-  Neurological: ALS, myasthenia gravis
-  Psy: anxiety, panic attacks

## Macroscopic

-  Anthracosis: the black deposits throughout the lung are most likely from carbon particles
-  Bullae: pockets of air that may have ruptured in the pathological slide.

## Pathophysiology of antracosilicosis

1.  Silica inhaled into lungs
2.  Alveolar macrophage engulf the particles
3.  Cytokines released and start process of inflammation
4.  Inflamm damages area of the lung
5.  Fibroblast prolif occurs as part of wound healing
6.  Overtime, the silica particles outlive the macrophages and continues the process of inflamm.
7.  Fibrosis resulting in formation of nodules and calcification

## How does fibrosis cause SOB

Pulmonary fibrosis is scarring of the lung tissue caused in this circumstance, from the presence of foreign body silica. The accumulaion of excess fibrous tissue thickens the walls of the alveoli and impairs oxygen diffusion into the pulmonary capillaries and overall oxygenation of blood leading to the sensation of shortness of breath.

## Ix

CT, CXR, Spirometry, ABG, tuberculin skin test, bronchoscopy. Don't have to use fine needle biopsy unless you are examining for cancers.

## Histopathology

-  Silicosis: has silicotic nodules which are densely fibrotic nodules within lymphovascular spaces in connective tissue septa and in subpleural locations. Many of the nodules contain significant amounts of carbon (anthrosilicosis). Refractile material would be visible under polarising light.
-  Asbestosis:
    -  Interstitial fibrosis: the alveolar walls and septa are markedly thickened and contain large amounts of fibrous connective tissue. Residual airspaces are distorted and may contain exudate and neutrophils, suggesting secondary pneumonia in this area.
    -  Pleural plaque: the pleura is thickened by extensive fibrosis which is up to 5mm thick in this section.
-  CWP: the coal macule consists of dust-laden macrophages. The coal nodules is a larger conglomeration, characterised by additional collagen fibers arrayed in a delicate network.
-  Hypersensitivity pneumonitis: The interstitial areas are expanded and contain numerous lymphocytes, macrophages and some plasma cells. Non-caseating granulomas: They are seen in the alveolar walls and consist predominantly of epithelioid macrophages.

# CBL main case

## Trigger 1

Even if man is "light-hearted" it is difficult to generally bring men into consultation. Men are generally less proactive about their health so if they are dragged in then it is likely to be serious. Men tend to minimise what is happening and he could be scared about what it is.

Cues: diesel fitter, works open cut mine, maintains heavy machinery. SOB whilst fishing, 58M.

**DDx for SOB:**

-  Resp: asthma, COPD, pulmonary hypertension, coal workers pneumoconiosis, pulmonary fibrosis, interstitial lung disease, pneumonia, cystic fibrosis $\rightarrow$ bronchiectasis
-  CVS: PE, IHD, heart failure
-  MSK: diaphragmatic atrophy/rupture/malfunction/palsy (nervous damage), trauma, scoliosis, ankylosing spondilosis (spine stiffening)
-  Autoimmune: Myasthenia gravis (autoimmune disorder that blocks transmission of nervous impulses; could have insidious onset)
-  GIT: ascites, liver carcinoma, hernia
-  CNS: phrenic nerve palsy, stroke
-  Psych: anxiety

**Further Hx**: Resp Qs = {smoking, dyspnoea, cough (productive/dry), what is in sputum (*red flag*: haemoptysis), audible breathing (wheezing/stridor), snoring (looking for obstructive sleep apnoea), chest discomfort (pain/tightness = can be something different e.g. asthma), fever (infective illnesses of the lungs)}. CVS Qs. SOCRATES of PC (if pain = TRUE).

## Trigger II

Cues: 18mo, no CP, no palpitations (*always ask cardiac questions with resp* as they tend to go together), no wheeze, few yrs cough worse in the mornings, brownish sputum (stuff breathe in is brown such as pollutants), no haemoptysis, no weight loss, shirt getting tighter (barrel chested $\leftarrow$ hyperinflated lungs $\leftarrow$ air trapping), pants looser (cachexia: muscle wasting as a result of hypoxaemia), smoking: 42 pack years, alcohol 2 standards a day is fine, no medications (less likely as ACE-I can cause dry hacking cough whilst he has productive cough), normal flu took him out for 7 weeks last year (likely he has underlying lung disease resulting in his increased susceptibility to a simple lung infection).

Fam Hx: father died MI 79 which is fine as it is quite old.

Hx: worked in mines since 18, never worked underground, exposure to gold silver and copper, often finishes work covered in dust and grime, never been exposed to uranium, asbestos, nickel.

COPD is the overall diagnosis for {emphysema, chronic bronchitis} and it is usually a combination of the two at once. Dx criteria for chronic bronchitis: productive cough for $\geq3$ months in two consecutive years; emphysema: loss of lung parenchyma. 2 separate pathological mechanisms.

Emphysema: perfusion stays the same whilst lung alveoli are damaged: decreased ventilation, and can form large air sacs called *bullae*. These expand during inspiration and compress other normal alveoli.

Chronic bronchitis: pollutant damage irritates resp epithelium $\rightarrow$ excessive mucin production and smooth muscle proliferation which obstructs terminal bronchioles $\rightarrow$ inhibits air movement into it $\rightarrow$ mucin in the alveoli prevents oxygen diffusion. Therefore treatment involves: positive airway pressure to force airways open, draining the mucous. People bring air in and can't get it out due to airway trapping arising from *dynamic airway closure*.

**Learn COPD mech**: hx of smoking

**2 important genetic conditions**

-  Cystic fibrosis
-  $\alpha_{1}$ anti-trypsin deficiency: livers cannot produce the eponymous protein $\rightarrow$ trypsin is unopposed $\rightarrow$ widespread fibrosis + liver cirrhosis. Is *autosomal codominant*

**2 parts to cough**

-  What triggers the cough
-  What activates the cough (which nerves, muscles)

**Examination**

-  Inspection
-  Vitals
-  General inspection {hands, arm, chest, head, neck; clubbing, work of breathing, cyanosis}
-  Feel for a few things {symmetrical expansion during inspiration, vocal fremitus [vibrations upon phonation]}
-  Auscultate {inspiration, expiration, diff lobes, axilla} [snake pattern to compare equivalent sides of the lungs]

## Trigger III

Cues: O/E: RR 20 is high normal/elevated, increased expiratory phase, barrel shaped chest, thin legs, using accessory muscles, percussion: hyper-resonant, inferior displacement of diaphragm ($\therefore$diaphragm is working *fine*), breath sounds reduced due to less air movement (not likely infn due to lack of crackles [i.e. pneumonia]), heart sounds normal, JVP not elevated, normal abdomen, no lymphadenopathy (argues against infn and cancer). Sats O~2~: 89%

Obstructive: prolonged expiration phase as they are trying to get gas out ($\Downarrow$FEV~1~).

Restrictive: decrease lung compliance from loss of lung elasticity (fibrosis) $\rightarrow$ more work during inspiration/can't breathe in as much.

Ix to order:

-  Blood: FBC ($\neg$ anaemic, white cell count to check inflamm), ABG (PCO2, PO2, pH), CRP/ESR (inflammation), CHEM20 (baseline liver/renal fn)
-  CXR: flattening of the hemi-diaphragms consistent with hyperinflation in COPD, bullae, no consolidation (white full of blood/pus/sputum; as we expect to see full of air)
-  Spirometry: FEV~1~, RV

## Trigger IV

Cues: ABG results: polycythaemic many cells in the blood referring to the Hb. *Chronic* and has been compensated by the high bicarbonate.

CXR findings: flattening of the hemidiaphragms due to hyperinflation of the lungs, narrowing of the mediastinum, no evidence of fluid in the costophrenic angles (no blunting; which is the first place fluid *would* accumulate)

$\frac{\text{FEV}_{1}}{\text{FVC}} < 0.7$ = COPD

## Trigger V

-  Although his findings is consistent with *coal miners lung*, his *Hx suggests not*.
-  What would suggest occupational disease: relevant Hx
-  Disability support pension: twice the dole

Cycle of change for quitting smoking

-  Pre-contemplative
-  Contemplative
-  Action
-  Change
-  Relapse

Mx strategies: Goals = {risk reduction, optimise fn, improve symptoms, delay progression, minimise comorbidities, manage/prevent exacerbations}

-  Pharmacological: nicotine replacement therapy (gum, lozenges, patches, spray, champix or varenicline [nicotonic-R antag; helps prevent cravings])

\textbf{Learn Pharm for asthma and COPD:}

-  SABA: salbutamol
-  SAMA: ipratropium

Maintainence:

-  LABA: salmeterol
-  LAMA: tiatropium

## Trigger VI

He is better as he is taking medications and quit smoking.

People watching out for COPD action plan stuff.

PO2 < 55mHG: O~2~ therapy

# Short Case I

Someone comes in with SOB what to ask them:

-  SOCRATES for SOB
-  Cough + types?
-  Job Hx: possible exposures

Exposure to insulation material: likely to have asbestos exposure.

Ex:

-  Resp: crackles, wheezes, barrel chest, deformities, altered percussion, use of accessory muscles
-  CVS: HF, arrhythmias, murmurs, displaced apex beat. Assess for signs of *anaemia* such as pallor, tachycardias...

Ix: to determine nature of resp condition

-  Spirometry: $\Downarrow$FEV1 and FVC to 70% of normal and normal FEV1/FVC is consistent with mild restrictive. No improvement after bronchodilator is not reversible airway disease.
-  CXR:

Mx of asbestosis:

-  Referral to thoracic physician
-  Maximise QOL and lung fn
-  Assess severity and fn impairment: give supp O~2~ if necessary.
-  Give early antimicrobials if there is an infn

# SCII

Pt comes in with SOB

Hx (same as above)

DDx: most likely = {resp, CVS, haem/lymph}

-  Sig hypotheses to consider: COPD, lung Ca, interstitial lung diseases
-  Anaemia, CVS

Ix:

-  FBC, electrolytes, CHEM20
-  Lung fn including Spirometry
-  CXR
-  CT scan?!

Dx: idiotpathic pulmonary fibrosis. Causes persistent and progressive scarring of the alveoli $\rightarrow$ SOB(OE) + dry cough. Rate of disease progression is variable, more common in smokers, elderly, possible genetic. Limited therapies.

### 2017-05-24 RESP3.rmd
yes

---
title: "Resp III"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# Path

-  Benign vs malignant
    -  Look for how well defined the borders of the tumours are

# Dumb Ca genetics?!

## Oncogenes

SNP: KRAS (colorectal cancer)

Gene amplification (CNV): HER2/neu, MYCN

Translocation: Bcr/Abl

## TSG

-  MSH2
-  MLH1: lynch syndrome; predisp to colon cancer
-  APC: adenomatous polyposis coli
-  TP53: Li Fraumeni
-  BRCA1: familial breast and ovarian cancer
-  VHL: von hippel lindau syndrome
-  RB: hereditary retinoblastoma

# Pneumonia

> including various classifications e.g. CAP vs. HAP, histological distribution, causative organism/agent etc.

Atypical pneumonia zoonotic cause is Chlamydia psittaci.

CAP: bacterial pneumonias often follow a viral URTI. Strep. pneumoniae is most common. Bacterial pneumonia has 2 patterns of anatomical distribution = {lobular bronchopneumonia = patchy consolidation (solidification of the lung due to exudate replacing air in the alveoli), lobar pneumonia = consolidation of a large portion or entire lobe}.

Symptoms for CA(Bacterial)P: abrupt onset of high fever, shaking chills, and cough producing mucopurulent sputum; occasional patients have hemoptysis

CA(Viral)P: from the RNA virus *influenzae*.

HAP: *defined as pneumonia that develops in a patient hospitalised for longer than 48 hours (and is not incubating at the time of admission to hospital)*. Common in hospitalised individuals with severe underlying disease, immunosuppressed, or those on prolonged antibiotic regimens. Patients on mechanical ventilation are at high risk, and infections called ventilator-associated pneumonia. Gram-negative rods (members of *Enterobacteriaceae* and *Pseudomonas spp.*) and *S. aureus* are the most common isolates. S. pneumoniae is not common.

Assess severity = {**C[U/O]RB-65** (confusion, urea/O2 sats < 90%, RR > 30, systolic BP < 90, age >65), SMART-COP is more predictive of need to go into ICU}.

Ix: CHEM20, ABG, VBG, FBC, serum free cortisol, serum lactate levels. Sputum eval, imaging with CXR, chest CT, bronchoscopy, thoracentesis (essential in pt with parapneumonic pleural effusion), histo Ex.

Mx: ANTIMICROBIALS then possibly analgesia, antipyretics, chest physiotherapy, IV fluids if indicated, pulse ox, O2 supplementation, resp therapy, suctioning, ventilation

# Quiz

Antibiotics = {isoniazid, cephalosporin, penicillin, tetracycline, tobramycin}

-  Narrows spectrum = isoniazid
-  Broadest spectrum = tetracycline
-  Glycopeptides: Inhibition of cell wall synthesis.

S. pneumoniae penecillin resistance is caused by mutations in the genes encoding penicillin-binding proteins.

Hemagglutinin and Neuraminidase proteins determine influenza virus subtype

Empirical therapy: Answer: antibiotic therapy commenced before the identification of the causative micro-organism is available.

Antibiotic stewardship: the optimal selection, dosage, and duration of antimicrobial treatment that results in the best clinical outcome for the treatment or prevention of infection, with minimal toxicity to the patient and minimal impact on subsequent resistance

Ix for pneumonia (justification is the most important):

-  Radiology (chest XR, CT scan); confirms presence of pneumonia, and pattern (bronchopneumonia or lobar)
-  Blood count: white cell count (host response); underlying anaemia
-  Serum biochemistry: metabolic effects of respiratory disease, indicating severity
-  Blood gases: assess severity, eg respiratory failure
-  Sputum and blood cultures: determine causative organism.
-  Serology:  determine causative organism
-  Urinary antigens: determine causative organism.

# Readings

-  Atelectasis is air in the alveoli
-  Pneumothorax is air in the pleural space
-  Chronic interstitial (restrictive) lung diseases: heterogeneous group of disorders characterised by bilateral pulmonary fibrosis affecting walls of the alveoli. The hallmark of these disorders is reduced compliance (stiff lungs), which in turn necessitates increased effort to breathe (SOB = dyspnoea)
    -  Idiopathic pumonary fibrosis: unknown, but probably genetic cause leading to fibrosis. Crackles on inspiration and progressive dyspnoea
    -  Pneumonoconioses = breathing in mineral particles with the 3 most common agents = {coal dust, silica, asbestos}

# CBL main case

## Trigger 1

69M, pleuritic (sharp on breathing in) CP, diaphoretic, has the "shivers", coughing and cranky.

Hypo:

-  Resp: pneumonia, pleuritis, empyema (pus in pleural space) [abscess is in parenchyma], ARDS, lung Ca (mesothelioma, adenomocarcinoma (glandular origin), SCC (most common in non-smokers), small cell carcinoma), PTX (pneumothorax), PE, trauma
-  MSK: costochondritis (Tietze's), rib #, muscle strain
-  CVS: ACS = {STEMI, NSTEMI, (un)stable angina}, IHD, pericarditis, aortic dissection, HF (can lead to SOB, coughing) $\rightarrow$ APO
-  Anaphylaxis

Reasons for refusing treatment

-  Minimising/underestimating
-  Negative experience with doctor
-  Wants to be the tough guy
-  Beliefs
-  Religious

## Trigger 2

Pt has: no confusion, no drowsiness, no meds, 30 pack years smoking but quit 10 years ago, no alcohol, no recent travel, no recent vaccinations, retired gardener. Doesn't believe in doctors as father died due to post-surgical complications.

Rule out: MSK, anaphylaxis

Mech for symptoms:

-  Sweating: sympathetic stimulation can occur during pain
-  Shivering: hypothalamic set point is raised $\rightarrow$ shivering as body tries to increase body temperature. ASsociated with focal infection or systemic infection such as sepsis.
-  Pleuritic chest pain: inflammation in the pleural space

Occupation

-  Gardener: more atypical infections due to shit in the soil such as pseudomonas
-  Office: air con is likely to be legionella

Resp Ex

-  Inspection (start at hands and move towards chess): resp distress (use of accessory muscles), chest deformities (pectus excavatum), wasting/cachexia(systemic), cyanosis, CO~2~ retention
-  Palpation: trachea to check if it is midline, ENT (look for lymph nodes), measure expansion of lung fields,
-  Auscultation: pectoriloquy, bronchophony, aegophony
-  Percussion

Have to differentiate if acute vs acute on chronic (acute presentation of chronic disease).

## Trigger 3

Ex

-  Vitals: tachycardic, tachypnoeic, hypotension, JVP not visible (not heart failure), febrile
-  Hemi thorax on inspiration: consolidation
-  Dullness to percussion: consolidation OR pleural effusion
-  Breath sounds harsh and loud: normally high f are filtered out due to aeration of the lung, if harsh and loud = consolidation
-  (No) Pleural rub suggests (no) inflammation of pleura

How does a healthy person get pneumonia: smoking, occupation, environmental exposures, lack of vaccines, travel, exposure to infected hosts.

CAP: Streptococcus pneumoniae (80%), haemopholius influenza (5%), mycoplasma, staph aureus  
HAP (nosocomial): Staph aureus, E. coli, pseudomonas

Pt is ill: Dx = pneumonia and early signs of septic shock

Ix:

-  Bloods (FBC, ABG, blood cultures (3 from 3 different sites whilst febrile), CRP/ESR, CHEM20)
-  CXR
-  Sputum samples: nasopharyngeal swab
-  Still do heart stuff because of CP: such as ECG and troponin levels
-  Urine samples

## Trigger 4

Type 1 resp failure: hypoxia *without* hypocapnoea

CXR results: significant opacity (not consolidation) in the lobe. Check lateral and see costophrenic angle suggests against pleural effusion

Microbial samples: sputum, nasopharyngeal aspirants (look for viruses)  
Pt effort

Low PaO2 due to shunting (V/Q mismatch) arising from blocked alveoli.

Mx for *severe* strep use ceftriaxone + IV azithromycin (mainly for the atypical causes in case), O~2~, analgesia titrated to pain

## Trigger 5

Empirical treatments = guidelines, Take into consideration liver/kidney fn

*Ceftriaxone* $\in$ Cephalosporin is a beta-lactam (along with penicillin); *azithromycin* $\in$ Macrolides which is a 50S inhibitor.

Follow up after discharge: send him to a GP in 1 week for clinical review, CXR 6 weeks to ensure that the acute PC is gone and to see if there is any chronic comorbidities going on, possibly do vaccinations again

## Trigger 6

Complications of pneumonia: sepsis, abscess (focal area), emphysematous change, scarring, empyema (pus in the pleural cavity)

# Short case I

Setting: med student in darwin at GP rotations.

Hx: 4F indigenous, woke up crying with right ear pain, paracetamol seems to settle it a bit

DDx for ear pain (otalgia): otitis externa, otitis media, infn of auricle, mastoiditis, inflamm of ear drum, throat jaw sinuses

What to ask about?

-  SOCRATES about PC
-  PMHx
-  PSHx
-  Medications
-  Allergies
-  Immunisation status
-  Family Hx including if anyone else is unwell

Ex:

-  General Inspection
-  Ex of ENT and full RESP as symptoms suggestive of URTI, and cervical lymph nodes
-  Auscultate chest for murmurs/general health
-  Assessment of hearing (Rinne and Weber testing or refer to audiologist)

Eustachian tube connects ear with URT.

Vaccine preventable causes of acute OM = {Streptococcus pneumoniae, Haemophilus influenzae, Influenza virus}

Acute otitis media with effusion vs chronic suppurative otitis media

-  Acute otitis media with effusion: no discharge, no perforation of tympanic membrane, ototoscope just looks red
-  CSOM: perforated tympanic membrane with persistent drainage from middle ear. Long term damage includes = {hearing loss}

# SC II

Pt: 42F, usually healthy.  
PC: Unwell 24 hours with fevers to 39, rigors, aches, sore throat, dry cough

DDx: acute nature suggests infectious cause (think organisms, virulence factors, host factors). Pt has no history suggesting immunocompromised but it should be considered. Influenza

Ex: vitals, whatever systems seem likely. Resp Ex.

Ix:

-  Bloods: FBC, CHEM20, cultures, sputum samples from nasopharyngeal swabs which get PCR'd

Mx:

-  Normally get lots of rest, sleep, fluids
-  Antiviral treatment is indicated in pt with established complications, severe or requiring hospital admission, elderly, down syndrome, chronic resp, immunocompromised. Otherwise DON'T as it causes adverse side effects
-  Get vaccine

### 2017-05-24 RESP4.rmd
yes

---
title: "Resp IV"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# Quiz

-  Lung carcinogens: Arsenic (inorganic), asbestos, bis(chloromethyl) ether and chloromethyl methyl ether, chromium compounds, gamma radiation, ionizing radiation (X rays), mustard gas, nickel compounds, polycyclic aromatic hydrocarbons, radon decay products, soots, tars, mineral oils, vinyl chloride.
-  A patient with a left-sided non-small cell lung cancer undergoes staging investigations which demonstrate the involvement of contralateral hilar lymph nodes (i.e. in the right hilum). According to the TNM system of classifying tumours, what 'N' classification would this receive? N3
-  Paraneoplastic syndromes are distant effects of a tumour NOT related to the local presence of cancer cells. They are due the production of hormones or other humoral factors by the tumour cells, or to an immune reaction to the tumour.
-  Methods of tissue sample in LC: Sputum / pleural fluid cytology, bronchoscopy, fine needle biopsy, mediastinoscopy, thoracoscopy, surgical biopsy.
-  2 mycobacterial infections: TB and MAC
-  Clinical Dx means that it does not need any testing.

# Path POW

Pt: 60M overweight presenting with gradually increasing SOB for 4/12. Worse on exertion, not when lying down. He has constant left-sided chest pain. A cough has lasted for 3/12. He has coughed-up blood several times.

SR (Systems Review)

-  Some increased hoarseness.
-  A little more fatigued than usual.
-  No dysphagia.
-  No fevers.
-  No abdominal pain.

O/E (on Ex)

-  Patient alert and oriented, but does not look well.
-  Clothes hanging a little.
-  No cervical or axillary lymphadenopathy
-  L lower lobe stony dull to percussion
-  L lung decreased expansion

3 key investigations

-  CXR
-  ECG: cardiac fn but to rule out MI
-  Spirometry: due to SOB
-  FBC

CXR demonstrated diffuse changes in the left lung and a single dominant nodule in the upper lobe of the left lung. A contrast CT chest was also ordered. Looks like pleura is most likely affected.

-  This points towards Dx of pleural mesothelioma

[Distinguishing between malignant and benign](malignant_vs_benign.png)

Asbestos greatly increases risk, smoking is not associated with increased risk.

Pathophys  
Begins as multiple small nodules in the *parietal* pleura which eventually coaelesce. Forms a rind of tumour fusing the parietal and visceral pleurae, going on to encase the lung. Extension into the lung may occur along interlobal fissures. Despite minimal direct invasion of the lung parenchyma, pt experience symptoms due to reduced relative functional capacity. Mediastinum may shift *towards* the lesion.

# Main case

## Trigger 1

51F, 7 weeks productive cough, change in sputum = {quality (type) = clear to cream to creamy brown, quantity (amount) = }. Afebrile, 33 pack year smoking Hx. Dx: COPD with acute bronchitis causing exacerbation. Bacterial superinfection (due to trashed epithelium) with the main causes Streptococcus and Staphylococcus.

Safety netting: order a follow up.

DDx:

-  Resp: acute on chronic bronchitis (even though afebrile it doesn't rule it out as she might have had paracetemol), pneumonia, bronchiectasis (when the large airways dilate), acute bacterial exacerbation of asthma, post nasal drip (wouldn't expect), occupational (pneumonoconioses), pertussis (6 weeks of coughing), sarcoidosis (autoimmune diseases of the lung)
-  CVS: LHF (pink frothy sputum)
-  GIT: GORD
-  Lung Ca

Red flags: according the the lung Ca gp guide reading includes persistent cough > 3 weeks.

Brown sputum: smoker (tar), blood, infections can cause funky coloured sputum,

Haemoptysis DDx: PE, lung Ca, infection, trauma (pulmonary contusion), AVM: vascular malformation, autoimmune disorders (Goodpastures syndrome; Ig against basement membranes), blood coagulation problem such as haemophilia, could be masquerading as haematemesis (vomiting blood) or epistaxis (blood coming from the nose)

Resp Hx: SOCRATES for PC, cough+sputum, SOB, noises (wheeze/stridor), pleuritic CP, trauma, vocal changes, URTI symptoms, travel, sick contacts, personal/fam Hx of atopy/asthma, B symptoms (night sweats, fever, weightloss) which are suggestive of malignancy

Initial Mx: ok for typical bacterial infns, atypical wouldn't work for it (mycoplasma use azithromycin (macrolide))

## Trigger 2

Works in air con office: legionnaires disease (legionella). Parents born in China could mean Tb.

SNAP = {smoking, nutrition, alcohol, physical activity (obstructive sleep apnoea)}

If the SOB changed then the SOB on two flights of stairs is important. Follow up with "Is that usual/different compared to normal"

DDx: rule out asthma (no fam Hx), pneumoconiosis (works in air con office), trauma, coagulopathies (has never seen blood in sputum before), Goodpastures (should be coughing out and peeing blood all the time), PE (no CP)

Changes due to smoking: squamous cell metaplasia, goblet cell hyperplasia (see Resp II GIFT)

5 A's of smoking

-  Ask
-  Assess (what stage of smoking): pt is pre-complentative
-  Advise stop smoking
-  Assistance (with shrink or nicotine patches)
-  Arrange follow up

Ex: inspection (fingers, cervical lymphadenopathies, ENT), auscultation (vocal resonance and tactile fremitus), percussion

## Trigger 3

Normal Ex but bad symptoms $\rightarrow$ have to do Ix

-  Normal HR suggests against CVS (particular PE which would be tachycardic)
-  Normal breath sounds suggests against pleural effusion
-  Normal percussion suggests against consolidation and pleural effusion
-  Abnormal

Ix:

-  Bloods: FBC ([Hb]), CHEM20, coags, CRP (inflamm response might suggests for malignancy), d-dimer to rule out PE
-  Sputum cultures
-  CXR
-  Spirometry
-  ECG

Case summary: Christine is 51 F presenting with 7 weeks of productive cough. On the background of a 33 pack year smoking Hx and COPD. Symptoms:sputum has changed,..., no haemoptysis, no SOB, no CP. Ex: breath sounds *equal bilaterally* (makes it sounds better than just saying normal). On examination things are unremarkable. At this stage I'd like to organise (list Ix)

## Trigger 4

DDx for single pulmonary nodule: sarcoidosis, Tb, neoplasms (primary/secondary), granuloma, hamartoma (tissue elements normally found at that site but growing in a disorganized manner), teratoma (multiple germ cell types), choristoma (mass of histologically normal tissue in an abnormal location), aspergiloma (ball of fungus), cicatricial tissue,

Lung Ca = {12% small cell, 70% are non-small cell, rest are metastases}. On CXR expect to see:

-  Size and repidity of growth
-  Irregular borders: some degree of asymmetry

Routes of Ca spread

-  Haematogenous
-  Lymphatics
-  Perineural spread: includes CSF
-  Transcoelemic
-  Direct invasion

Further Ix:

-  CT chest: confirm suspicions for lung Ca and location
-  Sputum cytology: pt coughs out cells
-  Biopsy
    -  CT guided FNA (fine needle aspirate): by an interventional radiologist for lateral tumours
    -  Bronchoscopy: look down vocal cords; useful for central tumours
    -  Bronchioalveolar lavage: wash out

Virtuous doctor: honesty, empathy, have to be transparent with pt "even though it is uncomfortable for me to say this, I have to be honest, it is possible you have cancer", not withhold information, courage to say that.

## Trigger 5

Adenocarcinoma most common with non-smokers.  
SCC most common in smokers.

Lung Ca usually presents as cough, haemoptysis, dyspnoea, pleuritic CP, change in voice (tumour obstructing recurrent laryngeal nerve). Pemberton's sign (raise hands and head turns red), paraneoplastic syndrome (malignant cell may produce hormones/peptides), horners syndrome (compression of sympathetic chain T1/T2 due to apical lung tumour).

Needle biopsy:

-  Risks: puncture an artery/vein $\rightarrow$ haemothorax, infn
-  Benefits: clinically diagnosing which affects treatment
-  Negative results does NOT exclude malignancy

Cancers

-  Grading:
-  Staging: spread of the cancer, where is it likely to spread to (lymph, surrounding, distant), TNM = {t = size, nodes, metastases}

## Trigger 6

Most of the time it is about encouraging pt to make a change rather than castigating them.

# Short case 1

## Trigger 1

48M homeless, alcoholic. Presenting feeling 'really bad', 2/7 Hx SOB slowly worsening, mild discomfort in his chest but only on deep inspiration. Denies recent weight loss and points to his abdo saying 'gut is getting bigger'.

Alcoholic likely to have *aspirational pneumonia*: atypical pathogens and is usually *polymicrobial*

DDx:

-  PE: risk factors are malignancy, pregnancy, female (oestrogen; the pill), DVT, immobility, fam Hx, personal Hx | Use Wells score
-  Pneumothorax: risk factor is tall and skinny, short onset pain and dyspnoea
-  Pleural effusion: haemothorax, empyema because irritating the parietal pleura
-  APO: less like as it wouldn't cause chest pain (also should have crackles)
-  MI:
-  GORD:

Ex:

-  Decreased breath sounds
-  Stony dull percussion note: pathognomic of pleural effusion (duller than *dull percussion* which could be consolidation, hyper resonant percussion = pneumothorax)
-  Decreased vocal resonance: not consolidation (which would amplify the sound)
-  Abdo distended
-  Ascites: fluid in the peritoneal space | caused by portal vein hypertension leading to fluid leaking into surrounding space likely to be from liver or RHF

Ix:

-  CXR: AP and lateral to locate it and distinguish between pneumonia or effusion
-  Bloods: FBC+FILM (platelet count), coag profile (more likely to have a coagulopathy as he is an alcoholic which causes cirrhosis depressing fn of the liver [producing clotting factors]), EUC $\subset$ ELFT (electrolytes and liver fn test) $\subset$ CHEM20 (liver and kidney fn = ELFT + Ca, Mg, P), ESR/CRP

## Trigger 2

Pleural fluid has come back as serous (yellowish straw coloured) transudate. Both visually look the same.

-  Transudate: serous fluid when congestion in capillaries to the point where fluid is pushed out of it. Junctions remain the same. Caused by increase in hydrostatic pressure in the venous system or a loss of oncotic pressure (proteins in the blood keeping fluid there; lack of protein can be caused by liver cirrhosis [hypoalbunaemia] or by kidney disease [nephrotic syndrome or hepatorenal syndrome which is leaky to protein] or by malnutrition).
-  Exudate: junctions open so more stuff can go out: proteins and other stuff can go out. Inflammatory response (changes in vascular permeability) so proteins and even some cells can eat through. Bit worse as this is a sign of pathology.

Pleuritic chest pain: worse on deep inspiration

# Short case 2

## Trigger 1

Pt: 73M regular comes to local GP presenting feeling 'a bit off', noticed SOB and a persistent dry cough over the last few/12. Feeling tired/fatigued(?), has lost 6kg in the past 3/12. He has a 60 pack year smoking Hx and quit 3 years ago.

-  Risk factors: smoking increases respiratory illnesses, age predisposes to Ca (telomeres shorten)
-  Red flags: weight loss, feeling very tired, signs of cachexia (foundin Ca, AIDS, coeliac disease, COPD, MS, RA, CHF, tuberculosis, poisoning)

DDx:

-  Resp:
    -  Exacerbation of COPD (emphysema)
    -  Lung Ca: smoking increases risk
    -  Asthma
    -  Interstitial lung diseases: idiopathic pulmonary fibrosis (unknown pathophys; increased incidence in smokers) would lead to fibrosed alveoli resulting in a decrease in gas exchange
    -  Viral infn
    -  Bronchospasms
    -  Allergies
-  CVS: PE, LHF
-  GIT: GORD
-  Sleep: OSA
-  ACE inhibitors

Ex: what to focus on

-  Inspection: is he looking well, work of breathing, mental state
-  Vitals
-  General inspection: hands, wrists, arms, face, ENT, neck (lymphadenopathies), chest
-  Ante:
    -  Palpate: expansion bilaterally, subcutaneous emphysema
    -  Percuss: resonance
    -  Auscultate: normal breath sounds, adventitia (wheezes, crackles), vocal resonance (bronchophony, whispering pectoriloquy, egophony)
-  Post:
    -  Palpate
    -  Percuss
    -  Auscultate
-  CVS Ex: JVP, heart sounds

## Trigger 2

Ex results: looks tired and thin face, Wt 65kg, afebrile, RR 20, HR 76, BP 155/95, heart sounds dual - no murmurs, no Cx/supraclavicular lymphadenopathy, trachea mid-line, breath sounds reduced at right base, fixed inspiratory wheeze right mid zone, percussion normal bilaterally, abdomen NAD (nothing abnormal detected?), nil peripheral oedema.

Ix ordered: CXR for chest signs and cough, bloods (FBC, electrolytes [U&E]?, CHEM20)

Interpretation of Ex results:

-  Positive: heart normal (awaiting bloods), afebrile = no infn, no lymphadenopathy (suggests against infn; Ca?), trachea midline (suggests against pneumothorax and atelactasis), abdo NAD = _____, nil peripheral oedema suggests against PVD and HF
-  Negative: tired, weight loss, breath sounds reduced at right base (atelectasis (collapse of lung), consolidation, pleural effusion = things in pleural space blocking sound (hydrothorax, pneumothorax, empyema, haemothorax)), fixed (probs tumour) inspiratory wheeze right mid zone (means that the bronchioles are stenosed or obstructed which can occur from a tumour compressing it, foreign bodies, fibrosis)

Appropriate imaging

-  CXR: used first due to safety and cost
-  CT: a lot more radiation. Can Dx nodules but low sensitivity in identifying malignancy.
-  PET: gives metabolic information, thus has high sensitivity for malignancies. Used for *staging* of the Ca, usually after the Dx is confirmed
-  Bone scan: area without PET
-  MRI: Dan (or use in the brain)

## Trigger 3

Ix results: slight hypokalaemia

Further Ix: bronchoscopy

Dx: SCLC, paraneoplastic syndrome with ectopic ACTH production (Cushing's) accounting for some symptoms such as hypokalaemia, elevated BP, elevated glucose.

SCLC: cell of origin are pulmonary neuroendocrine cells (why it has paraneoplastic effects). 12% of lung Ca, smoking related, extremely aggressive, and mestasises extremely early.

Staging shows *limited stage* disease (confined to 1 hemithorax and within 1 tolerable radiation field). Mx: chemo and radiotherapy. Prophylactic cranial radiotherapy (radio the brain because SCLC is HIGHLY aggressive, and ensure it doesn't metastasise to the brain). Resection is not done as the Ca is too aggressive and by the time it has Dx then it is likely to have metastasised.

2 years later Dx with *extensive stage* disease ($\neg$limited; Mx: chemotherapy alone) and dies.

Paraneoplastic syndromes are when tumour secretes substances or hormones into blood stream.

-  Cushings: [SCLC] ectopic production of ACTH | DDx: pituitary adenoma, cortical tumour of adrenal
-  Humoural hypercalcaemia of malignancy (HHM): [SCC] PTHrP release by tumour causing osteoclastic bone resorption | DDx: metastatic spread into bone
-  Lambert-Eaton myasthenic syndrome: [SCLC] auto-antibody production causing blockage of neuromuscular junction | DDx: myasthenia gravis
-  Syndrome of inappropriate antidiuretic hormone secretion: [SCLC] lots of water retention

Quitting smoking is tough:

-  Weight gain: 1st year is the most, then rate decreases
-  Psychoscocial effects: going against ingrained habits
-  Still increased risk for lung Ca, but it drastically reduces. Never returns to the never smoked level.

### 2017-05-24 renal1.rmd
yes

---
title: "Renal 1"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# Glomerulonephritis

3 cell populations in glomerulus = {endothelial, mesangial cells, epithelial cells (foot processes and parietal), } + 2 structural units = {basement membrane, mesangium}

Everything except for podocyte can regenerate

# Quiz (predominantly on UTIs)

-  Urine sample from 3/12 child: supra-pubic aspiration
-  Major mechanism causing stress incontinence: urethral hypermobility
-  Acute urinary retention requires urgen management
-  Non surgical Mx of BPH: $\alpha$-adrenergic-receptor antagonists
-  UTI w/ highly alkaline urine: *Proteus mirabilis*
-  Phimosis: foreskin cannot be retracted over the glans
-  Complicated UTI:
    -  First episode of acute cystitis in a sexually active 20 year old female.
	  -  *Acute pyelonephritis with fever in a 20 year old female.*
	  -  Urethritis in a 25 year old man who does not practice safe sex
	  -  Asymptomatic bacteriuria found on routine testing in a pregnant female
	  -  Bacteriuria associated with an indwelling catheter in a patient with neurogenic bladder
-  Lifetime prevalence of UTI in women: 50%
-  Trabeculation of bladder wall is hypertrophy
-  Medication associated with acute urinary retention: decongestant

# Random

## Targets of immune-mediated renal disease

-  Normal constituents of the glomerulus (e.g. Goodpastures syndrome; type II hypersensitivity)
-  Non-renal self antigens localised in globermuli => immune attack (e.g. SLE)
-  Exogenous antigens that localise in glomerular capillaries (e.g. PSGN)

# Main case

## Trigger 1

Allopurinol: xanthine oxidase inhibitor that reduces uric acid in the blood (treatment against gout)

What can cause urine to be red?

-  Blood (haemoglobin): test using urine microscopy/urinalysis = {ketones, villirubin / microscopy looks at RBCs, WBCs, epithelial cells {suggestive of non-midstream catch:not clean sample)} or urine dipstick test {protein (kidney disesae), pH, RBCs, leukocytes (WBCs), ketones(Diabetic Ketoacidosis; ketosis), specific gravity (osmolites/electrolytes in the urine), glucose (diabetes), bilirubin (cholestasis [nothing moving from gallbladder => spill into bloodstream]), nitrites (bacterial waste products)}
-  Beetroot/foods
-  Dehydration
-  Drugs/medications
-  Myoglobin
-  Menstruation
-  Porphyrins
-  Urobulinogen
-  Cholestasis: bile duct obstruction so it goes through the river (also has pale stools)

Haematuria: could be non disease cause such as trauma or exercise induced haematuria. >10E6/L is the defn cut-off.

Blood most likely to come from (trauma anywhere):

-  Glomerulus: leaky
-  Renal pelvis: due to calculi formation and trauma
-  Ureter: not so much as not very vascular
-  Bladder: cystitis, malignancy
-  Urethra: STDs (chlamydia clear discharge, gonorrhoea is purulent)

Pathological processes that can underlie haematuria:

-  Inflammation
-  Trauma
-  Infection
-  Malignancy
-  Vasculopathy
-  Autoimmune processes
-  Obstruction: of any hollow viscous
-  Necrosis/ischaemia: e.g. renal artery stenosis

Questions to ask

-  If pain => SOCRATES; ever had before
-  Fam Hx
-  Falls/trauma
-  Sexual Hx
-  Ever had before
-  Fam Hx

## Trigger 2

Triage: sick/unwell/well: briefly check red flags $\rightarrow$ send for some Ix

Focussed Ex: goes straight to the point/PC and deals with the less urgent stuff later, saves time

Hypertension $\rightarrow$ Hypertensive nephropathy (direct). More common is hypertension * (indirect) glomerular disease $\rightarrow$ haemoturia

Fenestrations in the foot processes of the podocytes. Allows small stuff and +ve ions.

DDx:

-  IgA nephropathy
-  Malignancy
-  Kidney stones
-  Infections
-  Trauma
-  Glomerulonephritis: glomerular disease
-  Coagulopathy such as haemophilia

Largest issue with low EGFR is hyperkalaemia: cardiac arrhythmias + death

Ix:

-  CHEM20: electrolytes, kidney, liver fn
-  Ultrasound KUB = {kidney, ureter, bladder}: see obstruction, fluid
    -  Contrast CT KUB is the only thing that shows about stones
-  Urinalysis
-  Urine cytology = looking for cells in urine

## Trigger 3

Membranous glomerulonephritis: common from Hep C

Dysmorphic red cells: have gone through the collecting tubules. Red cell casts: gone through the tubules  
Are pathognomic of glomerular disease (blood not coming from UTI or kidney stones)

Further Ix:

-  24h urine collection
-  Cystoscopy: puts camera into bladder, can argue against stones in the bladder, transitional cell carcinoma,...

## Trigger 4

Glomerulonephritis screen:

-  AMA: against autoimmune disease
-  dsDNA
-  Anti DNAase B: used as a screen for post infectious GN
-  Non nuclear auto-antibodies: Hep B, Hep C, and HIV
-  C3 C4 complement: found in post-infectious GN
-  ANCA: anti nuclear cytoplasm antibody
-  Anti-glomerular basement membrane: important for collagen disease

| Nephritic syndrome | Nephrotic syndrome |
| ------------------------ | --------------------------- |
| Blood in urine (RBC casts or dysmorphic cells) | Proteinuria >3.5g/24h |
| PHAROH = {proteinuria, haematuria, azotaemia, RBC casts (dysmorphic), oliguria, hypertension} | Proteinuria +++, hyperlipidaemia|
| Glomerulonephritis often a cause | ??? |

## Takeaways

-  Nephritic: PHAROH
-  Nephrotic: 3.5g/24h proteinuria AND oedema due to loss of hypoalbuminaemia. Children show it in the face.

Another way to look at it: glomerulonephropathies

-  Proliferative: too many cells (pink and purple) in glomerulus
-  Non-proliferative:

Biopsy look at it

-  H&E + microscopy
-  Immunofluorescence:
    -  Anti-GBM disease: the whole thing is lit up almost ribbon like
    -  IgA nephropathy: granular so lots of focal deposits
-  Electron microscopy

Focal vs segmental

-  Focal: whole glomerulus or part of glomerulus
-  Segmental: whole kidney, or just a part of a kidney

Watch the wikipedia [IgA nephropathy video](https://open.osmosis.org/). Tl;dr it is abnormally glycosylated IgA1 (malformed), don't get broken down and deposit in sites of filtration. Furthermore not recognised as self; breakdown in tolerance; immune complex deposition in mesangium. Immune response causes glomerular damage.

# Short case 1

## Trigger 1`

Glomerulonapathies:

-  Renal cause CVS: Decrease GFR => fluid retention => overloading the heart
-  CVS causing renal: vasculitis/renal artery stenosis => kidney failure => fluid retention => overloading the heart

Considering the timeframe: more like to have started a while ago and become protein ++++ in the past month.

Foamy protein = {proteins (such as albumin), increased force of urine, dehydration}

Glomerular damage => breach in the filtration system => heavy proteinuria.

Ultrasound KUB: doesn't say much about renal vasculature but looks at flow in ultrasound. Doppler shows blood flow through renal arteries.

Ix:

-  Urinalysis
-  24h urine collection: to see if it is actualyl nephrotic syndrome (by defn)
-  Bloods = {FBC, CRP, CHEM20, ddimer, lipid profile as }
-  Ultrasound KUB
-  Doppler ultrasound

UEC = urea, electrolytes, creatinine (EGFR)

UEG = UEC + glucose

LFTs = albumin

If need both LFTs and UEC then use CHEM20

## Trigger 2

DVT and kidney fn: nephrotic syndrome => losing proteins such as clotting factors (loses anti-thrombotic, loses pro-thrombotic but loses more anti-thrombotic due to the number?!) => hypercoagulable state

Hypertensive and kidney fn: decreased oncotic pressure

3 most common nephropathies for nephrotic syndrome

-  minimal change (children; bowmans space epithelial cells loses its negative charge, so not much morphological change)
-  Membranous glomerulonephritis:
-  Focal segmental glomerulosclerosis (FSGS):

## Trigger 3

Creatine clearance formula = Cockcroft-Gault Equation. Important as some drugs such as gentamicin have narrow therapeutic index.

Renal disease staging

1.  Normal eGFR, abno ultrasound
2.  eGFR > 60
3.  eGFR > 30
    a.  eGFR > 45
    b.  eGFR < 45
4.  eGFR > 15
5.  ESKD when eGFR < 15

Prognosis if you have no idea: young male, no comorbidities, proteinuria is alright

# Short case 2

## Trigger 1

-  Bilirubin breakdown clogged in liver, to avoid toxicity, body sends it ou bound to plasma in the urine (dark urine)
-  Assessing hydration status in children: mucous secretions, cap refill (centrally; push on chest 5s and see it refills in 2s)
-  Active urinary sedimen = {red cell casts, dysmorphic red cells}
-  GS = goodpastures syndrome
-  Henoch-Schonlein Purpura (HSP) = {Palpable purpura, abdo pain, arthralgia}

## Trigger 2

Hypertensive suggests renal disease.

Lymph nodes

-  Tender lymph node = infn/inflamm.
-  Non-tender lymph nodes = probs malignancy

Hypertension

-  Renal artery stenosis
-  Cushings
-  Renal disease
-  Malignancy: paraneoplastic effects (Conns, phaeochromocytoma)

Post infectious GN: circulating immune complexes form and clog up the glomeruli => inflamm response => damaged glomerular tufts => nephritic syndrome

-  Type III due to immune complex

Dysmorphic red cells pathognomic of glomerular disease

### 2017-05-24 renal2.rmd
yes

---
title: "Renal II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# Quiz

-  Small kidneys on ultrasound indicates kidney failure is *chronic* rather than *acute*. Incorrect answers = {metabolic acidosis, low eGFR, hyperkalaemia, resistant hypertension}.
-  Dipstick shows haematuria: repeat dipstick test
-  A 67 year old non-smoking man presents with an episode of fever for several days. Urine dipstick is positive for blood, nitrites, and leucocyte esterase. He reports several years of difficulty (hesitancy and dribbling) with urination. The most likely diagnosis is: UTI
-  HIV can cause glomerulopathy
-  Nephrotic syndrome most common cause
    -  Children: minimal change (loss of negative charge)
    -  Adults: focal segmental glomerulosclerosis

# Reading

## Acute kidney injury

AKI risk factors include older age, diabetes, hypertension, obesity and cardiovascular disease. Further, pre-existing chronic kidney disease (CKD) has been found to be the most important risk factor for the development of AKI (10 times more likely). Also AKI $\rightarrow$ CKD.

Prognosis: ranges from complete recovery through to mild CKD, severe kidney damage + long term dialysis, and death. Oliguria ($<400$mL/day) has worse prognosis. Anuria ($<100$mL/day) even worse.

Defn: KDIGO definition defines AKI by a $\geq0.3$ mg/dl increase in creatinine or a $50\%$ increase above baseline within 48 hours, or a $\geq50\%$ increase in creatinine above baseline within 7 days.

Categories

-  Prerenal - As an adaptive response to severe volume depletion and hypotension, with structurally intact nephrons. Aetiology = {volume depletion (diuretics, vomiting, diarrhoea), decreased CO (HF, PE), systemic vasodilation (sepsis), afferent arteriolar vasoconstriction (hypercalcaemia, NSAIDS)}.
-  Intrinsic/Intrarenal - In response to cytotoxic, ischaemic, or inflammatory insults to the kidney, with structural and functional damage. Aetiology = {vascular (renal artery/vein obstruction, malignant hypertension, transplant rejection), glomerular (anti-GBM, glomerulonephritis [RPGN]), tubular (acute tubular necrosis, crystals, drugs), interstitial (drugs, infn, systemic disease, acute interstitial nephritis)}.
-  Postrenal - From obstruction to the passage of urine. Aetiology = {mechanical (stones, structure, fibrosis)}

Complications

-  CVS: HF, arrhythmias, pericarditis (rare; think additional Dx such as SLE)
-  Resp: goodpastures, granlomatosis, sarcoidosis
-  GIT: nausea, anorexia, vomiting, uraemia?!, gastric bleeding, pacnreatitis, jaundice, hepatitis

Pathophysiology: tl;dr glomerular filtration depends on pressure gradient from glomerulus to bowman space.  
Main determinant is renal blood flow (RBF). Reductions in RBF no matter the cause of AKI are a common pathological pathway for $\Downarrow$GFR. Prerenal = compromised renal perfusion. Intrinsic = usually glomerulus or tubule; ischaemic injury is the most common cause. Obstruction = increase in tubulur pressure decreases filtration driving force. This gradient equalises and renel efferent vasoconstriction. Depressed RBF leads to ischaemia and cell death. Inflammation comes in and cell injury occurs even after restoration. During this time tubular cell damage fucks up tight junctions => leakage of glom filtrate + dyaing cells enter tubules forming casts => decrease in GFR => oliguria.

Restoration of RBF  
Early RBF recovery has better prognosis. Once it is restored remaining nephrons increase filtration and hypertrophy, however if not enough nephrons left then these sclerose => more nephron loss. Hyperfiltration theory of renal failure: as nephrons die then puts more pressure on the rest which are then more prone to dying.

# Path: revision

Lack of oxygen to myocardial cells lead to myocardial infarction (cell death), which weakens the wall causing a rupture. This in turn results in blood entering the pericardial sac (haemopericardium). When enough blood accumulates, cardiac tamponade will occur and subsequently death.

## Pot 1

Hypertension is a pre-disposing factor for => hypertrophy of wall => MI => blood leaking => cardiac tamponade

Histo: normal part can see nuclei in the myocardial cells. Infarcted area (no nuclei, can make out the cell; coagulative): blue stuff is neutrophil infiltrate and red stuff = RBCs (congestion). 1-2 weeks: fibroblasts come in; granulation tissue = {fibroblasts, new vessels, macrophages, xxxxxxx}.

The following can all affect the severity of an infarct.

-  Location, severity and rate of occlusion
-  Duration of occlusion
-  Metabolic/oxygen needs of the myocardium at risk
-  Extent of collateral blood supply
-  Coronary arterial spasm
-  Other factors – heart rate, alteration in blood pressure.

The sequence of events leading to artery occlusion are:

1.  Sudden change in plaque morphology – haemorrhage, rupture, ulceration.
2.  Exposed subendothelial collagen and plaque contents causing platelet adhesion and aggregation.
3.  Vasospasm stimulated by platelet aggregation.
4.  Other mediators activate the extrinsic pathway of coagulation leading to increased thrombus.
5.  Occlusion.

Subendocardial collagen/tissue needs to be exposed to trigger the coagulation cascade.

MI outcomes:

-  Cardiac tamponade
-  Arrhythmias
-  Ventricular aneurysms
-   df

## Pot 2

A 27-year-old female patient presents to her GP one week after a tooth extraction.  She complains that she feels tired and has been having recurrent fevers.  On examination she is febrile and has peripheral oedema. Her pulse is rapid and thready.  On auscultation, a mitral stenosis murmur is heard.  She is immediately sent to the emergency department.

Further history taking reveals that she had rheumatic fever as a child. RHD has smaller vegetations, fibrin brings about stenosis of the valve

Urgent investigations are carried out including a chest Xray.  Unfortunately she dies soon after these are taken.

Vegetations on heart valve - Rheumatic fever (acute immunologically mediated multisystem inflammatory disease that occurs a few weeks following an episode of group A Strep pharyngitis) leads to acute chronic rheumatic heart disease, which leads to chronic RHD (mitral stenosis commonest lesion) – fibrotic valvular disease – Left atrium dilates – congestion of lungs – Right ventricular hypertrophy (boot shaped heart) – dental extraction – infective endocarditis (bulky friable vegetations composed of thrombotic debris and organisma) – septicaemia and death.

RVH leads to the boot shaped heart in the CXR.

In regards to Non-infective endocarditis; deposition of small masses of fibrin, platelets and other blood components form on valve leaflets. The lesions are sterile.

It is usually seen in debilitated patients eg. Cancer, where there is a hypercoaguable state which leads to ‘marantic vegetations’.

## Pot 3

Restrictive lung diseases: FEV1 and FVC both fall so no change in the ratio. Would be caused by secretion of fibrogenic cytokines by macrophages.

Smoking => emphysema => centriacinar (not hereditary). $\alpha_{1}$-anti-trypsin would cause pan acinar

CXR: Flattening of the diaphragm => lung spaces have enalarged. Heart looks tubular (narrowing)

The loss of alveolar surface area due to emphysema makes it difficult for the patient to take in sufficient oxygen and excrete carbon dioxide at a sufficient rate. Since both the terminal bronchioles and alveoli along with the capillary bed have been destroyed, a matched defect exists between the ventilation and perfusion; areas of low ventilation also have poor perfusion.

Despite the “matched” V/Q defect, overtime hyperventilation develops and cardiac output (CO) drops which leads to areas of poor blood flow in relatively well oxygenated areas. Due to this poor CO, the rest of the body suffers from tissue hypoxia.

Pathogenesis of RHF: Matched V/Q defect --> Overtime hyperventilation develops --> Cardiac output drops --> Hypoxaemia --> Tissue hypoxia (this may cause polycythaemia)  and Increased CO2 retention (this may cause respiratory acidosis) --> Arterial vasoconstriction in lungs --> pulmonary hypertension --> Retrograde pressure build-up --> Elevated right ventricular pressure --> right ventricular hypertrophy --> right ventricular failure (Cor pulmonale)

## Pot 4

Bronchopneumonia: comorbidities predispose.

List three possible complications of pneumonia.

Abscess formation, empyema, septicemia, bronchiectasis.

## Pot 5

This x-ray shows a coin lesion in the right upper lobe.a

Hamartoma: excess tissue in a normal location

Choristoma: excess tissue in an abnormal location.

## Pot 6

 cavitating lesion

Abscess vs cavity

-  Abcsess is an area of necrosis that is well sealed
-  Cavity: leaves behind a hole

# Quiz

# Main case

## Trigger 1

22F, fell from horse 1:45h ago, at rural hospital. Closed (within the skin; open means it breaks the skin and ready for osteomyolitis) spiral fracture of the 1/3 mid femur: neurovascular compromise suspected (scared of haematoma => compartment syndrome). Diastysis of pelvic symphysis. No haemoperitoneum. Circulating blood volume depletion estimated at 2L. Central venous line (so can pump lots in) and urinary catheter (measure urine output).

Post-renal: stricture (narrowing), fistula

$\0.5mL/kg/hr$ is the bare \textit{minimum} urine output.

If the vasa recta are not perfused adequately => ischaemia of the tubules => necrosis. It does not have a compensatory mechanism so if perfusion drops then it just cannot perfuse the tubular cells.

Ix: renal challenge: give a bolus of fluid, and examine if urine output increases. If so then it would be due to hypoperfusion.

Hydration status:

-  Underhydrated
    -  Symptoms: dry mouth/lips, thirsty, sweating
    -  Signs: postural tachycardia as there is not enough fluid reserve to bolster it, pallor, decreased skin turgor, decreased capillary refill, cold
-  Overhydrated
    -  Symptoms: pittal oedema, moist cough due to APO, SOB due to pleural effusion
    -  Signs: added heart sounds, elevated JVP

## Trigger 2

Hydration status can be quantified by % volume loss.

Issues with kidney damage

-  Hyperkalaemia => ECG ASAP
-  Uraemia (increase in urea)

Ix to order:

-  Bloods: FBC (anaemia), CHEM20 (electrolytes), coags, ABG (pH)
-  Urine: dipstick (proteinuria, haematuria)
-  Imaging: renal ultrasound (obstruction or not)

## Trigger 3

-  Urine osmolality = 302 $\approx$ 298 which is the blood osmolality. Means that kidneys are not doing anything.
-  Urea is generally a marker for dehydration
-  Renal ultrasound normal rules out post-renal causes

Fn of the kidney

A.  cid/base
B.  P
C.  (s)alt+$H_{2}O$
D.  etox
E.  PO: hormone to produce RBCs
F.  luid balance

Anion gap = [Na^+^]+[K^+^]-[Cl^-^]+[HCO~3~^-^]

-  Small proteins have negative charge but we can't measure them. $\therefore$ gives indication that there is decreased kidney fn.-  \textbf{do some reading on this shit??}

Principles of Mx: supportive management by intervening when necessary. This means maintaining the ABCDEF kindey fn see if she gets better. Also maintain fluids, electrolytes.

## Trigger 4

Serum biochemistry (CHEM20): creatinine: increased which is a marker that kidney fn is BAD

Hyperkalaemic metabolic acidosis due to hypovolaemia.

Deteriorating biochemical profile: tubular epithelial cells can't secrete K^+^, and they are starting to lyse which also spills K^+^ into the blood.

Mx: e.g. acid/base: give bicarbonate. Compensate for metabolic acidosis.

Histology: loss of tubular epithelial cells, inflammatory infiltrate

## Trigger 5

Dialysis: passive fluid movement across membrane. Has risks of clotting, infn. Types:

-  Peritoneal: tenkoff catheter surgically implanted into ante abdo wall. Can do it at home, much greater $\mathbb{P}(Infection)$, not as effective as the other one.
-  Haemodialysis: surgical induced fistula between radial artery and cephalic vein (AV fistula; vibrates/buzzes) which provides a large target for a cannula.
-  CVVH: used in the ICU, can be in response to SIRS (systemic inflammatory response syndrome).

## Trigger 6

Histo: Loop of Henle sets up the concentration gradient in the filtrate, DCT ends up concentrating it. **Learn about countercurrent**.

Takes time for re-epithelialisation to occur.

# Short case 1

## Trigger 1

Failure to thrive: lower percentile of child weight growth. Causes:

-  Malnutrition
-  Neglect
-  Chronic/severe infns

Polydipsia: increased water intake

-  Hyperglycaemia: glucose being spilled off and water is dragged along (Diabetes)
-  Neurological

## Trigger 2

Low creatinine is never significant, we just care about high creatinine. Hyponatraemic, hypokalaemic, hyperchloraemic, metabolic acidosis (bicarbonate; guess). Serum biochemistry means there is losing ions, but the increased specific gravity means it is escaping through that. The level of ions suggests it is proximal tubule.

Diabetes = high sugar in blood, high in urine

Kidney disease = normal in blood, high in urine

Rickets = failure to lay down bone. Laying bones requires Ca^2+^ or PO~4~^3-^, Ca^2+^ requires Vit D. So any of the three could be fucked.

Ix extra to order:

-  24h urine: analyse for eGFR, electrolytes, amino acids

## Trigger 3

Nephrotic syndrome means glomerulus is defective.

Cystinosis => cysteine build up in proximal tubules => they start failing => lose electrolytes => Fanconi syndrome => decreased phosphate => decreased Ca absorption => rickles

# Short case 2

## Trigger 1

32M playing rugby knee to left flank. Coliky abdo pain.

Colicky pain = comes and goes but stays bad most of the time = {pregnancy, renal stones, biliary stones}. Is associated with peristalsis in hollow organ (bowel, ureter, bile duct, oesophagus) which results in the wave of pains.

Exercise injury: probably not frank (macroscopic) haematuria but instead retroperitoneal bleeding.

Infn of kidney = pyelonephritis (usually always bacterial and is quite painful).

DDx for flank pain:

-  Kidney: pyelonephritis, tumour, trauma
-  Appendix: appendicitis
-  MSK: trauma, unlikely for haematuria
-  GIT: diverticulitis (diverticulum gets infected),  
-  Fistula
-  Could be referred pain (bladder: cystitis) that goes to kidneys

Ix: to order

-  Urinalysis and 24h
-  FBC: check anaemia
-  COAGS
-  CHEM20
-  Renal ultrasound
-  Abdo x-ray

Mx: if pain is rated >7 then give opiates. Usually give paracetemol + NSAID.

## Trigger 2

Dx: polycystic kidney disease. Requirements for Dx: a combination of fam Hx + cysts identified.

PKD usually presents with pain and haematuria. Usually also have hypertension (due to fluid retention + activation of RAAS) and diastolic BP would increase. It causes stasis => kidney stone formation.

Types

-  Autosomal dominant PKD: (PKD1, PKD2; presents in adults) cysts form in organs such as liver, seminal vessels, vasculature and can lead to berry aneurysms
-  Autosomal recessive PKD: (PKHD1; presents in children) gets both copies so presents in *utero*.

Ascending cholangitis: name is poor but it means *infection* of biliary tree. Symptoms triad = {fever, jaundice, }.

### 2017-05-24 renal3.rmd
yes

---
title: "Renal III"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

Path

Dx of CKD requires:

-  Decline of kidney function for 3 months or more \textbf{AND}
-  Evidence of kidney damage (e.g. albuminuria or abnormal biopsy) textbf{OR}
-  GFR <60 mL/min/1.73 m2

Staging CKD

-  Stage 1: Kidney damage with normal or increased GFR (>90 mL/min/1.73 m2)
-  Stage 2: Mild reduction in GFR (60-89 mL/min/1.73 m2)
-  Stage 3: Moderate reduction in GFR (30-59 mL/min/1.73 m2)
-  Stage 4: Severe reduction in GFR (15-29 mL/min/1.73 m2)
-  Stage 5: Kidney failure (GFR <15 mL/min/1.73 m2 or dialysis)

### 2017-05-25 renal3.rmd
added short case 1

@@
 
-Path
+# Readings
 
@@
 -  Decline of kidney function for 3 months or more \textbf{AND}
--  Evidence of kidney damage (e.g. albuminuria or abnormal biopsy) textbf{OR}
--  GFR <60 mL/min/1.73 m2
+-  Evidence of kidney damage (e.g. albuminuria or abnormal biopsy) \textbf{OR}
+-  GFR <60 mL/min/1.73 m^2^
 
@@
 -  Stage 5: Kidney failure (GFR <15 mL/min/1.73 m2 or dialysis)
+
+See Dan's autoregulation of kidney [GIFT](https://drive.google.com/drive/folders/0B0QeJC_rDeMIYjZMU2VfQVR2RU0)
+
+# Main case
+
+**Attach the image from the whiteboard**
+
+High anion gap is an indicator for metabolic acidosis.
+
+Parenteral replacement is $\neg$oral (avoiding enteric system).
+
+Hypocalcaemia $\rightarrow$ hyperPTH and hypophosphataemia + hyperphosphaturia (lack of PO4 reabsorption in the nephrons).
+
+# Short case 1
+
+## Trigger 1
+
+35M presenting with increasing lethargy and weight loss. No fevers, night sweats, or pain.
+
+Past Hx: 12 years old Dx with T1D.
+
+Pathophysiology of diabetic nephropathy: glycosylation of basement membrane => leaky => proteins flow out => thickening of arteriole (mainly efferent) [vasoconstriction] => decreases blood flow to peritubular capillaries => they start dying.
+
+-  ACEi dilate efferent which renoprotects against this efferent constriction at the cost of decreased GFR
+-  Scarring due to HTN, glomerulopathy usually leads to atrophy due to fibrosis and it shrivels up, whereas due to diabetes has normal sized kidneys: sugar gets deposited in the kidney with sugar crystals and glycosylation (shit kidneys, low GFR, normal sized)
+
+Presentation for CKD does not have obvious symptoms.
+
+## Trigger 2
+
+Antimicrobials
+
+-  Co-trimoxazole = trimethoprim + sulphamexazole (folic acid inhibitors at different parts in the pathway; synergistic). General prophylaxis against bacteria. Good against gram(+) such as Staph.
+-  Valganciclovir = antiviral. Targeting CMV
+-  Amphotericin lozenges = antifungal
+
+HLA-DR3,4 is *extremely* common with T1D.
+
+Mechanisms of rejection
+
+1.  HLA incompatibility
+2.  ABO: blood type
+3.  Rh: rhesus factor
+4.  Cross-matching: check if recipient has any miscellaneous antibodies (anti-{Kell, Duffy, White} are some common ones)

### 2017-05-25 renal3.rmd
completed short case

@@
 4.  Cross-matching: check if recipient has any miscellaneous antibodies (anti-{Kell, Duffy, White} are some common ones)
+
+Immunosuppresive agents:
+
+-  Calcineurin inhibitors: tacrolimus and cyclosporin act and prevent IL-2 formation => no T cells
+-  Antimetabolites: mycophenalate prevent B and T lymphocyte production
+-  IL-2 R antag: basilixamab
+
+Side effects: opportunistic infections, cushings, malignancy, osteoporosis (w/ glucocorticoids)
+
+# Short case 2
+
+## Trigger 1
+
+42M severe right flank pain and is now all the way to his groin. Pain is coming in waves and came in 1 hour ago.
+
+DDx: renal colic caused by kidney stones (aka nephrolithiasis), biliary colic (gall stones biliary tree), pyelonephritis, PKD => cyst rupture, (Female can be ovarian stuff), renal capsule stretched, malignancy
+
+Colicky pain in hollow viscous with peristalsis.
+
+Pain radiation: referred pain due to nerves such as iliohypogastric
+
+Pathophysiology of HTN: build up of fluid => Down GFR => RAAS to increase BP. HTN is a risk factor for kidney stones
+
+Ix to order:
+
+-  Xray KUB: Ca stones (80-90%) are visible since it is a metal
+-  Ultrasound KUB: visualise hydronephrosis (swollen proximal ureter and thin distal ureter) due to an obstruction, can see calcification, see if kidney is swollen, bladder = {stones in there?, vesicoureteral junction reflux}
+-  CT KUB: if ultrasound is not diagnostic
+-  CHEM20 can view the PO4, Ca2+, Mg2+, uric acid  which may cause the stone | measure eGFR which can indicate for surgery
+-  Urine MCS: bacteriuria, casts/dysmorphic red cells, pyelonephritis, septic cause, maybe see crystals (oxalate)
+-  FBC: white cells can indicate pyelonephritis
+-  Urinalysis: proteinuria or haematuria, nitrites pathognomic of UTI (which includes pyelonephritis)
+
+## Trigger 2
+
+Kidney stones often made up of: Ca, oxlate, NH4+, PO4, cysteine, nitrogen compounds
+
+Risk factors urolithiasis: Hx of gout, foods high in oxalate (beetroot, strawberries), alkaline urine (risk for uric acid stones), diabetes, immobility (stasis; such as from obstructive disease like **PKD**), males, fam Hx, drugs, dehydration
+
+Stones can lead to infection, hydronephrosis, decreased GFR, AKI => CKD, shitload of pain, rupture of ureter
+
+Lifestyle Mx: hydration, change diet (less of some shit, less beer), exercise
+
+Tx:
+
+-  Medical: lithotripsy can break the stone down if it is small enough (<2mm), alpha blockers or CCB (?!; is possible)
+-  Surgery: urological surgery (by putting a stent in and allowing it to pass through
+
+Key factor: if the stone is >10mm then operation. Not then try medical treatment

### 2017-05-27 hsr-is-fucking-stupid.rmd
fuck public health

---
title: "MEDI7101 Individual Assigment"
author: "Aaron Khoo | 43577367 "
date: "27 May 2017"
output: pdf_document
---

ATSI controlled health organisations vs private GP

-  Similarities
-  DIfferences
-  DIscuss differences, explore why they exist, impact,

Aboriginal community controlled health service (ACCHS) sector

#  Health service delivery

Delivery model={ usually private care

Some deliver specialised stuff e.g. skin care gp

GPs usually reactive models

Multidisciplinary: physio, nutricianists, dieticians to imrpvoe service delivery. Limited number of agencies use this

Private practices can subsidise mental health nurse from state and federal gov.

Subsidises vaccinations provided reach criteria for state gov

## Safety and quality

set outside the practise: accreditation sets up this

## demand for care

Abos: up to 70s no specific community controlled services. Driven by community with community board structures. Only white medical staff, limited abo nurses|doctors. Demand for abo care not delivered by the state.

Universal health coverage cover 3 elements: universality/populations (which popns covered: medicare doesn't cover asylum seekers/refugees/visitors | many abos due to lack of documentation have not been recognised in far north qld. Changing to proactive model and pushing for abos to become clients) ||| Financial risk:freezing of medicare gap

Deadly choices:wells person check

#  Health workforce

Available: are staff there when they're needed.

Health staff feelign superior to pt?! Major driver of abos going to ACHSS is that other places treat them with disrespect/racism. *Responsiveness*.

ACCHS hjave dynamic workforce too: staff then transition elsewhere, or come in from overseas.

Need to register to database.

ACCHSs workforce devn

#  Health information and research

Systems to the cloud, linkage between labs, and gps. Have to make use of this info at the practice level.

GP: clinical data

GPs tend to deal one to one basis diff to community controlled which is interested in the whole abo population. ACCHSs trying to target social determinants of health just starting to use thsi information.

Social determinants of health poorly recorded such as employment, family status, accomodation.

#  Medical products and technologies

Seems to refer to national level.

In aus most of these decisions are made at commonwealth level ensuring quality, pbs ensures availability. Procurement isn't really an issue. New products are the issue: new products extremely expensive such as cancer drugs. Equitable access?! ACCHSs seek to cover gap so there is no financial disincentive for pt with healthcare card as they have to cover gap.

#  Healthcare financing

Seems to refer to national level. Try to eliminate possibility that family pushed into financial hardship due to medical issues.

Medicare system good for both. Medicare returns. Mental health reimbursements, cervical smears

GP funding driven by small business models usually, some being bought up by big GP.

ACCHSs funding is tougher. Requires pt to be registed with medicare to get reimbursed but many abos are NOT registered. More exaggerated risk as people trained in smoking cessation will have trouble transitioning as commonwealth funding for this project dries up. Dentistry available which isn;t for GP.

#  Leadership and governance

GPs now run as networks as they are being bought up by conglomerates. Policy comes from top (commonwealth, state) and also comes from how services are run.

ACCHSs have clear formal and informal ways of doing things which are different. Accountability is also clearly different,

ACCHS more community represented

GPs don't have much info of individual pt, not a strong way of analysing data (hospital, demographic). Continuity of care represents single individual || ACCHSs deal with population as a whole. Some potential clients are already happy with GP, but there is much of the population that does not have adequate medical contact.

### 2017-05-28 renal1.rmd
updated to helen healy's lectures

@@
 
-3 cell populations in glomerulus = {endothelial, mesangial cells, epithelial cells (foot processes and parietal), } + 2 structural units = {basement membrane, mesangium}
+3 cell populations in glomerulus = {endothelial, mesangial cells, epithelial cells (foot processes and parietal)} + 2 structural units = {basement membrane, mesangium}
+
+Proliferative:
+
+-  Endothelial cells: Post infectious GN
+-  Mesangial cells: IgA nephropathy
+-  Parietal epithelial cells (lining bowmans capsule): crescentic GN (RPGN + nephritic syndrome)
+-  GBM: Anti-GBM autoimmune response against Type IV collagen
 
@@
 -  Normal constituents of the glomerulus (e.g. Goodpastures syndrome; type II hypersensitivity)
--  Non-renal self antigens localised in globermuli => immune attack (e.g. SLE)
+-  Non-renal self antigens localised in globermuli $\rightarrow$ immune attack (e.g. SLE)
 -  Exogenous antigens that localise in glomerular capillaries (e.g. PSGN)
@@
 
--  Blood (haemoglobin): test using urine microscopy/urinalysis = {ketones, villirubin / microscopy looks at RBCs, WBCs, epithelial cells {suggestive of non-midstream catch:not clean sample)} or urine dipstick test {protein (kidney disesae), pH, RBCs, leukocytes (WBCs), ketones(Diabetic Ketoacidosis; ketosis), specific gravity (osmolites/electrolytes in the urine), glucose (diabetes), bilirubin (cholestasis [nothing moving from gallbladder => spill into bloodstream]), nitrites (bacterial waste products)}
+-  Blood (haemoglobin): test using urine microscopy/urinalysis = {ketones, villirubin / microscopy looks at RBCs, WBCs, epithelial cells {suggestive of non-midstream catch:not clean sample)} or urine dipstick test {protein (kidney disesae), pH, RBCs, leukocytes (WBCs), ketones(Diabetic Ketoacidosis; ketosis), specific gravity (osmolites/electrolytes in the urine), glucose (diabetes), bilirubin (cholestasis [nothing moving from gallbladder $\rightarrow$ spill into bloodstream]), nitrites (bacterial waste products)}
 -  Beetroot/foods
@@
 
--  If pain => SOCRATES; ever had before
+-  If pain $\rightarrow$ SOCRATES; ever had before
 -  Fam Hx
@@
 
-## Trigger 1`
+## Trigger 1
 
-Glomerulonapathies:
+Glomerulopathies:
 
--  Renal cause CVS: Decrease GFR => fluid retention => overloading the heart
--  CVS causing renal: vasculitis/renal artery stenosis => kidney failure => fluid retention => overloading the heart
+-  Renal cause CVS: Decrease GFR $\rightarrow$ fluid retention $\rightarrow$ overloading the heart
+-  CVS causing renal: vasculitis/renal artery stenosis $\rightarrow$ kidney failure $\rightarrow$ fluid retention $\rightarrow$ overloading the heart
 
@@
 
-Glomerular damage => breach in the filtration system => heavy proteinuria.
+Glomerular damage $\rightarrow$ breach in the filtration system $\rightarrow$ heavy proteinuria.
 
@@
 
-DVT and kidney fn: nephrotic syndrome => losing proteins such as clotting factors (loses anti-thrombotic, loses pro-thrombotic but loses more anti-thrombotic due to the number?!) => hypercoagulable state
+Hyperlipidaemia: lose proteins $\rightarrow$ liver detects this and compensates by producing more apolipoproteins/cholesterol  
+DVT and kidney fn: nephrotic syndrome $\rightarrow$ losing proteins such as clotting factors (loses anti-thrombotic, loses pro-thrombotic but loses more anti-thrombotic due to the number?!) $\rightarrow$ hypercoagulable state (antithrombin-3 mentioned in path tute)
 
@@
 
-3 most common nephropathies for nephrotic syndrome
+3 most common GN for nephrotic syndrome
 
@@
 -  Membranous glomerulonephritis:
--  Focal segmental glomerulosclerosis (FSGS):
+-  Focal segmental glomerulosclerosis (FSGS): adults
 
@@
 
-Post infectious GN: circulating immune complexes form and clog up the glomeruli => inflamm response => damaged glomerular tufts => nephritic syndrome
+Post infectious GN: circulating immune complexes form and clog up the glomeruli $\rightarrow$ inflamm response $\rightarrow$ damaged glomerular tufts $\rightarrow$ nephritic syndrome
 

### 2017-05-28 renal2.rmd
typos for week ii

@@
 
-Pathophysiology: tl;dr glomerular filtration depends on pressure gradient from glomerulus to bowman space.  
+Pathophysiology: tl;dr glomerular filtration depends on pressure gradient from glomerulus to bowmans space.  
 Main determinant is renal blood flow (RBF).
@@
 Inflammation comes in and cell injury occurs even after restoration.
-During this time tubular cell damage fucks up tight junctions => leakage of glom filtrate + dyaing cells enter tubules forming casts => decrease in GFR => oliguria.
+During this time tubular cell damage fucks up tight junctions $\rightarrow$ leakage of glom filtrate + dying cells enter tubules forming casts $\rightarrow$ decrease in GFR $\rightarrow$ oliguria.
 
@@
 Early RBF recovery has better prognosis.
-Once it is restored remaining nephrons increase filtration and hypertrophy, however if not enough nephrons left then these sclerose => more nephron loss.
+Once it is restored remaining nephrons increase filtration and hypertrophy, however if not enough nephrons left then these sclerose $\rightarrow$ more nephron loss.
 Hyperfiltration theory of renal failure: as nephrons die then puts more pressure on the rest which are then more prone to dying.
@@
 
-Hypertension is a pre-disposing factor for => hypertrophy of wall => MI => blood leaking => cardiac tamponade
+Hypertension is a pre-disposing factor for $\rightarrow$ hypertrophy of wall $\rightarrow$ MI $\rightarrow$ blood leaking $\rightarrow$ cardiac tamponade
 
@@
 
-Vegetations on heart valve - Rheumatic fever (acute immunologically mediated multisystem inflammatory disease that occurs a few weeks following an episode of group A Strep pharyngitis) leads to acute chronic rheumatic heart disease, which leads to chronic RHD (mitral stenosis commonest lesion) – fibrotic valvular disease – Left atrium dilates – congestion of lungs – Right ventricular hypertrophy (boot shaped heart) – dental extraction – infective endocarditis (bulky friable vegetations composed of thrombotic debris and organisma) – septicaemia and death.
+Vegetations on heart valve $\rightarrow$ Rheumatic fever (acute immunologically mediated multisystem inflammatory disease that occurs a few weeks following an episode of group A Strep pharyngitis) leads to acute chronic rheumatic heart disease, which leads to chronic RHD (mitral stenosis commonest lesion) – fibrotic valvular disease $\rightarrow$  Left atrium dilates $\rightarrow$ congestion of lungs $\rightarrow$ Right ventricular hypertrophy (boot shaped heart) $\rightarrow$ dental extraction $\rightarrow$ infective endocarditis (bulky friable vegetations composed of thrombotic debris and organisma) $\rightarrow$ septicaemia and death.
 
@@
 
-It is usually seen in debilitated patients eg. Cancer, where there is a hypercoaguable state which leads to ‘marantic vegetations’.
+It is usually seen in debilitated patients eg. Cancer, where there is a hypercoaguable state which leads to 'marantic vegetations'.
 
@@
 
-Smoking => emphysema => centriacinar (not hereditary). $\alpha_{1}$-anti-trypsin would cause pan acinar
+Smoking $\rightarrow$ emphysema $\rightarrow$ centriacinar (not hereditary). $\alpha_{1}$-anti-trypsin would cause pan acinar
 
-CXR: Flattening of the diaphragm => lung spaces have enalarged. Heart looks tubular (narrowing)
+CXR: Flattening of the diaphragm $\rightarrow$ lung spaces have enalarged. Heart looks tubular (narrowing)
 
@@
 
-Despite the “matched” V/Q defect, overtime hyperventilation develops and cardiac output (CO) drops which leads to areas of poor blood flow in relatively well oxygenated areas.
+Despite the 'matched' V/Q defect, overtime hyperventilation develops and cardiac output (CO) drops which leads to areas of poor blood flow in relatively well oxygenated areas.
 Due to this poor CO, the rest of the body suffers from tissue hypoxia.
 
-Pathogenesis of RHF: Matched V/Q defect --> Overtime hyperventilation develops --> Cardiac output drops --> Hypoxaemia --> Tissue hypoxia (this may cause polycythaemia)  and Increased CO2 retention (this may cause respiratory acidosis) --> Arterial vasoconstriction in lungs --> pulmonary hypertension --> Retrograde pressure build-up --> Elevated right ventricular pressure --> right ventricular hypertrophy --> right ventricular failure (Cor pulmonale)
+Pathogenesis of RHF: Matched V/Q defect $\rightarrow$ Overtime hyperventilation develops $\rightarrow$ Cardiac output drops $\rightarrow$ Hypoxaemia $\rightarrow$ Tissue hypoxia (this may cause polycythaemia)  and Increased CO2 retention (this may cause respiratory acidosis) $\rightarrow$ Arterial vasoconstriction in lungs $\rightarrow$ pulmonary hypertension $\rightarrow$ Retrograde pressure build-up $\rightarrow$ Elevated right ventricular pressure $\rightarrow$ right ventricular hypertrophy $\rightarrow$ right ventricular failure (Cor pulmonale)
 
@@
 
-This x-ray shows a coin lesion in the right upper lobe.a
+This x-ray shows a coin lesion in the right upper lobe.
 
-Hamartoma: excess tissue in a normal location
-
-Choristoma: excess tissue in an abnormal location.
+-  Hamartoma: excess tissue in a normal location
+-  Choristoma: excess tissue in an abnormal location.
 
@@
 
--  Abcsess is an area of necrosis that is well sealed
+-  Abscess is an area of necrosis that is well sealed
 -  Cavity: leaves behind a hole
@@
 22F, fell from horse 1:45h ago, at rural hospital.
-Closed (within the skin; open means it breaks the skin and ready for osteomyolitis) spiral fracture of the 1/3 mid femur: neurovascular compromise suspected (scared of haematoma => compartment syndrome).
+Closed (within the skin; open means it breaks the skin and ready for osteomyolitis) spiral fracture of the 1/3 mid femur: neurovascular compromise suspected (scared of haematoma $\rightarrow$ compartment syndrome).
 Diastysis of pelvic symphysis.
@@
 
-$\0.5mL/kg/hr$ is the bare \textit{minimum} urine output.
+$0.5mL/kg/hr$ is the bare *minimum* urine output.
 
-If the vasa recta are not perfused adequately => ischaemia of the tubules => necrosis. It does not have a compensatory mechanism so if perfusion drops then it just cannot perfuse the tubular cells.
+If the vasa recta are not perfused adequately $\rightarrow$ ischaemia of the tubules $\rightarrow$ necrosis.
+It does not have a compensatory mechanism so if perfusion drops then it just cannot perfuse the tubular cells.
 
@@
 
--  Hyperkalaemia => ECG ASAP
+-  Hyperkalaemia $\rightarrow$ ECG ASAP
 -  Uraemia (increase in urea)
@@
 
-Cystinosis => cysteine build up in proximal tubules => they start failing => lose electrolytes => Fanconi syndrome => decreased phosphate => decreased Ca absorption => rickles
+Cystinosis $\rightarrow$ cysteine build up in proximal tubules $\rightarrow$ they start failing $\rightarrow$ lose electrolytes $\rightarrow$ Fanconi syndrome $\rightarrow$ decreased phosphate $\rightarrow$ decreased Ca absorption $\rightarrow$ rickles
 
@@
 -  MSK: trauma, unlikely for haematuria
--  GIT: diverticulitis (diverticulum gets infected),  
+-  GIT: diverticulitis (diverticulum gets infected)
 -  Fistula
@@
 
-PKD usually presents with pain and haematuria. Usually also have hypertension (due to fluid retention + activation of RAAS) and diastolic BP would increase. It causes stasis => kidney stone formation.
+PKD usually presents with pain and haematuria.
+Usually also have hypertension (due to fluid retention + activation of RAAS) and diastolic BP would increase.
+It causes stasis $\rightarrow$ kidney stone formation.
 

### 2017-05-29 renal3.rmd
updated quiz

@@
 ---
+
+```{r setup, include=FALSE}
+setwd("~/Documents/uni_work/4_sem_1/7111/lectures/15_renal_iii/readings")
+```
 
@@
 
-See Dan's autoregulation of kidney [GIFT](https://drive.google.com/drive/folders/0B0QeJC_rDeMIYjZMU2VfQVR2RU0)
+See Dan's autoregulation of kidney [**GIFT**](https://drive.google.com/drive/folders/0B0QeJC_rDeMIYjZMU2VfQVR2RU0)
+
+# Quiz
+
+-  Which of the following drugs must be used with particular caution in elderly patients with an existing low eGFR? 
+    -  **Aspirin**
+-  Which of the following indicates that kidney failure is chronic rather than acute? 
+    -  Small kidneys
+-  Pathophysiologic mechanism implicated in the development of diabetic nephropathy?
+    -  Glomerulopathy due to glucose toxicity.
+-  Epithelial cells on urine microscopy indicates contamination.
+-  The commonest cause of death in severe chronic kidney disease is
+    -   vascular disease
+-  Greatest contributor to increase in CKD is diabetes mellitus NOT essential HTN/ageing population.
 
@@
 
-**Attach the image from the whiteboard**
+![Was scribe](scibe-is-shit.jpg)
 
@@
 
-Pathophysiology of diabetic nephropathy: glycosylation of basement membrane => leaky => proteins flow out => thickening of arteriole (mainly efferent) [vasoconstriction] => decreases blood flow to peritubular capillaries => they start dying.
+Pathophysiology of diabetic nephropathy: glycosylation of basement membrane $\rightarrow$ leaky $\rightarrow$ proteins flow out $\rightarrow$ thickening of arteriole (mainly efferent) [vasoconstriction] $\rightarrow$ decreases blood flow to peritubular capillaries $\rightarrow$ they start dying.
 
@@
 
--  Calcineurin inhibitors: tacrolimus and cyclosporin act and prevent IL-2 formation => no T cells
+-  Calcineurin inhibitors: tacrolimus and cyclosporin act and prevent IL-2 formation $\rightarrow$ no T cells
 -  Antimetabolites: mycophenalate prevent B and T lymphocyte production
@@
 
-DDx: renal colic caused by kidney stones (aka nephrolithiasis), biliary colic (gall stones biliary tree), pyelonephritis, PKD => cyst rupture, (Female can be ovarian stuff), renal capsule stretched, malignancy
+DDx: renal colic caused by kidney stones (aka nephrolithiasis), biliary colic (gall stones biliary tree), pyelonephritis, PKD $\rightarrow$ cyst rupture, (Female can be ovarian stuff), renal capsule stretched, malignancy
 
@@
 
-Pathophysiology of HTN: build up of fluid => Down GFR => RAAS to increase BP. HTN is a risk factor for kidney stones
+Pathophysiology of HTN: build up of fluid $\rightarrow$ Down GFR $\rightarrow$ RAAS to increase BP. HTN is a risk factor for kidney stones
 
@@
 
-Stones can lead to infection, hydronephrosis, decreased GFR, AKI => CKD, shitload of pain, rupture of ureter
+Stones can lead to infection, hydronephrosis, decreased GFR, AKI $\rightarrow$ CKD, shitload of pain, rupture of ureter
 

### 2017-05-30 renal4.rmd
added cbl main case

---
title: "Renal IV"
author: "Aaron Khoo"
date: "`r Sys.Date(format('%d %M %B'))`"
output: pdf_document
---

# Main case

## Trigger 1

> ED, 72M severe lower abdo pain, NPU (not passed urine) 1/7, distressed, sweating and very restless.

Sources of pain: urinary bladder, GIT (rectum), prostate, kidney radiating pain, urethra, distal ureter, radiating from gonads, Males: hernia,

Pathophysiological processes for pain: infn, inflamm, ischaemia, trauma, obstruction (could be benign thing), malignancy, herniation, fistulisation (abnormal connection between 2 epithelialised connections [pierced ears are a fistula]), autoimmune destruction (loss of tolerance)

Nature of the pain:

-  Somatic vs visceral: things affecting visceral peritoneum would be very sharp due to lots of nerves
-  Nociceptive (pain due to nociceptors detecting tissue damage. Cna be blocked with NSAIDS, opiods; sharp, burning) vs neuropathic (irritated/diseased nerves e.g. peripheral/autoimmune/diabetic neuropathy or obstructions; described as tingling, burning, pins and needles)
-  Types of abdo pain: epigastric (foregut), umbilical (midgut), lower abdo (hindgut)?!?!: most likely to be hindgut pain (2/3 of transverse colon and down)

Pathophysiological processes for decreased urine output: obstruction of {urethra, bladder, VUJ, ureter, renal pelvis}, obstruction due to {calculi, tumour (prostate, urothelioma, bladder), thrombus is rare, PKD, extraluminal compression symptoms e.g. from severe constipation}, drugs, hypovolaemia secondary to {dehydration, haemorrhage}, shocks, strictures (narrowing of ureter, urethra; more often congenital but severe UTI), neurological

Mech behind restlessness and sweating: SNS activation, UTI could lead to delirium (acute state that can be resolved by correcting underlying cause).

Micturition reflex triggered at 400mL: get uncomfortable:

-  Bladder: PS stimulation
-  Internal sphincter: sympathetic from hypogastric N
-  External sphincter: somatic

## Trogger 2

> No past Hx of note, no regular meds, bit of a cold recently and has taken some OTC decongestants. Decongestants block $\alpha$-adrenoceptors. Afebrile (suggests against UTI but it could be a trough in his temperature and peak later), hypertensive. He is hypertensive due to pain.

Risks of cathetirsation: urosepsis (septicaemia due to infection in the genitourinary system)

$V(Urine)$: high is more likely acute obstruction, smaller could be from chronic urinary retention. Bladder can be percussed >150mL, palpable >200mL, acute obstruction V usually <1L, chronic V can be larger as they are desensitised.

Medications interfering with micturition such as NSAIDS, decongestants for URTI

Dipstick provides info about: proteinuria, haematuria, leukocytes, pH, ketones, nitrites, bilirubin. Due to the low sensitivity of urine dipstick we are unable to rule out specific pathologies. Dipstick must be considered in the full clinical context, and further analysis such as urinalysis would be necessary.

## Trigger 3

> Urinary hesitancy, urgency, nocturia, has weak stream. Has had pre-syncopal episode whilst straining to empty bladder.

Neurogenic bladder: problems with the nervous system causing urinary retention / incontinence. Have to empty this to prevent stasis => infn.

Nocturia: waking up to piss at night. Pathophysiology: BPH, obstruction, polydipsia (hyperdynamic state; diabetes), if kidney can't concentrate urine, irritable bladder.

Prostate: inferior to bladder and anterior to seminal vesicles. Secretes stuff into the ejaculate. Prostate converts testosterone to DST (by 5-$\alpha$-reductase) which is 5 times more potent making it undergo hyperplasia and enlarge.

Bladder outlet obstruction anatomical causes: peyronie syndrome, rectum (faeces or tumour) can cause extraluminal compression, lymph nodes could cause commpression, abscess. Natural Hx: infn, VUR, hydronephrosis, bladder rupture.

Clarifying dizziness: distinguishing between:

-  Pre-syncope: feeling like about to lose consciousness; people often sit down. "Feel like you're going to pass out?"
-  Vertiginous: "feel like the room is spinning around you?"

Mechanism for dizziness when straining to void: increasing abdominal pressure => stimulation of vagal nerve => micturition syncope (emptying large amounts of V from bladder).

Medications associated with micturition problems: NSAIDs, anti-cholinergics (biggest ones as the inhibit the ACh induced PS stimulation; oxybutynin, opiods, anti-depressants, amphetamines (SNS activation))

Need to monitor to prevent him from going into retention and to prevent the urinary bag from going overflowing (prevent AKI and infn).

## Trigger 4

Trabeculated bladder: hypertrophy due to increased resistance (overcoming the obstruction of the enlarged prostate).

PSA elevation above normal range is significant (even if the real world it may be a really shit test). However the slow rise is reassuring compared to a massive sudden increase.

Reduced kidney fn: could be due to age, possibly due to HTN.

## Trigger 5

UTIs: not common in males. Risk factors = urinary stasis.

Common bacteria are coliforms (from the colon; gram neg rods such as E. coli, Klebsiella)

Symptomatic relief for prostatic hyperplasia:

-  $\alpha$-blockers: prazosin
-  5-$\alpha$-reductase inhibitors: finastemide
-  Anti-cholinergics: oxybutynin

### 2017-06-01 renal4.rmd
added hsr
added template for short case
updated sc1

@@
 ---
+
+# Quiz
+
+-  Clavulanic acid contains beta-lactam, gets preferentially broken down by resistant bacteria. Therefore penicillin and cephalosporins can have an effect.
+-  Receptors
+    -  M~3~: ACh from parasympathetic pelvic nerve to contract detrusor
+    -  B~3~: hypogastric N (sym): noradrenaline relaxation of detrusor
+    -  $\alpha_{1}$ in the int sphincter: hypogastric N: contraction of ext sphincter
+    -  Nicotinic in the ext sphincter: somatic from pudendal nerve to contract it
+    -  Also sensory N AFFERENT in detrusor that is stimulated when bladder is stretched
 
@@
 -  Anti-cholinergics: oxybutynin
+
+# Short case 1
+
+## Trigger 1
+
+> 18F severe burning pain on urination. Offensive urine = smelly. Sexually active. No fever, sweats, nausea, or back pain.
+
+Organ systems involved = {urinary system, reproductive}
+
+Sexual Hx: STI, pregnancy, infn. Facilitates trauma for bugs.
+
+Absence of systemic symptoms $\rightarrow$ localised: not urosepsis, pelvic inflammatory disease (sepsis within the abdomen)
+
+Well person Ix: don't really have to FBC since there is no education that there is systemic effects.
+
+## Trigger 2
+
+Sensitivity and specificity of dipstick: low for both
+
+Empirical treatment: don't want disease progression, also provide some relief from symptoms (stinging during urination due to acidity of urine and also infn+inflammation)
+
+Trimethoprim: works most of the time for *E. coli*, low side effects, generally well tolerated (low allergies)
+
+## Trigger 3
+
+Antibiotic selection = {narrow spectrum, lowest dose, shortest time}
+
+Medication: empaglutlozin blocks the SGLT-2 transporter so can't reabsorb sugar. Makes people piss out 80g urine, helpful for diabetes, obese, bacteria in urine (increased risk of UTIs).
+
+# Short case 2

### 2017-06-01 renal4.rmd
added ryan's farewell

@@
 # Short case 2
+
+## Trigger 1
+
+> 22/12M child, febrile, vomited, diarrhoea
+
+GIT symptoms from systemic things such as: thyroid problems, infn in urinary tract cytokines (IL-1, TNF-$\alpha$, IL-5), sick from URTI can give diarrhoea.
+
+Pathophysiological processes causing fever: malignancy, infn, inflammation (autoimmune), medications/immunisations,
+
+Pathophys involved with diarrhoea and vomiting: fluid shifts associated with fever, inflamm, infn from {protozoa, bacteria, viral}, toxicity, thyrotoxicosis, pain, overeating, motion sickness, allergic response
+
+NSAIDS and paracetamol make fever abate.
+
+Most common cause of fever in child: URTI such as otitis media, viral, skin infections
+
+Paediatrics is like veterinary science => have to ask random q's since child can't explain things themselves. \textbf{Always ask more and examine more in children}.
+
+Conditions not to miss in a febrile child (potentiall fatal or life altering)
+
+-  Ca
+-  Septic shock
+-  Meningococcal disease (Neiserria meningitis (gram neg coccus); meningitis): neck stiffness/rigidity, fever, koenig's sign, Brudzinski's sign (more for adults), rash (late sign; caused by septic emboli getting into tissue and causing microinfarctions => non-blanching rash which separates it from dermal rashes), decreased level of consciousness
+-  Measles: C's = {koplic spots (white spots in mouth), conjuctivitis, coryza}, fever
+
+## Trigger 2
+
+Lethargy is basically saying the child is unwell and needs some help.
+
+Fever
+
+-  Duration: $\geq5$ days: have to consider Kawasaki disease which is an autoimmune disease
+-  Frequency: cyclical (malaria,...)
+
+Asymptomatic UTIs: no dysuria = {young, elderly}
+
+Take samples **before** starting antibiotic treatment.
+
+## Trigger 3
+
+Asymptomatic bacteriauria (bacteria in urine) which isn't a problem unless symptoms, pregnant, or transplant.
+
+High fever usually indicates pyelonephritis (also common in first year of life) rather than cystitis.
+
+Ureters implant obliquely on the bladder which is a functional valve (not anatomical).

### 2017-06-02 pathreview.rmd
I'm dead now

# Hints

-  Learn path from tutes and go there to view it
-  Learn clinical vignette, pathophys processes, sequelae, risk factors, associations, etc.
-  Histo: learn fn of the tissues you are studying

# Micro

Clostridium: rod | Strep are NOT cat(+) | $\beta$-haemolytic are MRSA

Beta is complete, alpha is partial, gamma is almost non-existent

# Atherosclerosis

High yield. Location is important too. SO abdo aorta is most likely to occur.

Thrombus from rubture in plaque | Embolus: plaque moves around

Pathophys: smooth muscle from tunica media. Exposure of subendothelial collagen => platelet aggregation and activation

## Histo

Foam cell: clear, circular, and has a nucleus. Chloesterol clefts are the sword looking imprints. Smooth muscle does exist.

Plaque ruptured => traps RBCs. Pale area: fibrous-looking, contains RBCs and vessels = granulation tissue. Type III collagen is deposited => switches to Type I.

# Arteriosclerosis

3 types. Hyaline in general refers to pink proteinaceous material. Learn about AD-PKD since that is a fairly common Qs.

# Aneurysms

Circle of Willis; no internal elastic lamina => saccular (berry) aneurysms

Learn ehlers-danlos

# Aortic dissection

Blood goes from lumen into intima. Elastic tissue not holding up. Vasa vasorum are vessels that supply other vessels.

## Histopath

Intima is the top, because vasa vasorum is visible at the bottom.

# DVT

Phlebothrombosis are thrombosis in veins | Has lines of zahn which are aggregates of fibrin and platelets, means it is ante-mortem (pre-death).

# MI

-  B = blue cells are neutrophils so it is probably early hours
-  C = nuclei starting to disappear so probably still early
-  D = still massive amounts of neutrophilic shit everywhere so probably day 2 or 3
-  E =  fibroblasts and vessels = granulation tissue

Acute Inflamm: first cell is neutrophils, after day 2 macrophages come in along with granulation tissue, deposit type III collagen, then replaced by type I collagen

NTK: what time points are MI shit happening (histo)

Evolution: coagulative necrosis is when nuclei start to fade. Early granulation tissue is Type III collagen which can still rupture.

Pathology: aneurysm in ventricle

Know Virchow's Triad

# PE

E.g. Someone completed surgery and had a PE. List some risk factors:.:

# Infarcts

Red infarcts: haemorrhagic => due to collaterals. Lungs, liver, intestines

Pale: => end arteries such as heart, kidney, spleen

# Infective endocarditis

Gram stain pink = gram neg

# RHD

Pathognomic are aschoff bodies. Not often zoomed in either. Causative organ Strep pyogenes: E (purple cocci). D is probs Staph. A is enteroccosus?..

Tammy likes molecular mimicry.

Learn Jones criteria

Histo: aschoff nodules have caterpillar nuclei

Sequelae of viral myocarditis (a short case): dilated cardiomegaly | Probs should learn

# Asthma

2 features: mucous plugging, dilation of bronchial wall. UQ likes asthma. Reversible bronchoconstrictionHisto changes: mucous producing cells are goblet cells.

## Histo

Really pink granulated cells are eosinophils in the centre of the bronchus. Basement membrane is thickened (too pink). Cells have infiltrated (blue) are lymphocytes. "See blue cells in inflammatory setting => LYMPHOCYTES".

Drugs and pathophysiology are high yield. Type I hypersensitivity <= Tammy likes this. Th2 subtype is involved in sensitisation. ACh in the lung => bronchoconstriction.

# Emphysema

Bullae => pneumothorax

Types: centriacinar = mostly restricted to resp bronchioles | panacinar: cna consume whole alveolar space. Smoke causes centriacinar as it gets stuck there and can't progress further (usually upper lobe). Panacinar is seen in AAT.

Know lung fn tests:

# Bronchiectasis

Top lobe looks good, bottom has dilation of bronchioles. Is obstructive. Resp epithelium = pseudostratified ciliated columnar.

# Restrictive disease: silicosis

Sarcoidosis would appear earlier than 60.

Fibroblasts come in and try to contain the silica => produce collagen. Nodules build up and cause inflammation.

# Pneumoconiosis

Simple CWP no cancer. COmplicated CWP: higher chance of cancer. Caplan's syndrome

# Asbestosis

## Histo

Extremely thick pleura filled with connective tissue (fibrosis; collagen) and there is some inflammatory infiltrate (lymphocytes).

# Granulomatous inflamm

Activation of macro by Th1 cells secreting IL12. Eat everything then form a giant cell.

# Sarcoidosis

Non-caseating granuloma + hilar lymphadenopathy. Macro: hypercalcaemia due to hyperphosphate?!

# Pneumonia

Greay hepatisation: full of neutrophils exudate and fibrin. Lobar more acute presentation. Learn the stages of pneumonia.

Microbiology in short cases

Gram stain | Shape | Feature
E.coli, strep, Learn the atypicals that cause interstitial pneumonia.

# Neoplasms

HIGH Yield

Keratin pearl = squamous cell.

Learn what metaplasia is UQ likes that.

# Lung tumours

## Adenocarcinoma

Glandular on histo. Scar see collagen.

Paraneoplastic: lambert blocks presynaptic neuron => flaccid paralysis.

## Path/histo

Benign = well circumscribed mass.

# Kidney

# Acute pyelonephritis

# Renal cell carcinoma

carcinoma in-situ: has not yet invaded basement membrane

Chromophobe stains pinkish. Need to know clear cell one

VHL Genetics: signal to produce EPO is hypoxia.

Sarcomas: tumour of connective tissue (fat,...) spread through blood

Carcinomas generally spread lymphatically.

## Wilms tumour

UNdifferentiated embryological. Genes WT1 and WT2.

### 2017-06-02 CVS1.rmd
added old notes

---
title: "CVS I"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
header-includes: \usepackage{wrapfig}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/lectures/4_cvs_i/readings")
library(png)
library(grid)
library(gridExtra)
```

# ECG

```{r images, out.width="350px", fig.align="center", fig.pos="H"}
img1 <-  rasterGrob(as.raster(readPNG("heartwiring.png")), interpolate = FALSE)
img2 <-  rasterGrob(as.raster(readPNG("ecg.png")), interpolate = FALSE)
grid.arrange(img1,
             img2,
             ncol = 2)
```

SA node in RA, then depolarisation spreads through atrial muscle fibres. Delay when it reaches the AV node. Then goes down bundle of His which divides into right and left branches. Purkinje fibres through ventricular muscle. The normal heart rhythm with electrical activation starting from the SA node is known as "sinus rhythm".

The electrical change depends on the muscle mass, $\therefore$ electrical change for the contraction of the atria is small. Parts of ECG:

-  \textbf{P}-wave: contraction of atria (depolarisation)
-  \textbf{QRS}-complex: ventricle depolarisation and atria repolarisation (hidden by the massive ventricles)
-  \textbf{T}-wave: ventricle repolarisation
-  \textbf{U}-wave: origin uncertain, but it might represent repolarisation of papillary muscles. If it follows a normal T-wave then it is assumed to be normal.

The **R-R interval** gives the heart rate. Each square on the ECG printout is 0.2s, so 5 squares is 1s. **PR interval** is the time taken for excitation to spread from SA node through atrial muscles and AV node, and into the ventricular muscles. Normal interval is 120-220ms with the delay taken up by the AV node.

\pagebreak

**Six 'standard' leads**.

-  I, II, VL look at the left lateral surface of heart
-  III, VF at the inferior surface
-  VR examines RA
-  6 V leads look at the heart in a horizontal plane.
    +  V~1~ and V~2~ RV
    +  V~3~ and V~4~ septum between ventricles and ante wall of LV
    +  V~5~ and V~6~ ante and lat walls of LV

**ECGs**

```{r normalecg, out.width="380px", fig.align="centre"}
knitr::include_graphics("ecgnormal.png")
```

Causes of different heart rates

| **Sinus bradycardia** | **Sinus tachycardia** |
| --------------------- | --------------------- |
| Physical fitness | Exercise, pain, fright |
| Vasovagal attacks | Obesity |
| Hypothermia | Pregnancy |
| Hypothyroidism | Anaemia |
| | Thyrotoxicosis |
| | CO~2~ retention |

Tall P-waves may indicate atrial hypertrophy which can be significant if there is evidence of ventricular hypertrophy.

# Cardiovascular risk assessment

1.  Ascertain history of established CVD.
2.  Check for clinically high conditions such as BG, BP, eGFR lipid levels.
3.  Calculate cardiovascular risk based on Framingham risk charts (using Australian ones).
4.  Exclude underestimation of risk. Family history, smoking status, exercise, and diet.
5.  Management: Diet or drug therapy.

**Relative risk** is comparing groups such as smoker vs non-smokers. **Absolute risk** is the $\mathbb{P}(\text{Getting disease})$

# Septic shock

Bacterial infections can become generalised leading to severe sepsis and septic shock.

Pathogen leads to specific responses as well as generalised host response mechanisms resulting in a mix of pro- and anti-inflammatory responses.

**Pro-inflammatory:** Activation of vascular endothelial cells leads to thrombosis, increased vascular permeability, and vasodilation. This leads to a warm shock where patient is warm and flushed but not perfusing organs adequately. The thrombosis can cause disseminated intravascular coagulation => bleeding factors.

**Anti-inflammatory:** inhibition of pro-inflammatory cytokines and suppressing immune cells. It may control the inflammation but increase secondary infection susceptibility.

Organ disfunction arises from reduced tissue perfusion from the combination of {hypotension, interstitial oedema, microvascular thrombosis}. Also metabolic disturbances such as hyperglycaemia and cortisol release.

To stabilise a patient in shock?

-  Oxygen
-  Insert IV and take full bloods for culture
-  Start fluids – normal saline bolus at 20ml/kg/hr

## Infective endocarditis

-  Main organism is: *Streptococcus viridans*
-  First line treatment: penicillin; IV drug users: vancomycin or gentamycin.

# CVS exam

-  Bruit $\subset$ murmur. Bruit is a murmur in the arteries
-  Thrills - palpable heart murmurs
-  Heaves - praecordial movements felt with the heel of the palm on the sternal border

# Pulse Pressure

-  **Narrow**: decreased SV
-  **High**: high SV such as after exercise

# Heart sounds

> S1 -----Systole----- S2 --Diastole-- S1

Thus S1 is closure of mitral and bicuspid valves. S2 is closure of aortic and pulmonary valves.

## Is there a murmur?

When does murmur occur?

-  Systole
-  Diastole

Using S1 and S2 to orient ourself, we can figure out when it occurs. If this is too difficult to figure out, we can feel the carotid pulse which corresponds with systole (sound heard corresponding should be S1).

## Split second heart sound

During inspiration causes more blood to go to the lungs, $\therefore$ pulmonary valve closes *second* (i.e. after the aortic valve).

## What valves correspond with murmur?

Murmur during **systole**

-  **Open (A & P)**: could be stenosed
-  **Closed (T & M)**: regurgitation

## Location

1.  **Aortic area**: *right* upper sternal border in the 2nd ICS (intercostal space)
2.  **Pulmonary area**: *left* upper sternal border in 2nd ICS
3.  **Tricuspid area**: 5th ICS inferior to pulmonary area (+ **erbs** which is 3rd ICS)
4.  **Mitral area**: 4th ICS mid-clavicular line

If murmur is loudest in the aforementioned areas then it is highly like that that specific valve is involved

## Systolic murmurs

-  **Aortic stenosis**: **crescendo-decrescendo (diamond shaped murmur)**. Makes the most sound when blood is pumped hardest (i.e. the middle of systole). The sound also radiates outwards towards the carotid because the carotid is a branch of the aorta. **Ejection click**.
-  **Pulmonary stenosis**: basically the **same** thing that doesn't radiate to the carotids. Heard in the pulmonary area.
-  **Mitral regurgitation**: **pan-systolic murmur (flat murmur)** meaning it occurs throughout the whole of systole. This is as the $\text{P}_{\text{LV}} > \text{P}_{\text{LA}}$ throughout the duration of systole. This radiates to the axilla and is heard from the mitral area duh.
-  **Tricuspid regurgitation**: same thing but heard from tricuspid area. Causes a *pulsatile liver*
-  **Mitral valve prolapse**: **mid-systolic click** with a **late systolic murmur** (as it doesnt close properly) when the valve reopens in the middle of S1 and S2. Non-ejection click.

## Diastolic murmurs

-  **Aortic regurgitation**: heard from the left sternal border (bit inferior to pulmonary area). **Early diastolic decrescendo murmur** because right when the aortic valve is supposed to close at S2, it doesn't fully meaning that blood rushes in instantly. The LV soon fills from the LA which makes it decrescendo as the $P_{LV}$ increases.
-  **Pulmonary regurgitation**: same shit as above but on the right sternal border.
-  **Mitral stenosis**: LV rapidly fills at the beginning of diastole ($\Delta P$ is maximal) and the stenotic valve leaflets *snap* open. Leads to an **opening snap followed by mid diastolic rumble (decrescendo-crescendo)**
-  **Tricuspid stenosis**: **same** thing in tricuspid area.

## Extra heart sounds

The systolic clicks and opening snaps are extra heart sounds covered above.

These are heard when patient rolls onto their left so the heart is closer to the rib cage.

-  **S3**: **volume overload** typically heard right after S2 (diastole) during the rapid filling period as ventricle fills up really quickly tensing up the chordae tendinae which makes the sound. **Ventricular gallop** tensing the chordae tendinae
-  **S4**: **pressure overload** such as during hypertension. Concentric hypertrophy in the LV making it stiffer. Sound heard is when the atrium contracts to get the last bit of blood out into a hypertrophied ventricle. Therefore it is heard just **before S1**. This and S3 are heard best in mitral. **Atrial gallop**. This is always pathological.

# CBL

## Short case II

Risk factors for vascular disease (ABCDE)

-  Age
-  BP
-  Cigs
-  DM (diabetes mellitus)
-  Excess cholesterol
-  Family Hx
-  Gender
-  Hx of personal vascular disease

Hypertension

-  Vascular
    -  LV hypertrophy $\rightarrow$ LV failure.
    -  LV cardiomyopathy as LV gets too big it can't accept enough blood into it.
    -  Microtears in the endothelium
-  Brain
    -  Increased intracranial pressure $\rightarrow$ headaches
    -  $\Uparrow$ risk of aneurysm rupture
    -  Vessel can just burst $\rightarrow$ haemorrhagic stroke (ischaemic stroke is the other one)
-  Eyes - contains small vessels
-  Kidney
    -  Glomeruli explode $\rightarrow$ nephron

# Quiz

## Sympathetic receptors

-  $\alpha_{1}$ - vasoconstriction in many vascular beds. Pupil dilation.
-  $\beta_{1}$ - inotropy and chronotropy in the *heart*
-  $\beta_{2}$ - vasodilation to *sk muscle* as you have to fight things, decreased motility in *GIT*, relaxation of tracheal/bronchial smooth muscle in *lungs* allowing more air in.

## Antihypertensive drugs

$\beta$-blocker AND non-dihydropyridine calcium channel blocker is BAD and increasese risk of heart block.

ACE inhibitors are usually appropriate first line treatments for 53 year old, no evidence of target organ damage.

Avoid $\beta$ blocker with someone that has asthma. If given: this would cause bronchoconstriction - asthmatics already have this.

## Standing after being in a supine position

When a person stands from lying, blood pools in the legs so that venous return decreases. This results in $\Downarrow$ CO and arterial pressure $\rightarrow$ baroreceptors detect the fall and the vasomotor centre is activated to increase sympathetic and decrease parasympathetic outflow.

# Thrombosis

## Virchows triad (risk factors)

-  Endothelial damage
-  Hypercoagubility of the blood
-  Altered blood flow (stasis or turbulence)

## What happens to thrombus

-  Resolution: lysis by fibrinolytic system
-  Organisation: granulation tissue ingrowth from vessel wall
-  Recanalisation: development of new vessels through occlusion
-  Embolism:

# Embolism

-  Thrombi, bone marrow, fat, air, N~2~, can all embolise.
-  Car crash with compound fracture of femur: fat and amniotic fluid emboli

# Haemostasis

-  Vascular phase
-  Platelet phase: activation and aggregation $\rightarrow$ primary platelet plug
-  Coagulation phase: tissue factors = {I,..., XIII}-{VI}. {II, VII, IX, X} are vit-k dep. Goal is to convert Prothrombin $\rightarrow$ Thrombin which catalyses Fibrinogen $\rightarrow$ Fibrin.
-  Fibrinolytic phase: so it doesn't clot everywhere

## Again from a quiz though

1.  Vasoconstriction
2.  Platelet adhesion to exposed tissue
3.  Platelet activation to form platelet/haemostatic plug
4.  Reinforcement of plug with fibrin

Virchows triangle: endothelial damage is the most important.

Platelets can be activated by thrombin through: protease activated receptor (PAR)

# Infarcts

Necrosis due to ischaemia from occluded vessel.

Types:

-  Red (haemorrhagic) vs white (anemic): end organs
-  Septic (infn): inf endocarditis vs bland

MI:

-  Transmural: full thickness $\rightarrow$ path q waves
-  Subendocardial: just partial section of myocardium

### 2017-06-02 CVS2.rmd
added old notes

---
title: "CVS 2"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/lectures/5_cvs_ii/reading")
```

# Alveolar air equation

$$ P_{A}O_{2} = P_{I}O_{2}- \frac{P_{A}CO_{2}}{R}$$

where $P_{I}O_{2}$ = 150mmHg unless using oxygen mask and R = 0.8. $P_{A}CO_{2}=P_{a}CO_{2} \approx 40mmHg$.

## Haemoglobin

O~2~ binding is affected by:

-  $pO_{2}$
-  pH: Bohr effect
-  2,3-BPG: binds and stabilises deoxy form

## Shunts

-  Anatomical not sensitive to enriched O2 in inspired air
-  Physiological are responsive
-  Severity usually measured by shunt equation or alveolar-arterial oxygen difference.

$$ P_{(A-a)}O_{2}$$

# Double density sign

Seen on frontal chest radiograph in the presence of LA enlargement. From ante view, LA is hidden behind the other chambers, however enlargement can cause this to be seen as an extra shadow parallel to the right heart border (double density sign)

# Silent MI

Commonly occurs in older patients with diabetes. Diabetes puts them at risk of autonomic denervation in the heart leading to a silent MI.

# Pain in lower limb

> A 50-year-old man with a history of atrial fibrillation presents with sudden onset of pain in his right leg. On examination, his right leg is cold and pale and no pulses are palpable.

Probably femoral artery *embolism.* Patients with femoral artery *thrombosis* will usually have a history of claudication and long history of ischaemic symptoms including rest pain in the later stage.

# CXR

Will not be able to examine for PE. Good to exclude other causes of dyspnoea such as collapse lung, pneumonia,...

# Congenital heart defects

## Acyanotic

Most common. May have left-to-right shunts, such as various septal defects which result in additional flow from the systemic to the pulmonary circulation. This extra flow increases the normal work of the heart and may lead to chamber dilation, increase in pulmonary blood flow, arrhythmia, and in some cases, heart failure resulting from increased cardiac work.

## Obstruction?

Other patients may have valvular and/or arterial obstruction (such as pulmonary or aortic stenosis, or coarctation of the aorta). These may produce pressure load on the upstream ventricles, and in severe cases impair cardiac muscle function. Valve regurgitation may be seen in some patients as well.

## Cyanotic

Least common. E.g. tetralogy of Fallot, transposition of the great arteries, and the wide variety of single ventricle lesions). These patients have a right-to-left shunt, with desaturated blood entering the systemic circulation without first passing through the lungs, and producing lower than normal systemic

# CBL Short case 1

Prolapse (just move further and further apart. As it worsens then regurg worsens) vs regurgitation

## Trigger 1

Damaged valve $\rightarrow$ turbulent flow $\rightarrow$ **platelet and fibrin deposition on valves** $\rightarrow$ better spot for bacteria to grow (vegetation)

6 months ago became symptomatic of mitral valve prolapse $\rightarrow$ mitral valve regurgitation.

Mitral regurgitation $\rightarrow$ LV hypertrophy and LA dilation as compensatory mechanisms. Common cause is weakening of connective tissue which gets worse with age. Murmur is systolic which radiates to the axilla (armpit). Might feel a thrill and a hyperdynamic apex beat.

Peripheral stigmata of infective endocarditis:

-  Oslers nodes
-  Janeway lesions
-  Splinter haemorrhages

## Trigger II

BP bit low $\rightarrow$ possible hypotensive shock. $\Delta$Blood Pressure $\rightarrow$ So he doesn't have orthostatic (straight/standing) hypotension. Defined by change in 15 systolic and 10 diastolic is a sign of hypotension.

## Trigger III

Drugs

-  Benzyl penicillin - IV penicillin
-  Flucloxacillin has increased activity against staph and strep
-  Gentamycin - gram neg

ECG can pick up from echo

-  LV hypertrophy - elevated R wave
-  LA hypertrophy - bifid p wave
-  A hypertrophy -  Elevated p wave

Diagnosing IE

-  Dukes criteria - 2 major, 0 major 5 minor, 1 major 3 minor.
    -  Major: Positive blood cultures at multiple times (12 hour apart), vegetation on echo
    -  Minor: splinter haemorrhages

# SC II

DDx of exclusion: aging.

DDx

-  pulmonary hypertension, lung cancer, COPD due to smoking, asthma.
-  cardiac: worsening stenosis, heart failure

Risk factors for AS: ABCDE cardiovascular risk factors as athero causes AS

-  Stiffening due to athero
-  Thickening

Symptoms - angina, SOBOE, weak pulse pressure (difference between systolic and diastolic; waterhammer pulse), weak and slow pulse, syncope reduced blood to brain

# Quiz

## Thrombolysis + contraindications

Thrombolysis (clot buster; t-PA enhances plasmin production which degrades fibrin) used in MI, PE, stroke (ischaemic). **Absolute contraindications** = {hx of haemorragic stroke, hx internal bleeding, head trauma, suspected aortic dissection}

## Sinus rhythm

SA node (= sinus node) is the main pacemaker

## D-dimer

Is a fibrin degradation product present after a blood clot is degraded by fibrinolysis. Can be used to help diagnose thrombosis: if negative + low pre-test probability then it rules out thromboembolism.

## Troponins

Integral to muscle contraction in cardiac muscle. Used as diagnosis to differentiate between unstable angina and MI. Post-MI troponin may be high for 2 weeks.

## Endocardium

Serous membrane covering **inner surface of the heart** is endocardium NOT visceral pericardium

# Mitral regurg

Extremely common with an extremely heterogeneous aetiology. Damage, ruptured chordae tendinae, infarction, infection, trauma, iatrogenic,...

### 2017-06-02 CVS3.rmd
added old notes

---
title: "CVS 3"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

# ECG changes from infarct

| Location | Coronary Artery | ST elevation | Reciprocal depression |
| ------------ | ------------ | ------------ | ------------ |
| Anterior | LAD | V1-V6 | None |
| Septal | LAD-septal branches | V1-V4; disappearance of septum Q in V5-V6 | none |
| Lateral | LCX | I, aVL, V5, V6 | II, III, aVF |
| Inferior | RCA | II, III, aVF | I, aVL |

# Angina

-  Variant/printzmetal's angina: occurs without provocation, usually at rest due to coronary artery spasm.
    -  ECG: ST elevation during the pain; hyperventilation and cold-pressor testing required.
-  ECG for angina: transient ST depression + enlarged QRS complex

# MI

-  **STEMI**: transmural infarct; **NSTEMI**: subendocardial infarct
-  Noninfarcted segments undergo hypertrophy and dilation: known as ventricular remodelling. Dilation may outweigh hypertrophy $\rightarrow$ $\Uparrow$O~2~ demand and depressed cardiac fn.
-  Myocardial stunning is the reversible reduction of heart contractility fn
-  On the spot ECG at presentation CANNOT exclude MI
-  ECG ST elevation in II, III, aVF are suggestive of *inferior* MI.
-  Unstable angina and STEMI are frequently indistinguishable at initial evaluation.
-  First treatments and considerations: morphine, oxygen, aspirin, and nitrates (remember MOAN).

## Cardiac markers

-  Lactate dehydrogenase: isoenzyme LD1 is used, LD2 NOT used
-  Creatine kinase in the heart:  40% CK-MB and 60% CK-MM
-  Troponins:  TnT and TnI are used
-  Myoglobin is used alongside primary markers to act as a negative indicator of MI

# Venous thromboembolism

-  Patients age has no effect on duration and intensity of DVT therapy
-  Maintaining normal venous haemodynamics
    -  Thoracic and cardiac pump
    -  Respiratory pump
    -  Valvular competence
-  Post thrombotic syndrome: long term complication of DVT
-  Paget Von Schroetter Syndrome: DVT but for the upper limb (is rare).
-  Well's score: risk of DVT
-  Thrombus has one of four outcomes
    1.  Propagation towards heart
    2.  Embolisation
    3.  Dissolution through fibrinolysis: e.g. tPA = tissue plasminogen activator
    4.  Organisation + recanalisation: ingrowth of smooth muscle, fibroblasts, and endothelium into thrombus. Makes vessels in thrombus allowing blood to flow through again.

# Cholesterol

-  LPL activated by apoC2 which makes them smaller.
-  Bile acid binding resins can $\Downarrow$[Cholesterol] along with the statins (HMG-CoA reductase inhibitors). Bile acid resins also work such as cholestyramine.
-  Steps of cholesterol biosynthesis
    1.  Acetyl-CoA $\rightarrow$ HMG CoA
    2.  HMG CoA $\rightarrow$ melanovate by HMG CoA reductase
    3.  Mevalonate produces intermediates
    4.  Cholesterol synthesis
-  HDL carries cholesterol away from the tissues to the liver.

# Path

## Atherosclerosis

Morphological features:

1.  Irregular, yellow to white, raised deposits called 'plaques'
2.  Multiple, widely distributed and coalescing atherosclerotic plaques
3.  Areas of ulceration of plaques, with thrombosis

Pathogenesis flow

1.  Endothelial injury
2.  Platelet adhesion
3.  Accumulation of lipoproteins in the vessel wall.
4.  Recruitment and adhesion of monocytes and their differentiation to macrophages
5.  Lipid accumulation within the macrophages - Foam cells
6.  Smooth muscle cell recruitment
7.  Smooth muscle cell proliferation and release of extracellular matrix

Likely pathological event

-  Impaired renal fn: progressive stenosis
-  Thrombotic stroke: plaque rupture
-  AMI: plaque rupture

Complications of atherosclerotic plaque

-  **Plaque rupture**: Pro-coagulant molecules within the plaque are exposed to circulating blood $\rightarrow$ lead to formation of a thrombus and infarction of the supplied tissue $\rightarrow$ Unstable angina, myocardial infarction, thrombotic stroke
-  **Haemorrhage**: An intramural haematoma can form secondary to plaque rupture $\rightarrow$ Myocardial infarction, thrombotic stroke
-  **Embolisation** of the plaque fragments: Disrupted fragments from the atherosclerotic plaque can embolize to distal vascular sites $\rightarrow$ Embolic stroke (and not thrombotic stroke!)
-  **Aneurysm**: the fibrous plaque increases pressure in the vessel media and can provoke weakening of elastic tissue, with subsequent dilatation of the artery $\rightarrow$ Abdominal aortic aneurysm

Developed plaque formed from: fibrosis of vascular intima and *low* grade chronic inflammation around lipid deposits.

# CBL

**SOCRATES is a better for emergencies**

## Main case

### ACS

Overall disease until we decide whether it is:

-  Ischaemia - lack of O~2~ which is reversible. **ST depression is ischaemia**.  
`CP <- "chest pain"`; `OE <- "on exertion"`
    -  **stable angina** (CP OE)
    -  **unstable angina** (CP $\pm$ OE): caused due to *dynamic* obstruction of one or more coronary arteries.
-  Infarction - necrosis due to lack of O~2~
    -  **NSTEMI** (ST normal, troponin +) - minor troponin leak from a more minor infarct
    -  **STEMI** (ST elevation, troponin doesn't matter) - full thickness/transmural infarct

If diagnosed with (N)STEMI then straight to cath lab for PTCA (= percutaneous transluminal coronary angioplasty = STENT). CABG = coronary artery bypass graft.

### Oxygen mask

Uncomplicated MI - nope, but if there is evidence of shock, dyspnoea then give O~2~.

### DDx for chest pain

-  Cardiac
-  GIT
-  Respiratory
-  Musculoskeletal
-  CNS

## Short case I

Right leg pain 3 days after plane flight

-  DDx for leg pain: DVT, claudication (ischaemia to the lower extremities), muscle cramps, myopathy, myocitis (inflammation of muscle (not)infective + (not)auto-immune), trauma, bone diseases (osteosarcoma, metabolic bone disease), CNS (sciatic neuropathy = pain in the sciatic nerve AND below the knee), infection in the skin (cellulitis more likely on just one leg), varicose veins, inflammation of the lymph system, bakers cyst (cyst in the popliteal fossa which may burst)

In some occlusions the warm blood from deep veins is shunted to the superficial veins making the superficial veins WARM which is not inflammation.

D-Dimer is a fibrin degradation product. During blood clot fibrinolysis (degradation) occurs which D-Dimer forms as a product. However D-Dimer is present in a range of conditions so it is not specific. Diseases associated wtih elevated D-Dimer: thrombosis anywhere, blood transfusion, pregnancy (pro-thrombotic state), cancers,

**Pre-test probability** = probability of diagnosis before the results from the test come back. So if they both show the same thing then it is likely. Sorts out false positives and negatives.

**Sensitivity and specificity** = *Sp In Sn Out meaning* that specificity rules things in, sensitivity rules things out. Sensitivity detects that something is wrong.

Phrm

-  Antiplatelets - stop aggregation - aspirin, clopidogrel, ticagrelor
-  Anticoagulants - heparin, warfarin, clexane
    -  Warfarin - vit k reductase inhibitors and vit k is necessary for (2, 7, 9, 10) [australian tv channels]. Lots of variation in metabolism so people have to keep on getting blood tests each week to find the necessary amount. Is reversible so can quickly remove it's effects which is faster than NOACs
    -  Novel oral anti-coagulants (**NOACs**) - thrombin antagonists (which converts fibronigen $\rightarrow$ fibrin). {Dabigatran, rivaroxaban, apixaban}. These are orals, but chronic renal failure, *less intra-cranial haemorrage, more GIT haemorrage*.

## Short Case II

SMOKING IS A RISK FACTOR FOR VASCULAR DISEASES

Vasculopath - high risk for vascular diseases

Temperature

-  Arterial insufficiency - cold as blood is WARM
-  Venous insufficiency - warm as blood is shunted from deep veins to superficial veins

## PVD

-  Ankle/brachial index measurement is a screen for PVD
-  Drugs used are cilastazol amd naftidrofuryl
-  PAD $\subset$ PVD

### 2017-06-02 CVS4.rmd
added old notes

---
title: "CVS 4"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Arrythmias

## Bradycardias

Caused by **2** things:

1.  Reduced automaticity of SA node. Can occur during periods of increased vagal tone (sleep, carotid sinus massage), and secondary to drugs ($\beta$-blockers, Ca channel blockers).
2.  Blocked conduction. AV block has 3 degrees = {first = abnormally long AV conduction time (PR interval > 0.22s), second = some atrial impulses are conducted to ventricles (skips a QRS once in a while), third = no association between atria and ventricular activities}.

Often results in need for implantable device such as pacemaker.

## Tachycardias

Caused by **3** basic mechanisms:

1.  Increased automaticity arising from more rapid phase 4 depolarisation.
2.  Repolarisation is delayed (longer plateau phase) can trigger spontaneous depolarisations caused by reactivation of Na or Ca channels.
3.  Most commonly, tachycardias from a reentrant circuit. Any condition that gives rise to parallel by electrically separate regions with different conduction velocities (e.g. border zone of MI, accessory AV connection)
    -  Atrioventricular reentrant tachycardia (AVRT): accessory pathway such as WPW
    -  AV nodal reentrant circuit (AVNRT): there is a slow pathway and fast pathway in the AV node. Premature beat may come at a time when fast is refractory but slow isn't $\rightarrow$ causes reentrant circuit inside AV node leading to SVT.

Management depends on **QRS complex width**. If:

-  Narrow: then ventricular depol is occuring normally and arrhythmia is arising at or above AV node (supraventricular).
-  Wide: ventricular conduction is $\neg$ normal. Either from ventricular tissue or SVT with aberrant conduction over His-Purkinje system or an accessory pathway.

# Atrial fibrilation

Most common sustained arrhythmia worldwide. High risk of **stroke**. Four categories: Paroxysmal (< 48 hours), persistent (>7 days), permanent (>1 year), lone (without evidence of structural heart disease).

CHADS score gives risk for major AF-related thromboembolism. **CHADS** = {Cardiac failure, hypertension, age, diabetes mellitus, previous stroke}. If CHADS $\geq$ 2: long term oral anticoagulant therapy recommended (warfarin). This is as risk of GIT and intracranial haemorrhage from warfarin is outweighed by risk of stroke. **CHA~2~DS~2~-VASc** = {Congestive heart failure, hypertension, age (>75=2), diabetes mellitus, stroke Hx = 2, vascular disease, age (65-74 = 1), sex category (female = 1)}.

Novel anticoagulants: using direct thrombin inhibitors. Thrombin converts fibrinogen $\rightarrow$ fibrin, factor Xa also a acceptable target as one molecule of fXa generates more than 1000 thrombin molecules. Thrombin inhibition leads to reduced fibrin formation, reduces thrombin generation, and inhibits platelet activation. Direct thrombin inhibitor also can inactivate fibrin-bound as well as free thrombin in thrombus expansion. Direct thrombin inhibitor = dabigatran (twice-daily). Not used in valvular AF and if patient has mechanical heart valves. fXa inhibitor = {rivaroxaban (once-daily), apixaban (twice-daily)}. However these might be more expensive, no routine coagulation assay available, and no antidote for anticoagulation. Not recommended for those with renal disease.

HAS-BLED: risk factors for bleeding in people with AF. \textbf{HAS-BLED} = {Hypertension, Abnormal renal or liver function, Stroke (Hx), Bleeding (Hx or predisp), Labile INRs, Elderly (age >65 years), Drugs (antiplatelets, NSAIDs, alcohol)}

-  Rate control: if pt has persistent AF (e.g. 1 week). These drugs are used to control rapid ventricular HR.
-  Rhythm control: using anti-arrhythmic drugs to get back into sinus rhythm: used if pt is highly symptomatic.
-  Post anticoagulation give *combination of digoxin, atenolol, verapamil*

# Quiz

-  Arrhythmia with prolonged QT segment is Torsade de pointes. Often caused by antipsychotic medications (which prolong QT interval)
-  Delta wave is wolfe parkinson white syndrome
-  WARFARIN TO AF.
-  Elevated JVP can suggest obstructive shock. Post pacemaker insertion if it happens then complication in cardiac perforation can lead to cardiac tamponade (build up in pressure in the pericardial sack). Infection would NOT lead to elevated JVP
-  CXR: L and R main pulmonary arteries form hilar shadows (not bronchi). Abnormal cardiothoracic ratio > 50% is cardiomegaly. Curve of the aortic arch above the L hilum is called the aortic knucle.

ECG:

-  1mv signal moves stylus 10mm vertically
-  QRS represents V *depolarisation* whereas ST segment is when the ventricles are *depoalrised*
-  Lead II represents potential diff between L leg and R arm

Respiratory sinus arrhythmia is normal and HR increases on inspiration.

Cardiac tissue with slowest rhythmicity is ventricular muscle NOT purkinje fibres.

Digoxin inhibits Na-K ATPase indirectly reducing the inward transport of sodium ions and outward exchange of calcium ions in myocytes.

Triggered activity: abnormal action potentials triggered by preceding AP. Seen as spontaneous APs during phase III or IV (*afterdepolarisations*) and usually occurs in situations where there is an abnormally long AP (such as long QT syndrome).

JVP should decrease on inspiration. If it *increases*: kussmauls sign.

# CBL

# Short Case 2: Slow and steady

### Trigger 1

**What could it be?**:

-  High prob: CVS related since it is CVS week = {arrhythmias, AF (especially since older people can't tell onset?), valvular issues (aortic stenosis leads to decrease blood and hence decreased perfusion to brain, also mitral regurg), CAD}
-  Serious not to be missed: cancer (leukaemia)
-  Pitfalls: anaemia
-  Masquerades: old age, she has been drugged by someone else and is going to die, dehydration (hypovolaemia), low BSL, UTI,

Think of wide DDx for presyncopal

-  Low bsl
-  Dehydration
-  UTI
-  Vestibular infection

**Further Hx**: Vitals from more recently + SOCRATES (Site, onset, character, radiations, associations, time course - does it follow a pattern, exacerbating/relieving factors, severity) + impact on daily life + what does she think it is. Ask specific stuff about if she is unwell, medications, has she collapsed and lost consciousness.

**Approach**: old lady with some risk of cardiovascular disease and she has had presyncope. Corridor thoughts: what you think when walking to pt. Get relevant Hx (+further) and check for red flags (chest pains, dyspnoeic, new onset confusion = dropped GCS, low urine output).

Ex: vitals $\rightarrow$ systems (CVS $\rightarrow$ resp $\rightarrow$ GIT $\rightarrow$ CNS) $\rightarrow$ urinalysis + ECG + spirometry + pulse oximeter + BSL + (fundoscopy)

### Trigger 2 - note that ALL heart block are defined by PR > 0.2s.

**ECG findings systematically**: Narrow then ventricular depol is occurring normally and arrhythmia is arising at or above AV node (supraventricular).

-  Name/date: assume true for (n in 1:2)
-  Rate: $\frac{300}{7.5} = 40$, P-wave rate $\approx$ 300/6 = 50/min
-  Rhythm: says sinus rhythm but there doesn't seem to be relationship between P and QRS, look at RR only??. Regular PP, RR rhythm.
-  Axis: seems legit
-  Wave morphology: no pathological q waves. No U waves (path >0.5mm deflection after T wave seen in V2 and V3 - found in hypothermia, antiarrhythmic drugs). There are some weird QRS because the P pops up.
-  Intervals: PR is variable (abnormal), QRS = 0.100s (normal/narrow), RR = 7.5*0.20 = 1.5s, QT = 500m (long).
-  Segment: ST is not elevated.

**Narrowed DDx**

-  Cardiac:
    -  Complete heart block as the AV node is not letting stuff through. Therefore ventricular escape rhythm is coming from bundle branch-purkinje system. Aetiology: MI, could be congenital but unlikely because she is so old. AV-nodal blocking drugs (e.g. calcium-channel blockers, beta-blockers, digoxin). Ischaemic heart disease.
    -  Murmur radiates to carotids is aortic stenosis.
    -  [\textbf{\textit{Cannon waves}}](https://www.youtube.com/watch?v=gMBZMQwoi_I): when the V contract the AV valve is closed, but if the A contracts at the same time then blood goes out back into the jugular. Found in arrythmias.
-  Non-cardiac: cancer, hyperkalaemia

**Mx**: Probably urgent management as she is looking quite unwell and scared of cardiogenic shock. Needs to be paced.

**Further Ix**: bloods (FBC & film, electrolytes (CHEM20), TSH, troponin). TTE (Transthoracic echo), CXR

### Trigger 3

**Haemodynamic stability**: defined with a systolic blood pressure (SBP) >90 mmHg, a heart rate <120 beats/min and without clinical signs of shock.

[\textbf{\textit{Medications}}](https://tgldcdp-tg-org-au.ezproxy.library.uq.edu.au/viewTopic?etgAccess=true&guidelinePage=Cardiovascular&topicfile=arrhythmias-acute-management&guidelineName=Cardiovascular&sectionId=toc_d1e106#toc_d1e175): is atropine as it is a negative chronotrope. For most bradycardias whilst waiting for transcutaneous pacing and specialist intervention use: 0.5-1.5 mg IV, repeated 15 mins if necessary. Atropine is muscarinic antag. Or it could be isoprenaline positive inotropic and chronotropic effects.

**Possible reversible causes**: medications.

**Complications**: decrease CO, inadequate perfusion (syncope, pre-syncope, SOB), SCD, heart failure

**Permanent pacemakers**: pacing spikes mostly visible in the unipolar leads.

-  Unipolar pacemakers where there is only one lead connecting it to the body. Used in cases where AV conduction is likely to return. E.g. used if normal AV conduction but SA disorder then paces RA.
-  Dual chamber pacemakers: have pacing electrodes in both the RA and the RV. Allows maintenance of A and V relationship - can increase rate during exercise. Dual-chamber pacemaker implantation in the RV outflow tract is good for third degree AV block.
-  Dual site A pacing: 2 A leads, 1 in RAA and other in coronary sinus.
-  Biventricular pacemakers: leads in RA RV LV. Useful in patients with heart failure who have evidence of abnormal intraventricular conduction (LBBB etc.)

-  Single chamber V pacing - RV associated with increased mortality in patients with impaired V fn.
-  Dual chamber pacing
-  Single chamber

Abnormal rhythm can be hypomagnesium/calcaemia (mg, Ca)

# Short Case 1 - VT can lead to VF $\rightarrow$ die

Extremely high tachyarrhythmias mean that decreased SV. Therefore decreased perfusion to the coronary arteries (during diastole) leading to ischaemia. ANS via adrenaline ($\beta$-agonist)

Pathological increases in HR

-  Hypovolaemic shock
-  Stress mediated by adrenaline
-  Drug use
-  Pain
-  Fever
-  Sepsis
-  Hyperthyroid state
-  PE

Hyperdynamic praecordium = heaves.

### Trigger 2

Someone goes into ED do ECG, IV access and take bloods.

Valsalva manoeuvre: increased intrathoracic pressure pushes blood away from pulmonary veins to the the LA increasing pressure on Vagus nerve. Carotid sinus massage: can cause vagus to start firing $\rightarrow$ $\Uparrow$PS.

People with Hx of MI are more likely to get MI and arrhythmias.

Peaked T waves: bigger than expected gives the earliest sign of STEMI (hyperacute). Hyperkalaemia $\rightarrow$ DO ECG (massive risk factor for VT and death).

Broad QRS Ddx: ventricular origin (VT or VF) or above + something else.

ST depression = ISCHAEMIA. Can be without athero (due to occlusion, bronchospasms, sepsis as blood pressure gets too low)

Treatment of SVT as VT is fine. VT as SVT is not fine. VT until proven otherwise usually in broad QRS.

### Trigger 3

Guy has had cardiac arrest due to rapid deterioration and lack of perfusion to the heart. Has ischaemia.

Cardiac catheterisation is where they put a catheter through vein towards aortic valve. Coronary angiography then releases radioactive dye when AV closed. This allows for real time fluoroscopy. Electrophysiology then goes in further to looks at electrical impulses.

Shock pt for pulseless VT and any VF. Only used for some haemodynamically unstable SVT. VF has no synchrony, shock for VT has to be in synchrony with the cardiac cycle to prevent R on R which is bad.

Angina is restricted to poor pipes (coronary vessels).

RBBB leads to right axis deviation and has qRSR' in V1.

BBB:

-  Right: MARROW shows M shaped thing in V1 and W in V6
-  Left: WILLIAM shows the opposite of RBB

# Conduction system of heart

SA $\rightarrow$ **internodal** $\rightarrow$ AV node

### 2017-06-02 CVS5.rmd
added old notes

---
title: "CVS 5"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/lectures/8_cvs_v/readings")
knitr::opts_chunk$set(echo = FALSE)
```

# Quiz

## Medications

-  Dobutamine: Is a $\beta_{1}$ agonist so it will increase myocyte contractility and HR.
-  Diltiazem: reduces afterload and hence myocardial work.
-  $\beta$-blockers used in HF are only: metoprolol, bisprolol, and carvedilol

## Acute Rheumatic Fever

Symptoms:

-  Subcutaneous nodules
-  Carditis
-  Polyarthritis
-  Chorea
-  **Pathognomic**: aschoff bodies

## Aschoff nodules

Aschoff nodules are composed of T lymphocytes (not B lymphocytes) and multinucleated giant cells. Not seen until the subacute or chronic phase of rheumatic carditis.

## Ventricular assist device

Given to advanced HF pt with good prognosis.

## Complications of advanced heart failure

-  Hyponatremia
-  Impaired liver fn
-  Thromboembolism
-  Arrhythmias
-  NOT \textbf{hyper}natremia

## Pericardial tap

aka pericardiocentesis: needle and catheter remove fluid from pericardium.

Pericardial fluid has an exponential pressure volume relationship.

# Readings

## CXR

```{r image, out.width="500px"}
knitr::include_graphics("cxr.png")
```

[\textbf{\textit{url}}](http://www.radiologymasterclass.co.uk/tutorials/chest/chest_pathology/chest_pathology_page8#top_1st_img)

-  Septal lines are a specific sign of interstitial oedema in suspected heart failure. If not suspicious of HF then conditions causing lymphatic obstruction can be a cause.
-  Alveolar oedema: general shadowing of the lungs. Dense airspace due to fluid filling alveoli and small airways radiating from the hilar regions.
-  LV aneurysm: it just sticks out on the side
-  Consolidation of the lingula: poorly defined L heart border due to consolidation in the adjacent lingula.

# Random path terms

-  Petechiae: pinpoint areas of haemorrhage on skin or mucous membranes reflecting bleeding from capillaries.
-  Purpura: haemorrhages of skin or mucous membranes larger than petechiae.
-  Suppuration: formation/discharge of pus
-  Empyema: pus in cavity or hollow organism
-  Coagulative necrosis (heart): 3 characteristics = {loss of nuclei, increased eosinophilia of the cytoplasm, maintenance of the remnants of the cell outline}
-  Liquefactive necrosis: complete dissolution of cells and other structures
-  Fat necrosis: seen in adipocytes as a result of trauma or release of lipases
-  Karryorrhexis: fragmentation of the nucleus seen in necrotic and apoptotic cells
-  Exudate: protein rich fluid resulting from the selective extravasation of plasma in response to changes in vascular permeability during inflammation

# POW Week 8

-  Measure someones BP and it is 160/90. Not hypertension unless it is *measured again*
-  Cardiac hypertrophy (not hyperplasia as mature myocytes cannot divide) is a response to increased BP
-  LVH response to increase in systemic BP; RVH response to increase pulmonary BP
-  Types of hypertrophy
    -  Concentric: increase pressure causes thickening of walls but lumen roughly the same
    -  Eccentric: volume overload (increase in end diastolic volume) causes increase in lumen size whilst walls may/not hypertrophy
-  Causes of cardiac hypertrophy
    -  MI: loss of some of the cardiac muscle causes remaining muscle to adapt to increased workload
    -  Hypertrophic cardiomyopathy: genetic mutation causing thickening
    -  Cor pulmonale: pulmonary arterial hypertension leading to RV failure
    -  Aortic stenosis: valvular disease causing increased pressure load on RV
-  Outcomes of LVH:
    -  Increased tissue pressure in wall, harder for blood to perfuse during diastole
    -  Increased muscle mass requires more oxygen
    -  Combination of above 2 predisposes to ischaemic conditions

# Random notes on ECG

-  AF loses *atrial kick* and is a dribbly pattern
-  AFlutter is sawtooth baseline and occurs after the formation of a reentrant circuit
-  SVT: very fast rate with narrow QRS and
-  Second degree heart block 2:1 means that every alternative P wave doesn't conduct a QRS
-  Junctional (escape) rhythm: pacemaker from AV node or bundle of His. Absent or inverted P waves in lead II. Due to not much atrial contraction there is also loss of atrial kick
-  Accelerated junctional: loses P and T waves
-  Junctional tachy: (my own words) shark tooth QRS
-  Premature ventricular complex (PVC): increase in autorhythmicity or reentry
-  If P waves abnormal but still in sinus rhythm then it is NSR with PAC (premature atrial complex)

# CBL main - waking up breathless

## Trigger 1

68 F: Dyspnoea, orthopnoea (dysponea whilst being straight), PND, panicky, easy fatigueability, reduced exercise tolerance (e.g. can't push her buggy around anymore), chest tightness

-  SOB at night
    -  PND: peripheral venous blood redistributed to lungs. Increased venous return $\rightarrow$ more blood in pulmonary circulation $\rightarrow$ increase preload $\rightarrow$ LV can't keep up pumping into systemic circulation (LV failure) $\rightarrow$ congestion in the lungs $\rightarrow$ transient oedema
    -  Difference between PND and orthopnea: **PND**: go to sleep easily but then they wake up. Often decompensated orthopnea vs **Orthopnea**: difficulty falling asleep as SOB as soon as lying down. Ask by "do you get SOB when trying to sleep"
    -  COPD, asthma, OSA
-  SOBOE: typically symptomatic of IHD (CAD is just related to state of arteries).
-  **Chest pain DDx $\Leftarrow$ MEMORISE**:
    -  Cardiovascular - pericarditis, IHD
    -  Respiratory: PE, pneumonia $\rightarrow$ pleuritis
    -  MSK: costochondritis (Tietze syndrome: palpation causes pain),
    -  GIT: gastritis $\pm$ ulceration (peptic ulcer), reflex, oesophageal spasm
    -  NS: thoracic nerve root irritation (compression of nerve root), herpes zoster (shingles)
    -  $\psi$: anxiety
-  Tiredness: infection, anxiety, depression, anaemia, cancer, IHD, vavlular disease, thyroid disease
-  Hypoxaemia $\rightarrow$ sympathetic drive $\rightarrow$ panic attack (/disorder: repeated unpredictable panic attacks or spending lots of time about the impending panic attacks. Need to treat AND educate patients)
-  What do next? Hx (HPC,  systems review, meds, surgical Hx, family Hx, Allergies) $\rightarrow$ Ex $\rightarrow$ blah blah

## Trigger II

-  Acute on chronic: chronic disease that suddenly becomes more acute
-  Chest discomfort/**anything**: cardiac until proven otherwise - do ECG. Something that can't be missed.
-  Narrow Dx to cardiovascular: due to family Hx of vascular disease, hypertension. Probs not respiratory: has no palpitations, fever, cough or wheeze, does not smoke.
-  Calcium channel blockers
    -  Dihydropyridines: amlodipine, nifedipine $\Leftarrow$ ACTS PERIPHERALLY
    -  Non-dihydropyridines: verapamil & diltiazem $\Leftarrow$ acts centrally. Slows conduction through the AV node, negative inotropic and chronotropic activity.
-  Potential Dx: heart failure $\rightarrow$ crackles of lung bases, pitting oedema, elevated JVP, displaced apex beat, S3 (increased V filling pressures), might feel heaves
-  Pulsatile liver is pathognomic of tricuspid regurg

## Trigger III - where did her HF come from? Cardiomyopathy

-  Turbulent flow in the femoral arteries: stenosis, atherosclerosis
-  Bibasal crackles: APO $\rightarrow$ oedema fluid could be pus (pneumonia), blood, interstitial fluid. Lots of things that cause it, but in this pt other symptoms suggests oedema fluid
-  AV nipping and silver wiring: signs of hypertensive retinopathy (signs of hypertension in the eye)
-  She needs treatment but not that urgent as sats O~2~ is not that bad
-  Ix: ECG, bloods (FBC, CHEM20 - kidney(could lead to fluid retention $\rightarrow$ overload)/liver(late sign)/electrolytes(retention of Na due to RAAS leads to water retention + hyper/hypo-kalaemia lead to arrhythmias), **BNP** (peptide released from ventricles; use if not sure about HF: high = HF), troponin, lipid profile), CXR, echo
-  Hypercholesterolaemia $\subset$ Dyslipidaemia (too much or too little anything)

## Trigger IV

**ECG**:

-  Right axis deviation suggests RBBB
-  Path Q waves in V1, V2, V3: previous anterior MI
-  Lead III: the notching in the QRS complex suggests BBB
-  S waves normal in I and II
-  PR interval long = first degree heart block
-  Broad QRS: either ventricular or atrial with aberrant conduction through AV node (this is more likely since she has BBB)
-  V1: T wave is the inversion and sort of cuts the S wave off (change in gradient)

**Blood test results**:  Has dyslipidaemia. Thyroid fn measured as TH affects HR and contractility by the SNS. BNP is a specific test: can help see if treatment is working.

**CXR** signs of HF: may not see all 5, but quick to identify

-  A: alveolar oedema (bat wings) - cloudiness near the hilum
-  B: kerley B lines: septal thickening
-  C: Cardomegaly (Cardio thoracic ratio)
-  D: diverted upper vasculature: blood gets shunted to the top of the lungs since there is too much fluid. Basically see what you usually see inferior, see it superior
-  E: effusion: blunting of the costophrenic angles due to fluid build up

Pansystolic murmur means not surprising that there is tricuspid regurg (radiates to axilla?)

Echo gives valves and LV EF

Silent MI: most common in diabetics, can happen in women

## Trigger V

-  All Ca2+ channel blockers are contraindicated in HF as it decreases contractility of the heart.
-  Frusemide keeps Na and loses K, so give with K supplement (Span K)
-  Don't start medications that decrease work of the heart until *euvolaemic* - oedema is gone.
-  ACE inhibitor: dry cough due to bradykinin. Decreases BP and also helps with cardiac myocyte remodelling after an infarct.
-  Aspirin: low dose acts as antiplatelet agent by working on ADP.

## Trigger VI

-  Myocardial perfusion scan: uses radioactive contrast and get stressed through drugs (usually dobutamine which is a sympathetimimetic)
-  Multivessel disease means it is good for CABG. If just a single vessel then it is not worth
-  Stress testing is to examine perfusion. Usually stress echo $\rightarrow$ something not moving $\rightarrow$ angiogram
-  PTCA = percutaneous transluminal coronary angioplasty is an alternative to the heart surgery
-  Bypass surgery can result in death, haemorraghe, clotting, infn
-  LV failure can lead to RV failure leading to congestive HF.

[\textbf{\textit{Good ecg website}}](http://www.skillstat.com/tools/ecg-simulator)

# Midsem quizzes

## CVS anat and physiology

-  If SA node fails then AV node takes over. Internodal pathways don't do anything. 
-  Long saphenous vein is *superficial*
-  Aortic valve has 3 cusps; right coronary, left coronary, and *posterior*/non-coronary cusps.
-  LAD is most prone to obstruction
-  Paradoxical embolism: taken from venous system to arterial (such as through patent foramen ovale)
-  VSD is a L$\rightarrow$R shunt; tricuspid atresia is a R$\rightarrow$L shunt - an ASD *and* VSD must be present for blood to be pumped due to the RV atrophy. Hence it is a shunt.
-  Azygous vein is on the *right* and *posterior* thoracic wall
-  Opthalmic artery is a branch of the *internal carotid*
-  **$\Uparrow$Pulse pressure caused by:
    -  $\Downarrow$aortic compliance
    -  $\Uparrow$stroke volume
    -  Hypertension

## Embryo

-  Just after endocardial tubes have fused this is the order of blood flow: sinus venosus, primordial atrium, primordial ventricle, bulbus cordis, aortic sac
-  Development of foramen ovale: common atrium starts by growing septum primum downwards $\rightarrow$ ostium primum randomly forms and disappears $\rightarrow$ above an ostium secundum forms $\rightarrow$ septum secundum forms which is slightly offset forming a one-way valve.
    -  Tl;dr Septum primum, ostium secundum, foramen ovale, septum secundum
-  Ductus arteriosus closes due to
    -  Increased arterial oxygenation
    -  Decreased pulmonary vascular resistance
    -  $\Downarrow$Prostaglandin E~2~

## Reperfusion injury

-  O~2~ is incompletely reduced by damaged mt $\rightarrow$ $\Uparrow$O~2~ free radicals
-  Influx of activated leukocytes increases inflamm damage

# Short case 1

## Trigger 1

-  Cardiac pain (ischaemic) is not usually sharp (instead crushing, heavy, dull).
-  Chest pain to ignore MI: 3 P's = {palpable, pleuritic (breathing makes it worse), positional (pain changes on changing position)}
-  Positional chest pain likely pericarditis: pericarditis is inflammation of pericardium, moving around causes friction which irritates the already inflammed pericardium, gravity puts pressure on the pericardium.
-  Chest pain is usually urgent: DDx = cardiovascular, resp, neuro, musculoskeletal
-  What other info is needed: SOCRATES, PMHx (could be used to rule out oesophageal reflux,...), PSHx, medications, allergies

## Trigger 2

-  ECG
    -  Widespread ST elevation can be found in pericarditis
    -  Look for deep S waves in V1, V2
-  Pulsus paradoxus: drop in pressure during inspiration where you'd usually expect an increase due to increased venous return.
    -  Pericardial effusion is fluid around the heart
    -  Pericardium filled with fluid is known as tamponade.
-  Becks triad to Dx tamponade: hypotension, raised JVP, soft heart sounds
-  Ix: bloods (FBC for anaemic, CRP/ESR, troponin, cultures, CHEM20 especially because have to check renal/kidney fn before giving them medication, coagulation panel in case he is just bleeding into it, BNP in case HF)
-  Pericardial tap to drain the fluid

## Trigger 3

-  CXR: round heart is a sign of pericardial effusion
-  3 layers of the heart: endocardium, myocardium, epicardium
-  Troponin leak from the myocytes adjacent to the epicardium because of inflammation
-  BNP raised: decreased aortic BP (hypotension) causes $\Uparrow$BNP
-  ISBAR = {identify, situation, background, assessment, recommendation}
    -  "On Ex"
    -  "Ix shows...typical balloon sign of pericarditis. In addition, echo performed shows 2cm pericardial fluid" $\leftarrow$ which is the most important thing rather than just LVH etc.

## Trigger 4

-  Myopathy: diseased myocardium
-  Viral myocarditis $\rightarrow$ $\Uparrow$troponin $\rightarrow$ pericardial effusion
-  Hyperaemia: increased blood supply
-  Oedema: increased interstitial fluid
-  Rheumatic heart disease: Group A Strep pharyngitis $\rightarrow$ autoimmune response
    -  Jones criteria

# Short case 2

## Trigger 1

-  SOB at rest: NYHA 4
-  Tricuspid regurg $\rightarrow$ pulsatile liver
-  Lung disease $\rightarrow$ pulmonary hypertension $\rightarrow$ RHF/heart dysfn
    -  known as *cor pulmonale* (heart disease due to lung pathology)
-  Aspects of physical Ex and why: CVS Ex: focus on the heart and lower limbs.
    -  Elevated JVP
    -  Palpation of the chest: parasternal heaves for [L,R]VH, lateral displaced apex beat due to LVH
    -  Auscultation: valvulopathy as a cause of this, extra heart sounds (S3 due to HF)
    -  Lower limb Ex: oedema (pitting)
    -  lung auscultation(?): bibasal crackles suggestive of fluid from____(pneumonia)

## Trigger 2

-  Ex for work of breathing
-  Loud P2 from pulmonary hypertension
-  Ix:
    -  ABG
    -  Bloods (FBC for [Hb] to *exclude anaemia*, CHEM20 ensure kidney+liver fn before giving drugs, COAG panel looking for thrombosis, ESR/CRP: see if infn is causing),
    -  ECG: hypertrophy of ventricles in HF
    -  CXR: signs of HF, ABCDE={alveolar oedema, kerley \textbf{b}-lines, cardiomegaly, dilated upper vessels, pleural \textbf{e}ffusion}}
    -  Echo (EF)

## Trigger 3

-  Haematocrit is volume percentage of RBCs in the blood.
-  Slightly elevated red cell count (RCC) and haematocrit is a compensatory mechanism as a result of chronic hypoxaemia
-  COPD with high CO~2~ (CO~2~ retainers): the chemoceptor changes from peripheral CO~2~ detector changes to central receptors which are *dependant on O~2~*. Therefore giving O~2~ will decrease respiratory drive $\rightarrow$ death.
-  ECG
    -  Right axis deviation: high V1
    -  Inverted T-waves normal in V1 (upside down), whereas not normal in V2, V3 due to RV strain (ischaemia)
-  Mx
    -  Decrease peripheral oedema: *frusemide* to get the fluid off the lungs (alveolar oedema) and legs $\rightarrow$ euvolaemic
    -  Oxygen therapy: contraindicated with smoking
    -  Pulmonary and cardiac rehab that improves the muscles

### 2017-06-02 I&I.rmd
added old notes

---
title: "I&I 1 & 2"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
---

```{r setup, include=FALSE}
setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/lectures/3_i&i_ii/readings")
```

# Bacteria

-  Gram(-) rod: *E. coli*
-  Gram(-) coccus (round): *Neisseria meningiditis*
-  Gram(-) rod after burn with green and foul smelling exudate: *Pseudomonas aeruginosa*
-  Gram(+) bacili can form *spores* in order to resist environmental factors

# Drugs

-  Efficacy: maximal response
-  Potency: how fast it starts working
-  In the presence of competitive antagonist, the agonist dose-response curve is:  shifted to the right without changing slope or maximum effect
-  Selective toxicity is NOT: Clavulanic acid acting as a target for beta-lactamase

## Receptor

-  Desensitisation: decrease in potency (curve shifts right) usually due to partial uncoupling of receptor from second messenger
-  Down-regulation: decrease in efficacy (maximal effect) usually due to receptor internalisation
-  Not often seen for drugs with spare receptors

## Anti-bacterials

-  Interfere with folic acid synthesis so DNA synthesis cannot occur: sulfonamides (e.g. trimethoprim)
-  $\beta$-lactams: attach to penicillin-binding proteins on bacteria and interfere with peptidoglycan (cell wall) synthesis. Also cause lysis
    -  E.g. Penicillins, cephalosporins, glycopeptides (vancomycin)
-  Interfering with bacterial protein synthesis: 
    -  50S: macrolides (erythromycin)
    -  30S: tetracyclines (used for intracellular bacteria such as Chlamydia),  aminoglycosides (streptomycin, gentamicin, neomycin against gram(-) aerobes)
-  Affecting topoisomerase: quinolones

## Antiviral

Gancyclovir is against CMV. Acyclovir prevents nucleic acid synthesis. Enfuvirtade used in HIV: inhibits viral cell penetration/fusion.

## Anti-protozoan

Malaria $\in$ {Protozoa}. Therefore quinine and chloroquinine are used to target the schizont stage. Metronidazole inhibits nucleic acid synthesis and is also antibacterial.

## Anti-helminth

Its location means that Combantrin (non-spec depolariser) is used with poor absorption through GIT.

## Anti-fungals

Fluconazole and clotrimazole inhibit ergosterol production used against Candidiasis. Amphotericin used for life threatening (probably deep mycoses).

# Anti inflamm pharm

-  SAIDS: glucocorticoids indirectly inhibit phospholipase A by upregulating TXN of annexin-1
-  NSAIDS: inhibit COX which prevents synthesis of PG~E2~ which is a pyrogen. $\therefore$ they are anti-{inflammatory, pyretic} and analgesic. E.g. aspirin, ibuprofen, celecoxib, 
-  Paracetemol: NOT an NSAID. MOA unknown.

# Immunology stuff

\textbf{M}ost \textbf{D}octors \textbf{G}et \textbf{E}xtra \textbf{A}ss - IgA is in **secreted fluids**

\textbf{Hot} \textbf{T}-\textbf{B}one st\textbf{EAK}

-  IL-1: increases temperature (fever) making it an endogenous pyrogen. It is secreted from *macrophages* and also increases adhesion factors on the endothelial wall during inflammation.
-  IL-2: T cell stimulator
-  IL-3: B cell stimulator
-  IL-4: IgE (class switching from IgG) and B cell growth. Secreted from T~H~2 cells.
-  IL-5: IgA and eosinophils. Also secreted from T~H~2 cells.

T~H~1 secrete: (leading to macrophage mediated chronic inflammatory response)

-  IFN-$\gamma$: MOST SIGNIFICANT activator of macrophages along with IL-2(?)
-  TNF-$\alpha$: helps macrophages and is important in autoimmune disorders
-  IL-2: T cell stimulator

T~H~2 secrete: (main effectors are eosinophils, basophils, mast cells, and B cells)

-  IL-4: IgE + B-cell growth
-  IL-5: IgA
-  IL-6: Ac(K)ute phase protein synthesis
-  Supports B cell antibody responses

T~H~17: secretes proinflamm cytokines such as IL-22 which stimulate production of bacteriocidal and fungicidal proteins by epithelial cells.

Types of hypersensitivities: **ACID**

-  Type I: \textbf{A}naphylactic and atopic: mediated by basophils and mast cells
-  Type II: \textbf{C}ytotoxic (antibody mediated)
-  Type III: \textbf{I}mmune complex
-  Type IV: \textbf{D}elayed (Cell-mediated)

Fever

-  Temperature is raised by pyrogens: exogenous and endogenous. Exogenous includes LPS from gram(-) bacteria. This triggers response leading to IL-1 secretion which migrates to the hypothalamus signalling for increase in body temperature.
-  Types: Intermittent (returns to normal and goes back), remittent (varies but never goes to normal),
-  Pyrogens: IL-1, IL-6, TNF-$\alpha$

PAMPs

-  LPS is found in gram(-) bacteria which is detected by TLR-4
-  Lipoteichoic acid from gram(+), peptidoglycan, viral nucleic acid (such as dsRNA) are detected by TLR-3.
-  Microorganism binding to PAMP-R leads to cytokine activation.

Random notes

-  B cell receptor is bivalent whilst TCR is NOT
-  Most important ion for determining RMP is K^+^
-  Volume of distribution is most important in determining loading dose of a drug
-  Neutrophils are necessary for umbilical cord separation.
-  NK cells are important *early* and cytotoxic t cells (CD8) are important *late* in the course of a viral infection.
-  During acute inflammation, blood vessels in the affected region of the body dilate leading to an overall increase in blood flow: most likely to be present is Increased adhesion of leukocytes to endothelial cells

Activation of complement leads to:

-  **Phagocytosis**: by opsonising antigens. C3b is most important here
-  **Inflammation**: attracts macrophages and neutrophils
-  **Membrane attack**: directly rupturing cell membrane of foreign pathogens
-  Does NOT activate Th cells

Macrophages:

-  Phagocytose bacteria
-  Present antigen on MHCII molecules
-  Recruit neutrophils to a site of infection
-  Responsible for controlling release of cytokines
-  Reside in tissues ($\neg$ in lymph nodes)
-  Different locations are known by different names
    -  Microglial in brain
    -  Kupffer in liver
    -  Histiocytes in connective tissue
    -  Osteoclasts in bone

Professional antigen presenting cells: constitutively activate MHC II

-  Macrophages
-  Dendritic cells
-  B lymphocytes

T-cell receptor: a surface bound receptor that is never secreted. Binds to processed antigens. CD3 sits adjacent to the alpha-beta complex.

Release of histamine and TNF is most associated with mast cells/basophils.

**Polymorphonuclear leukocytes** = {neutrophils, basophils, eosinophils}

3 fn of antibodies: ADCC, opsonisation, neutralisaton, (mast cell activation too?!)

cAMP is a second messenger that has the effect of activating protein A

Coupling of G-protein alpha subunit to an agonist occupied receptors initially leads to exchange of bound GDP with GTP

# GIFTs

## Arachidonic acid

```{r tol, echo=FALSE, out.width="500px"}
knitr::include_graphics("aa_pathway.png")
```

Phospholipids from cell membranes (when broken it is released into the cell triggering these pathways). Arachidonate metabolites $\in$ {Eicosanoids}. Glucocorticoids are steroids inhibiting COX-2 and includes cortisol, cortisone, prednisolone.

# CBL

## Drugs

### Acne

-  \textbf{Benzoyl peroxide} (topical) - applied topically is bactericidal against *P. acnes*.
-  \textbf{Doxycycline} (oral) $\in$ {Tetracyclines} which are broad spectrum antibiotics. Inhibits TLN by binding to bacterial 30S subunit.
-  \textbf{Roaccutane} (oral **isotretinoin** is also known as 13-cis-retinoic acid) $\in$ {Retinoid}. It is used as a treatment for severe acne however there are many side effects whilst it is a teratogen so must not get pregnant. Suppresses all the four aetiological factors responsible for acne

### Candidiasis

-  Fluconazole (oral) $\in$ {azole}. Azoles are antifungals that inhibit fungal cell membrane synthesis. Fluconazole inhibits lanosterol-C14-demethylase which inhibits lanosterol $\rightarrow$ ergosterol production. More expensive but only one oral dose is required.
-  Clotrimazole (topical) $\in$ {azole}. Inhibits ergosterol production as well. Is cheaper but it has to be administered for a week intra-vaginally.

### Transplant

-  Immunosuppresive therapy: Cyclosporin A, prednisolone, azathioprine $\in$ {immunsuppressive medication} given during RA, and kidney transplants. It gets incorporated into replicating DNA thus halting replication and it also inhibits purine synthesis. $\therefore$ it affects quickly proliferating cells such as B and T lymphocytes)
-  Ganciclovir is used to treat CMV. Inihibits viral DNA pol by being a competitive inhibitor of dGTP and is also a poor substrate for chain elongation.

### Vaccinations lul

## Pathogens

-  \textit{Aspergillus fumigatus}: infects immunocompromised hosts
-  *Staphyloccocus aureus*
-  *Candida albicans*

# Immunodeficiency diseases

**Cytomegaly virus (CMV)**

-  Part of herpes genus so virus remains latent and can reactivate if immune system is depressed
-  Common in AIDS patients, cancer patients, and organ transplant recipients
-  CMV has been found in most body fluids and thus can be spread in a variety of ways

**HIV**

-  Low CD4 T cell count, not undergone immunosuppressive therapy, not previously diagnosed with congenital immune system diseases, and has one or more AIDS defining conditions.
    -  CMV which disseminates. Usually just found in transplant patients.
    -  Tb is the leading cause of death in HIV(+) patients
    -  Pneumonia due to *Pneumocystis jiroveci* as this is a fungus and doesn't cause pneumonia in healthy individuals.
    -  Candidiasis which disseminates outside the usual locations of vagina and oral cavity.
    -  Cryptococcosis caused by yeast.
-  Saquinavir and ritonavir are antiretrovirals used to prevent and inhibit HIV. They are protease inhibitors.
-  Enfuvirtade inhibits viral cell penetration.

**Severe combined immunodeficiency (SCID)**

-  Diarrhoea, failure to thrive, and life-threatening infection

**Chronic granulomatous disease**

-  Phagocytes (neutrophils and macrophages) use a respiratory burst to destroy pathogens such as bacteria. This uses a ROS such as superoxide.
-  CGD - the phagocytes can't produce ROS meaning they can't kill bacteria. So patients have weakened immune system as their phagocytes are useless.
-  Leads to formation of granulomas in organs as they cannot phagocytose things.

**X-linked agammaglobulinaemia**

-  are more susceptible to pneumococcal infections

Selective IgA deficiency is the most common primary immune deficiency.

# Pathology

## Types of cells

-  Labile
-  Stable
-  Permanent: hypertrophy cannot hyperplasia

## Cellular changes

-  Atrophy: both $\Downarrow$Number and $\Downarrow$Size
-  Reversible: cellular swelling, fatty change, hyalinisation
-  Irreversible: rupture, mt damage, nuclear changes (e.g. karryohexis)

## Chronic inflamm

-  Tissue injury and repair co-exist
-  Mononuclear cells predominate: macro, lympho, plasma cells

## Granulomatous inflammation

-  \textit{Sometimes it happens that way, hyung.}
-  Things too big to be phagocytosed. E.g. Myobacterium tuberculosis, leprosy,...
-  T~H~1 detect antigen and secrete: IFN-$\gamma$ activating macrophages (epitheloid) leading to granuloma formation

## Wound healing

-  Regeneration: dead cells replaced by alive ones of same type (e.g. liver)
-  Repair: scar formation
-  Tissue repair: granulation tissue + wound healing
    -  Granulation tissue formation = {inflamm, clearance, ingrowth of granulation tissue}
    -  Wound healing = {first intention, second intention}
        1.  First intention: wounds with opposed edges (e.g. surgical incision). Not much fibrosis, not much scarring, scant inflamm, healing time is short. 
        2.  Second intention: fucked wound. Lots of fibrosis, scarring, extensive inflammation, takes a while.

### 2017-06-07 renal1.rmd
updated a couple of lines

@@
 -  minimal change (children; bowmans space epithelial cells loses its negative charge, so not much morphological change)
--  Membranous glomerulonephritis:
+-  Membranous glomerulonephritis: *sub-epithelial deposits* of immune complexes (type III hypersensitivity)
 -  Focal segmental glomerulosclerosis (FSGS): adults

### 2017-06-07 renal4.rmd
updated a couple of lines

@@
 
+-  Learn bladder Ca
+    -  Smoking is major risk factor. Can present as painless haematuria
+-  Major mech of stress incontinence: *urethral hypermobility*
+-  *Proteus mirabilis* is a gram(-) rod that leads to highly alkaline urine
+-  Medication associated with acute urinary retention: decongenstant
+-  Enuresis: bedwetting during sleep in child >6
+-  Pharm Mx of BPH: $\alpha$-adrenergic receptor antagonists
 -  Clavulanic acid contains beta-lactam, gets preferentially broken down by resistant bacteria. Therefore penicillin and cephalosporins can have an effect.
@@
     -  Also sensory N AFFERENT in detrusor that is stimulated when bladder is stretched
+-  Calculi:
+    -  Struvite likely to be from infn
 

### 2017-06-08 renal4.rmd
fixed some typos

@@
 -  Calculi:
-    -  Struvite likely to be from infn
+    -  Struvite likely to be from infn (such as *Proteus*)
 
@@
 -  Nociceptive (pain due to nociceptors detecting tissue damage.
-Cna be blocked with NSAIDS, opiods; sharp, burning) vs neuropathic (irritated/diseased nerves e.g.
+Can be blocked with NSAIDS, opiods; sharp, burning) vs neuropathic (irritated/diseased nerves e.g.
 peripheral/autoimmune/diabetic neuropathy or obstructions; described as tingling, burning, pins and needles)
@@
 
-## Trogger 2
+## Trigger 2
 
@@
 
-Risks of cathetirsation: urosepsis (septicaemia due to infection in the genitourinary system)
+Risks of catheterisation: urosepsis (septicaemia due to infection in the genitourinary system)
 
@@
 
-Medications interfering with micturition such as NSAIDS, decongestants for URTI
+Medications interfering with micturition such as NSAIDS, decongestants for URTI.
 

### 2017-07-03 msk1.rmd
msk1

```
title: MSK I
author: Aaron Khoo
date: `r format(Sys.Date(%d %B %Y))`
```

# Readings

## Describing a fracture

1.  Describe the film: who what when where why
2.  What type of fracture? Complete = {transverse, oblique, spiral, comminuted}, Incomplete = {bowing, bucle, greenstick}, Salter-Harris
3.  Where is fracture = {diaphysis, metaphysis, epiphysis}
4.  Is it displaced? {angulation, translation, rotation}
5.  Something else happening too?

## Trauma

### Primary survey

-  Circulation (including haemorrhage control)
-  Airway Maintenance (cervical spine control)
-  Breathing and Ventilation
-  Circulation (reassess again - this is IMPORTANT)
-  Disability (check neurological status)
-  Exposure (undress patient completely - strip and flip, examining both sides of the patient) - cover as soon as practical to prevent hypothermia.

### Ix

-  FBC - Hb, Hct, Platelets
-  ABG with Lactate  - check for ventilatory failure and marker of shock (lactate)
-  Coag's - evidence of early trauma associated coagulopathy.
-  Blood alcohol
-  EUC
-  Group and Hold +/- X-Match as indicated.
-  ECG
-  CXR, PXR
-  eFAST (extended FAST ultrasound scan)
-  CT scans as indicated (generally in the obtunded major trauma patient a panscan is performed if the patient is haemodynamically stable)
-  Urinalysis

### 2017-07-03 msk1.rmd
Finished reading, added to be completed

@@
 -  Urinalysis
+
+## Principles of fracture Mx
+
+### Initial
+
+Maintain/improve:
+
+-  Blood supply
+-  Soft tissue health
+
+Open fractures
+
+-  Prevent infn: antibiotics immediately
+-  Allow fracture to heal
+-  Restore fn in the limb
+
+### Non-operative therapy
+
+-  Casting: contraindicated if = {no displacement, displacement exists but not relevant to fn outcome, reduction impossible, reduction cannot be maintained, fracture produced by traction forces}
+-  Traction = {skin, skeletal}
+
+### Surgical therapy
+
+-  Kirschner wires
+-  Plates = {buttress, compression, neutralisation, tension-band, bridge} and screws
+-  External fixation: used with open fractures that have significant soft tissue disruption

### 2017-07-04 msk1.rmd
Finished cbl

@@
 date: `r format(Sys.Date(%d %B %Y))`
+output: pdf_document
 ```
@@
 -  Breathing and Ventilation
--  Circulation (reassess again - this is IMPORTANT)
+-  Circulation (reassess again - this is IMPORTANT; check for pulses, bleeding, cyanosis,...)
 -  Disability (check neurological status)
@@
 -  EUC
--  Group and Hold +/- X-Match as indicated.
+-  Group and Hold +/- X-Match as indicated. (Checks blood type)
 -  ECG
@@
 -  External fixation: used with open fractures that have significant soft tissue disruption
+
+# CBL Main case
+
+## Trigger 1
+
+58M 2 story fall, injuries right leg and arm, denies head or neck injury.
+
+Immediate Mx: primary survey (see above).
+
+Further info from pt:
+
+-  HPC: What caused him to fall such as seizure or slip? How did he fall? Land on anything? Remember the whole thing?
+-  Site, onset, character, radiation, associations, time course, exacerbating/relieving factors, severity (in-depth HPC later).
+-  See if anything else is wrong. Distracting injury could obfuscate.
+-  Who is next of kin?
+
+After primary survey => secondary survey
+
+-  Head to toe Ex
+-  Look for fractures
+-  Pulses, sensation, pain: neurovascularly intact
+-  Auscultate for breath sounds, heart sounds
+-  Palpate
+-  FAST ultrasound: see if there is something bleeding
+-  Log roll to turn them over and look for same stuff
+    -  Push down c-spine to check if there is any pain
+
+## Trigger 2
+
+Pt tachycardic when he came in and hypotensive. Pulse gone down for a bit probably due to morphine getting rid of pain.
+
+Imaging indicated: X-rays of right upper and lower limb. Can be difficult to assess for spinal injuries so it might be alright to image spine.
+
+## Trigger 3
+
+Fractures:
+
+-  Complete spiral fracture to distal diaphysis R tibia
+-  Comminuted fracture proximal diaphysis R fibula
+-  Compound comminuted (through the skin) middle diaphysis right ulna
+    -  Note: have to be methodical because if the ulna is fucked then the radius could be too
+
+Fracture Mx:
+
+-  Non-operative: closed reduction (pulling the bone back into alignment)
+-  Operative: open reduction
+
+Complications from open fracture:
+
+-  Infn
+-  Embolism
+
+## Trigger 4
+
+DDx post-op pain: 5W's of post-op fever = {wind [pneumonia], water [UTI], wound [infns especially after day 5], wonderdrugs [iatrogenic], walking [walking reduces DVT]}
+
+-  Compartment syndrome
+-  DVT
+-  Infn
+-  Nerve palsy
+
+On Ex for new leg pain:
+
+-  Inspection: signs on inflammation, cyanosis
+-  Palpate: neurovascular, loss/decrease pulses, temperature, cap refill (comparing both sides)
+-  Neuro assessment
+
+## Trigger 5
+
+Compartment system features:
+
+-  Parasthesia
+-  Tenderness
+-  Increased pain from passive stretching (e.g. toes)

### 2017-07-05 msk1.rmd
Added sc2

@@
 -  Increased pain from passive stretching (e.g. toes)
+
+# Short case 2
+
+## Trigger 1
+
+> GP in Sarina (popn 6,000). Called by nurse working night shift. 72M brought in by *ambulance* found on bathroom floor. COuldn't get up, pain left hip, wife along.
+
+Further info:
+
+-  HPC: What caused him to fall such as seizure or slip? How did he fall? Land on anything? Remember the whole thing?
+-  SOCRATES: site, onset, character, radiation, associations, time course, exacerbating/relieving factors, severity (in-depth HPC later).
+-  See if anything else is wrong. Distracting injury could obfuscate.
+-  Who is next of kin?
+
+DDx: why can't pt get off ground
+
+-  MSK: fracture, muscular issues
+-  GIT: constipation
+-  CVS: shock, MI, stroke
+-  Resp:
+-  Neuro: nerve palsy, parkinsons?, lots of pain
+-  Depression? Drugs? Alcohol
+
+DDx: hip pain
+
+-  Extrinsic
+    -  Abdo/pelvic causes = {appendicitis, ovarian pathology, bowel pathology, endometriosis}.
+    -  Local pathology such as inguinal/femoral herniae, lymph nodes.
+    -  Spinal or sacroiliac referred pain.
+    -  Meralgia paraesthetica (entrapment of lateral cutaneous nerve) or sciatica
+    -  Malignancy
+-  Intrinsic
+    -  Soft tissue = {tendonitis, bursitis, snapping hip [movement of ITB over greater trochanter]}
+    -  Joint = {osteoarthritis, rheumatoid, septic, synovitis}
+    -  Bone = {hip/femur fracture, avascular necrosis}
+
+Risk factors for falls in the elderly
+
+-  MSK: muscle weakness, pain in feet
+-  Neuro: balance, parkinsons, poor vision
+-  CVS: hypotension
+-  External: clothes, external hazards, medications
+
+## Trigger 2
+
+> Found before 5am lying on floor, crying in pain, unable to get up.
+No PMHx of chest pain, palpitations, "fits, faints or funny turns" [pre-syncope/syncope?].
+PMHx = {osteoarthritis, HTN, insomnia, hearing loss, BPH}.
+Meds = {atenolol [$\beta-1$ blocker; used for HTN], oxazepam [benzodiazepine; used for anxiety, insomnia, and alcohol withdrawal], panadol osteo [paracetamol; relief for osteoarthritis]}.
+Said he slipped on bath mat after waking up whilst going to the toilet.
+Falling $\rightarrow$ severe pain in left hip, felt unable to stand.
+
+Specific physical Ex:
+
+-  General inspection
+-  Vitals
+-  Cardio/resp Ex
+-  Secondary survey?
+-  Lower limb pulses
+
+Hip Ex?
+
+-  Inspect: deformities, swelling, scars, wasting
+-  Posture: spinal, leg, foot
+-  Gait: Trendelenburg sign or crossed adductor gait suggests advanced hip disease
+-  Feel: swellings (herniae)
+-  Move: flex, extend, int/ext rotate, ab/ad-duct
+-  Fabre test: tries to reproduce pain to localise?
+-  Impingement (fadir) test
+    -  Ante: joint path, tendonitis, bursitis
+    -  Post: sacroiliac or lumbar
+
+Priorities: Mx severe pain in left hip and get him walking again. Urgent: probs not
+
+PMHx and meds: osteoarthritis, HTN, atenolol, oxazepam (muscle relaxant)
+
+## Trigger 3
+
+> Man in pain.
+Given analgesia.
+Hypotensive, nil CVS/resp, no abnormalities other than left hip painful and limb appears shortened and externally rotated.
+Pedal pulses palpable bilaterally, sensation present bilaterally, dorsiflexion and plantar bilaterally.
+Too painful to proceed further.
+Hip X-ray ordered + FBC and eLFT.
+2.5mg IV morphine, pt is kept nil by mouth(?)
+
+Shortening
+
+-  Proximal to trochanter: loss of articular cartilage from infn or arthritis, dislocation
+-  Distal to trochanter: fractures of femur, tibia
+-  Lengthening of one limb: increased vascularity stimulating growth (such as long bone fracture/neoplasia)
+
+Mx priorities:
+
+-  Pain
+-  Fix fracture
+-  Prevent fracturing again
+
+Secondary survey: focused Hx and physical Ex. Assumed that life-threatening issues have been found and corrected. Includes Hx taking and vitals
+
+Pt is at risk of hypovolaemia if it is bleeding.
+
+Ix:
+
+-  Imaging: X-ray to view what is happening
+-  FBC: check for anaemia?
+-  CHEM20
+
+## Trigger 4
+
+> X-ray hip shows sub-capital fracture of the femur.
+Decreased bone density of the hip and femur.
+Normal biochemistry and haematology.
+Transferred to Mackay hospital where succesful surgery for fractured neck of femur.
+Bone mineral density scan confirms demineralisation of long bones and vertebrae with significant loss of trabeculae bone consistant with primary osteoporosis.
+
+A pathologic fracture is a bone fracture caused by disease that led to weakness of the bone structure.
+E.g.osteoporosis, Ca, infn, inherited bone disorders,....
+Fragility fracture $\in$ pathologic fracture, occurs as a result of normal activities such as fall from standing height or less.
+
+Bone mineral density (looks at calcium):
+
+-  $-2.5<\text{osteopaenia}\leq-1$
+-  $\text{osteoporosis}\leq-2.5$
+
+Osteoporosis Mx:
+
+-  Pharmacologic: bisphosphanates (alendronate), calcitonin, denosumab (RANKL antag)
+-  Vertebroplasty and kyphoplasty: for pt with persistent and severe focal back pain due to vertebral collapse
+-  Dietary measures: adequate calcium and vit D
+-  Physical therapy
+
+Blood supply: avascular necrosis and non-union are common in fractures of the neck of femur (sub-capital) without surgical intervention, because the blood supply is interrupted.
+Rare after intertrochanteric fractures.
+
+Prognosis: varied, but is complicated by advanced age, as hip fractures increase the risk of death and major morbidity in the elderly. QoL may be decreased due to mobility issues.
+
+Safely discharged:

### 2017-07-06 msk1.rmd
added sc

@@
 ```
-title: MSK I
-author: Aaron Khoo
-date: `r format(Sys.Date(%d %B %Y))`
+title: "MSK I"
+author: "Aaron Khoo"
+date: "`r format(Sys.Date(), '%d %B %Y')`"
 output: pdf_document
@@
 
+# Short case 1
+
+## Trigger 1
+
+> 43F bilateral shin pain, worse on L. Worsening past 2-3/12. Back from 1/12 training deployment army involving vigorous training.
+
+Deep fibula N supplies anterior compartment of lower limb.
+
+Pain origin: referred pain (sciatica can radiate to lower leg) [o/e could be no pain in the isolated area, but palpating the spine can demonstrate], dodgy gait, could be claudication, varicose veins, arthritis
+
+To think of DDx: VINDICATE = {vascular, inflamm, neoplastic, degenerative/deficiency, idiopathic, congenital, autoimmune/allergic, traumatic, endocrine}
+
+Questions to exclude red flags:
+
+-  Have you been feeling tired and fatigued (Cancer)
+-  Pain at rest or nocturnal, enough to disturb sleep
+-  Weight loss, night sweats, or chills (Ca)
+-  Hx of malignancy
+-  Neuro compromise
+-  Drugs or steroids (osteoporosis)
+-  Immunosuppressed (infn; osteomyelitis)
+
+Bone structure from exercise: Wolffs law, bones laid down where it is needed.
+
+Hx questions:
+
+-  HPC: SOCRATES, what do you think it is, why did you come in *today* as opposed to last week or yesterday,...
+-  PMHx/PSHx
+-  Medications (also gives indication for PMHx)/allergies/drugs
+-  Fam Hx
+-  Repro Hx
+-  Social Hx: lifestyle, food, exercise, smoking
+
+Menopause is a risk factor for osteoporosis.
+
+MSK Ex: look $\rightarrow$ feel $\rightarrow$ move whilst comparing both sides
+
+## Trigger 2
+
+> Pain before deploying, has gradually worsened and now occurs with lighter exercise. Present at night.
+
+Ex: Medial tibial stress syndrome has no swelling
+
+Pathophysiology processes: fractures, necrosis
+
+## Trigger 3
+
+> Normal X-ray. She gets MRI
+
 # Short case 2
@@
 
-> GP in Sarina (popn 6,000). Called by nurse working night shift. 72M brought in by *ambulance* found on bathroom floor. COuldn't get up, pain left hip, wife along.
+> GP in Sarina (popn 6,000). Called by nurse working night shift. 72M brought in by *ambulance* found on bathroom floor. Couldn't get up, pain left hip, wife along.
 
@@
 -  SOCRATES: site, onset, character, radiation, associations, time course, exacerbating/relieving factors, severity (in-depth HPC later).
+-  Did you hit head
+-  Remember everything, how did you fall?
 -  See if anything else is wrong. Distracting injury could obfuscate.
 -  Who is next of kin?
+-  Social Hx to see how healthy he is
 
@@
 -  Neuro: nerve palsy, parkinsons?, lots of pain
--  Depression? Drugs? Alcohol
+-  Depression? Drugs diuretics can make pt get up at night? Alcohol
 
@@
 -  Extrinsic
-    -  Abdo/pelvic causes = {appendicitis, ovarian pathology, bowel pathology, endometriosis}.
-    -  Local pathology such as inguinal/femoral herniae, lymph nodes.
-    -  Spinal or sacroiliac referred pain.
+    -  Abdo/pelvic causes = {appendicitis, ovarian pathology, bowel pathology, endometriosis}
+    -  Local pathology such as inguinal/femoral herniae, lymph nodes
+    -  Spinal or sacroiliac referred pain
     -  Meralgia paraesthetica (entrapment of lateral cutaneous nerve) or sciatica
@@
 -  MSK: muscle weakness, pain in feet
--  Neuro: balance, parkinsons, poor vision
+-  Neuro: balance, parkinsons, poor vision, inner ear pathology
 -  CVS: hypotension
--  External: clothes, external hazards, medications
+-  Renal: urinary incontinence can make people rush to the bathroom with increased risk of falling
+-  External: clothes, external hazards, medications (polypharmacy)
 
@@
 
-Priorities: Mx severe pain in left hip and get him walking again. Urgent: probs not
+Priorities: Mx severe pain in left hip and vitals. Urgent: probs not
 
@@
 -  CHEM20
+-  ECG
+-  Group and hold
+-  COAGs
 
@@
 -  Vertebroplasty and kyphoplasty: for pt with persistent and severe focal back pain due to vertebral collapse
--  Dietary measures: adequate calcium and vit D
--  Physical therapy
+-  Dietary measures: adequate calcium and vit D, smoking and alcohol
+-  Physical therapy: weight bearing exercise
+-  Hormone therapy
 

### 2017-07-06 histo_template.rmd
Added histo template script

---
title: "Histo of ____"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
  -  \usetheme{metropolis}
output:
  beamer_presentation:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/pracs/histo/week_")
# rename 's/test-this/REPLACESTRING/g' *
```

```{r image_loop2, results = 'asis'}
dickpics <- gtools::mixedsort(list.files("images/"))
for(i in 1:length(dickpics)){
  cat("### ", strsplit(dickpics[i], "\\.")[[1]][1], "\n", '\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{"', file.path("images", strsplit(dickpics[i], "\\.")[[1]][1]), '"}.PNG}', "\n", "\n", sep = "")
}
```

### 2017-07-10 msk2.rmd
added msk2

```
title: "MSK II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
```

# Bone

## Physiology

-  Matrix composed of 35% osteoid (mainly type I collagen) and 65% mineral component (mainly hydroxyapatite).
-  2 types of matrix = {woven, lamellar}
-  Cells = {osteoblasts, osteoclasts, osteocytes}
-  Development:
    -  Long bones from endochondral ossification
    -  Flat bones from intramembranous ossification
-  Remodelling:
    -  RANKL: stimulates RANK => increased osteoclast differentiation (stimulated by PTH)
    -  OPG: inhibits RANKL RANK signalling => decreased osteoclast differentiation

## Pathology

### Congenital

-  Dysostoses: defects in migration and condensation of mesenchyme
-  Dysplasia: global disorganisation of bone/cartilage
    -  Achondroplasia: most common AD disorder from g.o.f *FGFR3* mutation => retarded cartilage growth
    -  Thanatophoric dysplasia: lethal dwarfism also g.o.f *FGFR3*
    -  Osteogenesis imperfecta: mutation in *type I collagen* => dominant negative lof mutation => too little bone => fragility
    -  Osteopetrosis: reduced bone resorption resulting from impaired formation or fn of osteoclasts. AR mutations in *CA2* or *TCIRG1*

### Metabolic

-  Osteopenia
-  Osteoporosis: age related, reduced physical activity, genetic factors, calcium nutritional state, hormonal influences (e.g. estrogen deficiency).  
Clinical course can lead to lordosis/kyphoscoliosis/PE/pneumonia.
-  Rickets and osteomalacia: manifestations of vit D deficiency or abnormal metabolism

### Paget's

Paget disease is condition of increased but disordered and structurally unsound bone. Both genetic and environmental causes (usually *SQSTM1*). Hallmark is mosaic pattern of lamellar bone.

1.  Initial osteolytic stage
2.  Mixed osteoclastic/osteoblastic stage => predominance of osteoblastic activity
3.  Burned out quiescent osteosclerotic phase

### Fracture healing

-  Rupture of vessles results in haematoma surrounding area of bone injury
-  Clotted blood provides fibrin mesh, sealing off fracture site, and providing scaffold for inflamm cells/fibroblasts/angiogenesis.
-  Degradnulated platelets and inflamm cells release PDGF, TGF-$\beta$, FGF to activate osteoprogenitor cells.
-  End of first week: mass of uncalcified tissue called *soft callus* provides anchorage between ends of fractured bone.
-  After two weeks: soft callus => bony callus due to deposition of woven bone, hyaline/fibro-cartilage, and endochondral ossification.

### Osteomyelitis

Inflamm of bone and marrow, usually secondary to infn.

-  Pyogenic: usually bacteria, rarely fungi. Normally *Staph aureus*
    1.  Haematogenous spread
    2.  Extension from contiguous site
    3.  Direct implantation after compound fractures
-  Mycobacterial osteomyelitis

### Bone tumours

| Category          | Behavior  | Tumor Type                    | Common Locations                           | Age (yr) | Morphology                                                                                               |
|-------------------|-----------|-------------------------------|--------------------------------------------|----------|----------------------------------------------------------------------------------------------------------|
| Cartilage forming | Benign    | Osteochondroma                | Metaphysis of long bones                   | 10–30    | Bony excrescence with cartilage cap                                                                      |
| —                 | —         | Chondroma                     | Small bones of hands and feet              | 30–50    | Circumscribed hyaline cartilage nodule in medulla                                                        |
| —                 | Malignant | Chondrosarcoma (conventional) | Pelvis, shoulder                           | 40–60    | Extends from medulla through cortex into soft tissue, chondrocytes with increased cellularity and atypia |
| Bone forming      | Benign    | Osteoid osteoma               | Metaphysis of long bones                   | 10–20    | Cortical, interlacing microtrabeculae of woven bone                                                      |
| —                 | —         | Osteoblastoma                 | Vertebral column                           | 10–20    | Posterior elements of vertebra, histology similar to osteoid osteoma                                     |
| —                 | Malignant | Osteosarcoma                  | Metaphysis of distal femur, proximal tibia | 10–20    | Extends from medulla to lift periosteum, malignant cells producing woven bone                            |
| Unknown origin    | Benign    | Giant cell tumor              | Epiphysis of long bones                    | 20–40    | Destroys medulla and cortex, sheets of osteoclasts                                                       |
| —                 | —         | Aneurysmal bone cyst          | Proximal tibia, distal femur, vertebra     | 10–20    | Vertebral body, hemorrhagic spaces separated by cellular, fibrous septae                                 |
| —                 | Malignant | Ewing sarcoma                 | Diaphysis of long bones                    | 10–20    | Sheets of primitive small round cells                                                                    |

## Joints

-  Solid (nonsynovial: synarthroses): structural integrity
    -  Fibrous synarthroses: cranial structures and bonds between roots of teeth and jawbone
    -  Cartilaginous synarthroses (synchondroses): symphyses between sternum and ribs
-  Cavitated (synovial joints): have joint space for wide range of movement. Lined by type A synoviocytes (specialised macros with phago activity) and type B synoviocytes (sim to fibroblasts and synthesise hyaluronic acid). Hyaline cartilage serves as elastic shock absorber.

### Arthritis

-  Osteoarthritis: degeneration of articular cartilage leading to structural and fn failure of synovial joints.
-  Rheumatoid arthritis: chronic inflamm of autoimmune origin attacking the joints
-  Juvenile idiopathic arthritis: affects before age 16
-  Seronegative spondylarthropathies:
    -  Absence of rheumatoid factor
    -  Path changes in ligamentous attachments rather than synovium
    -  Involvement of sacroiliac joints $\pm$ others
    -  Association with HLA-B27
    -  Bony proliferation leading to ankylosis
-  Infectious arthritis
-  Suppurative arthritis: bacterial infns that cause acute suppurative usually enter joints from distant sites by haematogenous spread. Restricted ROM
-  Lyme disease: systemic infn by *Borella burgdorferi*
-  Crystal induced arthritis:
    -  Gout: urate crystal deposition in/around joints.
    -   Calcium Pyrophosphate Crystal Deposition Disease (Pseudogout)

### Tumours

Primary are rare.

-  Ganglion and Synovial Cysts: small (1–1.5 cm) cyst that is almost always located near a joint capsule or tendon sheath.
-  Tenosynovial Giant Cell Tumor

## Soft tissue tumours

Soft tissue refers to non-epithelial tissue excluding the skeleton, joints, central nervous system, hematopoietic and lymphoid tissues

### Tumours of adipose tissue

-  Lipoma: benign
-  Liposarcoma: malignant

### Fibrous tumours

-  Nodular fasciitis: self limiting fibroblastic and myofibroblastic proliferation occuring in upper extremities of young adults
-  Superificial fibromatosis: infiltrative proliferation causing local deformity but has innocuous clinical course
-  Deep fibromatosis: desmoid tumours. Large infiltrative masses that recur frequently but do not metastasise.

### Sk muscle tumours

-  Rhabdomyosarcoma: malignant mesenchymal tumour with sk muscle differentiation. Three main subtypes are recognized: alveolar [fusions of FOXO1 to PAX3/PAX7] (20%), embryonal (60%), and pleomorphic (20%).

### Smooth muscle tumour

-  Leiomyoma: benign most common in the uterus but can arise in any soft tissue site.
-  Leiomyosarcoma: They occur in adults and affect women more frequently than men.

### Tumours of uncertain origin

-  Synovial sarcoma
-  Undifferentiated pleomorphic sarcoma: ncludes malignant mesenchymal tumors with high-grade, pleomorphic cells that cannot be classified into another category by a combination of histomorphology, immunophenotype, ultrastructure, and genetics.

### 2017-07-11 histo_template.rmd
added mc, have to complete histo_template

@@
 ```
+
+```{r image_loop2, results = 'asis'}
+png_images <- gtools::mixedsort(list.files("images/"))
+for(i in 1:length(png_images)){
+    name <- strsplit(png_images[i], "\\)  # complete this here
+    cat("\begin{frame}", "\n", 
+  cat("### ", strsplit(dickpics[i], "\\.")[[1]][1], "\n", '\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{"', file.path("images", strsplit(dickpics[i], "\\.")[[1]][1]), '"}.PNG}', "\n", "\n", sep = "")
+}
+```

### 2017-07-11 msk2.rmd
added mc, have to complete histo_template

@@
 -  Synovial sarcoma
--  Undifferentiated pleomorphic sarcoma: ncludes malignant mesenchymal tumors with high-grade, pleomorphic cells that cannot be classified into another category by a combination of histomorphology, immunophenotype, ultrastructure, and genetics.
+-  Undifferentiated pleomorphic sarcoma: includes malignant mesenchymal tumors with high-grade, pleomorphic cells that cannot be classified into another category by a combination of histomorphology, immunophenotype, ultrastructure, and genetics.
+
+# Main case
+
+## Trigger 1
+
+> GP suburban, 32F, wrist pains over last few weeks, started after netball, R>L and right hand dominant
+
+Pathophysiological processes: VINDICATE
+
+-  Vascular: infarction
+-  Inflammation/infn: RA, reiters syndrome (after $\Delta$sex Hx especially: males; hurts to pee, can't see, hurt my knee), crystal arthropathies, lyme disease, septic arthritis, viral arthralgia
+-  Degenerative: OA (could be secondary), RSI, carpal tunnel syndrome (nerve entrapment) 
+-  Iatrogenic: drugs (hydrolazine)
+-  Congenital: osteogenesis imperfecta
+-  Autoimmune: RA, SLE
+-  Trauma: overuse injury, trauma [did you fall weirdly?]
+-  Endocrine:
+
+Arthritis (inflamm) vs arthralgia (pain in jts)
+
+-  Mono: 1
+-  Oligo: 2-4
+-  Poly: $\geq5$
+
+Things masquerading as joint pain: muscle, nerve
+
+Imp systemic symptoms for DDx: fever (timecourse, associated with other symptoms), malaise (might suggest high metabolic demand; chronic inflamm)
+
+-  HPC: SOCRATES + have you had before + impacting on daily life, have you been sick recently, did you fall?
+-  Systems review:
+    -  CVS: palpitations, chest pain
+    -  Neuro: fits, faints, funny, numbness, headaches, vision/hearing
+    -  Resp: SOB goes with CVS
+    -  GIT: nausea/vomiting, changes in bowel habit
+    -  Genito/urinary: frequency/pain/bleeding on urination, women = menstrual Hx
+-  PMHx, sexual Hx, FHx + SHx + travel Hx
+-  Any neurological symptoms?
+-  Any skin changes/rashes
+-  Red flags: Waking at night? Constant pain or does it ever go away? Fevers?
+-  Pt says they feel tired: are you always tired? Sleeping well? Did it come on same time as pain? Can you remember when you last fell well? (is better than "how long have you been feeling tired")
+-  Drugs/medications/smoking
+-  Social Hx
+
+## Trigger 2
+
+> Feeling very tired for few months, dull ache worse in mornings, has not woken at night, full time secretary 2 young kids, no relevant PMHx, FHX: 55 aunt has some sort of arthritis, systems review unremarkable, no other jt problems
+
+Pathophysiological process between fatigue and joint pain: infn $\Uparrow$metabolic demand, pain can be taxing and have to compensate, think of cytokines for inflamm as they can cause fatigue.
+
+Autoimmune diseases: $\text{genetic}\times\text{environment}$
+
+## Trigger 3
+
+> Inflammation around prox interphalangeal joints of 2, 3, 4.
+
+Mechanical vs inflamm:
+
+-  Mechanical: just painful when moving it, gets worse as using it during the day, wouldn't expect systemic symptom from a knee abrasion
+-  Inflamm: movement makes better, morning stiffness, systemic symptoms,
+
+Ix:
+
+-  Bloods: FBC to check white cells, CHEM20 to check for liver fn because viruses (EBV, RRV), CRP, RA factor, CCP (citric citrulinated peptide)
+-  Imaging: X-ray
+
+## Trigger 4
+
+> Ix results
+
+According to criteria she'd have 6 points => definite RA
+
+Natural Hx: progressive deformities of jt, 10% increased risk of CVS mortality within 10 years of Dx (inflamm => vasculitis/athero)
+
+## Trigger 5
+
+Biological agents: counteract body

### 2017-07-12 histo_template.rmd
Added clean histo template but unusable in rmarkdown

--- 2017-07-06 histo_template.rmd
@@
 ```
+
+```{r image_loop2, results = 'asis'}
+png_images <- gtools::mixedsort(list.files("images/"))  # Stores all the files in the dir images/
+for(i in 1:length(png_images)){
+    name <- strsplit(png_images[i], "\\.")[[1]][1]  # Splits up the string by "." and stores the first part (to get rid of .png)
+    cat("\begin{frame}", "\n", 
+  cat("### ", strsplit(dickpics[i], "\\.")[[1]][1], "\n", '\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{"', file.path("images", strsplit(dickpics[i], "\\.")[[1]][1]), '"}.PNG}', "\n", "\n", sep = "")
+}
+```

### 2017-07-12 latex_histo_template.tex
Added clean histo template but unusable in rmarkdown

\begin{frame}

\begin{frame}
\frametitle{1 vincunt}
\includegraphics[width=\linewidth, height=\textheight, imagekeepaspectratio]{images/1 vincunt.png}
\end{frame}

\begin{frame}
\frametitle{2 full stops are aids}
\includegraphics[width=\linewidth, height=\textheight, imagekeepaspectratio]{images/2 full stops are aids.png}
\end{frame}

\begin{frame}
\frametitle{image}
\includegraphics[width=\linewidth, height=\textheight, imagekeepaspectratio]{images/image.png}
\end{frame}

\begin{frame}
\frametitle{rofllmao}
\includegraphics[width=\linewidth, height=\textheight, imagekeepaspectratio]{images/rofllmao.png}
\end{frame}

\end{frame}

### 2017-07-13 histo_template_clean_leaves_extra_line.rmd
added clean histo template

---
title: "Histo of ____"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
  -  \usetheme{metropolis}
output:
  beamer_presentation:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
setwd("~/Documents/histo-test/")
#setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/pracs/histo/week_")
# rename 's/test-this/REPLACESTRING/g' *

### Requirements ###
# LaTeX engine with \usetheme{metropolis} (TeXlive-full?!)
# Directory called "images"
# ONLY image files in images that \includegraphics will allow for
# Pandoc/knitr (comes with RStudio by default)
```

```{r image_loop2, results = 'asis'}
png_image_ext <- gtools::mixedsort(list.files("images/"))
png_image <- c()
for(i in 1:length(png_image_ext)){
  png_image[i] <- strsplit(png_image_ext[i], "\\.")[[1]][1]  # removing the extension

  cat("### ", png_image[i], "\n",  # name of the image
      "\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{",
      file.path("images", png_image_ext[i]), "}}", "\n", "\n", sep = "")
}
```

### 2017-07-13 histo_template_latex_ver.rmd
added latex rmd flie that doesn't really work

---
title: "Histo of ____"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
  -  \usetheme{metropolis}
output:
  beamer_presentation:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
setwd("~/Documents/histo-test/")
# rename 's/test-this/REPLACESTRING/g' *
```

```{r image_loop, results = 'asis'}
image_name_ext <- gtools::mixedsort(list.files("images/"))  # Stores all the files in the dir images/ as a sorted character vector
image_name <- c()
for(i in 1:length(image_name_ext)){
    image_name[i] <- strsplit(image_name_ext[i], "\\.")[[1]][1]  # Splits up the string by "." and stores the first part (to get rid of file extension)

    # Outputting using LaTeX syntax to make new beamer slides, using double \\ 
    cat("\\begin{frame}", "\n", 
        "\\frametitle{", image_name[i], "}", "\n",
        "\\includegraphics[width=\\linewidth, height=\\textheight, imagekeepaspectratio]{",
        "images/", image_name_ext[i], "}", "\n",
        "\\end{frame}", "\n", "\n",
        sep = "")  # Make sure no unwanted spaces in between strings
}

# Obtaining second sub element of list sapply(strsplit(image_name_ext, "\\."), "[[", 1)
```

### 2017-07-13 histo_template.rmd
cleaned the histo_template

--- 2017-07-06 histo_template.rmd
@@
 knitr::opts_chunk$set(echo = FALSE)
-setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/pracs/histo/week_")
+setwd("~/Documents/histo-test/")
+#setwd("/home/khoo/Documents/uni_work/4_sem_1/7111/pracs/histo/week_")
 # rename 's/test-this/REPLACESTRING/g' *
+
+### Requirements ###
+# LaTeX engine with \usetheme{metropolis} (TeXlive-full?!)
+# Directory called "images"
+# ONLY image files in images that \includegraphics will allow for
+# Pandoc/knitr (comes with RStudio by default)
+
+### Notes ###
+#    \includegraphics{{"../Current folder/1.This file"}.png}
+# will not include the file name at the same time the quotations
+# The extra curly brackets are to ensure we can choose it from a directory
+# Below had to switch the ' and " around because LaTeX uses "
 ```
@@
 ```{r image_loop2, results = 'asis'}
-dickpics <- gtools::mixedsort(list.files("images/"))
-for(i in 1:length(dickpics)){
-  cat("### ", strsplit(dickpics[i], "\\.")[[1]][1], "\n", '\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{"', file.path("images", strsplit(dickpics[i], "\\.")[[1]][1]), '"}.PNG}', "\n", "\n", sep = "")
+png_image_ext <- gtools::mixedsort(list.files("images/"))
+png_image <- c()
+image_ext <- c()
+for(i in 1:length(png_image_ext)){
+  png_image[i] <- strsplit(png_image_ext[i], "\\.")[[1]][1]  # removing the extension
+  image_ext[i] <- strsplit(png_image_ext[i], "\\.")[[1]][2]  # keeping the extension
+
+  cat("### ", png_image[i], "\n",  # name of the image
+      "\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{",
+      '"images/', png_image[i], '"}.', image_ext[i], "}", "\n", "\n", sep = "")
 }

### 2017-07-13 histo_template_clean_leaves_extra_line.rmd
cleaned the histo_template

@@
 # Pandoc/knitr (comes with RStudio by default)
+
+### Notes ###
+#    \includegraphics{{"../Current folder/1.This file"}.png}
+# will not include the file name at the same time
 ```
@@
 png_image <- c()
+image_ext <- c()
 for(i in 1:length(png_image_ext)){
   png_image[i] <- strsplit(png_image_ext[i], "\\.")[[1]][1]  # removing the extension
+  image_ext[i] <- strsplit(png_image_ext[i], "\\.")[[1]][2]  # keeping the extension
 
@@
       "\\includegraphics[width=\\linewidth,height=\\textheight,keepaspectratio]{{",
-      file.path("images", png_image_ext[i]), "}}", "\n", "\n", sep = "")
+      file.path("images", png_image_ext[i]), "}.", image_ext[i], "}", "\n", "\n", sep = "")
 }

### 2017-07-13 msk2.rmd
added clean histo template
cleaned the histo_template

@@
 Biological agents: counteract body
+
+# Short case 1
+
+## Trigger 1
+
+> Michael Lee 24M back pain 4 years worse in the morning
+
+Anatomical structures:
+
+-  Tendons, ligaments, muscles
+-  Nerves, spinal cord
+-  Kidneys: stones/cyst
+-  Repro stuff iff female
+
+Questions to ask
+
+-  PC: SOCRATES: have you had any bladder weakness (neuro question)?
+
+## Trigger 2
+
+> Pain gets better on exercise.
+
+Types of back pain
+
+| Type | Character | Time course | Radiation | NSAID response | Onset | Age |
+| ------ |---------------  | ---------- | --------- | ------- |  ------ | ----- |
+| Mechanical | Thinking muscular: diffuse/achey | Worse with movement PM AM | Lumbar to back of thigh | Moderate/good | Post trauma/gradual | Mid-late |
+| Inflammation | Something pushing: stabbing | AM more likely than PM | To the butt | Really good | Insidious/gradual | 20s |
+| Radicular | Sharp/stabbing | Worse with activity | Goes down the leg towards ankle and foot | Limited | Usually sudden | Any |
+| Neuropathic | Burning | Constant, no association with activity | Towards ankle and foot | Limited | Insidious |
+| Malignant | Deep gnawing contionuous ache | Constant, wakes up at night | Limited | Insidious |
+
+## Trigger 3
+
+What are the types of spondyloarthropathies (Rheumatoid factor negative)
+
+-  Ankylosing spondylitis: $M>F$,
+-  Psoriatic arthritis: nail pitting, sausage fingers
+-  Reactive arthritis: after an infn (think UTIs, STIs), $M>F$
+-  IBD associated: 
+-  
+-
+
+Range of symptoms (don't forget extra-articular symptoms):
+
+-  Oseopenia
+-  LOTS of CVS issues
+
+## Trigger 5
+
+Natural Hx:
+
+-  Inflamm of tendons (such as calcaneal tendon)
+-  Oedema
+-  Ossification
+-  Fusion of spine => bamboo spine
+
+Treatment:
+
+-  Pain management: NSAIDs
+-  Physiotherapy: also helps manage pain whilst preventing fusion
+
+Effects:
+
+-  Depression
+-  Affect work: maybe think about re-skilling
+
+# Short case 2
+
+## Trigger 1
+
+DDx for mono-articular pain:
+
+-  Gout 
+-  Septic arthropathies
+-  Trauma
+-  Osteoarthritis
+-  Psoriatic arthritis

### 2017-07-13 msk2.rmd
deleted old

@@
 
--  Gout 
--  Septic arthropathies
+-  Gout: first MTP joint
+-  Septic arthritis: one large joint
 -  Trauma
--  Osteoarthritis
+-  Osteoarthritis: knee/hips (weight bearing)
+-  RA: little joints
 -  Psoriatic arthritis
+-  Bleeding into the joints: haemophilia
+
+Other Qs to ask
+
+-  Skin conditions: psoriatic
+-  Fever/inf: septic arthritis
+-

### 2017-07-13 msk2.rmd
updated readme and added sc

--- 2017-07-13 msk2.rmd
@@
 
--  Gout 
--  Septic arthropathies
+-  Gout: first MTP joint
+-  Septic arthritis: one large joint
 -  Trauma
--  Osteoarthritis
+-  Osteoarthritis: knee/hips (weight bearing)
+-  RA: little joints
 -  Psoriatic arthritis
+-  Bleeding into the joints: haemophilia
+-  Bite
+
+Other Qs to ask
+
+-  Skin conditions: psoriatic
+-  Fever/inf: septic arthritis
+
+Ex:
+
+-  General inspection: how she looks, gait, how she is walking
+-  Look: inflammed
+-  Touch
+-  Move
+-  Maybe CVS and look at peripheral pulses
+
+## Trigger 2
+
+Pathophysiology of gout:
+
+-  Hyperuricaemia: underexcretion (90%) or overproduction (10%)
+-  Excess urate => crystalisation => inflamm => phago of crystals => inflamm: joints get affected
+-  Can clear up in a couple of days
+
+Risk factors
+
+-  M
+-  Obesity
+-  Kidney disease
+
+Diagnostic finding in pt with gout: urate crystals
+
+## Trigger 3
+
+Mx of gout:
+
+-  Acute
+    -  NSAIDs
+    -  Prednisone
+    -  Colchicine for acute gout: interrupts cycle of mono-Na crystal deposition. Stops the inflammatory reaction to the crystals. 
+-  Chronic
+    -  Lower urate to $<0.36\text{mmol/L}$
+    -  Lifestyle modifications: low purine diet, low sugar (fructose => hyperuricaemia), stay hydrated
+    -  Allopurinol: inhibits xanthine oxidase => inhibits urate production. Not really given during acute flare up because it *may* exacerbates.
+
+Synovial fluid in following disorders:
+
+-  Normal: clear straw like colour
+-  Septic arthritis: elevated WCC, culture positive, purulent exudate
+-  Crystal deposition disease: see crystals, elevated WCC, negative culture, high urate/Ca
+-  RA: elevated WCC, sterile, normal urate, neutrophil predominance
+-  OA: joint mice
+
+## Trigger 4
+
+Long term gout damage:
+
+-  Joint damage
+-  Recurrent gout: deposits
+-  Skin nodules
+-  Extra-articular damage: 
+    -  Kidney injuries/stones
+    -  HTN
+    -  Hypercholesterolaemia/hyperuricaemia
+
+Mx in this pt:
+
+-  Check medications, maybe can't use the thiazide
+-  Lifestyle mentioned above

### 2017-07-16 msk3.rmd
added new msk week

---
title: "MSK III"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
---

# Readings

# Common paediatric hip diseases in primary care

## Developmental dysplasia of the hip

Most common in F firstborn infants. Inherited polygenic, acetabulum is shallow, and may contain pulvinar or fat.

Tests are Ortolani, Barlow, or Galeazzi signs.

## Slipped capital femoral epiphysis (SCFE)

Displacment of the capital femoral epiphysis from the rest of the femur through the cartilaginous growth plate known nas proximal femoral physis. Occurs  due to forces + acetabulum holding femoral head in place. This has sig proactive surgical methods to prevent severe hip arthritis. Risk factors = {obesity, male, african-american, pacific islander, endocrine disorders}

## Legg-Calve-Perthes disease

-  Synovitis (couple of weeks): reduced ROM
-  Osteonecrosis (6-12/12): contour of femoral head is maintained but part of it becomes ischaemic
-  Fragmentation phase (up to 3 years): pain, limping, reduced ROM
-  Revascularisation and regeneration (1-3 years): resorption of necrotic bone and dev of new bone. Permanent deformity can occur here.

# General approach to MSK

-  Inspection: SEADS = {swelling, erythema, atrophy, deformity, scars}
-  Palpation
-  ROM
-  Neurovascular status

## Treatment

-  Modification of activities
-  Ice
-  Compression
-  Elevation

# Main case

# Short case 1

# Short case 2

### 2017-07-18 msk3.rmd
added main case and added linkurl to be blue

@@
 classoption: a4paper
+urlcolor: blue
 ---
@@
 
-## Developmental dysplasia of the hip
+## Developmental dysplasia of the hip (DDH)
 
@@
 
+## Trigger 1
+
+> 32M, crossfit 4/12, sudden sharp pain in groin
+
+DDx groin pain:
+
+-  CVS: avascular necrosis, 
+-  Resp: 
+-  MSK: muscle strain, tendonitis, trochanteric bursitis, overuse injury, fracture, labral tear, ligamentous injury, 
+-  GIT: bowel obstruction
+-  Neuro: nerve impingement
+-  Genitourinary: kidney stones, testicular torsion, ovarian cyst
+-  Lymphadenopathy, cellulitis
+
+Anatomical structures referring pain: knee, back, hip, intraabdominal = {repro, kidneys}
+
+Questions to ask: also about shooting/numbness/sharp or stabbing/...
+
+## Trigger 2
+
+Catching = joint, giving way under him = ligamentous
+
+[RACGP article on hip pain](http://www.racgp.org.au/afp/2014/april/hip-pain-in-young-adults/)
+
+## Trigger 3
+
+If worried about adductor pain: can use adductor squeeze.
+
+Imaging:
+
+-  X-ray: only see bone
+-  CT: loads of radiation
+-  Ultrasound:
+-  MRI: lots of soft tissue, fractures
+
+## Trigger 4
+
+Ix:
+
+-  Bloods: FBC, 
+-  More imaging
+
+Hip impingement:
+
+-  CAM lesion: non-smooth (aspherical) femoral head: 14% in normal population.
+-  Pincer: non-smooth acetabulum
+-  Combined: both CAM and pincer
+
+Prognosis:
+
+-  Increased risk of OA
+
+## Trigger 5
+
+Arthrogram: injection of contrast followed by MRI.
+Allows intraarticular structures such as cartilage to be found.
+Risks = {anaphylaxis, infn, hitting a nerve}.
+Doubles sensitivity compared to normal MRI without arthrogram (0.60).
+
+Labral tears are common, often asymptomatic.
+
+Mx:
+
+-  Rest
+-  NSAIDs
+-  Activity modification
+-  Physiotherapy exercises
+-  Arthroscopy
+-  Debridement of bone
+
+## Trigger 6
+
+Physio exercises
+
 # Short case 1

### 2017-07-18 msk3-gift.rmd
added gift

---
title: "Imaging in MSK"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
  word_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

Info from [this presentation](https://meds.queensu.ca/central/assets/modules/msk_imaging/attach/imaging_slides.pdf).

| Type | Shortened | Relative cost (2000 Canada) |
| -------- | -------- | -------- | -------- | -------- |
| Plain radiographs | X-ray | 1 (hand) |
| Computed tomography | CT | 8 (lumbar spine) |
| Magnetic resonance imaging | MRI | 18 (knee) |
| Ultrasound | USS | 2 (knee) | 
| Bone scan | | 7? |

### 2017-07-18 msk3-gift.rmd
finished gift

@@
 
-| Type | Shortened | Relative cost (2000 Canada) |
-| -------- | -------- | -------- | -------- | -------- |
-| Plain radiographs | X-ray | 1 (hand) |
-| Computed tomography | CT | 8 (lumbar spine) |
-| Magnetic resonance imaging | MRI | 18 (knee) |
-| Ultrasound | USS | 2 (knee) | 
-| Bone scan | | 7? |
+|  Type                        |  Relative cost (Canada; 2000)  | Radiation           | What can see                                                                           | What can't one see                 |
+|------------------------------|--------------------------------|---------------------|----------------------------------------------------------------------------------------|------------------------------------|
+|  Plain radiographs           |  1 (hand)                      | Low (in comparison) | Bone                                                                                   | Soft tissues                       |
+|  Computed tomography         |  8 (lumbar spine)              | High                | Bone high detail Soft tissues Brain aneurysms Chest, abdo, pelvis                      | Soft tissues not AS good as MRI    |
+|  Magnetic resonance imaging  |  18 (knee)                     | None                | Soft tissues amazingly.
+E.g.
+ligaments, tendons, spinal cord injury, brain tumours,...
+| Bone is not as clear as xray or CT |
+|  Ultrasound                  |  2 (knee)                      | None                | Soft tissues, joint effusions Intra-abdominal organs Flow of blood Foetus              | Bones                              |
+|  Bone scan                   |  7?
+| Medium              | Metabolically active bone                                                              |                                    |
+
+## Plain radiographs
+
+Usually first line imaging as they are inexpensive, quick, specific, and readily available.
+
+Advantages:
+
+-  Inexpensive
+-  Widely available
+-  **Most specific** modality
+-  Provide excellent initial evaluation of bone and anatomical relationships. E.g. visualising a dislocation or fracture.
+
+Disadvantages:
+
+-  **Not very sensitive** ($\mathbb{P}(\text{Not having disease | Negative diagnosis})$ is low)
+-  Less sensitive to bone marrow pathology than MRI and bone scan
+-  Less sensitive to cortical bone pathology than CT
+-  Not optimal for examining soft tissue
+-  Contrast $\Uparrow$ risk of allergies/infn
+-  Radiation risk. E.g. Adolescent female with scoliosis will require frequent x-rays over a period of several years $\rightarrow$ exposes breast and reproductive tissues to radiation.
+
+## CT
+
+Often a follow up to abnormal radiograph. Rotating X-ray beam passes through slices of the body, computer reconstructs these.
+
+Advantages
+
+-  Very good bone detail and improved soft tissue resolution
+-  Takes a short amount of time (compared to MRI)
+-  Most systems can be scanned = {chest, abdomen, pelvis, brain,...}
+
+Disadvantages
+
+-  Highest **radiation risk**; uses multiple x-ray scans
+-  Requires breath holding
+
+## MRI
+
+Gold standard of soft tissue imagnig, no radiation required.
+
+Advantages
+
+-  Excellent soft tissue contrast
+-  Multi-planar
+-  No radiation
+
+Disadvantages
+
+-  Expensive and not widely available
+-  Poor bone detail
+-  Uncomfortable procedure (claustrophobic, noisy, takes time; children may need anaesthetics)
+-  Some patients cannot undergo it. E.g. have pacemaker, prosthetic valves, tattoos (?)...
+
+## USS
+
+Uses sound waves, commonly used in soft tissues, identifying joint effusions, and obstetrics.
+
+Advantages
+
+-  Highly available
+-  Real time
+-  No ionising radiation
+-  Good soft tissue contrast
+
+Disadvantages
+
+-  Operator dependent
+-  Ultrasound waves cannot penetrate bone or air
+
+## Bone scan
+
+Detects metabolically active bone.
+Good for occult(?) fractures, infns, bony mets.
+Radioactive tracer is injected IV, accumulates in area of bone with high metabolic activity, detector detects some of this radiation.
+
+Advantages
+
+-  Early detection of mets and infn
+-  High sensitivity
+-  Detects fractures not visible on xrays
+
+Disadvantages
+
+-  Moderate radiation risk $\text{xrays}<\text{bone scan}<\text{CT}$
+-  Tough to interpret in children due to presence of metabolically active growth plates
+-  Poor specificity

### 2017-07-19 msk3-gift.rmd
updated gift

@@
 Info from [this presentation](https://meds.queensu.ca/central/assets/modules/msk_imaging/attach/imaging_slides.pdf).
+More up to date information on the *cost* of these modalities is available [here](https://www.treasury.gov.au/~/media/Treasury/Consultations%20and%20Reviews/Consultations/2016/2017%20PreBudget%20submissions/Submissions/PDF/Australian%20Diagnostic%20Imaging%20Association.ashx) however it was too hard to understand so just treat cost as qualitative.
 

### 2017-07-19 msk3-gift.rmd
added more info

@@
 More up to date information on the *cost* of these modalities is available [here](https://www.treasury.gov.au/~/media/Treasury/Consultations%20and%20Reviews/Consultations/2016/2017%20PreBudget%20submissions/Submissions/PDF/Australian%20Diagnostic%20Imaging%20Association.ashx) however it was too hard to understand so just treat cost as qualitative.
+
+More info on specifics for these modalities is available from the [WA Imaging Pathways](http://imagingpathways.health.wa.gov.au/index.php/about-imaging/common-procedures).
 
@@
 Gold standard of soft tissue imagnig, no radiation required.
+The most frequent indications for MRI are for diseases of the neurological and musculo-skeletal systems.
+Other indications include tumour staging (for example rectal, prostatic and gynaecological malignancies), hepatobiliary disease, assessment of the breast in patients of high-risk for cancer, and of breast implants.
+Increasingly, MRI is substituting for CT scans particularly in children and young adults because of the lack of ionizing radiation.
+However, in Australia the indications that attract a Medicare rebate are limited.
 
@@
 
-Uses sound waves, commonly used in soft tissues, identifying joint effusions, and obstetrics.
+Uses sound waves.
+Useful for imaging the abdomen, pelvis, small parts (scrotum, thyroid, etc), breast, musculoskeletal applications, and transthoracic echocardiography.
+It is the main modality used in obstetric scanning.
+Endoluminal ultrasound applications include transrectal prostate ultrasound, transvaginal ultrasound, endoscopic ultrasound, and transoesophageal echocardiography.
+Doppler techniques are used to image vascular structures (arteries and veins) to assess flow characteristics and to detect thromboses and stenoses.
+Also useful as real-time guiding procedure for fine needle aspiration biopsy or core biopsy and drainage of fluid collections.
 

### 2017-07-20 msk3-gift.rmd
Added short case x | Typose in the gift

@@
 |  Plain radiographs           |  1 (hand)                      | Low (in comparison) | Bone                                                                                   | Soft tissues                       |
-|  Computed tomography         |  8 (lumbar spine)              | High                | Bone high detail Soft tissues Brain aneurysms Chest, abdo, pelvis                      | Soft tissues not AS good as MRI    |
+|  Computed tomography         |  8 (lumbar spine)              | High                | Bone high detail.
+Soft tissues.
+Brain aneurysms.
+Chest, abdo, pelvis                      | Soft tissues not AS good as MRI    |
 |  Magnetic resonance imaging  |  18 (knee)                     | None                | Soft tissues amazingly.
@@
 
-Gold standard of soft tissue imagnig, no radiation required.
+Gold standard of soft tissue imaging, no radiation required.
 The most frequent indications for MRI are for diseases of the neurological and musculo-skeletal systems.

### 2017-07-20 msk3.rmd
Added short case x | Typose in the gift

@@
 
+## Trigger 1
+
+Knee structures
+
+-  Bones: patella, tibia, fibula, femur
+-  Soft tissue: tendons, menisci, ligaments
+
+Contact sport: ACL (cruciates: snap and then "ow" | non-contact change in direction), lateral collateral ligament, med meniscus
+
+Questions:
+
+-  HPC
+-  Did you hit your head? Remember a snap/something? How exactly did you fall? Allergies?
+
+Bucket list for knee pain
+
+-  Acute: trauma
+-  Chronic: knee pain without trauma
+
+## Trigger 2
+
+Ligaments responding to stress
+
+-  Special tests may give *false negatives* straight away
+
+## Trigger 3
+
+Effusion: in a joint space | potential space
+
+Oedema: outside a joint
+
+Mx of ACL injury: (rupture/tear)
+
+-  Initial: RICE = {rest, ice, compression, elevation}
+-  Non-operative: physio, balance
+-  Operative: surgical repair
+
 # Short case 2
+
+## Trigger 1
+
+How are hips checked:
+
+-  Barlow: can dislocate the hip
+-  Ortolani: pull it out
+
+Galeazzi test: look at height of knee
+
+## Trigger 2
+
+## Trigger 3
+
+Tx of DDH:
+
+-  Van Rosen splint: QLD + Scandinavia
+-  Pavlik harness
+-  If instability at 10 years then surgery
+
+## Trigger 4
+
+Leg length discrepancy: dislocation (moves posteriorly and shortening)
+
+-  SUFE: 10-16
+-  Perthes: 4-10

### 2017-07-23 msk4.rmd
added msk4

---
title: "MSK IV"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Readings

## Cancers

Described as a dull and aching pain in the location. Worse at night and sometimes increases with activity. Common symptoms also fever and night sweats. Many pt do not have any symptoms but notice a painless mass.

Primaries:

1.  Multiple myeloma: tumour of bone marrow.
2.  Osteosarcoma: most common in teenagers and children. Found femur, tibia, hip, shoulder.
3.  Ewings sarcoma: age between 5-20. Common locations are upper and lower leg, pelvis, upper arm, ribs. 
4.  Chondrosarcoma: 40-70 years of age. Found hip, pelvis, shoulder.

Common sites for secondaries: breast, lung, thyroid, renal, prostate

# Main case

# Short case 1

# Short case 2

### 2017-07-24 msk4.rmd
updated path powpowpow

@@
 
+# Path POW
+
+## Arthritis
+
+If the PIP extended and DIP flexed: swan neck deformity
+
+| OA | RA | Both |
+| ------------ | ----------- | ----------- |
+| Minimal joint inflammation | Morning joint stiffness lasts >1 hour | Affects small joints of the hands |
+| Joint pain worsens after use | Presence of extra-articular manifestations  | Visible nodules on hands |
+| | Abnormal levels of autoantibodies  | | 
+| | Mostly symmetrical polyarthritis affecting young females | |
+
+Radiological
+
+-  Both: loss of jt space, malalignment of joints
+-  OA: subchondral cysts and sclerosis, formation of bony outgrowths at the end of joints
+
 # Main case

### 2017-07-27 msk4.rmd
re-added

@@
 
+## Trigger 1
+
+Ulcer: local defect of tissue or organ caused by sloughing of inflammatory or necrotic material.
+
+Pathogenesis of ulcer formation: possibly from kicking the bed, then repetitive trauma.
+
+Factors preventing ulcer from healing (learn this):
+
+-  Infn
+-  Smoking
+-  *Old* age
+-  Heavy alcohol consumption
+-  Medication: steroids, NSAIDs
+-  Obesity
+-  Repetitive trauma
+-  Diabetic: due to decreased blood flow, increased risk for infn due to high sugar, neuropathy
+-  Vascular disease (oxygen supply)
+-  Poor hygiene + improper wound care
+-  Malnutrition
+-  Super rare genetic diseases
+
+Wound present for long time:
+
+-  Can affect gait/joints
+-  Osteomyelitis
+-  Scar
+-  Could turn malignant (if wound not healing, always biopsy it)
+
 # Short case 2

### 2017-07-27 msk4-nerve-quiz.rmd
quick save

---
title: "Nerves of upper limb"
author: "Aaron"
date: "27 July 2017"
output: beamer_presentation
header-includes: 
-  \usetheme{metropolis}
---

# Kahoot link

I don't know how exactly it works

https://play.kahoot.it/#/k/c06b4397-ac92-46c8-8d9a-1d74db17dd34

# What nerve supplies trapezius?

-  Spinal part of accessory nerve
-  Nerve to trapezius
-  Long thoracic nerve
-  Medial & lateral pectoral nerves

# What nerve supplies trapezius? (ans.)

-  \alert{Spinal part of accessory nerve}
-  Nerve to trapezius
-  Long thoracic nerve
-  Medial & lateral pectoral nerves

# A lesion to which of these nerves is most likely to cause "winged scapula"?

-  Dorsal scapula nerve
-  Nerve to subclavius
-  Long thoracic nerve
-  Medial and lateral pectoral nerves

# A lesion to which of these nerves is most likely to cause "winged scapula"? (ans.)

-  Dorsal scapula nerve (rhomboids, levator scapulae)
-  Nerve to subclavius (subclavius)
-  \alert{Long thoracic nerve (serratus anterior)}
-  Medial and lateral pectoral nerves (pec major + minor)

# Which nerve does NOT supply the rotator cuff muscles?

-  Suprascapular nerve
-  Subscapular nerve
-  Axillary nerve
-  Lower subscapular nerve

# Which nerve does NOT supply the rotator cuff muscles? (ans.)

-  Suprascapular nerve (supra/infraspinatus)
-  Subscapular nerve (subscapularis)
-  Axillary nerve (teres minor)
-  \alert{Lower subscapular nerve (teres major)}

# What muscle does the thoracodorsal nerve innervate?

-  Latissimus dorsi
-  Deltoid
-  Pectoralis major/minor
-  Trapezius

# What muscle does the thoracodorsal nerve innervate? (ans.)

-  \alert{Latissimus dorsi}
-  Deltoid
-  Pectoralis major/minor
-  Trapezius

#  Coracobrachialis, biceps brachii, and brachialis are innervated by which nerve?

-  Median nerve
-  Radial nerve
-  Nerve to the *brachi*
-  Musculocutaneous nerve

#  Coracobrachialis, biceps brachii, and brachialis are innervated by which nerve? (ans.)

-  Median nerve
-  Radial nerve
-  Nerve to the *brachii*
-  \alert{Musculocutaneous nerve}

#  Which nerve supplies the posterior group of arm muscles (triceps brachii, anconeus)?

-  Radial nerve
-  Median nerve
-  Nerve to the posterior arm muscles
-  Musculocutaneous nerve

#  Which nerve supplies the posterior group of arm muscles (triceps brachii, anconeus)? (ans.)

-  \alert{Radial nerve}
-  Median nerve
-  Nerve to the posterior arm muscles
-  Musculocutaneous nerve

#  In the superficial layer of anteromedial forearm muscles, what is NOT innervated by the median nerve? (bonus: what is it innervated by?)

-  Palmaris longus
-  Flexor carpi ulnaris
-  Pronator teres
-  Flexor carpi radialis

#  In the superficial layer of anteromedial forearm muscles, what is NOT innervated by the median nerve? (bonus: what is the other innervated by?) (ans.)

-  Palmaris longus
-  \alert{Flexor carpi ulnaris}
-  Pronator teres
-  Flexor carpi radialis

# What is innervated by the ulnar nerve? (fyi: anterior interosseous muscle that supplies a couple things here is a branch of the median nerve)

-  Flexor digitorum profundus medial part
-  Flexor digitorum profundus lateral part
-  Flexor digitorum superficialis
-  Flexor pollicis longus

# What is innervated by the ulnar nerve? (fyi: anterior interosseous muscle that supplies a couple things here is a branch of the median nerve) (ans.)

-  \alert{Flexor digitorum profundus medial part}
-  Flexor digitorum profundus lateral part
-  Flexor digitorum superficialis
-  Flexor pollicis longus

# What are *all* the muscles of the posterior forearm muscles innervated by (including their branches)?

-  Radial nerve (branches to deep branch of radial nerve then posterior interosseous nerve)
-  Median nerve (branches into anterior interosseous nerve)

# What are *all* the muscles of the posterior forearm muscles innervated by (including their branches)? (ans.)

-  \alert{Radial nerve (branches to deep branch of radial nerve then posterior interosseous nerve)}
-  Median nerve (branches into anterior interosseous nerve)

#  Brachial plexus is easy - Conrad 2017. Roots > Trunks > Divisions > Cords > Branches. Which of the 5 main nerve branches only comes from the lateral cord?

-  Axillary
-  Musculocutaneous
-  Radial
-  Ulnar
-  Median

#  Brachial plexus is easy - Conrad 2017. Roots > Trunks > Divisions > Cords > Branches. Which of the 5 main nerve branches only comes from the lateral cord? (ans.)

-  Axillary (posterior)
-  \alert{Musculocutaneous}
-  Radial (posterior)
-  Ulnar (medial)
-  Median (both medial and lateral cords)

#  Which of these intrinsic thumb muscles is innervated by the deep branch of the ulnar nerve?

-  Adductor pollicis
-  Abductor pollicis brevis (these and below all by recurrent branch of the median nerve)
-  Flexor pollicis brevis
-  Opponens pollicis

#  Which of these intrinsic thumb muscles is innervated by the deep branch of the ulnar nerve? (ans.)

-  \alert{Adductor pollicis}
-  Abductor pollicis brevis (these and below all by recurrent branch of the median nerve)
-  Flexor pollicis brevis
-  Opponens pollicis

#  Which nerve innervates the intrinsic little finger (not Peter Baelysh) muscles (abductor digiti minimi, flexor digiti minimi brevis, opponens digiti minimi)?

-  Recurrent branch of median nerve
-  Deep branch of median nerve
-  Deep branch of ulnar nerve
-  Recurrent branch of ulnar nerve

#  Which nerve innervates the intrinsic little finger (not Peter Baelysh) muscles (abductor digiti minimi, flexor digiti minimi brevis, opponens digiti minimi)? (ans.)

-  Recurrent branch of median nerve
-  Deep branch of median nerve
-  \alert{Deep branch of ulnar nerve}
-  Recurrent branch of ulnar nerve

#  Lumbricals and interossei generally supplied by?

-  Deep branch of ulnar nerve except lateral 2 lumbricals?
-  Deep branch of ulnar nerve except medial 2 lumbricals?

#  Lumbricals and interossei generally supplied by? (ans.)

-  alert{Deep branch of ulnar nerve except lateral 2 lumbricals?}
-  Deep branch of ulnar nerve except medial 2 lumbricals?

# Match the "hand" and nerve lesion together

-  Median nerve lesion: claw hand
-  Ulnar nerve lesion: claw hand
-  Diego Maradona: hand of god
-  Ulnar nerve lesion: hand of benediction

# Match the "hand" and nerve lesion together

-  Median nerve lesion: claw hand
-  alert{Ulnar nerve lesion: claw hand}
-  Diego Maradona: hand of god
-  Ulnar nerve lesion: hand of benediction

### 2017-07-27 msk4.rmd
quick save

@@
 
+## Trigger 2
+
+Kids more likely for acute osteomyelitis (haematogenous route) whereas adults more likely for chronic (direct).
+
+Imaging to see osteomyelitis:
+
+-  Xray: 2 weeks after infn
+-  MRI: a couple of days
+
+Questions to ask for a possible primary cause:
+
+-  Any claudication
+-  Check for diabetes
+-  General inspection for malnutrition
+
+## Trigger 3
+
+Most common cause for osteomyelitis: *Staph*, *Pseudomonas* is stinky and common in diabetics,
+
+-  Haematogenous: staph and influenza B (common in kids)
+-  Open fractures: staph e coli
+-  Spinal: staph, strep, e coli, IV drug users = candida
+-  Diabetes/PVD: polymicrobial
+
+Pathogenesis:
+
+-  Infection
+-  Inflammatory response
+-  Phagocytosis attempt to contain infection by liberating enzymes
+-  Enzymes lyse the bone
+-  Pus forms
+-  Pus travels through Haversian system
+-  Increased intraosseous pressure
+-  End arteritis and impaired blood flow
+-  Ishaemic necrosis of bone
+-  Pus breaks through cortex
+-  Abscess
+-  Subperiosteal infection
+-  Elevated periosteum (involucrum) 
+-  Rupture of periosteum(cloacae)
+
 # Short case 2
+
+## Trigger 1
+
+Common causes of shoulder pain:
+
+-  Glenohumeral
+    -  RC muscles
+    -  Bursae
+    -  Tendons
+    -  Adhesive capsulitis
+    -  Labrum
+    -  Synovium
+    -  Articular cartilage
+    -  Articular bone surface
+-  Extraglenohumeral
+    -  MSK: OA common in AC joint, diaphragmatic irritation
+    -  CVS: MI
+    -  C spine
+    -  Malignancy: pancoast tumour (Horner's syndrome)
+    -  Herpes
+    -  PE
+    -  Depression
+    -  Polymyalgia
+
+DDx: also RA, malignancy, tendinopathy (-itis),
+
+Questions: Any particular movements that hurt? Have you started doing anything new recently? Do an ECG.
+
+## Trigger 2
+
+Hx indicative of pathology: probably acute, maybe traumatic
+
+Overhead work: puts pressure on RC muscles
+
+Abducting arms:
+
+-  10-20: supraspinatus
+-  20-90: deltoids
+-  90+: levator scapulae
+
+Smoking and shoulder pain:
+
+-  Decreased bone density
+-  Tendinopathy
+-  Risk factor for RC tears
+
+## Trigger 3

### 2017-08-01 msk4-nerve-quiz.rmd
Updated main case and typos in quiz

@@
 
--  alert{Deep branch of ulnar nerve except lateral 2 lumbricals?}
+-  \alert{Deep branch of ulnar nerve except lateral 2 lumbricals?}
 -  Deep branch of ulnar nerve except medial 2 lumbricals?
@@
 -  Median nerve lesion: claw hand
--  alert{Ulnar nerve lesion: claw hand}
+-  \alert{Ulnar nerve lesion: claw hand}
 -  Diego Maradona: hand of god

### 2017-08-01 msk4.rmd
Updated main case and typos in quiz

@@
 ## Trigger 3
+
+If glenohumeral cause then even passive ROM should be tender.
+
+Natural Hx of RC injury: little evidence for spontaneous repair. Once damaged a bit there is the same amount of force subjected to the same tendon so increased predisposition to further damage.
+
+Natural Hx of RC injury: little evidence for spontaneous repair. Once damaged a bit there is the same amount of force subjected to the same tendon so increased predisposition to further damage.
+
+Natural Hx of RC injury: little evidence for spontaneous repair. Once damaged a bit there is the same amount of force subjected to the same tendon so increased predisposition to further damage.
+
+Risk factors for RC tendinopathy
+
+-  Extrinsic: overhead repeated use, prior injuries, problems in one shoulder can affect the other, anything decreasing ability to heal
+-  Intrinsic: shoulder laxity/instability, smoking, Ca, old age
+
+Imaging necessary: not really. Unless red flags, fractures, super old age then don't need to image.
+
+## Trigger 4

### 2017-08-01 neuro1.rmd
Updated main case and typos in quiz

---
title: "Neuro I"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Readings

# Main case

## Trigger 1

Key observations:

-  Vitals
-  Oriented
-  Look at the wound: bleeding, discolouration, neurovascularly intact?
-  Any other injuries
-  Is he dangerous?

DRSABCDE: guy talking probably has patent airway and breathing. Circulation probably alright since his blood is trickling.

-  E = Exposure for secondary survey

MIST + AMPLE

M.  Mechanism of Injury
I.  Injuries Sustained
S.  Signs
T.  Treatment and Trends in the Vital Signs

A.  Allergies
M.  Medications
P.  Past Medical History
L.  Last Ate / LMP (in females)
E.  Events (This is usually the MIST)

Calming pt down:

-  "Why you so agitated?"
-  Just talk normally

Hx

-  HPC: how did you get injured?
-  Any numbness/tingling
-  What drugs are you on?
-  Which hand is dominant, when did you last eat?
-  Allergies

## Trigger 2

Info from physical

-  General inspection: mental state, conscious, oriented, body habitus, distress
-  Strength testing
-  Flexion/extension

## Trigger 3

Handover: include no signs of vascular compromise, mechanism of injury: fell through a window.

Numbness: damage to nerve which prevents sensation:

-  Compression
-  Laceration
-  Lesion
-  Demyelination
-  Ischaemia: pins and needles
-  Iatrogenic

Ix:

-  Xray: fracture/foreign body because worried about bleeding
-  FBC: check baseline haemoglobin in case he gets surgery

## Trigger 4

Delay before surgery: has to be nil-by-mouth, also theatre may be busy

Peripheral nerves can regenerate, CNS cannot. Process:

-  Needs to be realigned
-  Distal segment degrades, proximal grows through the sheath

## Trigger 5

Tinels sign: tap over nerve => if tingly/pins & needles => nerve irritation. Over time, expect the tingly feeling to go downwards more and more to show the nerve is healing.

Cock up splint to restore his finger flexion. Also so it doesn't atrophy.

# Short case 1

# Short case 2

### 2017-08-03 neuro1.rmd
added sc

@@
 
+## Trigger 1
+
+Anatomical site: bilateral is probably peripheral neuropathy.
+
+Hx:
+
+-  HPC: does it also occur in the hands? Also have to clarify if numb = tingling/weakness?
+
+Ex: where exactly does it occur? Systemic or not?
+
+DDx:
+
+-  Diabetic neuropathy
+-  Heavy metal toxicity
+-  Alcohol
+-  Nutritional disorders
+-  GBS: often feel like they're walking on cotton
+
+## Trigger 2
+
+Other ANS involvement
+
+-  Syncope/presyncope
+-  Feeling hot/sweaty
+
+Positive and negative symptoms:
+
+-  Positive: might feel burning/tingling
+-  Negative: loss of motor control, sensory, fine motor control
+
+Pathology of peripheral neuropathies:
+
+-  Axonal damage
+-  Loss of myelination
+
+Mononeuropathies:
+
+-  Carpal tunnel syndrome
+-  Tarsal tunnel
+-  Bells palsy
+-  Ulnar palsy: hitting funny bone
+
+## Trigger 3
+
+Ix: neuropathy screen (nerve conduction, biopsies, CSF (GBS)), bloods (FBC, CHEM20, antibodies (vasculitis)), fasting blood sugar
+
+Mx for diabetes:
+
+-  Diabetes: diet, metformin
+-  Treatment of pain itself: antidepressents, anticompulsants (capsaicin, gabapentin)
+-  Care of the foot:
+-  Quit smoking
+
 # Short case 2
+
+## Trigger 1
+
+Anatomical structures:
+
+-  Carpal tunnel
+-  Guillans tunnel
+-  Unlikely cerebral hemispheres
+
+## Trigger 2
+
+DDx:
+
+-  Carpal tunnel syndrome
+-  Ganglion
+-  Peripheral neuropathy
+-  Injured nerve
+
+## Trigger 3
+
+Radiculopathy: other muscular and sensory deficits
+
+Risk factors for carpal tunnel
+
+-  Repeated wrist movements with force
+-  Female: smaller carpal tunnel
+-  40-60
+-  Obesity
+
+Diagnosis for carpal tunnel: hyperflexion test. Flex for 1 min and see if symptoms increase. Median nerve slow could be carpal tunnel syndrome.
+
+## Trigger 4

### 2017-08-10 neuro2.rmd
Added neuro 2 with complete sc x

---
title: "Module II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Readings

# Main case

## Trigger 1

> 26M pipes went through window, hit neck, pain and electric shock throughout body, alert and orientated, pins and needles in hands and can't move legs, bradycardic and hypotensive, flaccid in triceps, gets worse and is intubated, indwelling catheter (IDC) is inserted.

Hypo:

-  Spinal cord injury: [this reading](http://asia-spinalinjury.org/wp-content/uploads/2016/02/International_Stds_Diagram_Worksheet.pdf) gives idea for spinal cord levels. 
-  Neurogenic shock = {bradycardia, hypotension, hypothermia} (due to loss of sympathetic tone to blood vessels; heart cannot compensate)
-  Spinal shock = is NOT any circulatory outcomes, it is just loss sensory/motor due to spinal cord injury
-  ANS dysfunction: bradycardic and hypotensive
-  Sensory dysfn: pins and needles

Classes of spinal cord injury:

-  Complete: complete loss
-  Incomplete: preservation of sensory OR spinal below level

Autonomically unstable: loss of sympathetic innervation

Breathing affected: possibly acute effects at C7 affecting C345 such as swelling | Note: always be on the lookout for breathing in spinal injuries as it can randomly deteriorate.

IDC used: patient go into acute urinary retention and also need to monitor it

## Trigger 2

Reflex:

-  GTO alerts when tendon is stretched too much. Is not involved in patellar tap reflex
-  Muscle spindle:

Shitting (both constipation)

-  UMN => increased external sphincter tone
-  LMN => bowel cannot physically move the stool along so it gets stuck

Autonomic dysreflexia: level of spinal cord injury at or above the T6 | potentially life threatening hypertension.

1.  Strong sensory input (such as bladder/bowel) go toward spinal cord.
2.  Evokes strong reflex so massive sympathetic response causing hypertension. 
3.  Brain detects through baroreceptors.
4.  Sends down descending inhibitory signals which cannot get past the lesion.

## Trigger 3

| | UMN | LMN |
| ------------ | ---------------------- | ---------------------- |
| Muscle | Clonus, spasticity | Fasciculations |
| Reflex | Hyper | Hypo |
| Bulk | Normal bulk | Atrophy of muscle |
| Tone | $\Uparrow$ | $\Downarrow$ |
| Plantar | Have (dorsiflexion; always abN) | No reflex present |
| Sensory | Associated disturbances | Localised at site |

Sensory tests:

-  Anterolateral tract: spinothalamic (pain and temp), spinoreticular, spinotectal = {pain, temperature, crude touch, firm pressure}
-  Dorsal lemniscal: check lecture 4 of this week (Bellingham)

## Trigger 4

Pressure sores: pressure on area of tissue => lack of perfusion => anaerobic metabolism => tissue ischaemia. Spinal injuries: can't feel it.

Poor mobility/altered sensation at risk of: DVT, infn (UTI), depression, obesity, erectile dysfunction, CVS (ischaemic heart disease), obstructive sleep apnoea, post traumatic syringomyelia (cyst or cavity forming in spinal cord), temperature regulation issues, muscle spasms (spasticity)

## Trigger 5

Anything above C5 is probably incompatible with independent life.

Would need to be catheterised and bowel management (enemas, manual evacuation, digital...)

# Short case 1

# Trigger 1

ACh:

-  Nicotinic
-  Muscarinic: in PS, $1000x$ more sensitive than nicotinic, so excess ACh will primarily affect PS over sym. 
    -  M2,4 = inhibitory heart
    -  M1,3,5 = excitatatory on GI
    -  Also found in sweat glands so you sweat even though it is sympathetic

ANS

| | Sympathetic | Parasympathetic |
| ------------ | --------------------------- | --------------------------- |
| Origin | Thoracolumbar region | Craniosacral region |
| Site of ganglia | Sympathetic ganglia close to spinal cord | Close to organs they are innervating (PARAPRELONG) |
| Primary NTs | Preganglionic: ACh, Post: (N)A | Pre: ACh, Post: ACh |

## Trigger 2

PS:

Muscarinic

-  D: diarrhoea/diaphoresis
-  U: urination
-  M: miosis
-  B: bradycardia, bronchoconstriction, bronchorrhoea
-  E: emesis (vomiting)
-  L: lacrimation
-  S: salivation/seizures

Nicotinic

-  Muscle cramps
-  Tachycardia
-  Weakness
-  Twitching
-  Fasciculations

High doses on muscle: fasciculations => inability to repolarise cell membranes => weakness + paralysis => ventilatory failure =>

## Trigger 3

Atropine: muscarinic antagonist. Actions would be the opposite of DUMBELS

Memorise the picture from the Eastern European lecturer.

# Short case 2

## Trigger 1

Acute back pain:

-  MSK: strain on erector spinae, infn of vertebrae, poor hip posture => functional scoliosis, spondylothesis (fracture), spondylitis, spinal stenosis, vertebral fracture, scoliosis, disc prolapse, osteomyelitis, epidural abscesses (IV drug users), OA
-  CVS: aortic aneurysm
-  GIT: pancreatitis, gall bladder infn, colon Ca, kidney stones, ischaemia intestines
-  Pregnancy stuff
-  Metabolic: cushings, acromegaly, paget disease
-  Miscellaneous: herpes zosta, sarcoidosis

Parts of spine:

-  Cervical: high mobility for the head
-  Thoracic: more rib injuries, less vertebral injuries
-  Lumbar: holds up all the weight

Red flags: fractures, Ca, osteomyelitis (recent infn, lumbar puncture, IV drug user), cauda equina syndrome, ankylosing spondylitis

-  Change in bowel habits: incontinence or acute urinary retention | indicates cauda equina syndrome
-  Night pain
-  Constant pain
-  Previous Hx of malignancy, fevers, weightloss
-  Have you been well recently, any infections (screen for immunocompromised)

Risk factors chronic lower back pain: occupation, old age, smoking, depression/anxiety, obesity, improper technique

Likely DDx: mechanical

## Trigger 2

Radicular pain (nerve compression) with straight leg raiseA

Sciatica is suggestive of something compressing the nerve.

Imaging is not usually indicated as it usually resolves, however patient has had red flag of trauma so probably should investigate.

Treatment is hard: don't want to get them hooked on endone. 1st line: simple analgesics and tell them to come back in a week. 2nd line: increased analgesia, hot packs, heat wraps, spinal manipulative therapy. Do not order imaging, avoid bed rest as have to strengthen muscles. Have to explain to pt that "there will be discomfort"

## Trigger 3

Normal fn intervertebral disk: shock absorber, vertebrae separated from each other, protect nerves

Intervertebral disk: nucleus pulposus loses viscocity, annulus fibrosis tears

Treatment:

-  Physio
-  Long term rehab
-  Surgery: fusion

6/52 usually can recover fn, but lots relapse at 12/12.

Required treatment: needs physio, possibly OT if affecting life. Explain proper dose paracetemol, must take it according to dose even if the pain isn't there.

### 2017-08-15 neuro3.rmd
Added neuro 3

---
title: "Module II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Readings

# Main case: online case with stroke

Types of speech deficits

-  Dysarthria: is a disorder of (motor) speech resulting from neurological injury of the motor component of the motor-speech
-  Dysphonia: hoarse voice
-  Dysphasia: disorder of language
    -  Expressive (Broca's): can't speak | damage in L frontal lobe reduced speech fluency with preserved comprehension. Fails to construct sentences well. Speech non-fluent.
    -  Reactive (Wernicke's): can't understand | L temporal-parietal damage leaves fluency of language but words are muddled. Cannot understand spoken or written words.

# Short case 1

# Short case 2

### 2017-08-17 neuro3.rmd
Added short cases

@@
 
+## Trigger 1
+
+> 22F sudden onset severe headache, jogging after work, max intensity instantly, sore and stiff up her neck, vomited twice, complained about brightness of light.
+
+Sudden onset symptoms:
+
+-  Trauma: haematoma
+-  Vascular: aneurysm, haemorrhage => $\Uparrow$ neck stiffness, hypertensive emergency
+-  Inflamm: meningitis (may not be infn), encephalitis, MS
+-  Malnutrition
+-  Migraine: first presentation often gets CT'd
+-  Hydrocephalus: accumulation of CSF
+-  Tumour => bleeding => mass effect
+-  Cyst
+-  Iatrogenic: medicines can cause headaches
+
+Further Hx:
+
+-  HPC: first time headache? How was she jogging, pushing herself? Any other symptoms that day or earlier on?
+-  Mental state: confusion, orientation, mental state, visual disturbances (migraine = hemi), altered sensation (parasthesia, numbness), loss of consciousness?, nausea/vomiting, midriasis (blown pupil)
+-  Medications/allergies
+-  PMHx, FHx
+-  Pregnant?
+-  Cushings triad: indicates increased ICP: hypertension,  bradycardia, widening pulse pressure
+-  Red flags: 
+    -  Systemic = fever, rash, weight loss, immunocompromised
+    -  Neuro = parasthesia, confusing, altered consciousness, meningism signs = {brudzinski's sign [passive flex neck and hip and knees reflexively flex], kernigs sign [straighten lower leg]}, seizures
+    -  Onset is new = hypertensive emergencies, acute something glycoma
+    -  Other associated conditions = headaches awakening from sleep, other drugs 
+    -  Previous Hx of headaches = is there change in character, frequency, onset
+
+## Trigger 2
+
+LOC: massive red flag, migraines don't usually cause loss of consciousness
+
+Photophobia: massive red flag
+
+DDx:
+
+-  Less likely to be stroke: pain is from meninges, she doesn't have the hemi loss of neural control
+-  Not dehydration, heat stroke
+-  Too serious to likely be migraine
+-  Hydrocephalus usually begins after something like this
+-  Likely things still being considered: ruptured aneurysm, infn [meningitis], AVM (arteriovenous malformation), HTN emergency
+
+Key features in physical Ex:
+
+-  Vitals
+-  Inspection: trauma?
+-  Eye reflex: pupils [PERRL = pupils equal round reactive to light]
+-  Cranial nerves
+-  Otoscopy: effusion, blood, perforation
+-  Fundoscopy: not likely in her since she is photophobic
+-  Neck
+
+## Trigger 3
+
+Imaging: check WA imaging pathway for thunderclap headache / headache
+
+1.  CT head: can see blood, is quick
+2.  Lumbar puncture: can see blood (if SAH), neutrophils (if meningitis)
+3.  CT Angiogram: treatable info for SAH
+
+Causes for haemorrhage:
+
+-  Trauma most common
+-  Berry aneurysm in circle of willis: most common in non traumatic
+-  AVM
+-  Tumour
+-  HTN is risk factor
+
+Veins in subdural space. Arteries in subarachnoid space.
+
+## Trigger 4
+
+Complications of SAH:
+
+-  *Repeat haemorrhages*
+-  Infn: surgical risks
+-  Brain damage: ischaemia/necrosis
+-  Hydrocephalus
+-  MRI becomes contraindicated
+-  Long term
+    -  *Seizures*
+    -  Epilepsy
+    -  Cognitive dysfn
+    -  *Increase risk of cardiac arrhythmias/MI due to damage in ANS*
+
 # Short case 2
+
+## Trigger 1
+
+> 29M falling off bike, head first into ground. Pain back of head and neck, neuro rapidly degenerates as he gets to the hospital going from responsive => unresponsive
+
+Initial Mx after falling: DRSABCD
+
+TBI: Traumatic brain injury (TBI) is a nondegenerative, noncongenital insult to the brain from an external mechanical force, possibly leading to permanent or temporary impairment of cognitive, physical, and psychosocial functions, with an associated diminished or altered state of consciousness.
+
+-  Mild: $GCS>12$, CT normal, no operative lesions, hospital stay less than 48h
+-  Moderate: $9\leq GCS\leq12$, abnormal CT, stay more than 48h, operative intracranial lesions
+-  Severe: $3\leq GCS\leq8$
+
+Learn GCS
+
+Mx of stable TBI
+
+-  DRSABCD
+-  CT head
+
+GCS 8 intubate
+
+## Trigger 2
+
+Signs of raised ICP
+
+-  Headache
+-  Neck stiffness
+-  LOC
+-  Pupils not reactive to light: optic nerve compressed
+-  Altered mental state
+-  Nausea/vomiting
+
+## Trigger 3
+
+Subdural: irregular as pressed against gyri. Extradural: crescent shape, lentiform shape.
+
+Consequences of TBI:
+
+-  Post traumatic seizures
+-  Hydrocephalus
+-  DVT
+-  Heterotopic ossification
+-  Aphasia/ataxia
+
+# Takeaway
+
+Use first aid textbook to actually learn shit.

### 2017-08-20 RESP2.rmd
sed -i 's/=>/$\rightarrow$/g' *

@@
 
-CO2 retainers: chronic increased PaCO2 => decreased sensitivity in the central chemoceptors => hypoxia becomes main breathing drive => DON'T GIVE O2
+CO2 retainers: chronic increased PaCO2 $\rightarrow$ decreased sensitivity in the central chemoceptors $\rightarrow$ hypoxia becomes main breathing drive $\rightarrow$ DON'T GIVE O2
 
@@
 
-Loss of lung tissue => decreases lung recoil => increased intrapleural pressure => early airway closure, decreased FVC, FEV1. It INCREASES PIF due to decreased recoil.
+Loss of lung tissue $\rightarrow$ decreases lung recoil $\rightarrow$ increased intrapleural pressure $\rightarrow$ early airway closure, decreased FVC, FEV1. It INCREASES PIF due to decreased recoil.
 

### 2017-08-20 msk2.rmd
Updated yaml header to make it a4. Stuff american letter paper
sed -i 's/=>/$\rightarrow$/g' *

@@
-```
+---
 title: "MSK II"
@@
 date: "`r format(Sys.Date(), '%d %B %Y')`"
-output: pdf_document
-```
+output: 
+  pdf_document
+documentclass: article
+classoption: a4paper
+---
 
@@
 -  Remodelling:
-    -  RANKL: stimulates RANK => increased osteoclast differentiation (stimulated by PTH)
-    -  OPG: inhibits RANKL RANK signalling => decreased osteoclast differentiation
+    -  RANKL: stimulates RANK $\rightarrow$ increased osteoclast differentiation (stimulated by PTH)
+    -  OPG: inhibits RANKL RANK signalling $\rightarrow$ decreased osteoclast differentiation
 
@@
 -  Dysplasia: global disorganisation of bone/cartilage
-    -  Achondroplasia: most common AD disorder from g.o.f *FGFR3* mutation => retarded cartilage growth
+    -  Achondroplasia: most common AD disorder from g.o.f *FGFR3* mutation $\rightarrow$ retarded cartilage growth
     -  Thanatophoric dysplasia: lethal dwarfism also g.o.f *FGFR3*
-    -  Osteogenesis imperfecta: mutation in *type I collagen* => dominant negative lof mutation => too little bone => fragility
+    -  Osteogenesis imperfecta: mutation in *type I collagen* $\rightarrow$ dominant negative lof mutation $\rightarrow$ too little bone $\rightarrow$ fragility
     -  Osteopetrosis: reduced bone resorption resulting from impaired formation or fn of osteoclasts. AR mutations in *CA2* or *TCIRG1*
@@
 1.  Initial osteolytic stage
-2.  Mixed osteoclastic/osteoblastic stage => predominance of osteoblastic activity
+2.  Mixed osteoclastic/osteoblastic stage $\rightarrow$ predominance of osteoblastic activity
 3.  Burned out quiescent osteosclerotic phase
@@
 -  End of first week: mass of uncalcified tissue called *soft callus* provides anchorage between ends of fractured bone.
--  After two weeks: soft callus => bony callus due to deposition of woven bone, hyaline/fibro-cartilage, and endochondral ossification.
+-  After two weeks: soft callus $\rightarrow$ bony callus due to deposition of woven bone, hyaline/fibro-cartilage, and endochondral ossification.
 
@@
 
-According to criteria she'd have 6 points => definite RA
+According to criteria she'd have 6 points $\rightarrow$ definite RA
 
-Natural Hx: progressive deformities of jt, 10% increased risk of CVS mortality within 10 years of Dx (inflamm => vasculitis/athero)
+Natural Hx: progressive deformities of jt, 10% increased risk of CVS mortality within 10 years of Dx (inflamm $\rightarrow$ vasculitis/athero)
 
@@
 -  Ossification
--  Fusion of spine => bamboo spine
+-  Fusion of spine $\rightarrow$ bamboo spine
 
@@
 -  Hyperuricaemia: underexcretion (90%) or overproduction (10%)
--  Excess urate => crystalisation => inflamm => phago of crystals => inflamm: joints get affected
+-  Excess urate $\rightarrow$ crystalisation $\rightarrow$ inflamm $\rightarrow$ phago of crystals $\rightarrow$ inflamm: joints get affected
 -  Can clear up in a couple of days
@@
     -  Lower urate to $<0.36\text{mmol/L}$
-    -  Lifestyle modifications: low purine diet, low sugar (fructose => hyperuricaemia), stay hydrated
-    -  Allopurinol: inhibits xanthine oxidase => inhibits urate production. Not really given during acute flare up because it *may* exacerbates.
+    -  Lifestyle modifications: low purine diet, low sugar (fructose $\rightarrow$ hyperuricaemia), stay hydrated
+    -  Allopurinol: inhibits xanthine oxidase $\rightarrow$ inhibits urate production. Not really given during acute flare up because it *may* exacerbates.
 

### 2017-08-20 neuro2.rmd
sed -i 's/=>/$\rightarrow$/g' *

@@
 
--  UMN => increased external sphincter tone
--  LMN => bowel cannot physically move the stool along so it gets stuck
+-  UMN $\rightarrow$ increased external sphincter tone
+-  LMN $\rightarrow$ bowel cannot physically move the stool along so it gets stuck
 
@@
 
-Pressure sores: pressure on area of tissue => lack of perfusion => anaerobic metabolism => tissue ischaemia. Spinal injuries: can't feel it.
+Pressure sores: pressure on area of tissue $\rightarrow$ lack of perfusion $\rightarrow$ anaerobic metabolism $\rightarrow$ tissue ischaemia. Spinal injuries: can't feel it.
 
@@
 
-High doses on muscle: fasciculations => inability to repolarise cell membranes => weakness + paralysis => ventilatory failure =>
+High doses on muscle: fasciculations $\rightarrow$ inability to repolarise cell membranes $\rightarrow$ weakness + paralysis $\rightarrow$ ventilatory failure $\rightarrow$
 
@@
 
--  MSK: strain on erector spinae, infn of vertebrae, poor hip posture => functional scoliosis, spondylothesis (fracture), spondylitis, spinal stenosis, vertebral fracture, scoliosis, disc prolapse, osteomyelitis, epidural abscesses (IV drug users), OA
+-  MSK: strain on erector spinae, infn of vertebrae, poor hip posture $\rightarrow$ functional scoliosis, spondylothesis (fracture), spondylitis, spinal stenosis, vertebral fracture, scoliosis, disc prolapse, osteomyelitis, epidural abscesses (IV drug users), OA
 -  CVS: aortic aneurysm

### 2017-08-20 neuro3.rmd
sed -i 's/=>/$\rightarrow$/g' *

@@
 -  Trauma: haematoma
--  Vascular: aneurysm, haemorrhage => $\Uparrow$ neck stiffness, hypertensive emergency
+-  Vascular: aneurysm, haemorrhage $\rightarrow$ $\Uparrow$ neck stiffness, hypertensive emergency
 -  Inflamm: meningitis (may not be infn), encephalitis, MS
@@
 -  Hydrocephalus: accumulation of CSF
--  Tumour => bleeding => mass effect
+-  Tumour $\rightarrow$ bleeding $\rightarrow$ mass effect
 -  Cyst
@@
 
-> 29M falling off bike, head first into ground. Pain back of head and neck, neuro rapidly degenerates as he gets to the hospital going from responsive => unresponsive
+> 29M falling off bike, head first into ground. Pain back of head and neck, neuro rapidly degenerates as he gets to the hospital going from responsive $\rightarrow$ unresponsive
 

### 2017-08-20 renal4.rmd
sed -i 's/=>/$\rightarrow$/g' *

@@
 
-Neurogenic bladder: problems with the nervous system causing urinary retention / incontinence. Have to empty this to prevent stasis => infn.
+Neurogenic bladder: problems with the nervous system causing urinary retention / incontinence. Have to empty this to prevent stasis $\rightarrow$ infn.
 
@@
 
-Mechanism for dizziness when straining to void: increasing abdominal pressure => stimulation of vagal nerve => micturition syncope (emptying large amounts of V from bladder).
+Mechanism for dizziness when straining to void: increasing abdominal pressure $\rightarrow$ stimulation of vagal nerve $\rightarrow$ micturition syncope (emptying large amounts of V from bladder).
 
@@
 
-Paediatrics is like veterinary science => have to ask random q's since child can't explain things themselves. \textbf{Always ask more and examine more in children}.
+Paediatrics is like veterinary science $\rightarrow$ have to ask random q's since child can't explain things themselves. \textbf{Always ask more and examine more in children}.
 
@@
 -  Septic shock
--  Meningococcal disease (Neiserria meningitis (gram neg coccus); meningitis): neck stiffness/rigidity, fever, koenig's sign, Brudzinski's sign (more for adults), rash (late sign; caused by septic emboli getting into tissue and causing microinfarctions => non-blanching rash which separates it from dermal rashes), decreased level of consciousness
+-  Meningococcal disease (Neiserria meningitis (gram neg coccus); meningitis): neck stiffness/rigidity, fever, koenig's sign, Brudzinski's sign (more for adults), rash (late sign; caused by septic emboli getting into tissue and causing microinfarctions $\rightarrow$ non-blanching rash which separates it from dermal rashes), decreased level of consciousness
 -  Measles: C's = {koplic spots (white spots in mouth), conjuctivitis, coryza}, fever

### 2017-08-23 neuro4.rmd
added neuro4

---
title: "Neuro IV"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Readings

# Main case

## Trigger 1

Causes of irritability: high pitched cries

-  VINDICATE for the DDx

Parent Hx

-  Good Hx: associated Sx (vomiting, diarrhoea,...)
-  Outbreaks at the daycare
-  Any complications during birth (obstetric Hx)
-  Developmental Hx
-  How does Mad act usually?
-  How has her feeding changed
-  Nappy change rate
-  Rash?
-  Sleep habit change
-  Activity/interactivity level
-  Family Hx/sickness/who is living at home?

Assessment:

-  Paediatric assessment triangle
-  Primary survey (ABCDE)
-  Secondary survey
-  Ongoing assessment

## Trigger 2

Mx plan: address "what's wrong with her": you've brought her in early, but too early to figure out exact cause. Make sure to come back if it gets worse, come back even if not for vaccinations. Have to really figure out how the parents feel.

## Trigger 3

Physical signs

-  Poor cap refill in child

Meningitis: causes Hib, Neisseria, Strep

Under 3/12 & fever: send to ED

$\text{Septic screen} = \text{urine}+\text{bloods (FBC)}+\text{blood cultures} \pm \text{LP} $\pm$ \text{xray (if resp symptoms)}$

Meningitis:

|                   | Viral                      | Bacterial                                                            |   |   |
|-------------------|----------------------------|----------------------------------------------------------------------|---|---|
| Clinical symptoms | Headaches, increased ICP,… | Viral + confusion (much worse, progressesfaster as bacteria has pus) |   |   |
| Rate              | Way more common            | Less common                                                          |   |   |
| CSF               |                            |                                                                      |   |   |
|                   |                            |                                                                      |   |   |

Pathophysiology:

1.  Inflammation
2.  Infn => ventricles
3.  Necrosis, impaired CSF flow, CNS fn

Increases permeability of BBB

Strep pneumoniae is most common cause of MOPS = {meningitis, otitis media, pneumonia, sinusitis}

| Test                       | Bacterial                 | Viral                        |
|----------------------------|---------------------------|------------------------------|
| Opening pressure           | Elevated                  | Usually normal               |
| White blood cell count     | $\geq1000\text{ per }mm^{3}$| <100 per mm3            |  
| Cell differential          | Predominance of PMNs      | Predominance of lymphocytes |
| Protein                    | Mild to marked elevation  | Normal to elevated           |
| CSF-to-serum glucose ratio | Normal to marked decrease | Usually normal               |

## Trigger 4

Risk with LP: removal of CSF can suck brain through foramen magnum (coning)

# Short case 1

# Short case 2

### 2017-08-23 neuro4-encephalitis.rmd
added encephalitis gift

---
title: "Encephalitis"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# References

-  [Harrison's internal medicine](http://emedicine.medscape.com/article/791896-overview)
-  [Medscape](http://accessmedicine.mhmedical.com.ezproxy.library.uq.edu.au/content.aspx?sectionid=79734335&bookid=1130&jumpsectionID=98713820&Resultclick=2)
-  [Australian info](http://www.health.gov.au/internet/main/publishing.nsf/Content/cda-cdi3302e.htm)

# Tl;dr

Inflammation of the brain parenchyma, presents as diffuse/focal neurological dysfunction. Meninges are frequently involved (meningoencephalitis). Most cases are viral in origin with HSV\footnote{Herpes simplex virus} being the most common cause. Mx (pre-ED): seizures (lorazepam), shock, hypotension, airway protection if altered mental status, oxygen. Mx (ED): beyond HSV and varicella, viral encephalitides are non-treatable beyond supportive care. $\text{Acyclovir} \pm \text{antibiotics} \pm \text{steroids}$

# Differentiating meningitis and encephalitis

Encephalitis usually presents with altered mental state, altered level of consciousness, fever, headache, and sometimes focal neurological symptoms like seizures. Meningitis generally presents with the triad of fever/photophobia, headache and nuchal rigidity.

# Pathophysiology

Viral infn. Portals include mosquitos/ticks (arboviruses), animal bites (rabies), immunocompromised state (VZV\footnote{Varicella zoster virus}, CMV\footnote{Cytomegalovirus}).

Generally virus replicates outside CNS and gains access via *haematogenous spread* or along *neural pathways* (rabies, HSV, VZV). Once past the BBB\footnote{Blood brain barrier}, infects neural cells.

Infection results in cell death, perivascular congestion, haemorrhage, and the inflammatory response that preferentially affects grey > white matter. Region specific effects associated with certain viruses is because certain neuronal cell membrane receptors are found within specific brain portions, e.g. HSV mainly impacts inferior and medial temporal lobes.

Acute disseminated encephalitis (EBV\footnote{Epstein barr virus}, CMV) result in multifocal demyelination of white matter.

# Causative agents of death (in Australia)

-  Unknown (51%)
-  HSV
-  Measles
-  Zoster
-  Listeria

# Signs and symptoms

Usual presentation of viral infn:

-  Fever
-  Headache
-  Nausea
-  Vomiting
-  Lethargy
-  Myalgia

Also specific presentations according to the cause. E.g. CMV, measles, VZV, EBV: rash, lymphadenopathy, hepatosplenomegaly and parotid enlargement.

Typical findings also include:

-  Altered mental status
-  Personality changes (very common)
-  Focal findings (eg, hemiparesis, focal seizures, and autonomic dysfunction)
-  Movement disorders (eg, St Louis encephalitis, eastern equine encephalitis, and western equine encephalitis)
-  Ataxia
-  Cranial nerve defects
-  Dysphagia, particularly in rabies
-  Meningismus (less common and less pronounced than in meningitis)
-  Unilateral sensorimotor dysfunction (postinfectious encephalomyelitis)

Complications:

-  Seizures
-  SIADH
-  ICP
-  COMA

# Dx

-  Bloods and urine: FBC, CHEM20 (electrolytes, blood urea nitrogen, creatinine), glucose, urinalysis (electrolytes).
-  LP: need CSF and can run cultures or PCR
-  Imaging: head CT ($\pm\text{contrast}$ should always be done in most patients with encephalitis, to find info on ICP, obstructive hydrocephalus or mass effect. Helps exlude brain haemorrhage/infarction)., MRI is more sensitive earlier in disease course. 
-  Brain biopsy is the diagnostic standard (96% sensitivity, 100% specificity)

# Mx

### Pre-hospital

DRSABCD, seizure precautions, give oxygen and secure IV access.

### In-hospital

-  $\text{Acyclovir} \pm \text{antibiotics} \pm \text{steroids}$
-  Collect blood samples prior to IV therapy
-  Neuro-imaging (MRI > head CT) before LP
-  Considerations: hydrocephalus, increased ICP, systemic complications

# Prognosis

Depends on the causative agent. Outcomes are usually not so great, worsened outcomes if YO(P [not mentioned])I. High mortality, learning disabilities, motor and mental disabilities, memory problems, epilepsy, seizures.

HSV with actclovir reduces mortality from 70% to 25-30%.

Rabies has mortality close to 100%.

### 2017-08-24 neuro4.rmd
Added short cases, should try to start new sentence on new line from now on

@@
 
+## Trigger 1
+
+One eye: probably not occipital lobe as at optic chiasm, left optic tract gets info from both eyes (left visual field), sim for right.
+
+Questions to ask:
+
+-  Fam Hx
+-  What was she doing at the time?
+-  Any headaches when she does move her eye?
+-  Associated symptoms: numbness, tingling, incontinence?
+-  Nature of pain
+-  Better or worse
+-  Onset: there all day?
+-  Allergies, drugs
+
+## Trigger 2
+
+Ex for pupilliary reflex
+
+-  Direct
+-  Consensual
+
+## Trigger 3
+
+Optic neuritis: inflammed myelin sheaths of optic nerve.
+85% of cases permanent damage to optic nerve.
+Tx: steroids for severe or chronic cases.
+
+## Trigger 4
+
+Causes of demyelinating disease: autoimmune, Vit B12,
+
+Dx for MS: McDonald criteria: 2 diff lesions in 2 diff places and 2 diff times
+
 # Short case 2
+
+## Trigger 1
+
+Peripheral nerves is site.
+
+Breathless: up to 30% of GBS require ventilation.
+
+## Trigger 2
+
+Main cause is campylobacter causing GIT infns. 
+Bacteria have LPS which are similar to myelin gangliocytes => creating molecular mimicry. 
+Early lymphocytic infiltrates than macrophage mediated stripping of myelin. 
+Segmental demyelination then infiltrates. 
+Most repair normally.
+
+| Mnemonic    | UMN                                 | LMN           |
+|-------------|-------------------------------------|---------------|
+| Strength    | $\Downarrow$                        | $\Downarrow$  |
+| Tone        | $\Uparrow$                          | $\Downarrow$  |
+| Others      | Superficial reflexes. Absent clonus | Fasiculations |
+| Reflexes    | $\Uparrow$                          | $\Downarrow$  |
+| Muscle mass | $\sim N$                            | $\Downarrow$  |
+| Babinski    | $(+)$                               | $(-)$         |
+
+Mx:
+
+-  Plasmaphoresis or Ig but NOT both
+-  20% mortality if ventilated
+-  Long term complications DVT, paralysis, death, peripheral neuropathy
+-  Recovery takes a *long* time: 6/12 to *years*

### 2017-09-02 msk3.rmd
Updated POWPOW for path week3

@@
 -  Elevation
+
+# POW: Pagets
+
+-  Osteoclasts: multi-nucleated bad boys that resorb bone
+-  Osteoblasts: found on the surfaces of bone and deposit new bone, produce RANKL, produce OPG (inactivates RANKL)
+
+Lesions of pagets phases
+
+1.  Lytic: osteoclasts more numerous, increased size and number of nuclei
+2.  Mixed: osteoblasts increase, loose conn tissue replaces bone marrow, sign neovascularisation occurs
+3.  Sclerotic: osteoblasts predominate and new bone is deposited with pathognomic mosaic pattern, thick cement lines between randomly arranged lamellar bone
+
+Aetiology: paramyxoviruses, pagetic osteoclasts secrete excess IL-6
+
+Symptoms: most asymptomatic, wide range of symptoms including deafness, neuropathy, bone pain (shortening and bowing of legs), hip pain, kidney stones.
+Imp Dx are serum alkaline phosphatase and urinary pyridinoline (markers of bone turnover).
+Treatable with bisphosphanates.
 

### 2017-09-03 stat1201.R
removed old stuff

@@ -1,13 +1,0 @@
-x <- seq(-3, 3, length=1000)
-y <- dnorm(x, mean = 0, sd = 1)
-z_20 <- dt(x, df = 20)
-z_10 <- dt(x, df = 10)
-z_5 <- dt(x, df = 5)
-z_1 <- dt(x, df = 1)
-
-plot(x, y, type="l", lwd=1, col = "black")
-
-lines(x, z_1, col = "red")
-lines(x, z_5, col = "purple")
-lines(x, z_10, col = "pink")
-lines(x, z_20, col = "green")

### 2017-09-04 page-remover.sh
added bash scripts that are not complete

@@ -1,0 +1,32 @@
+#!/bin/bash
+
+echo -n "Hello $USER, how many pages does the bullshit lecture have in total? "
+read MaxPage
+
+
+echo What pages do you want to remove?
+read page1 page2 page3 page4 page5
+
+
+
+# Multiple numebr of inputs 
+
+# Check they are all numbers else echo Type in some numbers you twat
+
+
+
+# echo "What is your .pdf called? Leave out the .pdf and it is caps sensitive. "
+#read pdfdoc
+
+#pdfseparate -f 1 -l $MaxPage $pdfdoc.pdf output-%d.pdf
+#rm output-$page1.pdf
+#rm $pdfdoc.pdf
+#pdfunite $(ls -v *.pdf) complete.pdf
+#rm output*
+
+echo "page1 $page1"
+echo "page2 $page2"
+echo "page3 $page5"
+
+echo Should be right
+exit

### 2017-09-04 render2.sh

@@ -1,0 +1,23 @@
+#!/bin/bash
+
+read -p 'Enter the .rmd file you wish to render: ' File
+read -p 'Do you wish to clean it and remove the "=>"? Answer as y/n ' Clean
+
+if [[ -e "$File" ]] ; then
+    if [["$Clean" == 'y']]; then
+        sed -i 's/=>/\$\\rightarrow\$/g' $File
+        cp $File tmp.rmd
+        echo 'you said yes to clean'
+    elif [["$Clean" == 'n']]; then
+        cp $File tmp.rmd
+    else
+        echo "You didn't type y or n correctly..."
+        exit
+    fi
+
+    R -e 'rmarkdown::render("tmp.rmd")'
+    rm tmp.rmd
+else
+  # if it does not exist:
+  echo 'File does not exist, better luck next time'
+fi

### 2017-09-04 render2.sh
s

@@ -1,23 +1,0 @@
-#!/bin/bash
-
-read -p 'Enter the .rmd file you wish to render: ' File
-read -p 'Do you wish to clean it and remove the "=>"? Answer as y/n ' Clean
-
-if [[ -e "$File" ]] ; then
-    if [["$Clean" == 'y']]; then
-        sed -i 's/=>/\$\\rightarrow\$/g' $File
-        cp $File tmp.rmd
-        echo 'you said yes to clean'
-    elif [["$Clean" == 'n']]; then
-        cp $File tmp.rmd
-    else
-        echo "You didn't type y or n correctly..."
-        exit
-    fi
-
-    R -e 'rmarkdown::render("tmp.rmd")'
-    rm tmp.rmd
-else
-  # if it does not exist:
-  echo 'File does not exist, better luck next time'
-fi

### 2017-09-04 render-doc-2.sh
finished up 1

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+# Modify this by making conditional based on if extension or not (if there is a ".")
+read -p 'Enter the .rmd file you wish to render (without the extension): ' File
+echo "$File".rmd
+
+if [[ -e "$File".rmd ]]; then
+    echo "Do you wish to clean the tmp .rmd file? "
+    select yn in "Yes" "No" "Cancel"; do
+        case $yn in
+            Yes ) cp "$File".rmd tmp.rmd; sed -i 's/=>/\$\\rightarrow\$/g' tmp.rmd; R -e 'rmarkdown::render("tmp.rmd")'; rm tmp.rmd; mv tmp.pdf "$File".pdf; break;;
+            No ) cp "$File".rmd tmp.rmd; R -e 'rmarkdown::render("tmp.rmd")'; rm tmp.rmd; mv tmp.pdf "$File".pdf; break;;
+            Cancel ) exit;;
+        esac
+    done
+    echo "Succesfully knitted the document (hopefully)"
+else
+    echo "File does not exist"
+fi
+

### 2017-09-04 render-doc.sh

@@ -1,12 +1,18 @@
 #!/bin/bash
 
-read -p 'Enter the .rmd file you wish to render: ' File
-read -p 'Do you wish to clean it and remove the "=>"? Answer as y/n ' Clean
+# Modify this by making conditional based on if extension or not (if there is a ".")
+read -p 'Enter the .rmd file you wish to render (without the extension): ' File
 
-if [[ -e "$File" ]] ; then
-    if [["$Clean" == 'y']]; then
-        sed -i 's/=>/\$\\rightarrow\$/g' $File
-        cp $File tmp.rmd
+# Cleaning refers to making a tmp file and:
+#    - Converting => to $\rightarrow$ for LaTeX output
+#    - Whatever else I decide to add in
+read -p 'Do you wish to clean the .rmd file? Answer as y/n: ' Clean
+
+if [[ -e "$File".rmd ]] ; then
+    # Cleaning commands go in this conditional here
+    if [[ echo "$Clean" | grep -iq "^y" ]]; then
+        cp $File.rmd tmp.rmd
+        sed -i 's/=>/\$\\rightarrow\$/g' tmp.rmd
         echo 'you said yes to clean'
     elif [["$Clean" == 'n']]; then
         cp $File tmp.rmd

### 2017-09-04 render-doc.sh
removed old v

@@ -1,29 +1,0 @@
-#!/bin/bash
-
-# Modify this by making conditional based on if extension or not (if there is a ".")
-read -p 'Enter the .rmd file you wish to render (without the extension): ' File
-
-# Cleaning refers to making a tmp file and:
-#    - Converting => to $\rightarrow$ for LaTeX output
-#    - Whatever else I decide to add in
-read -p 'Do you wish to clean the .rmd file? Answer as y/n: ' Clean
-
-if [[ -e "$File".rmd ]] ; then
-    # Cleaning commands go in this conditional here
-    if [[ echo "$Clean" | grep -iq "^y" ]]; then
-        cp $File.rmd tmp.rmd
-        sed -i 's/=>/\$\\rightarrow\$/g' tmp.rmd
-        echo 'you said yes to clean'
-    elif [["$Clean" == 'n']]; then
-        cp $File tmp.rmd
-    else
-        echo "You didn't type y or n correctly..."
-        exit
-    fi
-
-    R -e 'rmarkdown::render("tmp.rmd")'
-    rm tmp.rmd
-else
-  # if it does not exist:
-  echo 'File does not exist, better luck next time'
-fi

### 2017-09-04 doc-render.sh
Cleaned up

@@ -1,8 +1,19 @@
 #!/bin/bash
 
-# Modify this by making conditional based on if extension or not (if there is a ".")
+###############################################################
+####      Renders .rmd file using R using knitr/pandoc     ####
+###############################################################
+
+# Dependencies
+#   R
+#   TeX (if you decide to output a pdf_document)
+#   R packages: knitr
+
+# To do
+#   making conditional based on if extension or not (if there is a ".")
+#   Make the cleaning stuff functions so it is easier to examine
+
 read -p 'Enter the .rmd file you wish to render (without the extension): ' File
-echo "$File".rmd
 
 if [[ -e "$File".rmd ]]; then
     echo "Do you wish to clean the tmp .rmd file? "

### 2017-09-05 neuro5.rmd
main case

---
title: "Neuro V"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Readings

# Main case

## Trigger 1

Symptoms due to old age: unilateral, stiff and weak, only 60 so probs not. Age related degeneration is typically bilateral, older,...

CNS

-  Motor power
-  Movement
-  Coordination: cerebellum

DDx:

-  Neuro: stroke, parkinsons, huntingtons, MS, GBS, MND (motor neuron disease), myasthenia gravis, cerebellar alcohol, vertigo, essential tremor, visual impairment that is not being picked up, dementia, peripheral neuropathy
-  Malig: primary or secondary tumours

Sensory for coordination:

-  Propioception
-  Visual
-  Vestibular apparatus

Hx taking:

-  Flesh out PC
    -  Onset: how did symptoms come in
    -  Fumbling at rest or moving
    -  Dizziness/falling over: vertigo, inner ear
    -  Visual fn: reading or not
    -  Headaches/faints: tumours
-  Anything else
-  PMHx
-  PSHx
-  FHx
-  Sexual Hx
-  Medications, allergies, adverse drugs
-  Social Hx: smoking, drinking, drugs, rest of it, how is it affecting life, dependents

## Trigger 2

Key symptoms

-  Slowing up: bradykinesia
-  Tremor: increases when trying to hurry or agitated
-  Mumbling
-  Episode of vertigo in the past: stemetil

Parkinsons: bradykesia, rigidity, tremor, postural instability

Basal ganglia: disinhibition and inhibition of differnet planned movements. Therefore has issues starting a movement.

| Extrapyramidal             | Corticospinal             | Cerebellar           |
|----------------------------|---------------------------|----------------------|
| Parkinsonism               | Increased tendon reflexes | Intention tremor     |
| Dystonia                   | Fasiculations             | Troubles with speech |
| Restlessness, urge to move | Increased tone            |                      |
| Gait: shuffling            |                           |                      |
| Lack of normal arm swing   |                           |                      |
| Walking heel to toe        |                           |                      |
| Cogwheel rigidity          |                           |                      |

## Trigger 3

Parkinsons: loss of 60-80 of dopaminergic neurons in pars compacta

Extrapyramidal disorders: anything disrupting the pathway. Lesions in basal ganglia, drugs

## Trigger 4

Side effects of levodopa:

-  Arrhythmias
-  Hypotension
-  Nausea
-  GI bleeding
-  Hallucinations
-  End of dose deterioration of fn
-  Oscillations
-  Dyskinesea

There are no disease modifying agents for parkinsons. These drugs only cure symptoms for a bit, disease will still progress.

## Trigger 5

On/off symptoms: try reduce dose but give more frequently so it still has effects without adverse effects

Natural Hx: loss of motor fn, dementia, death, autonomic effects

# Short case 1

# Short case 2

### 2017-09-05 neuro5.rmd
half added sc1

@@
 
+## Trigger 1
+
+> 53M weakness and clumsiness of right hand 3/12. 6/12 began dropping things with R hand, gradual deterioration in handwriting, losing grip strength.
+
+Anatomical sites (DDx for muscle weakness; unilateral/focal tends to be neurological))
+
+-  Muscles:
+    -  Primary myopathies: hereditoary/congenital (dystrophies), finalmm, metabolic
+    -  Secondary myopathies: infn (trichanosis), conn tissue disease, thyroid/adrenal dysfn, paraneoplastic, drugs
+    -  Trauma
+-  NMJ: myasthenia gravis, lambert eaten, toxins such as snake venoms/ticks/pesticides
+-  LMN: peripheral nerve entrapment, polyneuropathy, GBS
+-  Spinal cord: radiculopathy from spinal stenosis, neoplasms, stuff damaging peripheral nerve roots, infn from CMV/lyme can damage anterior horn => LMN damage, MS
+-  UMN: motor neuron diseases such as amyotrophic lateral sclerosis, 
+-  Cerebral hemisphere: CVA (usually +sensory)
+-  Systemic: Vit B12 deficiency, hypercalcaemia/hyperparathyroidism (+abd pain, excessive urination, renal stones)
+
+Hx:
+
+-  HPC
+    -  Tell me more...
+    -  What does weakness mean?
+Sometimes pt confuse weakness with fatigue, often part of systemic disorders, psych, MS, parkinsons.
+Pt might describe rigidity/incoordination as weakness, ask "what could you do that you no longer can" fishing for "lifting up milk",...
+    -  Site
+    -  Onset: insidious onset suggests degenerative or chronic (metabolic ,hereditary), intermittent (e.g.
+using it makes it worse suggests myasthenia gravis), particular positions making it bad then mechanical issue such as entrapment neuropathy, radiculopathy, localised vascular occlusion, particular times of day suggests ion channel disorder, random symptom timing with quick recovery suggests transient ischaemia of CNS
+    -  Character, radiation
+    -  Associated symptoms: 
+        -  Atrophy/fasciculations suggest LMN
+        -  Systemic symptoms (probs not unilateral arm) can be from endocrine (thyroid, diabetus)
+        -  Neoplasias/infn
+        -  *Sensory* such as parasthaesia, vision, vertigo suggests NS
+        -  Change in character of voice, problems swallowing (pharyngeal muscles)
+        -  Increased urinary urgency can suggest spinal damage (myelopathy)
+    -  How is it affecting daily life, what does he think it is?
+-  PMHx: can get clues for infn cause or neoplastic
+-  SHx
+-  FHx
+-  Medications, allergies, ADR
+-  Social Hx: drugs, alcohol, cigarettes
+
+**finish up the DDx**
+
+## Trigger 2
+
+Cues:
+
+-  Generally weak
+-  No sensory or paraesthesia
+-  Head, neck, or shoulder pain
+-  No recent injuries
+-  No pattern to weakness, not aware of stuff that makes it better or worse
+-  No fatigueable element
+-  No bladder or bowel disturbances, no double vision
+-  No difficulties with swallowing or changes to voice/speech
+
+Ex:
+
+## References
+
+-  [Good DDx for weakness](http://www.dartmouth.edu/~dons/part_2/chapter_12.html)
+-  [AFP article: evaluation of pt with muscle weakness](http://www.aafp.org/afp/2005/0401/p1327.html)
+-  [Medscape MND](http://emedicine.medscape.com/article/2111360-overview#a1)
+-  [AFP article of MND](http://www.racgp.org.au/download/documents/AFP/2011/December/201112zoing.pdf)
+-
+
 # Short case 2

### 2017-09-06 neuro5.rmd
not completed yet

--- 2017-09-05 neuro5.rmd
@@
 
+## Trigger 1
+
+> 53M weakness and clumsiness of right hand 3/12. 6/12 began dropping things with R hand, gradual deterioration in handwriting, losing grip strength.
+
+Anatomical sites (DDx for muscle weakness; unilateral/focal tends to be neurological))
+
+-  Muscles:
+    -  Primary myopathies: hereditary/congenital (dystrophies), finalmm, metabolic
+    -  Secondary myopathies: infn (trichanosis), conn tissue disease, thyroid/adrenal dysfn, paraneoplastic, drugs
+    -  Trauma
+-  NMJ: *myasthenia gravis*, lambert eaten, toxins such as snake venoms/ticks/pesticides
+-  LMN: peripheral nerve entrapment, polyneuropathy, GBS, *post-polio syndrome* (in sufferers of polio), *multifocal motor neuropathy*
+-  Spinal cord: *spinal cord compression* (radiculopathy from spinal stenosis), neoplasms, stuff damaging peripheral nerve roots, infn from CMV/lyme can damage anterior horn => LMN damage, *MS*, 
+-  UMN: tumour
+-  Cerebral hemisphere: CVA (usually +sensory)
+-  Neurodegenerative diseases like *motor neuron diseases (MND)* = {amyotrophic lateral sclerosis (ALS), primary lateral sclerosis (PLS), hereditary spastic paraparesis (HSP), spinal muscular atrophy (SMA), x-linked spinobulbar muscular atrophy (SBMA)}
+-  Systemic: *Vit B12 deficiency*, hypercalcaemia/hyperparathyroidism (+abd pain, excessive urination, renal stones), depression, *heavy metal poisoning*, *lyme disease*
+
+Hx:
+
+-  HPC
+    -  Tell me more...
+    -  What does weakness mean?
+Sometimes pt confuse weakness with fatigue, often part of systemic disorders, psych, MS, parkinsons.
+Pt might describe rigidity/incoordination as weakness, ask "what could you do that you no longer can" fishing for "lifting up milk",...
+        -  Fatigue describes the inability to continue performing a task after multiple repetitions
+        -  Primary weakness is unable to perform the first repetition of the task
+        -  Asthenia is a sense of weariness or exhaustion in the absence of muscle weakness.
+Common in people who have chronic fatigue syndrome, sleep disorders, depression, or chronic heart, lung, and kidney disease
+    -  Site
+    -  Onset: insidious onset suggests degenerative or chronic (metabolic ,hereditary), intermittent (e.g.
+using it makes it worse suggests myasthenia gravis), particular positions making it bad then mechanical issue such as entrapment neuropathy, radiculopathy, localised vascular occlusion, particular times of day suggests ion channel disorder, random symptom timing with quick recovery suggests transient ischaemia of CNS
+    -  Character, radiation
+    -  Associated symptoms: 
+        -  Atrophy/fasciculations suggest LMN
+        -  Systemic symptoms (probs not unilateral arm) can be from endocrine (thyroid, diabetus)
+        -  Neoplasias/infn
+        -  *Sensory* such as parasthaesia, vision, vertigo suggests NS
+        -  Change in character of voice, problems swallowing (pharyngeal muscles)
+        -  Increased urinary urgency can suggest spinal damage (myelopathy)
+    -  How is it affecting daily life, what does he think it is?
+-  PMHx: can get clues for infn cause or neoplastic
+-  SHx
+-  FHx
+-  Medications (amiodarone, chemotherapeutic agents, interferon, cocaine, corticosteroids, penicilin, sulfonamides, statins, NSAIDS), allergies, ADR
+-  Social Hx: drugs, alcohol, cigarettes
+
+## Trigger 2
+
+Cues:
+
+-  Generally weak
+-  No sensory or paraesthesia: do exam first, but suggests against 
+-  Head, neck, or shoulder pain
+-  No recent injuries: not trauma
+-  No pattern to weakness, not aware of stuff that makes it better or worse
+-  No fatigueable element: less likely NMJ
+-  No bladder or bowel disturbances, no double vision
+-  No difficulties with swallowing or changes to voice/speech: not cranial nerves?
+-  PMHx relatively uneventful: no post-polio
+
+Ex clinical signs:
+
+-  
+-  Neuro upper limb
+
+## Trigger 3
+
+MND: often Dx of exclusion
+
+-  ALS: combination of UMN and LMN in the limbs and bulbar muscles
+-  Progressive muscle atrophy with only lower motor neuron signs initially; some patients develop later lower motor neurone signs
+-  Primary lateral sclerosis, with predominantly upper motor neuron features
+-  Progressive bulbar palsy, with degeneration of bulbar nuclei and preservation of anterior horn cell and upper motor neuron function.
+
+## References
+
+-  [Good DDx for weakness](http://www.dartmouth.edu/~dons/part_2/chapter_12.html)
+-  [AFP article: evaluation of pt with muscle weakness](http://www.aafp.org/afp/2005/0401/p1327.html)
+-  [Medscape MND](http://emedicine.medscape.com/article/2111360-overview#a1)
+-  [AFP article of MND](http://www.racgp.org.au/download/documents/AFP/2011/December/201112zoing.pdf)
+
 # Short case 2

### 2017-09-06 neuro5.rmd
trialling images, also cbf 100% completing sc1, should be rihgt

@@
 
--  Generally weak
--  No sensory or paraesthesia: do exam first, but suggests against 
--  Head, neck, or shoulder pain
--  No recent injuries: not trauma
--  No pattern to weakness, not aware of stuff that makes it better or worse
+-  Generally weak: something systemic, or just feeling shit
+-  No sensory or paraesthesia: do exam first, but suggests against radiculopathies
+-  No head, neck, or shoulder pain
+-  No recent injuries: against trauma
+-  No pattern to weakness, not aware of stuff that makes it better or worse: probably not mechanical/orthopaedic such as nerve entrapment
 -  No fatigueable element: less likely NMJ
 -  No bladder or bowel disturbances, no double vision
--  No difficulties with swallowing or changes to voice/speech: not cranial nerves?
+-  No difficulties with swallowing or changes to voice/speech: less likely cranial nerves involved
+    -  CN V: muscles for mandible, tensor veli palatini
+    -  CN VII: motor muscles of facial expression + stapedius, sensory ante $\frac{2}{3}$ tongue
+    -  CN IX: motor middle pharyngeal constrictor + stylopharyngeus, sensory post $\frac{1}{3}$ tongue, velum, tonsils
+    -  CN X: motor majority of viscera, sends motor commands to pharyngeal constrictors, cricothyroid for changes in pitch, glossopalatine and levator veli palatine muscles
+    -  CN XI: trapezius and SCM of the neck
+    -  CN XII: motor controlling tongue movement (intrinsic + extrinsic). Lesions here can result in fasciculations.
 -  PMHx relatively uneventful: no post-polio
@@
 
+Clin findings suggestive of MND
+
+-  Asymm weakness in upper or lower limbs
+-  Brisk reflexes in wasted limb
+-  Weakness in absence of pain or prominent sensory symptoms
+-  Relentless progression of symptoms and signs
+
 ## Trigger 3
 
-MND: often Dx of exclusion
+https://i.imgur.com/fUq88Wh.png <!-- image to be knit into the doc, under images/ -->
 
--  ALS: combination of UMN and LMN in the limbs and bulbar muscles
--  Progressive muscle atrophy with only lower motor neuron signs initially; some patients develop later lower motor neurone signs
--  Primary lateral sclerosis, with predominantly upper motor neuron features
--  Progressive bulbar palsy, with degeneration of bulbar nuclei and preservation of anterior horn cell and upper motor neuron function.
+| Dx                    | Positives                   | Negatives                   |
+|-----------------------|-----------------------------|-----------------------------|
+| Peripheral neuropathy | Explains weakness           | Hyperreflexia               |
+| Cervical cord lesion  |                             | Nothing below the level     |
+| Muscular dystrophy    | Explains bilateral weakness | Reflexes increased, PAINFUL |
+| R ulnar nerve lesion  | Weakness R hand             | Doesn't explain bilateral   |
+| Motor neuron disease  | Vague as fuck               |                             |
+
+MND: often Dx of exclusion.
+Generally occurs when age $\in[30,60]$, with degeneration of ante horn cells in spinal cord, motor nuclei of lower cranial nerves, and corticospinal/bulbar pathways.
+Usually sporadic, some familial.
+5 varieties based on clinical grounds:
+
+1.  Progressive bulbar palsy: bulbar involvement predominates owing to disease processes affecting primarily the motor nuclei of the cranial nerves
+2.  Pseudobulbar Palsy: bilateral corticobulbar disease reflecting UMN dysfn. "Pseudo-bulbar effect" with uncontrollable laughing/crying to stimuli
+3.  Progressive spinal muscular atrophy: LMN deficit in limbs due to degen of ante horn cells
+4.  Primary lateral sclerosis: purely UMN deficit in the limbs
+5.  Amyotrophic lateral sclerosis: mix of U and LMN in the limbs, sometimes associated with cognitive decline (frontotemporal dementia), psudobulbar effect, or parkinsonism
+
+Differences between bulbar and pseudobulbar palsy (signs and symptoms):
+
+-  Bulbar: LMN lesion to cranial nerves IX, X, XI, XII, in the medulla or lesons of the lower cranial nerves outside the brainstem.
+Symptoms = {tremulous lips, tongue weak and wasted and fasculations, drooling + dysphagia (unable to swallow), dysphonia, articulation issues (cant pronounce *r*)}
+-  Pseudobulbar: lesion to corticobulbar tracts.
+Bilateral damage must be present as many muscles have bilateral innervation.
+Presentation = {paralysed tongue and no fasciculations, dribbling, facial muscles paralysed, exaggerated reflexes, dysphonia, dysphagia, emotional lability may be present}
+
+Further Ix
+
+-  Electromyography: may show changes of chronic partial denervation, ab~N spontaneous activity in resting muscle.
+Dx should not be made with confidence unless change are found in $\geq3# spinal regions, or $2$ regions + bulbar
+-  Motor conduction velocity is usually normal
+-  Sensory conduction studies normal
+-  Biopsy of wasted muscle
+-  Serum CK may be elevated, but not as high as muscular dystrophies
+-  CSF normal
+
+What are possible future complications
+
+-  Sialorrhoea: unintentional loss of saliva through the mouth
+    -  Choking
+    -  Pneumonia
+-  Dermatological: rashes
+-  Resp failure
+-  Death
+
+Principles of Mx for MND
+
+-  Refer
+-  Admit for initiation or titration of noninvasive ventilation for those with low inspiratory pressure, may prolong survival
+-  Riluzole: reduces presynaptic release of Glu, increases short term survival for ALS
+-  Symptomatic/supportive measures: trials of anticholinergics
+    -  Braces/walker
+    -  Physical therapy to prevent contractures
+-  Relieve pseudobulbar affect with combination dextromethorphan/quinidine
+-  Gastronomy tube if dysphagia is severe
+-  Palliative
 
@@
 -  [AFP article of MND](http://www.racgp.org.au/download/documents/AFP/2011/December/201112zoing.pdf)
+-  [Categories of MND](http://accessmedicine.mhmedical.com.ezproxy.library.uq.edu.au/content.aspx?sectionid=135716257&bookid=1843&jumpsectionID=135717169&Resultclick=2)
 

### 2017-09-06 neuro5.rmd
added table from racgp

@@
     -  Physical therapy to prevent contractures
+    -  Mx saliva
+    -  Dyspnoea
+    -  Constipation
+    -  Muscle cramps/spasms/spasticity
+    -  Depression, emotional lability
+    -  Choking sensation
+    -  Panic claustrophobia
 -  Relieve pseudobulbar affect with combination dextromethorphan/quinidine

### 2017-09-07 neuro5.rmd
completing week 9

@@
 -  Electromyography: may show changes of chronic partial denervation, ab~N spontaneous activity in resting muscle.
-Dx should not be made with confidence unless change are found in $\geq3# spinal regions, or $2$ regions + bulbar
+Dx should not be made with confidence unless change are found in $\geq3$ spinal regions, or $2$ regions + bulbar
 -  Motor conduction velocity is usually normal
@@
 # Short case 2
+
+## Trigger 1
+
+Proceed with consultation: direct Q at pt, ask to have a chat with husband
+
+Dx: can get advanced health directive, clin trials, gives pt time to do stuff before dying, interventions can help
+
+Depression, dementia, delirium have to d/dx
+
+## Trigger 2
+
+Commonest causes of memory impairment: important to know ddx as some are treatable
+
+-  Dementia
+-  Alzheimers
+-  Parkinsons
+-  Thyroid
+-  Depression
+-  Cerebral infn
+-  Sleep deficiencies
+
+Classification of dementia: 4 main types = {alzheimers, lewy body (concentration, distances, *visual hallucinations, parkinsonism, fluctuation in mental state*), vascular (step deterioration, sudden onset), frontotemporal}
+
+Specific cognitive tests:
+
+-  MMSE: out of 30, few questions testing working, visualspacial awareness,... Below 28 get concerned. This is biased
+-  GPCOG: piece of paper for both pt and their SO
+
+Ex and Ix: to mainly rule stuff out
+
+-  Bloods + Vit B12
+-  LP
+-  Prions
+-  MRI
+-  PET
+
+## Trigger 3
+
+Aged care assessment, talk to both people at once, discuss prognosis, mental exercises
+
+## Trigger 4
+
+Medications: cholinesterase inhibitors, NMDA-R antagonists, symptomatic stuff for sleeping, craziness, antidepressants

### 2017-09-08 neuro2.rmd
updated some typos, hopefully table goes towards the \section

@@
 ---
-title: "Module II"
+title: "Neurology II"
 author: "Aaron Khoo"
@@
 ---
-
-# Readings
 
@@
 
--  Anterolateral tract: spinothalamic (pain and temp), spinoreticular, spinotectal = {pain, temperature, crude touch, firm pressure}
--  Dorsal lemniscal: check lecture 4 of this week (Bellingham)
+-  Anterolateral tract: spinothalamic (pain and temp), spinoreticular, spinotectal = {pain, temperature, crude touch, firm pressure}. Goes up or down 1 or 2 levels then decussates.
+-  Dorsal lemniscal (column): fine touch and proprioception, decussates up in the brainstem.
 
@@
 
-High doses on muscle: fasciculations $\rightarrow$ inability to repolarise cell membranes $\rightarrow$ weakness + paralysis $\rightarrow$ ventilatory failure $\rightarrow$
+High doses on muscle: fasciculations $\rightarrow$ inability to repolarise cell membranes $\rightarrow$ weakness + paralysis $\rightarrow$ ventilatory failure $\rightarrow$ death
 
@@
 Atropine: muscarinic antagonist. Actions would be the opposite of DUMBELS
-
-Memorise the picture from the Eastern European lecturer.
 

### 2017-09-08 neuro3.rmd
Fixed name of document...

@@
 ---
-title: "Module II"
+title: "Neurology III"
 author: "Aaron Khoo"

### 2017-09-08 neuro4.rmd
Added POWPOW

@@
 
-# Readings
+# POW
+
+Presentation: 30F left lag dragging, difficulties walking long distances, feels tingling in hands and it seems to have come back.
+
+Ix: HLA-DR2 haplotype associated with MS, oligoclonal bands in CSF suggest inflamm process, normal metabolic panel and thyroid excludes other explanations such as hypothyroidism.
+
+Explanation of oligclonal bands: each band represents IgG produced by activated B cell.
+These result in the autoimmune attacks.
+Being present in CSF but NOT plasma means it originates in CNS.
+If was in blood, then systemic immune response could have crossed leaky BBB into CSF.
+
+Pathophysiology: Multiple sclerosis is an example of a Type IV hypersensitivity.
+Initially, T cells cross the blood brain barrier and attach to ligands on the myelin sheath and oligodendrocytes .
+The T cells mistake these cells as being foreign and release cytokines, resulting in inflammation that damages the myelin.
+Normally, myelin increases the speed of electrical signals via saltatory conduction.
+When the myelin is gone, the signals can’t propagate along the axons as well.
+This can eventually lead to conduction block .
+Hence, we see the neurological deficits in this patient.
+At first, through the process of remyelination some of the symptoms will resolve.
+However, as time goes on, this process is less effective and symptoms get worse.
+
+Histo: fried egg looking cells are oligodendrocytes.
+
+| Symptom       | Site of damage                                                      |
+|---------------|---------------------------------------------------------------------|
+| Blurry vision | Optic nerve                                                         |
+| Double vision | Oculomotor nerve                                                    |
+| Limb weakness | Pyramidal tract                                                     |
+| Limb numbness | Posterior columne medial lemniscus pathway  (fine touch and propio) |
+| Ataxia        | Cerebellum                                                          |
 
@@
 
-$\text{Septic screen} = \text{urine}+\text{bloods (FBC)}+\text{blood cultures} \pm \text{LP} $\pm$ \text{xray (if resp symptoms)}$
+$\text{Septic screen} = \text{urine}+\text{bloods (FBC)}+\text{blood cultures} \pm \text{LP} \pm \text{xray (if resp symptoms)}$
 

### 2017-09-11 msk1.rmd
Updated yaml header to make it a4. Stuff american letter paper
sed -i 's/=>/$\rightarrow$/g' *
fixed typo

@@
-```
+---
 title: "MSK I"
@@
 date: "`r format(Sys.Date(), '%d %B %Y')`"
-output: pdf_document
-```
+output:
+  pdf_document
+documentclass: article
+classoption: a4paper
+---
 
@@
 
-After primary survey => secondary survey
+After primary survey $\rightarrow$ secondary survey
 
@@
 PMHx = {osteoarthritis, HTN, insomnia, hearing loss, BPH}.
-Meds = {atenolol [$\beta-1$ blocker; used for HTN], oxazepam [benzodiazepine; used for anxiety, insomnia, and alcohol withdrawal], panadol osteo [paracetamol; relief for osteoarthritis]}.
+Meds = {atenolol [$\beta_{1}$ blocker; used for HTN], oxazepam [benzodiazepine; used for anxiety, insomnia, and alcohol withdrawal], panadol osteo [paracetamol; relief for osteoarthritis]}.
 Said he slipped on bath mat after waking up whilst going to the toilet.

### 2017-09-11 neuro6.rmd
added week 10

---
title: "Neurology VI"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
urlcolor: blue
---

# Main case

# Short case 1

# Short case 2

### 2017-09-11 msk2.rmd
fixed a few typos

@@
 | Type | Character | Time course | Radiation | NSAID response | Onset | Age |
-| ------ |---------------  | ---------- | --------- | ------- |  ------ | ----- |
+| --------- |---------------  | ---------- | --------- | ------- |  ------ | ----- |
 | Mechanical | Thinking muscular: diffuse/achey | Worse with movement PM AM | Lumbar to back of thigh | Moderate/good | Post trauma/gradual | Mid-late |
@@
 
-What are the types of spondyloarthropathies (Rheumatoid factor negative)
+What are the types of spondyloarthropathies (Rheumatoid factor and other antibody negative; many carry HLA-B27))
 
@@
 -  IBD associated: 
--  
--
+-  Undifferentiated:
 
@@
 
--  Oseopenia
+-  Osteopenia
 -  LOTS of CVS issues

### 2017-09-12 neuro6.rmd
added epilepsy main case

@@
 
+## Trigger 1
+
+DDx seizure/unconsciousness:
+
+-  Vascular: CVA/TIA
+-  Infn: encephalitis, meningitis, abscess
+-  Trauma: head trauma
+-  Neuro: MS
+-  Medications/drugs/alcohol/sleep deprivation
+-  Systemic: diabetes (hypo/hyper-glycaemia), hyperthyroidism, vit B12 deficiency, ion deficiency, young adult female = PREGNANT, dehydration (syncopal episode)
+
+Initial assessment
+
+-  DRSABC(DEFG = don't ever forget glucose)
+-  MIST/AMPLE: mechanism of injury, injury sustained, vitals and signs/symptoms, treatments, / , allergies, medications, past illness, last meal, events related to injury
+
+Seizures: inapproiate excitability/synchrony => increased bursts/decreases inhibition.
+Variety of genetic and structural defects that can contribute.
+Idiopathic epilepsy: bilateral generalised discharge.
+
+Hx:
+
+-  HPC: can she remember, onset: eating, how quickly LOC, drugs, previous episodes before, associated symptoms = {headaches, nausea, vomiting, aura}, history of head trauma, drugs/alcohol, [have to get collateral]
+-  PMHx/SHx: genetic/congenital, Ca, mental disabilities, seizures before!, previous head trauma, neurosurgery
+-  FHx
+-  Repro Hx: sexually active
+-  Allergies, medications, natural therapies + 
+-  Social Hx: sleep, diet, if she lives alone, hobbies, drives for work,... % maybe don't need to find right *now* but do have to consider these
+
+Seizures:
+
+-  Focal: 
+    -  Retained awareness
+    -  Impaired awareness
+-  General: tonic-clonic
+
+Postictal state is the altered state of consciousness after an epileptic seizure
+
+Ex: usually unremarkable for epilepsy, have to do if considering structural/infn
+
+-  General inspection: checking for trauma
+-  Neuro
+-  CV
+
+## Trigger 2
+
+[Status epilepticus](https://lifeinthefastlane.com/ccc/status-epilepticus/):
+
+-  Continuous seizure activity for 5 minutes or more without return of consciousness, or
+-  Recurrent seizures (2 or more) without an intervening period of neurological recovery
+
+Medications
+
+-  Medazolam: benzodiazopine increases GABA so it is inhibitory
+-  Phenytoin: anticompulsant that blocks V-gated Na channel
+
+## Trigger 3
+
+Hx:
+
+-  [Febrile convulsions](http://www.rch.org.au/clinicalguide/guideline_index/Febrile_Convulsion/) are *common*
+-  Focal seizures with impaired awareness leading to secondary generalised seizures: suggests structural abnormality
+-  Absence seizures
+
+Classification: still clasify with motor, non-motor, unclassified
+
+-  Focal onset = {aware or impaired awareness}
+-  Generalised = {tonic-clonic, absence, myoclonic (one muscle), clonic, tonic, atonic}
+-  Unknown = {inadequate info, unusual presentation}
+
+Further Hx:
+
+-  Find the reason why she's had another seizure: triggered? forgot meds?
+-  Fever, missing meals, pregnant,...
+
+Ix:
+
+-  EEG (+ sleep deprived to try provoke it)
+-  MRI
+-  Thyroid
+-  Bloods: FBC
+
+## Trigger 4
+
+Temporal lobe: most common place for it to start. Temporal lobe epilepsy most common, mesial temporal sclerosis pathological finding. Fits with findings as she has focal => general stuff happening.
+
+## Trigger 5
+
+Medications: side effects generally drowsiness, learn which ones are for general vs focal epilepsy
+
+-  Sodium valproate: increases production and inhibits breakdown of GABA | side effect = birth defects, tremor, thinning of hair
+-  Carbamazepine: sodium channel blocker
+
+Surgery: uncontrollable from medications, too severe. Can remove lobe or cut corpus callosum.
+
 # Short case 1

### 2017-09-12 neuro6.rmd
changed fontsize

@@
 classoption: a4paper
+fontsize: 11pt
 urlcolor: blue

### 2017-09-13 neuro1.rmd
sed -i 's/=>/$\rightarrow$/g' *
not sure what I did

@@
 
-Tinels sign: tap over nerve => if tingly/pins & needles => nerve irritation. Over time, expect the tingly feeling to go downwards more and more to show the nerve is healing.
+Tinels sign: tap over nerve $\rightarrow$ if tingly/pins & needles $\rightarrow$ nerve irritation. Over time, expect the tingly feeling to go downwards more and more to show the nerve is healing.
 

### 2017-09-14 neuro6.rmd
sc

@@
 
+## Trigger 1
+
+DDx for 1st seizure:
+
+-  Acute vascular event/trauma: haemorrhage
+-  Infn: encephalitis > meningitis
+-  Systemic: ionic imbalances, hypoglycaemia, hyperthyroidism
+-  Drugs, medications
+-  Tumour: glioma
+-  Is she pregnant?
+-  Epilepsy: might not have been apparent previously
+
+## Trigger 2
+
+Headache from space occupying lesion: worse on lying down (sleep/morning), coughing, sneezing. Can come from tumour, abscess,..
+
+5 most common secondaries:
+
+1.  Breast
+2.  Lung
+3.  Melanoma
+4.  GIT/liver
+
+Intrinsic: glioma,...
+
+Extrinsic: meningioma, pituitary adenoma
+
+Ix:
+
+-  *Don't LP (yet): signs of increased ICP*
+-  MRI
+-  Bloods: blood glucose, FBCs (platelets, red cells), CHEM20 (ions, underlying kidney fn)
+-  EEG 
+-  PET later on to see if it was a secondary met
+
+## Trigger 3
+
+Mx:
+
+1.  Surgical
+2.  Radiotherapy
+3.  Adjuvant chemotherapy: some passes BBB, some not
+
 # Short case 2
+
+## Trigger 1
+
+Anatomical sites:
+
+-  Incontinent of urine: could be generalised seizure
+-  Right motor dysfn: left hemisphere
+
+Hx:
+
+-  HPC: ever had incontinence before, trauma, how was he right before, general health in the lead up, sitting or standing, 
+    -  Systems neuro: changes in vision, headaches, dizziness/faint, nausea, memory problems, concentration, personality changes
+-  PMHx/PSHx: had a brain surgery before, post CVA
+-  Medications/allergies/adr/drugs/cigarettes/alcohol
+
+## Trigger 2
+
+1.  Can't maintain airway during a seizure: oxygen
+2.  Stop seizure (status epilepticus): benzodiazepine
+3.  IV access
+
+## Trigger 3
+
+Bloods:
+
+-  FBC: RBCs, white cells
+-  ESR/CRP: inflamm
+-  CHEM20: underlying kidney/liver fn, specificaly Ca and Mg
+-  Drug screen
+-  ABG: due to resp symptoms
+
+Ix:
+
+-  Head CT
+-  CXR
+-  ECG
+
+PDx: metastatic brain cancer
+
+## Trigger 4
+
+Further Ix:
+
+-  Biopsy/chest CT to examine Ca better
+-  PET
+
+## Trigger 5
+
+5 most common mets to brain:
+
+-  Lung
+-  Breast
+-  Melanoma
+-  Renal cell
+-  Bowel

### 2017-09-18 git1.rmd
im so tired

---
title: "Module II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
fontsize: 11pt
urlcolor: blue
---

# Main case

## Trigger 1

> 30s M distressed, bending over around abdomen, speaks no English

Anatomical systems causing abdo pain: (everything can)

-  CVS: ruptured AAA, MI
-  GIT: oesophagus, stomach, bile ducts, liver, large intestine, pancreas, spleen, small intestine, *peritoneum*
-  Neuro: radicular pain
-  MSK: rectus abd
-  Resp: 
-  Renal: nephrolithiasis, pyelo
-  Repro: testicular torsion
-  Endocrine:

DDx: (consider his age, is he YOPI?)

\includegraphics{images/git1-ddx-abdo.png}

GIT usually present with one of 4 signs:

1.  Abdo/chest pain
2.  Altered ingesting of food (from nausea, vomiting, dysphagia, odynophagia [painful swallowing], anorexia)
3.  ALtered bowel movements (diarrhoea or constipation)
4.  GI bleeding

Pathological processes: inflamm, neoplasm, ischaemia, obstruction, infn, haemorrhage, distension, chemical irritation (toxicity), movement, pressure

Visceral vs somatic pain:

-  Somatic = structures supplied by spinal nerves, includes the parietal pleura and peritoneum. Thus perforated duodenal ulcer and generalised peritonitis can cause diffuse abdo pain (T6-T12). 
-  Visceral: sympathetic component of the ANS (rami communicans => sympathetic chain => sphlancnic nerves [preganglionic efferent stimulatory signals; greater 59, lesser 1011, least 12, lumbar 12] => prevertebral vascular ganglia and plexues => postganglionic efferent fibres follow individual arteries to viscus

|                       | Somatic                                    | Visceral                                                       |
|-----------------------|--------------------------------------------|----------------------------------------------------------------|
| Origin                | Parietal peritoneum/adjacent musculature   | Viscera                                                        |
| Pain pathway          | Spinal nerves: dermatomes and myotomes     | Sympathetic NS                                                 |
| Perception            | Localised to region                        | Poorly localised or diffuse, often midline, distant  site      |
| Stimuli               | Inflamm, laceration, heat, chem irritation | Distension, ischaemia, inflamm                                 |
| Accompanying features | Local muscle spasm (guarding)              | Nausea, vomiting, tachycardia, $\Uparrow$BP, sweating, anxiety |

Location of abdo pain:

\includegraphics{images/git1-location-abdo.png}

Hx:

-  HPC: nature of the pain, how long it lasts, where is it located,...
-  GIT: $\Delta${Appetite, weight (malabsorption, malignancy, diabetes, thyroid, IBD)}, dysphagia/odynophagia, indigestion or heartburn, nausea or emesis, haematemesis (vomiting blood), abdo pain, jaundice, bowel habits, per rectal blood
-  PMHx, PSHx
-  Medications
-  Social history
-  FHx
-  Red flags in acute abdo pain: sudden onset pain, increasing severity, (pre-)syncope, emesis/haematemesis, abdo distension, pallour and diaphoresis, tachycardia or AF, hypotension, fever, rebound tenderness, decreased urine output, positive pregnancy test

## Trigger 2

> Afghan refugee, 3 years refugee camp with poor diet and unsanitary conditions, massive psychological and physical trauma, heavy smoker, does not drink alcohol. Abdo pain first occured 4/52 and recurring, always upper abdo, deep gnawing pain relieved by food, occasionally wakes at night, similar to something 3 years ago. Responsive to antacids, no Hx of diarrhoea or vomiting, black foul-smelling stool about 1/52, no Hx of NSAID usage.

Most common causes of recurrent epigastric pain: GORD and peptic ulcer disease (duodenal and gastric; can't separate based on Hx alone): gnawing and burning sensation, occuring after meals gastric (shortly after) and duodenal (2-3 hours afterwards)

Antacids: contain alkaline ions that neutralise the stomach acid

Normal: balance between gastric acid secretion and gastroduodenal defense (mucus and bicarbonate secretion, ion pumps to remove excess H+, tight junctions, cellular restitution, epithelial renewal). Patho: agressive factors such as NSAIDs or H pylori, alcohol, bile salts, acid, pepsin can alter this balance by allowing back diffusion of H+ => epithelial damage.

Gastric acid: important for digestion of proteins. From parietal cells which are regulated by

1.  Cephalic phase: sight, smell, taste, thoughts => medulla => vagal ouput => submucosal plexus => increases secretion from gastric glands and stomach motility
2.  Gastric phase: food reaching stomach => stretch and chemoreceptors => secretion of gastrin by G cells
3.  Intestinal phase: stimulation of intestinal receptors => secretion of duodenal hormones: *secretin* reducers gastric secretion, cholecystokinin inhibits gastric emptying

NSAIDs: Non-selective can damage the gastric and duodenal mucosal barrier and are associated with an increased risk of upper gastrointestinal ulceration, bleeding and perforation.

## Trigger 3

Ix:

-  Bloods: FBC, CHEM20, amylase, lipase
-  Endoscopy

Upper gastro intestinal endoscopy indicated: signs of bleeding, weight loss, vomiting, dysphagia, progression of symptoms, >55, FHx of gastrointestinal carcinoma. Helps rule out carcinoma, visualise ulcer, determine if bleeding

## Trigger 4

## Trigger 5

13C urea: high sensitivity and specificity, requires mass spec for Dx of *H. pylori*

# Short case 1

# Short case 2

### 2017-09-18 git1.rmd
should be finished

@@
 
-Normal: balance between gastric acid secretion and gastroduodenal defense (mucus and bicarbonate secretion, ion pumps to remove excess H+, tight junctions, cellular restitution, epithelial renewal).
-Patho: agressive factors such as NSAIDs or H pylori, alcohol, bile salts, acid, pepsin can alter this balance by allowing back diffusion of H+ => epithelial damage.
+Normal: balance between gastric acid secretion and gastroduodenal defense (mucus and bicarbonate secretion, ion pumps to remove excess H+ *intracellularly*, tight junctions, cellular restitution, epithelial renewal), restitution where healthy cells migrate to epithelium.
+Patho: agressive factors such as NSAIDs or *H.
+pylori*, alcohol, bile salts, acid, pepsin can alter this balance by allowing back diffusion of H+ => epithelial damage.
+Smoking convers increased risk of gastric ulcer
 
@@
 
+Ex: (image not attached)
+
+-  Intro
+-  General inspection: mental state (withdrawal symptoms), colour (jaundice, pallour, bronzed skin), hydration
+-  Inspect hands and upper limb:clubbing, nail changes (iron deficiency, chronic liver disease), hepatic flap (asterixis from hepatic encephalopathy or CO2 retention or renal failure)
+-  Radial pulse, RR, BP
+-  Face inspection: eyes, glands, mouth
+-  Neck lymph nodes
+-  Inspection chest and abdomen: chest, abdomen = {masses/distension, stretch marks, colour changes, prominent veins, pulsations, scars}
+-  Palpate the abdomen: 9 locations, look for tenderness at McBurneys point (acute appendicitis), Rovsing's sign is rebound tenderness (localised peritonism), Murphy's sign is tenderness under right costal margin during inspiration (acute cholecystitis), Courvoisier's law is palpable non-tender gallbladder under right costal marging + pt jaundiced (gallstones unlikely so more likely cancer of head of pancreas)
+-  Liver Ex: palpate, percuss
+-  Spleen: palpate
+-  Kidney: palpate
+-  Abdo aorta
+-  Ascites
+-  Auscultate bowel signs: active gurgling signs called borborygmi (diarrhoeal states), shake pt and here succussion splash which is liquid+air within luminal organ (normal if eaten, path if pt is nil bu mouth for >6 hours and clear fluids >2 hours), auscultate for bruits
+-  Lower limbs: inspection and palpation
+
 ## Trigger 3
 
-Ix:
+Ix
 
 -  Bloods: FBC, CHEM20, amylase, lipase
--  Endoscopy
+-  ECG?
+-  *H. pylori tests*: 
+-  Digital rectal exam
+-  Endoscopy: preferred as ulcers may be malignant, also follow up to ensure healing
 
@@
 
+Peptic ulcers are defects in the gastric or duodenal mucosa that extend through the muscularis mucosa.
+Mucosal injury (+ PUD) occurs when balance is disrupted.
+Factors such as NSAIDs, *H.
+pylori*, alcohol, bile salts, acid, pepsin can alter the mucosal defence allowing *back diffusion* of H+ => epitheliel injury.
+*Acute* = no signs of fibrosis vs chronic, erosions do *not* penetrate the msucularis mucosae.
+
+*H.
+pylori*: gram-neg helix/spiral with multiple flagella => motility.
+Burrows and lives beneath the mucus layer adherent to epithelial cells.
+pH is close to neutral here and acidity is buffered by production of urease (from *H.
+pylori*).
+Exclusively colonises gastric-type epithelium and found in duodenum.
+*Chronic gastritis* from the local inflamm response, predominantly from bacterial cagA and vacA genes.
+cagA+ more likely to cause disease than cagA-.
+Triggers increased gastrin release from G cells => increased acid production by parietal cells => mostly does nothing, some people experience duodenal ulceration.
+
+cont: 1% infected leads to pangastritis => gastric atrophy and hypochlorhydria => other bacteria can proliferate in stomach => secrete mutagenic nitrites => predisposing gastric cancer
+
+Detecting *H. pylori*
+
+-  Urease test: during gastroscopy mucosa is biopsied and placed in indicator, will be less acidic if urea is produced
+-  Serology: lacks specificity, cannot d/dx past infn
+-  13C urea: high sensitivity and specificity, requires mass spec for Dx of *H. pylori*
+-  Faecal antigen test: cheap specific, acceptability is a disadvantage
+-  Histology, rapid urease tests, microbiological culture [all invasive]
+
+Mx: aim is to relieve symptoms, induce healing, and prevent recurrence.
+
+-  All pt with proven ulcers and H.
+pylori positie have eradication as primary therapy.
+PPI taken simultaneously with two antibiotics (from amoxicillin, clarithromycin and metronidazole) for 7 days.
+Success in 80-90% of people, depends on antibiotic resistance, compliance due to side-effects (diarrhoea common, flushing and vomiting, nausea, abdo cramps, headache, rash).
+-  General: stop cigarettes, aspiring, and NSAIDs
+-  Surgery if indicated: persistent ulceration, recurrent ulceration, gastric outflow obstruction, perforation/haemorrhage
+
 ## Trigger 5
 
-13C urea: high sensitivity and specificity, requires mass spec for Dx of *H. pylori*
+Drugs
+
+-  PPI
+    -  Esomeprazole: blocks H+ secretion so stomach pH is increased, reducing irritation of stomach epithelium, allowing healing
+-  Antibiotics
+    -  Amoxycillin: 3rd generation aminopenicillin, $\beta$-lactam which inhibits cell wall synthesis. 
+    -  Clarithromycin: macrolide so it is a 50s inhibitor, inhibiting bacterial protein synthesis
+
+Other medications for dyspepsia:
+
+-  Empiric PPI
+-  Antidepressesants (refractory functional dyspepsia)
+-  Prokinetics: 
+-  Antibiotics
+-  Psychological treatments: classic psyc experimental design (i.e. shit)
+
+## Reference
+
+-  [random article](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3002574/)
+-  Rest were the three asterixed readings from the week 11 reading list
 

### 2017-09-19 git1.rmd
mc

@@
 ---
-title: "Module II"
+title: "Gastrointestinal I"
 author: "Aaron Khoo"
@@
 -  MSK: rectus abd
--  Resp: 
+-  Resp: pneumonia can sometimes present
 -  Renal: nephrolithiasis, pyelo
@@
 2.  Altered ingesting of food (from nausea, vomiting, dysphagia, odynophagia [painful swallowing], anorexia)
-3.  ALtered bowel movements (diarrhoea or constipation)
+3.  Altered bowel movements (diarrhoea or constipation)
 4.  GI bleeding
@@
 
--  Somatic = structures supplied by spinal nerves, includes the parietal pleura and peritoneum. Thus perforated duodenal ulcer and generalised peritonitis can cause diffuse abdo pain (T6-T12). 
+-  Somatic: structures supplied by spinal nerves, includes the parietal pleura and peritoneum. Thus perforated duodenal ulcer and generalised peritonitis can cause diffuse abdo pain (T6-T12). 
 -  Visceral: sympathetic component of the ANS (rami communicans => sympathetic chain => sphlancnic nerves [preganglionic efferent stimulatory signals; greater 59, lesser 1011, least 12, lumbar 12] => prevertebral vascular ganglia and plexues => postganglionic efferent fibres follow individual arteries to viscus
+-  Example of visceral => somatic: appendicitis starts off as general then localises as it is visceral then starts affecting peritoneum (somatic pain)
 
@@
 
+*Look at patient, not translator during Hx taking*
+
 ## Trigger 2
@@
 Responsive to antacids, no Hx of diarrhoea or vomiting, black foul-smelling stool about 1/52, no Hx of NSAID usage.
+
+Pancreatitis: usually caused by heavy drinking
+
+Black foul smelling stool: malaena from blood
 
@@
 pylori*, alcohol, bile salts, acid, pepsin can alter this balance by allowing back diffusion of H+ => epithelial damage.
-Smoking convers increased risk of gastric ulcer
+Smoking confers increased risk of gastric ulcer
 
@@
 2.  Gastric phase: food reaching stomach => stretch and chemoreceptors => secretion of gastrin by G cells
-3.  Intestinal phase: stimulation of intestinal receptors => secretion of duodenal hormones: *secretin* reducers gastric secretion, cholecystokinin inhibits gastric emptying
+3.  Intestinal phase: stimulation of intestinal receptors => secretion of duodenal hormones: *secretin* reduces gastric secretion, cholecystokinin inhibits gastric emptying
 
 NSAIDs: Non-selective can damage the gastric and duodenal mucosal barrier and are associated with an increased risk of upper gastrointestinal ulceration, bleeding and perforation.
+Prostaglandins => decrease gastric acid
+
+Melaena: black malodourous stool is suggestive of upper GI bleeding
 
@@
 -  Palpate the abdomen: 9 locations, look for tenderness at McBurneys point (acute appendicitis), Rovsing's sign is rebound tenderness (localised peritonism), Murphy's sign is tenderness under right costal margin during inspiration (acute cholecystitis), Courvoisier's law is palpable non-tender gallbladder under right costal marging + pt jaundiced (gallstones unlikely so more likely cancer of head of pancreas)
+    -  Use stethoscope if you wanna try get rid of guarding
 -  Liver Ex: palpate, percuss
@@
 
+He doesn't look cachexic (malignancy, malabsorptive)
+
 Ix
@@
 -  Endoscopy: preferred as ulcers may be malignant, also follow up to ensure healing
+-  USS: more likely if he has cholecystitis or biliary colic
 
@@
 Factors such as NSAIDs, *H.
-pylori*, alcohol, bile salts, acid, pepsin can alter the mucosal defence allowing *back diffusion* of H+ => epitheliel injury.
-*Acute* = no signs of fibrosis vs chronic, erosions do *not* penetrate the msucularis mucosae.
+pylori*, alcohol, bile salts, acid, pepsin can alter the mucosal defence allowing *back diffusion* of H+ => epithelial injury.
+*Acute* = no signs of fibrosis vs chronic, erosions do *not* penetrate the muscularis mucosae.
 
@@
 -  All pt with proven ulcers and H.
-pylori positie have eradication as primary therapy.
+pylori positive have eradication as primary therapy.
 PPI taken simultaneously with two antibiotics (from amoxicillin, clarithromycin and metronidazole) for 7 days.
 Success in 80-90% of people, depends on antibiotic resistance, compliance due to side-effects (diarrhoea common, flushing and vomiting, nausea, abdo cramps, headache, rash).
--  General: stop cigarettes, aspiring, and NSAIDs
+-  General: stop cigarettes, aspirin, and NSAIDs
 -  Surgery if indicated: persistent ulceration, recurrent ulceration, gastric outflow obstruction, perforation/haemorrhage
@@
 -  PPI
-    -  Esomeprazole: blocks H+ secretion so stomach pH is increased, reducing irritation of stomach epithelium, allowing healing
+    -  Esomeprazole: blocks H+ secretion so stomach pH is increased, reducing irritation of stomach epithelium, allowing healing. Irreversible blockage of H/K ATP-ase
 -  Antibiotics

### 2017-09-19 git1-discharge-letter.rmd
completed letter

---
output: pdf_document
documentclass: article
classoption: a4paper
fontsize: 11pt
---

\hfill `Sys.Date()`

\hfill `random(hospital)`

Dear GP,

Thank you for continuing the care of Abdul Rauf,

He is a 34 year old male who was admitted to the ED at metropolitan hospital following severe deep gnawing epigastric pain. Hx: pain first started approximately 4/52 prior and recurred intermittently, relieved by food, and occasionally waking him at night. It is similar to a previous pain 3 years ago which responded well to antacids, and he passed a likely melaena 1/52. He is a recent refugee from Afghanistan. He is a heavy smoker, does not drink alcohol, and there is no history of NSAID usage.

Ex revealed an anxious but otherwise healthy looking man. There was slight tenderness in the epigastrium with voluntary guarding, but an otherwise unremarkable GIT exam with no organomegaly and normal bowel sounds. There was also normal vitals and CVS exam.

On Ix, gastroscopy confirmed a *duodenal ulcer* with no perforation, and chronic gastritis in the stomach antrum. The urease test was positive for *H. pylori*.

He was diagnosed with peptic ulcer disease and treated on esomeprazole, amoxycillin, and clarithromycin for 7 days.

Plan:

We refer him to you for follow up as well as future long term care. We recommend ^13^C urea breath test in 1/12 to confirm *H. pylori* eradication, in addition to regular follow up. He speaks no English, so to aid in finding an interpreter, he is fluent in Afghan Dari and Pashto.

Kind regards,

Name (Position)

### 2017-09-21 git1.rmd

@@
 
+## Trigger 1
+
+DDx for retrosternal pain:
+
+-  CVS: pericarditis, angina, aortic dissection
+-  Resp: pneumonia
+-  MSK: trauma, costochondritis, oesophageus muscle spasms, hiatal hernia
+-  GIT: GORD, cholecystitis, peptic ulcer disease, gastritis, infn, eosinophilic oesophagitis (not very common; allergy)
+-  Cancers
+
+Hx:
+
+-  HPC: trouble eating/swallowing, anything making worse/better, SOB/emesis/coughing/wheezing/palpitations/syncope, does the pain ever come on with exercise? recent illness
+-  Systems: cardiac history
+-  Any medications/alcohol/smooking/drug
+-  Fam Hx of gastric, oesophageal cancers, CVS history
+
+## Trigger 2
+
+Dysphagia for solids: could be from lack of saliva or from a mass. Much more concerned if dysphagia for liquids.
+
+Red flags: dysphagia, weight loss, pain, emesis, age > 50, fam Hx, resp symptoms such as wheezing hoarse voice, evidence of GIT bleeding, anaemia, unresponsive to PPI: signal stricture/Ca/..
+
+Weight loss of 10% over 3/12. If someone has "unintentional weight loss" => bloods + CRP. "How do your clothes fit?" since not many patients weigh themselves.
+
+Ix:
+
+-  Bloods: FBC (anaemia), ELFT
+-  ECG
+-  Endoscopy: due to dysphagia
+-  13C urea breath test for *H. pylori* if weren't gonna endoscopy
+
+## Trigger 3
+
+GORD pathophys:
+
+Movement of stomach juice into oesophagus. Three main pathways:
+
+1.  Transient sphincter relaxations
+2.  Hypotensive lower oesophageal sphincter: smoking and alcohol affect this
+3.  Anatomical disruption: e.g. hiatal hernia
+
+Reflux => damages epithelium => reflux oesophagitis => scarring => predisposition for further reflux as the sphincter keeps weakening => granulocytes and eosinophils => more likely for sphincter incompentence
+
+Risk factors: smoking, dry mouth, obesity, diabetes, foods (carbonated, peppermint, fatty foods, chocolate)
+
+PPI: luminal surface of parietal cells H/K-ATPase which are irreversibly inhibited (lasts 24-48h)
+
+## Trigger 4
+
+Oropharyngeal dysphagia: difficulty initiating.  
+Oesophageal dysphagia: difficulty a couple seconds after starting and then things get stuck.
+
+Oesophageal dysphagia causes:
+
+-  Intrinsic: eosinophilic oesophagitis, oesophageal muscle spasms, nervous damage
+-  External: compression by tumour, lymphadenopathy/thyroid, radiation
+
+Barrett's oesophagus: metaplasia of squamous => columnar in lower oesophagus due to damage (produce mucus in response to acid).
+Massive predisposition to adenocarcinoma.
+Some patients get barrett's but just have to monitor it for progression to cancer.
+
+Oesophageal stricture due to scarring.
+
 # Short case 2
+
+## Trigger 1
+
+Epigastric: think of foregut structures: oesophagus => 1st part of duodenum, + liver + gallbladder + splenic
+
+DDx:
+
+-  GIT: GORD, PUD, IBD
+-  Psych: anorexia, bullimia
+-  Cancer: colon, oesophageal, gastric, pancreas
+
+Initial Ix:
+
+-  Bloods: FBC (anaemic), ELFT (), ESR/CRP (inflamm), if lots of bleeding/bruises (COAG)
+-  Endoscopy
+
+## Trigger 2
+
+Chronic disease: lead to anaemia
+
+## Trigger 4
+
+Gastric Ca
+
+-  Types: adeno (most) = {intestinal (chronic inflamm of mucosa => metaplasia => neoplasia; most common type), diffuse (normal mucosa, not assicated with peptic ulcers or *H.
+pylori*)}, lymphoma also associated with *H.
+pylori*
+-  Staging: TNM
+-  Prognosis: poor
+-  Mx: resection, could put a stent stopping it from not emptying, laser ablation

### 2017-09-25 git2.rmd
readings lol

---
title: "Gastrointestinal II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
fontsize: 11pt
urlcolor: blue
---

# Readings

[Explanation of diff git symptoms/signs](http://accessmedicine.mhmedical.com.ezproxy.library.uq.edu.au/content.aspx?bookid=1585&sectionid=96308009)

## Coeliacs

Coeliacs: gluten => gliadin (in the stomach; resists breakdown) => reaches the SI then binds to TFR (transferrin receptor; usually used to absorb iron, may be overexpressed) => transcytosed across enterocyte => lamina propia => gliadin is deamidated then phagocytosed by macrophages => macros are usually doing this for normal food anyway; abnormal reaction in coeliacs => Th2 (CD4+) => inflamm cytokines such as interferon-$\gamma$ and TNF-$\alpha$ => direct damage to epithelial cells in SI + killer CD8 t-cells stimulated => bad. 
Ab blood tests: anti-tTG (deamidates the gliadin) and anti-endomysial (EMAs). 
Symptoms = {children = [abd distension, failure to thrive, diarrhoea], adults = [chronic diarrhoea, bloating (distension),...]}. 
Tl;dr autoimmune.

## Diarrhoea

Diarrhoea: arises from four pathological processes

1.  Osmotic
2.  Secretory
3.  Altered motility
4.  Exudative/inflamm

Classified into chronic vs acute ($\leq2\text{week}$). 
Acute is usually infectious. 
Diarrhoea is bowel movements excessive in volume, frequency, or liquidity. 
It then interferes with nutrition. 
Causes can be infectiours or non-inf (i.e. drugs). Malabsorption: consistency reflects high nutrient and bacterial fermentation being bulky, pasty, or liquid, and tend to float in toilets (due to fats).

### Acute

Inflamm vs noninflamm

-  Noninflamm: watery, mild, self limited, Dx evaluation limited to pt with severe or persistent (>7 days) diarrhoea
-  Inflamm: blood or pus/fever, caused by invasive or toxin-producing bacterium, Dx requires routine stool bacterial cultures and testing as indicated for *C. difficile* toxin and ova and parasite

Mx:

\includegraphics{images/git2-acute-diarrhoea.png}

Rx:

-  Diet
-  Rehydration
-  Antidiarrhoeal agents: opioids, bismuth subsalicylate
-  Antibiotic therapy

# Main case

# Short case 1

# Short case 2

### 2017-09-26 git2.rmd
mc

@@
 
+## Trigger 1
+
+DDx:
+
+-  Motility: hyperthyroidism (hyper) or scleroderma (hypo)
+-  Osmotic: improves with fasting
+-  Exudative: damage to mucosa, e.g. shigella infn
+-  Secretory: cholera
+
+Most common cause in his age: infectious; chronic so thinking of some things like parasites, or immunocompromised
+
+Fluids: 8-9L per day, only 1L is from intake whereas rest is from salivary glands,... 
+SI = nutrients, LI = water
+
+Hx:
+
+-  HPC: onset, time course (how often, when, has it changed), consistency, how much he produces, how many wet nappies does he create, associated (systemicall unwell, any pain), was there a precipitating illness => antibiotics, who he comes into contact with (indigenous communities are a lot more communal)
+-  PMHx: vaccinations
+
+## Trigger 2
+
+Poor diet: can by itself cause diarrhoea, e.g. full fat diet
+
+## Trigger 3
+
+Pale conjunctiva and nail beds: anaemia either from nutrition (Fe or Vit B12) or haemorrhage
+
+Foul smelling, pale soft fluffy stool: malabsorptive. Smelly due to fermentation of the fats in it
+
+Ix:
+
+-  Bloods: FBC, ELFT, cultures (if he was febrile)
+-  Stool sample
+
+## Trigger 4
+
+Cysts: not 100% specific or sensitive
+
+Giardia:
+
+-  Flagellated protozoa
+-  Fecal-oral transmission
+-  Proximal small bowel
+-  Incredibly widespread: 70-80% of kids pass cysts but are asymptomatic
+
+Anaemic: iron deficiency is microcytic, B12/folate are macrocytic |
+
+-  Microcytic = {T = thalassaemia, I = iron deficiency, C = chronic disease, S = sideroblastic anaemia}
+-  Macrocytic = {A = alcohol (cirrhosis), B = B12 deficiency, C = compensatory reticulocytosis (blood loss and hemolysis), D = drug/dysplasia, E = endocrine (hypothyroidism), F = folate deficiency/foetus (pregnancy)}
+
+## Trigger 5
+
+Metronidazole: inhibits nucleic acid synthesis so it affects both bacteria and protozoa
+
+Long term consequences of parasitic infn:
+
+-  Damage to gut mucosa
+-  Can migrate to subarachnoid space => meningitis/encephalitis
+-  Malnutrition
+-  Failure to thrive
+
 # Short case 1

### 2017-09-26 doc-render-md.sh
added some stuff

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+###############################################################
+####      Renders .md file using R using knitr/pandoc      ####
+###############################################################
+
+# Difference from the other one
+#   This uses a .md file rather than a .rmd file. Main reason is because there are nice vim plugins for .md and these files are not heavy on R code
+
+
+# Dependencies
+#   R
+#   TeX (if you decide to output a pdf_document)
+#   R packages: knitr
+
+# To do
+#   Make a function so this thing isn't messy as hell
+#   Make the cleaning stuff functions so it is easier to examine
+
+read -p 'Enter the .md file you wish to render (without the extension): ' File
+
+if [[ -e "$File".md ]]; then
+    echo "Do you wish to clean the tmp .rmd file? "
+    select yn in "Yes" "No" "Cancel"; do
+        case $yn in
+            Yes ) cp "$File".md tmp.rmd; sed -i 's/=>/\$\\rightarrow\$/g' tmp.rmd; R -e 'rmarkdown::render("tmp.rmd")'; rm tmp.rmd; mv tmp.pdf "$File".pdf; break;;
+            No ) cp "$File".md tmp.rmd; R -e 'rmarkdown::render("tmp.rmd")'; rm tmp.rmd; mv tmp.pdf "$File".pdf; break;;
+            Cancel ) exit;;
+        esac
+    done
+    echo "Succesfully knitted the document (hopefully)"
+else
+    echo "File does not exist"
+fi
+

### 2017-09-28 git2.rmd
sc

@@
 
-Coeliacs: gluten => gliadin (in the stomach; resists breakdown) => reaches the SI then binds to TFR (transferrin receptor; usually used to absorb iron, may be overexpressed) => transcytosed across enterocyte => lamina propia => gliadin is deamidated then phagocytosed by macrophages => macros are usually doing this for normal food anyway; abnormal reaction in coeliacs => Th2 (CD4+) => inflamm cytokines such as interferon-$\gamma$ and TNF-$\alpha$ => direct damage to epithelial cells in SI + killer CD8 t-cells stimulated => bad.
+Epidemiology: caucasians of european ancestry.
+
+Coeliacs: gluten => gliadin (in the stomach; resists breakdown from gastric enzymes) => reaches the SI then binds to TFR (transferrin receptor; usually used to absorb iron, may be overexpressed) => transcytosed across enterocyte => lamina propia => gliadin is deamidated then phagocytosed by macrophages => macros are usually doing this for normal food anyway; abnormal reaction in coeliacs => Th2 (CD4+) => inflamm cytokines such as interferon-$\gamma$ and TNF-$\alpha$ => direct damage to epithelial cells in SI + killer CD8 t-cells stimulated => bad.
 Ab blood tests: anti-tTG (deamidates the gliadin) and anti-endomysial (EMAs). 
@@
 -  Macrocytic = {A = alcohol (cirrhosis), B = B12 deficiency, C = compensatory reticulocytosis (blood loss and hemolysis), D = drug/dysplasia, E = endocrine (hypothyroidism), F = folate deficiency/foetus (pregnancy)}
+-  RACH, BLOOD?
 
@@
 
+## Trigger 1
+
+DDx:
+
+-  Repro/gynae: pregnancy, ovarian cysts/cancer, fibromatosis
+-  GIT: hernia, malabsorption (lactose intolerance => osmotic diarrhoea), coeliacs, IBD, infective, peptic ulcer (sequelae include cancers)
+-  Psych: depression, chronic stress
+-  Systemic: anaemia (Fe deficiency can lead to diarrhoea)
+-  Endocrine: hypothyroidism, diabetic ketoacidosis
+
+Ex:
+
+-  Inspection: scarring, swelling, asymmetry, skin changes, guarding, general discomfort, also inspect eyes for jaundice/anaemia
+-  Palpation: masses (firm/fixed/flatulent), organs, belot (wiggle) the kideys, for pain
+-  Auscultate
+-  Percussion: if considering ascites
+-  Rectal
+
+## Trigger 2
+
+Aphthous ulcers: benign mouth ulcers associated with systemic stuff like coeliacs
+
+Ix:
+
+-  Bloods: FBC (Hb = anaemia), thyroid fn test, HLA typing, B12, ELFT
+-  Stool sample
+-  Tissue typing: ?
+-  Discuss possibility for endoscopy: skip lesion things like Crohns; wait for bloods  maybe
+
+## Trigger 3
+
+Ix: suggest of coeliacs disease (systemic autoimmune disease). IgA-tTG: deamidating gluten products
+
+Further Ix:
+
+-  Iron studies
+-  Gastroscopy and biopsy
+-  Diet
+
+## Trigger 4
+
+Coeliacs associated with FHx of thyroid issues and T1D, Trisomy-21, IBD, GORD
+
+Complications: bloating, chronic diarrhoea, nutritional deficiency, increased risk of malignancy, higher motility
+
+Mx:
+
+-  Gluten free diet
+-  Frequent monitoring of coeliacs and systemic nutrient levels
+
+Gluten found in wheat, rye, HLA-DQ2 (95%) and HLA-DQ-8. Can present at any age and common (1%). Learn the histopath => GIT systems + diverse systemic symptoms
+
 # Short case 2
+
+## Trigger 1
+
+DDx:
+
+-  Hernias: read medscape article
+-  Cancer
+
+Hx:
+
+-  Reducible/irreducible, strangulated/incarcerated (soft/hard)
+-  Fluctuants: cyst can look like hernia
+-  Worse
+
+Lump:
+
+-  Fixed is worse than mobile
+
+## Trigger 2
+
+Mx: surgery
+
+Natural Hx: will only get worse
+
+## Trigger 3
+
+The hernia became strangulated => ischamia => bowel necrosis
+
+Medical emergency
+
+Vomiting:
+
+-  Pain
+-  Bowel obstruction

### 2017-10-03 git3.rmd
mc

---
title: "Gastrointestinal III"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
fontsize: 11pt
urlcolor: blue
---

# Main case

## Trigger 1

DDx: upper vs lower, stuff higher up => melaena

-  Haemorrhoids: varicose vein of perianal region
-  Colitis: ischaemic, inflamm, infn
-  CRC
-  IBD: ulcerative colitis + crohn's disease (less common for blood)
-  Trauma
-  AVM
-  Diverticulitis
-  Fissure
-  Polyps/FAP
-  Arterioenteric fistula
-  Peptic ulcer

Hx:

-  HPC: onset, frequency, anything worse/better, change in diet, 
    -  Info about poo: offensive odour (fat/mucous), consistency (liquid/solid), watery, colour, volume
    -  How frequent is blood
-  Associated symptoms: pain, nausea/vomiting, systemic weight loss, fatigue, bleeding/bruising
-  PMHx/PSHx
-  FHx
-  Social Hx

## Trigger 2

Constipation: usually haemorrhoids or anal fissure especially in 27M. 27M shouldn't really be anaemic. Even in F shouldn't just say "menstruation"

Types of diarrhoea: (todo: whip up a table for these types)

-  Osmotic: retention of water in the bowel from accumulation of non-absorbable substances
-  Secretory: electrolyte imbalance => water diffuses into stool. Often from inflamm
-  Motility: changed motility from hyperthyroidism, IBS, partial gastrectomy
-  Inflammatory: colon is inflamed, typically causes blood diarrhoea, e.g. IBD
-  (Malabsorptive: stool will float due to fats)

This pt: exudative (blood and mucus) + inflammatory. Not malabsorptive as his stool

Mucous in GIT: secreted for protection. Pus (secreted from inflamed). IBS and IBD

Associated pain to rectal bleeding:

-  Rectal: fissures
-  Abdominal:
    -  Severe: ischaemic colitis
    -  Mild to moderate: enteritis

Loperamide: opiod receptor agonist that doesn't cross BBB, decreasing activity of myenteric plexus

## Trigger 3

Hospital admission: think holistically, this pt is young adult male who has a partner who can look after him, chance that he'll have to go to hospital at night => yes

## Trigger 4

Bloods: urea elevated due to dehydration; albumin elevated due to chronic bleed; Hb due to anaemia, but MCV normal so probs not iron deficiency(?); CRP+WCC due to inflamm

Sigmoidoscopy: uses shorter instrument which can be done a lot faster

Todo: Copy a table between UC and CD from Don's thing

Natural Hx IBD: extraintestinal symptoms such as eyes, => complications such as toxic megacolon,

Toxic megacolon: inflamm => dilation of bowel => increased permeability + perforation => $\Uparrow$ septic

Mx: principles start aggressive => less aggressive

-  Short term: steroids
-  Long term: immunosuppressants, resection for UC

## Trigger 5

IBD associated

-  Skin: psorosis
-  Jt: seroneg arthritis, ank spond
-  Liver: primary sclerosin cholangitis
-  Thromboses, metabolic bone diseases

Regular monitoring: if the disease is being managed, liver damage, females: pregnant and many drugs are teratogenic, common malignancy in immunosuppressed are skin, vaccines

# Short case 1

# Short case 2

### 2017-10-05 git3.rmd
sc

@@
 
+## Trigger 1
+
+[Haemorrhoids](http://emedicine.medscape.com/article/775407-overview): *common* disease, varicose veins of perianal region. 
+Causes = {pregnancy, constipation, obesity, portal hypertension, tumours}. 
+Present: usually painless unless thrombosis, bleeding, "feel something when wiping". Rx: usually conservative (diet, stool softening), surgery: rubber band resection
+
+-  Internal: above pectinate line, superior rectal
+-  External: below pectinate line, inferior rectal, usually painful and usually thrombosed
+
+Causes of GIT bleeding: split into upper, lower, anal (see: lecture 3)
+
+## Trigger 2
+
+Bowel cancer:
+
+-  Left sided: constipation and is found earlier, blood on top of the poo (as poo has already formed)
+-  Right sided: more liquidy so doesn't cause constipation, occult bleeding (blood mixed in)
+
+Pain on defaecation: think haemorrhoids (thrombosed), anal fissure, anal polyps, abscesses, rectal ulcers, rectal cancer
+
+[Laxatives](https://www.nps.org.au/australian-prescriber/articles/managing-constipation-in-adults):
+
+-  Osmotic: indigestible carbohydrates => pass through => relative osmotic diarrhoea to soften stool
+-  Bulking agents 
+-  Stool softeners 
+-  Stimulants
+-  Lubricants
+
+Diverticulosis: common, L in West, R in Asia, increases with age. 
+Most are asymptomatic (70-80%), diverticular bleeding (outpouch with the vasa recta which is the weak part => bleeding not uncommon) (5-15%), diverticulitis (4%) = {simple (85%), complicated (15%)}
+
+-  True: congenital, e.g. meckels
+-  False: not all the layers, usually around vasa recta as that is the weakest, e.g. zenkers
+
+## Trigger 3
+
+Top *X* differentials, where $X=\in\mathbb{Z}$:
+
+-  CRC: due to bleeding, and altered bowel habits
+-  Diverticulitis/diverticulosis
+-  Haemorrhoids
+
+Explaining: emphasise importance and that it can cure things inside it
+
+## Trigger 4
+
+Further Ix:
+
+-  Bloods: FBC (anaemia), ELFT, CRP/ESR, group+hold?, CEA (specific marker for Ca)
+-  CT for staging
+
+Risk factors for adenocarcinoma:
+
+-  Smoking
+-  Diet
+-  Obesity
+-  Genetic risk: FAP, lynch syndrome
+
+Presentation: bleeding, abdo pain, tenesmus (feeling of incomplete emptying), anaemia (right), complete bowel obstruction,
+
+Screening: important because catching it early has high survival rates
+
 # Short case 2
+
+## Trigger 1
+
+DDx:
+
+-  CVS: ruptured AAA => ischaemia of bowel, vasculitis
+-  Renal: UTI + symptoms of cephalexin, pyelonephritis
+-  GIT: ischaemic bowel, perforation, ulcer, CRC, diverticulitis, anal fissure, haemorrhoids, fistula, abscess, infectious colitis (*C. difficile* because in nursing home; everyone would get it)
+-  Genitourinary: pelvic inflamm disease
+
+Compared to Trevor (IBD, CRC), she is Old F in nursing home on antibiotics: infectious colitides, CRC (less likely presentation as her thing is so acute), diverticulitis, ischaemic bowel
+
+Mechanisms that antibiotics causing diarrhoea:
+
+-  Inflammatory: *C. difficile* secrete toxins
+-  Secretory: damage mucosa
+-  Osmotic: disturb carbohydrate balance
+-  Motility: increase gut motility
+
+## Trigger 2
+
+She's a vasculopath
+
+*Triple whammy for renal failure: ACE, diuretic, NSAID (aspirin)*
+
+Ex:
+
+-  CVS: important to check JVP
+-  GIT: 5 Fs
+
+## Trigger 3
+
+Rigidity, guarding, rebound tenderness (pain after moving hand away after slowly palpating): suggests peritonitis/inflammation
+
+Ix:
+
+-  Bloods: FBC (anaemia), ELFT (kidney fn; she is on triple whammy for renal failure), ESR/CRP
+-  ECG
+-  Urine: urinalysis, dipstick
+-  Stool: cultures (hard to grow *C. difficile*)
+-  Imaging: CT (ischaemia less enhanced), angiograms
+-  Colonoscopy
+
+## Trigger 4
+
+Ischaemic colitis (pain after eating; basically angina in GIT): global hypoperfusion => ischaemia. 
+Can also be caused by strangulation, embolism, infn. 
+Hers is an *acute on chronic* presentation.
+
+Ischaemic colitis: chronic issue, usually resolves on its own, Rx: bowel rest, broad spectrum IV antibiotics
+Acute mesenteric colitis: medical emergency

### 2018-02-22 endo2-sc1-ioslides.rmd
added short case - ioslides presentation

---
title: 'Endocrine 2: Short Case 1'
author: "Aaron Khoo"
date: "22 February 2018"
output: 
  ioslides_presentation:
    incremental: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PC

**(trigger 1) Sharon Leyton, a 32-year old woman, presents to you, her GP, with earache. After taking a history and examining her ear you diagnose acute otitis media, likely from her recent URTI. As you explain these findings you also notice a lump in her anterior neck that moves when she swallows.**

-  What is the significance of Sharon’s lump moving upon swallowing?
-  Review anatomy of the area

## Moving lump example

<iframe width="560" height="315" src="https://www.youtube.com/embed/ZLO6h8RsRSw" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

## Anat

<a href="https://i.imgur.com/Jzl8vrV.png">
  <img src="https://imgur.com/Jzl8vrVl.png" />
</a>

(from Talley and O'Connor)

## Anat (cont.)

<a href="https://i.imgur.com/TFVilav.png">
  <img src="https://imgur.com/TFVilavl.png" />
</a>

(from duckduckgo)

## Anat (cont.)

<a href="https://i.imgur.com/qVUMoCI.png">
  <img src="https://imgur.com/qVUMoCIl.png" />
</a>

## DDx (quickly)

Discuss possible hypotheses for a lump in this region (this is a good opportunity to consider the anatomy of the area!)

> - Superficial
>    -  Sebaceous cyst, lipoma (soft, lobulated, and not fixed to skin), abscess, dermoid cyst
> - Lymph nodes
>    -  Superficial and deep
> - Deep
>    -  Anterior triangle (moves on swallowing): thyroid and thyroglossal cyst (+ moves on tongue being poked out)
>    -  Ant. triangle (does not move on swallowing): salivary glands, branchial cyst, carotid body tumour, carotid aneurysm, sternomastoid "tumour" (appears only in babies and is result of trauma)
>    -  Post. triangle: cervical rib, subclavian artery aneurysm, pharyngeal pouch (may fill and empty following a meal), cystic hygroma (collection of dilated lymphatics presenting in infancy; compressible, soft)

## Hx

-  Returning to ‘history’ in light of this finding, what specific questions you would now want to ask Sharon?

- **(trigger 2) Sharon says that until you pointed it out, she had been unaware of any swelling in her neck. She says that she has been feeling anxious and stressed for some time, but she attributes this to having three young children and doing a PhD. She denies any weight loss, sweating, or palpitations. She has had no difficulties swallowing or any change in voice. There are no other significant symptoms and apart from the delivery of her three children and a miscarriage, she has no past medical or family history of note. She has never smoked and only rarely drinks alcohol.**

-  Do you understand the relevance of each of the specific questions that were asked? Were there any that you had not thought of when discussing the last trigger to note for the future?

## Ex

-  What are the general principles for examining a lump in any region? What is the significance of the specific features of the lump that you examine for? 
    -  Size, location, shape (well defined?), consistency (smooth/hard/rubbery/irregular/...), fluctuance (fluctuant suggests fluid filled lesion), trans-illumination, pulsatility (vascular origin), temperature (inflamm cause), overlying skin changes, relation to underlying/overlying tissues (does it move or not), auscultation
    -  [neck lump Ex](https://geekymedics.com/neck-lump-examination-osce-guide/)
-  Discuss what further examination you would perform on Sharon.

-  **(trigger 2) You then turn to examine Sharon further:**
-  **clinically euthyroid**
-  **neck palpation reveals a soft, 2cm mobile nodule in the right lobe of the thyroid gland with no fixation to adjacent tissues and no palpable lymph nodes.**

## Ix

-  **(trigger 2) You arrange for Sharon to have her TSH levels checked in the first instance. The test results come back the next day and demonstrate a TSH that falls within the normal range (1.8 mU/L [0.40-5.00mIU/L])**
-  On the basis of clinical findings so far, what would be your next line of investigation, and why?
    -  Bloods: FBC (Hb, WCC), ESR (increases in infection and malignancy), thyroid fn test
    -  USS (solid/cystic swellings, size, location)
    -  FNAC

## if thyrotoxic (still trigger 2)

-  What different approach might you have taken had Sharon been thyrotoxic?

-  <a href="https://i.imgur.com/U1vA1Se.png">
  <img src="https://imgur.com/U1vA1Sel.png" />
</a>

-  *In most cases, once thyrotoxicosis has been detected, patients should be referred to an endocrinologist for management* [BMJ 2006 Dx and Mx of thyrotoxicosis](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1476727/)

## Ix results

-  **(trigger 3) An ultrasound of Sharon’s thyroid gland demonstrates a single nodule within the lower pole of the right lobe of the thyroid, 2cm in diameter. There are no worrying or suspicious features on ultrasound examination, nor is there any evidence of abnormally enlarged cervical lymph nodes.**
-  **An aspirate is taken from the centre of the nodule and sent for cytological examination. The report comes back as ‘malignant features consistent with papillary thyroid cancer’.**
-  **You arrange to see Sharon to discuss this result and refer her to the thyroid cancer clinic at the hospital for further management.**

## Thyroid cancer epidemiology

-  How common is thyroid cancer? Why might the incidence be increasing?

-  <a href="https://i.imgur.com/s6NCprM.png">
  <img src="https://imgur.com/s6NCprMl.png" />
</a>

-  Increasing rapidly, almost universally amongst high income countries. Almost exclusively attributable to papillary thyroid cancer. **Why: ?** Increased detection of small tumours that are incidentally noted on diagnostic imaging studies. HOWEVER increase in prevalence of larger tumours, higher rates of aggressive papillary thyroid cancers and distant mets and increase in thyroid-spec mortality after around a decade of stability. [source](https://www-sciencedirect-com.ezproxy.library.uq.edu.au/science/article/pii/S0140673612622053)

## Thyroid cancer + Mx principles

-  What are the major pathological types of thyroid cancer? What kind of prognosis is expected with each? [2010 7th ed AJCC Cancer staging manual](https://www.cancer.org/cancer/thyroid-cancer/detection-diagnosis-staging/survival-rates.html)
    -  Papillary (80%): I = near 100%, II = near 100%, III = 93%, IV = 51%
    -  Follicular (15%): I = near 100%, II = near 100%, III = 71%, IV = 50%
    -  Medullary (3%): I = near 100%, II = 98%, III = 81%, IV = 28%
    -  Anaplastic/undifferentiated (2%): all IV = 7%
    -  Thyroid lymphoma (~1%; usually w/ hashimotos): IE = ~80%, IIE and IVE < 36% [wikipedia](https://en.wikipedia.org/wiki/Thyroid_lymphoma)
-  What are the principles of management of thyroid cancer?
    -  Multidisciplinary team, *patient focus*, surgery, pathology (staging + is pt high/low risk), I-131 ablation/therapy +/- radiotherapy, follow-up, [In-depth here by British thyroid association; royal college of physicians](http://www.baets.org.uk/wp-content/uploads/2013/02/management-_thyroid-cancer-guideline.pdf)

## Breaking bad news

-  How would you go about explaining to Sharon about her diagnosis having called her back to see you? 
-  `as.integer(runif(1,1,10))`

-  **SPIKES**
    -  Setting
    -  Perception
    -  Invitation
    -  Knowledge
    -  Explore and empathise
    -  Strategy and summary

## Mx {.smaller}

**(trigger 4) Sharon sees an endocrine surgeon who performs a total thyroidectomy, together with an ipsilateral central lymph node dissection when lymph node involvement is suspected intraoperatively. Post-operatively her serum calcium is monitored with no hypocalcaemia documented, and Sharon is later discharged home on replacement thyroxine therapy. Final pathology confirms a 2cm papillary thyroid cancer, with two central lymph nodes involved by metastatic cancer. Sharon is seen in the combined endocrinology/radiation oncology thyroid cancer clinic for radioiodine ablation and supervision of long-term follow-up. Her post-ablation whole body scan shows no evidence of metastatic disease and thyroglobulin levels are undetectable. Sharon remains on thyroxine and at 12 months has a further ultrasound scan and measurement of thyroglobulin levels. Both the scan and thyroglobulin level are negative, so the tumour is able to be classified as stage 1 (T1N1aM0).**

-  T1: Tumor ≤ 2 cm in greatest dimension limited to the thyroid 
-  N1: Metastasis to regional nodes. N1a: Metastasis to level VI or VII (pretracheal, paratracheal, prelaryngeal / Delphian or upper mediastinal) lymph nodes; this can be unilateral or bilateral disease
-  M0: No distant metastasis

-  Hypocalc**ae**mia: calcium blocks sodium channels and inhibits depolarization of nerve and muscle fibers => reduced calcium lowers the threshold for depolarization
    -  CATs go numb - convulsions, arrhythmias, tetany, and numbness in the hands and feet and around the mouth

## Trigger 4 questions

-  Discuss the relevance of each of the cues from this trigger.
-  Why might Sharon have become hypocalcaemic post surgery?
    -   Parathyroid gland trauma/devascularisation/removal
-  What are some of the other potential complications of thyroid surgery?
    -  postop bleeding, recurrent laryngeal nerve and superior laryngeal nerve damage, thyrotoxic storm (pre/intra/post-op), infection, hypothyroidism
-  Do you understand the purpose and principles of radioiodine ablation?
    -  [The radioactive iodine destroys any remaining cancer cells and any remaining healthy thyroid tissue.](https://www.medscape.com/viewarticle/763197). Destroy remaining thyroid tissue as the surgeon may have left some residual especially around key structures such as recurrent laryngeal nerve and parathyroid glands.
-  What is thyroglobulin and what is the purpose of measuring it here?
    -  Precursor of T3 and T4. In differentiated thyroid cancer (particularly papillary and follicular) [thyroglobulin] is usually increased to >100 ng/mL. Therefore post surgery and I-131 ablation it is a good prognositic indicator of disease status. A baseline initially can be taken, but it just adds info as to whether the tumour is producing thyroglobulin. 
-  What can you tell Sharon about her long-term prognosis?
    -  Good

### 2018-02-27 mfhd1-mc.rmd
added main case for week 4

---
title: mfhd1mc
output: ioslides_presentation
---

## Usual causes of abN by age

<a href="https://i.imgur.com/Tm5nXOb.png">
  <img src="https://imgur.com/Tm5nXObl.png" />
</a>

[source: uptodate](https://www-uptodate-com.ezproxy.library.uq.edu.au/contents/image?imageKey=OBGYN%2F61684&topicKey=OBGYN%2F3263&source=see_link)

## Menstrual cycle

<a href="https://i.imgur.com/d77w8GT.png">
  <img src="https://imgur.com/d77w8GTl.png" />
</a>

[source: osmosis](https://www.osmosis.org/learn/Menstrual_cycle)

## Menstrual cycle (hormones)

<a href="https://i.imgur.com/3NqfjYa.png">
  <img src="https://imgur.com/3NqfjYal.png" />
</a>

[source: wikipedia](https://upload.wikimedia.org/wikipedia/commons/f/f3/Figure_28_02_07.jpg)

## Vaginal (pelvic?) exam 4:38

<iframe width="756" height="426" src="https://www.youtube.com/embed/Z-O_JYtyQqE" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

## Cervical screening

[go to table](http://www.cancerscreening.gov.au/internet/screening/publishing.nsf/Content/healthcare-providers)

## position of uterus (1)

<a href="https://i.imgur.com/7TUGzmM.png">
  <img src="https://imgur.com/7TUGzmMl.png" />
</a>

[dunno](http://y7177.com/sa/anato/uterus/anteversion,anteflexion2.jpg)

## position of uterus (2)

<a href="https://i.imgur.com/OTJVfb3.png">
  <img src="https://imgur.com/OTJVfb3l.png" />
</a>

*Moore's Clinical Anatomy*

## endometrial thickness

[normal](https://radiopaedia.org/articles/endometrial-thickness)

[abnormal+USS](https://radiopaedia.org/articles/abnormally-thickened-endometrium-differential-1)

## endometrial biopsy

<iframe width="560" height="315" src="https://www.youtube.com/embed/5ymMtSJ_qKI?start=51" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

## the other one

<iframe width="560" height="315" src="https://www.youtube.com/embed/_Iv5IDmb5jQ?start=73" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

## Histology

<a href="https://i.imgur.com/7wGCZKL.png">
  <img src="https://imgur.com/7wGCZKLl.png" />
</a>

[not the source but watch this video](https://www.youtube.com/watch?v=cfLBJs2kKBA)

## classification of endometrial hyperplasia

<a href="https://i.imgur.com/XxXhh9n.png">
  <img src="https://imgur.com/XxXhh9nl.png" />
</a>

[from medscape](https://www.medscape.com/viewarticle/507187)

### 2018-03-27 endo2-ex-signs.rmd
sc2 and revision

---
title: "Inspection during the Endocrine exam"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
  -  \usetheme{metropolis}
  -  \usepackage{moresize}
output:
  beamer_presentation
---

# Review

## Thyroid

\scriptsize

```{r thyroidtable, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
thyroid_table <- "|           | Hyperthyroidism                                                                  | Hypothyroidism                                                                                                                                                  |
|----------|----------------------------------|----------------------------------|
| General   | Weight loss, anxiety, frightened facies of thyrotoxicosis                        | Mental and physical sluggishness, myxoedema madness, hypothroid speech (slow, nasal, deep)                                                         |
| Hands     | Tremor, onycholysis, thyroid acropathy (clubbing), palmar erythema, sweaty palms | Peripheral cyanosis, swelling, yellow discolouration (hypercarotenaemia), palmor crease pallor                                                                  |
| Arms      | Proximal myopathy, abnormal brisk reflexes                                       |                                                                                                                                                                 |
| Eyes/Face | Exopthalmos, thyroid stare, lid retraction, lid lag                              | Hypercarotenaemia skin not sclera, alopecia, vitililgo, loss of outer third of eyebrows, xanthelasmata, coolness, dryness, hair thinning, tongue swelling |
| Neck      | Thyroid enlargement                                                              | Thyroid enlargement                                                                                                                                             |
| Chest     | Gynaecomastia                                                                    | Auscultate for pericardial/pleural effusion                                                                                                                     |
| Legs      | Pretibial myxoedema, proximal myopathy, hyperreflexia                            | Non-pitting oedema, hung-up reflex (achilles tendon)                                                                                                            |
"
cat(thyroid_table)
```

## Pituitary

\footnotesize

```{r pittable, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}

pituitary_table <- "|                | Panhypothyroidism                                                          | Acromegaly                                                                                                   |
|----------------|----------------------------------------------------------------------------|--------------------------------------------------------------------------------------------------------------|
| General        | Short stature, pallor, fine wrinkled skin, lack of body hair               |                                                                                                              |
| Face           | Skin wrinkles (gonadotrophin deficiency), visual fields due to mass effect | Large supraorbital ridge, lips thickened, visual field defects, enlarged tongue, splayed and separated teeth |
| Chest          | skin pallor, decrease in nipple pigmentation                               | Coarse body hair and gynaecomastia, arrhythmias                                                              |
| Genital region | loss of pubic hair, testicular atrophy                                     |                                                                                                              |
| Ankle reflexes | Hung up jerks (pituitary hypothyroidism)                                   |                                                                                                              |
| Hands          |                                                                            | Wide shape, increased sweating, warmth, thickened skin, proximal myopathy                                    |"
cat(pituitary_table)
```

## Adrenals

-  Cushings: skin thickness of hands, moonlike facies, central obesity, bruising, excessive pigmentation, proximal myopathy, buffalo hump, striae, psychosis, visual fields, acne, hirsutism, oedema in legs, hypertension, increase in hepatic gluconeogenesis
-  Addinson's: cachexia, pigmentation in the palmar creases elbows gums and buccal mucosa, vitiligo, postural hypotension

### 2018-03-27 mfhd5-sc2.rmd
sc2 and revision

---
title: "MFHD5: SC2"
author: "Aaron Khoo"
date:  `r format(Sys.Date(), '%d %B %Y')`
output:
  ioslides_presentation:
    incremental: true
    smaller: true
---

## Trigger 1

In an urban general practice, your next patient is Mrs Reem Rashidi, aged 26 years.

After looking at her notes you remember initially seeing Mrs Rashidi with her husband last year for an upper respiratory tract infection. She had recently arrived in Australia from her home in Egypt. She had migrated here with her husband for his work. At the time she had spoken no English.

You invite Mrs Rashidi, who is unaccompanied this time, to sit down, at which point she apologises for not being able to speak English well. Her English is in fact very good, and while not fluent, you certainly have no difficulty in establishing that she has a urinary tract infection for which you prescribe antibiotics.

As you are running on time in your clinic (!), you ask how she is settling in to life in Australia. Given that you know from your discussion today that she is sexually active, you also decide to use this as an opportunity for health promotion. While printing off a leaflet in Arabic about the topic, you ask her if she has heard about cervical cancer and the test that is offered to screen for it in Australia.

## National cervical screening program

-  **What is the current cervical screening program in Australia? What is its purpose? Where is the data stored?**
-  Current:
-  Purpose
-  Data: "securely" stored in the Register, protected by latest "state-of-the-art-data security" measures, according to "strict" Australian government info security requirements. Stored on-shore in Australia. Commonwealth Department of Health will have control over all info in the Register.
-  **Who is eligible/should be targeted for screening**
-  **What are some reasons for women not taking up cervical screening?**
-  **What are the risk factors for development of cervical cancer?**
-  HPV infection (HPV protein E7 cuts Rb brakes; E6 silences and degrades *TP53*), cigarette smoking (increases genetic damage via carcinogens), lack of regular screening
-  **How is the test reported, and what would be the process for an abnormal result?**

-  **What symptoms may patients with cervical pathology present with?**
-  **What are the barriers to health promotion in culturally and linguistically diverse (CALD) patients? How might you go about addressing these?**

## Collecting a sample

-  **How in practice is the cervical sample actually taken? How would you explain this to a patient?**
1.  After the patient is prepared and comfortable, begin the speculum examination by examining the external genitalia for any abnormalities, then:

## Trigger 2

Mrs Rashidi takes the leaflet and you tell her that she can make an appointment to discuss it further with you and get the test done if she wishes to. You explain that although her English is very good, it may be wise to use an independent interpreter to discuss the matter further to ensure that she fully understands the test and what is involved. You reassure her that the interpreters are bound by confidentiality. A nurse will also be present for the test, and Mrs Rashidi requests that if this is the case she would like a female nurse.

Mrs Rashidi returns two weeks later and an interpreter has been arranged via telephone for the consultation. You ask her what she understands about the Cervical Screening Test and why it is done to which she responds clearly via the interpreter. You go through the practicalities of what the test involves and show her a speculum. When you ask if she would like to go ahead and have the test done today she replies,

‘Yes, I would, but…..I’m not sure whether you will be able to do it on me.’

You ask her to explain. ‘I was ‘cut’ when I was nine. It often hurts with sex and Isometimes have a little bleeding afterwards. I’m not sure that you will be able to get ‘this’ (pointing to the speculum) inside.

’You tell her that you can’t be sure until you have a look but that if it is difficult she can stop you at any time. You say that given her bleeding after sex, it is particularly important. (On further questioning she denies any other changes in menstrual pattern). She indicates that she would like to try and you introduce her to your practice nurse who guides her through undressing and positioning on the bed.

## Questions

-  **Why use an independent interpreter rather than her husband or other family member?**
-  Increased medical knowledge, better understanding of languages, covered under insurance, confidentiality
-  **Do you understand what is meant by the reference to being ‘cut’? What is female genital cutting / mutilation (FGM) (a term often best avoided), and what is its prevalence worldwide?** [more info](http://www.who.int/mediacentre/factsheets/fs241/en/)
-  Female genital mutilation (FGM) comprises all procedures that involve partial or total removal of the external female genitalia, or other injury to the female genital organs for non-medical reasons. It has no health benefits for girls and women. Procedures can cause severe bleeding and problems urinating, and later cysts, infections, as well as complications in childbirth and increased risk of newborn deaths.
More than 200 million girls and women alive today have been cut in 30 countries in Africa, the Middle East and Asia where FGM is concentrated. *It is a violation of  human rights*.
-  **Why might she have dyspareunia and post-coital bleeding?**
-  **What other complications are associated with the practice of FGM?**
-  Short term: pain, haemorrhage, shock, infn, HIV (possibly due to lack of sterlisation), urinary retention/dysuria, scarring, shock, psychological trauma
-  Long term: pain, infn (UTIs, chronica genital infns, chronic repro tract infns), dysuria, menstrual issues (due to obstruction of vaginal opening), keloids, HIV, female sexual Hx (removal of highly sensitive tissue, particularly clitoris, may decrease desire, pleasure, leading to dyspareunia, decreased lubrication, reduced frequency/absence of orgasm [anorgasmia], mental trauma), obstetric complications (such as caesarean, post-partum haemorrhage, tears/lacerations, prolonged labour), perinatal risks, psychological consequences
-  **What are the options if a speculum examination is too difficult in the GP practice?**
-   **What pathology of the cervix might be directly visualized on speculum examination?** [reading with some images](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3031439/)
-  Nabothian cysts: common & benign. May be translucent or opaque, whitish to yellow, mm to 4cm in diameter. Caused by transformation squamous metaplasia/inflamm blocking a gland orifice => endocervical columnar cells secrete mucus => covered by epithelium => cyst formation.
-  Leiomyomas
-  Cervical ectropion: eversion of the endocervix exposes columnar epithelium to vaginal milieu. Vaginal discharge most common symptom. Epithelium has reddish appearance sim to granulation tissue.
-  Cervical polyps: larger polyps should be evaluated and removed by a gynecologist; removal is typically straightforward office procedure
-  Endometriosis (present as red, blue, or black cervical lesions (“powder burns”)) and adenomyosis (may involve the endocervical canal or form a polypoid mass protruding into the endocervical canal)
-  Cervicitis: appearance depends on cause such as HSV (multiple small vesicular or ulcerative lesions), trichomonoiasis (punctate haemorrhages => strawberry cervix), chlamydial and gonococcal (mucopurulent discharge, cervical friability, cervical oedema)

## Trigger 3

Mrs Rashidi is nervous but your practice nurse helps to put her at ease. On examination of the external genitalia she has had a clitoridectomy and complete excision of her labia minora, leaving scarring.

You use a small speculum with a water-based lubricant and ask Mrs Rashidi to insert it a little way herself before you take over. Whilst she does wince sharply at one point and you stop immediately, she says she is ok and to carry on. You manage to open the speculum and view the cervix. The cervix is of a nulliparous appearance and you can see no ectropion or other evidence of abnormality. The walls of the vagina are also normal in appearance. You perform the test, obtaining cells from the transformation zone (there is no contact bleeding), before warning Mrs Rashidi that you are about to gently remove the speculum.

You explain to Mrs Rashidi once dressed that if any abnormalities are detected it does not necessarily mean that she has cancer but she will be contacted to have some further tests. You also explain if the test result is normal, she will receive a reminder from the National Cervical Screening Program in five years when another test is due.

## Questions

-  **Do you understand the terms ‘endocervix’, ‘transformation zone’, and ‘ectropion’?**
-  Transformation zone: between the endo and ectocervix which have two different epithelial linings. Endocervix. This is where HPV infects
-  **How does the transformation zone ‘move’ throughout reproductive life?**
-  Glandular epithelium moves down to ectocervix during female maturation
-  **What is colposcopy? Why and when is this necessary?**
-  **How would you explain the new cervical screening test to your patients who are used to having a Pap smear? What are the major changes you’d explain to them?**
-  Now it is every 5 years instead of 2 years (this is fine as the development is so slow), cheaper and safer, and pap cytology has been replaced by HPV molecular genotypinmg test
-  **Why do women who have had the HPV vaccine still need this screening? What is the theory behind the HPV cancer vaccine? When is it given in Australia?**
-  Gardasil targets 6/11/16/18, cervical cancer can also be caused by non-HPV 16/18 (25%) and non-HPV related (5%)

### 2018-03-29 mfhd5-sc2-ioslides.rmd
done sc2

--- 2018-03-27 mfhd5-sc2.rmd
@@
 author: "Aaron Khoo"
-date:  `r format(Sys.Date(), '%d %B %Y')`
+date:  "`r format(Sys.Date(), '%d %B %Y')`"
 output:
@@
 
+-  [Use this reading for full info](http://www.cancerscreening.gov.au/internet/screening/publishing.nsf/Content/C2058A7D155867ACCA2581C400082790/$File/CAN174-Understanding-the-National-Cervical-Screening-Program-Management-Pathway.pdf)
 -  **What is the current cervical screening program in Australia? What is its purpose? Where is the data stored?**
--  Current:
--  Purpose
+-  **How is the test reported, and what would be the process for an abnormal result?**
+-  Current: occurs every 5 years.
+HPV testing uses partial genotyping to classify into oncogenic HPV 16/18 OR oncogenic HPV not-16/18.
+If HPV positive, then reflex LBC (liquid based cytology: placing the cervical sample in liquid suspension, rather than directly applying the sample on a microscope slide like a conventional Pap test) performed.
+-  <img src="images/mfhd5-risk-cervical-ca.png" width="400" />
+
+## Questions (cont.)
+
+-  Purpose: to prevent cancer by detecting early and treating early
 -  Data: "securely" stored in the Register, protected by latest "state-of-the-art-data security" measures, according to "strict" Australian government info security requirements.
@@
 -  **Who is eligible/should be targeted for screening**
+  -  Anyone with a cervix (inc transgender men, intersex, females,...), 25 to 74 years, asymptomatic/symptomatic (e.g. abnormal vaginal bleeding, unusual vaginal discharge, dyspareunia)/pregnant
+  -  Self collecting: 30 years or older, cannot be/think they are pregnant, declined to have cervical sample taken by clinician, and either 1) overdue by 2 or more years, 2) have never screened
 -  **What are some reasons for women not taking up cervical screening?**
+-  Don't want to, low access to health services, don't know about it
 -  **What are the risk factors for development of cervical cancer?**
 -  HPV infection (HPV protein E7 cuts Rb brakes; E6 silences and degrades *TP53*), cigarette smoking (increases genetic damage via carcinogens), lack of regular screening
--  **How is the test reported, and what would be the process for an abnormal result?**
+
+## Questions (cont. 2)
 
 -  **What symptoms may patients with cervical pathology present with?**
+-  Vaginal bleeding, contact bleeding, loss of appetite, weight loss, fatigue, pelvic/back/leg pain, bone fractures
 -  **What are the barriers to health promotion in culturally and linguistically diverse (CALD) patients? How might you go about addressing these?**
+-  Language differences: refer to a clinician that speaks the language, use an interpreter, show the pamphlets in other language
+-  Cultural differences: cultural sensitivity courses
 
@@
 
--  **How in practice is the cervical sample actually taken? How would you explain this to a patient?**
-1.  After the patient is prepared and comfortable, begin the speculum examination by examining the external genitalia for any abnormalities, then:
+-  **How in practice is the cervical sample actually taken? How would you explain this to a patient?** [geekymedic is a great resource for OSCE stuff](https://geekymedics.com/cervical-smear-test/)
+1.  After the patient is prepared and comfortable, begin the speculum examination by examining the external genitalia for any abnormalities
+2.  Inspect cervix
+3.
+The objective of cervical screening is to sample cells from the transformation zone of the cervix, where HPV is present and cell abnormalities that precede the development of squamous cell carcinoma are usually found
+4.
+After, or before, the sample has been collected, record the patient name, DOB and ID number on the specimen vial and any other patient identifiers required by the laboratory, and fill out pathology request form
+5.  After sample collection, ensure the details of the consultation and procedure are accurately documented in the patient’s clinical record
 
@@
 
+## Trigger 2 (cont.)
+
 You ask her to explain.
@@
 -  **Why use an independent interpreter rather than her husband or other family member?**
--  Increased medical knowledge, better understanding of languages, covered under insurance, confidentiality
+-  Increased medical terminology knowledge, better understanding of language, covered under insurance and other legal stuff, confidentiality
 -  **Do you understand what is meant by the reference to being ‘cut’?
@@
 More than 200 million girls and women alive today have been cut in 30 countries in Africa, the Middle East and Asia where FGM is concentrated. *It is a violation of  human rights*.
+
+## Questions (cont.)
+
 -  **Why might she have dyspareunia and post-coital bleeding?**
@@
 -  **What are the options if a speculum examination is too difficult in the GP practice?**
+-  Refer
+
+## Questions (cont. 2)
+
 -   **What pathology of the cervix might be directly visualized on speculum examination?** [reading with some images](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3031439/)
@@
 
--  **Do you understand the terms ‘endocervix’, ‘transformation zone’, and ‘ectropion’?**
--  Transformation zone: between the endo and ectocervix which have two different epithelial linings. Endocervix. This is where HPV infects
+-  **Do you understand the terms ‘endocervix’, ‘transformation zone’, and ‘ectropion’?** [this site more info](https://www.gfmer.ch/ccdc/pdf/module1.pdf)
 -  **How does the transformation zone ‘move’ throughout reproductive life?**
--  Glandular epithelium moves down to ectocervix during female maturation
+-  Cervix = ectocervix (visible during speculum exam) + endocervix
+-  Endocervix: luminal cavity within the cervix forming a passageway between the external os and the internal os. Lined by simple columnar cells that secrete mucous
+-  Ectocervix lined by simple non-keratinizing stratified squamous epithelium
+-  The squamocolumnar junction (SCJ): junction between the squamous epithelium and the columnar epithelium.
+Located at the external os pre-puberty, repro age is variable, post menopausal the SCJ is not visible and has receded into the endocervix.
+-  Ectropion: eversion of squamocolumnar junction and large portions of columnar epithelium onto the ectocervix.
+This is not an abnormality and it can be caused by things such as hormonal changes, taking the OCP, pregnancy.
+However it is indistinguishable from early cervical cancer.
+When the everted columnar epithelium is exposed to acidic vaginal environment => squamous metaplasia => stratified squamous
+-  Transformation zone (TZ): cervical area between the old SCJ and new SCJ. >90% of precancerous lesions begin here
+
+## Questions (cont.)
+
 -  **What is colposcopy? Why and when is this necessary?**
+-  The cervix is inspected and acetic acid is applied to its surface – abnormal cells have increased surface proteins, which are coagulated by the acid and turn white, producing “acetowhite” areas which can be biopsied for histological assessment.
+Necessary: see above table
 -  **How would you explain the new cervical screening test to your patients who are used to having a Pap smear? What are the major changes you’d explain to them?**
--  Now it is every 5 years instead of 2 years (this is fine as the development is so slow), cheaper and safer, and pap cytology has been replaced by HPV molecular genotypinmg test
+-  Now it is every 5 years instead of 2 years (this is fine as the development is so slow), cheaper and safer, and pap cytology has been replaced by HPV molecular genotyping test
 -  **Why do women who have had the HPV vaccine still need this screening? What is the theory behind the HPV cancer vaccine? When is it given in Australia?**
--  Gardasil targets 6/11/16/18, cervical cancer can also be caused by non-HPV 16/18 (25%) and non-HPV related (5%)
+-  Gardasil targets 16/18, cervical cancer can also be caused by non-HPV 16/18 (25%) and non-HPV related (5%)

### 2018-05-02 haem2.rmd
add haem2

---
title: "Haematology II"
author: "Aaron Khoo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document
documentclass: article
classoption: a4paper
fontsize: 11pt
urlcolor: blue
---

# Main case

## Trigger 1

DDx:

-  Liver pathology can cause both thrombocytopaenia and loss of girth
-  Mets to bone marrow such as from lung
-  EBV
-  Lymphoma
-  Leukaemia
-  Multiple myeloma
-  Acquired haemophilia
-  Could be two separate pathologies

Ex:

-  Inspection: cachexia is hints of rapid weight loss (overall wt loss, loose skin, looks unhealthy)
-  Haem: lymph nodes, pallor, bony tenderness, scratch marks
-  Abdo: hepatosplenomegaly, jaundice/spider naevia/gynaecomastia/oedema for liver pathology
-  Resp: if worried about lung Ca

## Trigger 2

Top Dx

-  Lymphoma
-  Leukaemia
-  Metastatic Ca
-  Still consider HIV as generalised lymphadenopathy $\pm$ pain

Splenomegaly

-  Massive:
-  Large: malaria

## Trigger 3

Pancytopaenia

-  Lymphoma: pancytopaenia from bone marrow infiltration and crowding out so it cannot produce normal cells
-  Leukaemia: producing abnormal cells?
-  Aplastic anaemia: basically no more bone marrow left, cannot produce anything
-  Vitamin B12/folate deficiency
-  Medications

B symptoms: reserve for use when talking about lymphoma. Night sweats have to be drenching. If not quite B symptoms then it is called constitutional symptoms

## Trigger 5

-  Infiltration extramedullary haematopoeisis => hepatosplenomegaly => early satiety from pushing against it

# Short case 1

# Short case 2

### 2018-05-04 haem2.rmd
sc

@@
 
+Mech of fever: exogenous pyrogens => IL-1 and TNF => increased cyclooxygenase activity => $\Uparrow$ PG including PGE~2~ => reset temperature set point in hypothalamus
+
+## Trigger 1
+
+Fatigue in children + *infn*
+
+-  Acute: *URTI*, sinus *infn*, hepatitis, chickenpox, measles, meningitis
+-  More prolonged: *mononucleosis*, CMV, OSA
+-  Chronic: tb, diabetes, *IBD*, CKD, thyroid disease, heart failure, poor nutrition, *cancer*, abscesses
+
+Anaemia needs to be severe to cause persistent fatigue in children
+
+## Trigger 2
+
+"Growing pains": shouldn't affect the chest (sternum) but can affect children as young as 5
+
+If generalised bone pain => really should consider leukaemia
+
+## Trigger 3
+
+Pale during fever because of vasodilation and less blood going to skin
+
+Ix:
+
+-  Good idea to do uric acid, LDH, Ca, Phos as a baseline in case tumour lysis syndrome
+
 # Short case 2
+
+## Trigger 1
+
+Amoxiciliin + clavulanic acid: common for strep
+
+DDx:
+
+-  EBV
+-  HIV
+-  CMV
+-  Acute hepatitis: abdo symptoms
+-  Lymphoma: drenching night sweats
+-  Withdrawal symptoms from alc/drugs can cause night sweats
+-  Infective endocarditis
+
+Histoplasmosis can come from bat bites
+
+Pharyngitis is mostly viral
+
+## Trigger 2
+
+First line for tonsilitis is penicillin (amoxiclav)
+
+Ix:
+
+-  FBC
+-  ELFT + creatinine = liver + kidney
+-  CXR: tb in lungs
+-  USS: spleen size, but it won't be diagnostic of prognostic so a bit useless..
+
+## Trigger 3
+
+Bloods:
+
+-  Dehydration: osmolality high and urea high
+-  ALT and AST: ALT = hepatocellular, AST = alcohol (sometimes)
+-  Liver impairment

### 2018-05-10 haem3-sc1.rmd
sc1

---
title: Bad Formation (haem3sc1)
date: "10/05/2018"
author: Aaron Khoo
output:
  ioslides_presentation:
    incremental: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hx

Lorraine Cooper, is a new patient， 73F

<a href="https://i.imgur.com/QKrjYof.png">
  <img src="https://imgur.com/QKrjYofl.png" />
</a>

## Hx + Ex  (Trigger 1, 2)

You note on the chart that she is a 73 year old female with a local address. You greet Lorraine and ask her how you can help. She responds “To be perfectly honest with you doctor, I just want a second opinion. I have been **tired for some time now**, when I told my other doctor they did some tests which showed that my blood cells are no longer OK, especially my red cells– I think they call that anaemia. The doctor then said that my fatigue was due to my **anaemia** and this **should be investigated further with a number of tests**. Isn’t there something I could just take for this rather than more investigations? **I really don’t want any more poking and prodding**. You make a quick phone call to have Lorraine’s recent results sent to you and continue with your history.

You get some more background on Lorraine, and she tells you she is very well and in fact is **not on any medications**. Apart from some **osteoarthritis of her knees which she treats with hydrotherapy and physical exercise**, she has never had any serious illnesses. She has always been a **keen athlete** and attributes her good health to **regular physical activity**. She tells you she began **feeling fatigued over the course of the last year** and only brought it to her **usual doctor’s attention two weeks ago**. When she was told she was **anaemic**, Lorraine was partially surprised and didn’t realise anaemia could make her feel this tired. She describes **struggling to get out of bed in the morning and can feel a little stiff in her joints** but figured this was just aging. She has had to **give away bowls and tennis as she just doesn’t have the energy anymore**.

Her **examination is unremarkable**. Her **obs are stable** with a BP of 120/80mmHg and PR of 65. A haematological examination is unremarkable.

## Questions (Trigger 1)

-  **Given the story, what is the likely abnormality on Lorraine’s bloods?**
-  **What further aspects of history do you want to clarify?**
-  What is meant by fatigue? It can be used to describe
    -  Difficulty or inability to initiate activity (*subjective sense of weakness*)
    -  Reduced capacity to maintain activity (*easy fatigability*)
    -  Difficulty with concentration, memory, and emotional stability (*mental fatigue*)
-  Onset (ty uptodate)
    -  Acute fatigue: one month or less
    -  Subacute fatigue: between one and six months
    -  Chronic fatigue: over six months
        - Cause of chronic fatigue can be identified in approximately two-thirds of patients. Sometimes explained by overwork. Small number of cases are diagnosed with myalgic encephalomyelitis/chronic fatigue syndrome (ME/CFS) a disorder of unknown cause, but if the criteria is not met then it is known as idiopathic chronic fatigue. Both are diagnoses of exclusion.

## Fatigue

-  **When is fatigue pathological?**
-  Red flags: recent onset of fatigue in a previously well older patient, unintentional weight loss, abnormal bleeding, shortness of breath, unexplained lymphadenopathy, fever, recent onset or progression of cardio/gastro/neuro/rheum symptoms
-  **What are common causes for fatigue in this age group?**
-  **What are serious and not to be missed causes for fatigue in the elderly?**
-  The below is just general causes of fatigue (not spec to elderly)
-  *Probability diagnoses* include stress and anxiety, depression, viral/post-viral infection and sleep-related disorders
-  *Serious disorders* include malignancy, cardiac arrhythmias, cardiomyopathy, anaemia, bacterial infection, haemochromatosis, human immunodeficiency virus (HIV) infection and hepatitis C
-  *Common pitfalls* include coeliac disease, pregnancy, renal failure and metabolic disturbances. Medication-related side effects are another key consideration.

## Fatigue causes by system

-  **Cardiopulmonary**: congestive heart failure, chronic obstructive pulmonary disease, sleep apnoea.
-  **Endocrinologic/metabolic**: hypo/hyper-thyroidism, chronic renal disease, chronic hepatic disease, adrenal insufficiency, electrolyte abnormalities.
-  **Haematologic/neoplastic**: anaemia of diff causes, occult malignancy, myeloproliferative disorder, myelodysplastic syndromes, leukaemia/lymphoma, aplastic anaemia.
-  **Infectious**: Mononucleosis syndrome, viral hepatitis, HIV, subacute bacterial endocarditis, *tuberculosis*.
-  **Rheumatologic**: fibromyalgia, polymyalgia rheumatica, *systemic lupus erythematosus*, rheumatoid arthritis, Sjögren's syndrome, *sarcoidosis*.
-  **Psychological** (esp if prolonged fatigue): Depression, anxiety disorder, somatisation disorder.
-  **Neurological**: Multiple sclerosis.
-  **Medications**: Benzodiazepines, antidepressants, muscle relaxants, first-generation antihistamines, beta-blockers, opioids.
-  **Substance abuse**: EtOH, marijuana, opioids, cocaine/other stimulants.

## Bloods (Trigger 2)

<a href="https://i.imgur.com/WHAdzYZ.png">
  <img src="https://imgur.com/WHAdzYZl.png" />
</a>

## Simple blood cell lines

<a href="https://i.imgur.com/maYXUIT.png">
  <img src="https://imgur.com/maYXUITl.png" />
</a>

## Questions (Trigger 2)

-  Trigger 2 cont: You discuss with Lorraine that it would be worthwhile doing some further blood tests and considering referral to a haematologist. She is happy to oblige, anything to work out what's wrong.
-  **What pattern of abnormality is seen on the FBC?**
-  **What is the medical term used to describe this pattern of changes in her FBC and what are the causes?**
-  Pancytopenia (bicytopenia?): usually due to cancer, leukaemia, lymphoma, MM, TB, fungal, bone marrow aplasia (drug toxicities, nutrition, HIV, hepatitis, Parvovirus B19, autoimmune destruction)
-  **What other mechanisms can lead to a thrombocytopenia?**
-  *Decreased platelet production* (bone marrow failure/suppression, chronic EtOH, infn, MDS), *increased platelet consumption* (allo/auto-immune, DIC, sepsis, drug induced, infn, mechanical destruction by mech valves, HELLP, TTP/HUS), *sequestration/other* (chronic EtOH, hypersplenism, liver disease)
-  **What other possible causes of neutropenia are there?** [source](http://www.neutropenianet.org/wp-content/uploads/2013/07/publication-eval-treatment-of-scn.pdf)
-  Infn, drugs, autoimmune, nutritional, MDS, aplastic anaemia, cong bone marrow failure, neutropenia syndromes (GSD1b and other rare things)
-  **At what level of neutropenia does infection become a concern?**
-  0.2 - 0.5 increased risk of infn, < 0.2 risk of severe life threatening infn and opportunistic organisms

## Questions cont. (Trigger 2)

-  **What is the potential cause of high LDH level?**
-  Tissue damage: haemolytic anaemia, pernicious anaemia, infn, sepsis, AKI, acute liver disease, pacnreatitis, bone fractures, malignancy
-  **Is this anaemia proportionate to the degree of fatigue? How do the symptoms of anaemia evolve as the degree of anaemia increases?**
-  Age usually compounds fatigue, more anaemia = more fatigue, younger persons can tolerate more anaemia (no reputable source)
-  **What are your differential diagnoses?**
-  Pancytopenia: aplastic anaemia, mets to bone, MM, lymphoma, leukaemia, MDS, myelofibrosis, severe B12 folate deficiency, SLE, HIV/EBV, pernicious anaemia, medications, low dose arsenic poisoning, megaloblastic anaemia

## Further Ix (Trigger 3)

You repeat the blood tests which show a stable picture on the FBC, but specifically ask
for a reticulocyte count.

Additional results show that the reticulocyte count is 7 (10 - 100 x10^9^/L) indicating a
marrow cause for her anaemia.

Lorraine agrees to see a haematologist.

## Questions (Trigger 3)

-  **What is the significance of the abnormalities on the smear?**
-  **What are macrothrombocytes?**
-  Neutrophils: [normal](https://en.wikipedia.org/wiki/Neutrophil#/media/File:Neutrophils.jpg) vs [hypogranulated](https://en.wikipedia.org/wiki/Myelodysplastic_syndrome#/media/File:Hypogranular_neutrophil_with_a_pseudo-Pelger-Huet_nucleus_in_MDS.jpg)
-  [Oval macrocytes](https://i.imgur.com/IV5TEb7.png)
-  Macrothrombocytes: [smartsparrow](https://smartsparrow.global.ssl.fastly.net/repo/images/1d2066e379f598eeb9e9d4f6f101b90a.png)
-  **What are the causes of macrocytic anaemias?**
-  Macrocytic: increased reticulocytes (haemolytic anaemia, bone marrow recovery following chemo, recovery from bleed), megaloblastic (vit b12, folate, medications), multifactorial (EtOH, liver disease, HIV, MDS, aplastic anaemia, MM, preggers)
-  **How are the reticulocyte count and marrow involvement help in differentiating your hypotheses list?**
-  Aplastic anaemia, leukaemia, MDS, CHIP

## Trigger 4

Lorraine sees the haematologist who arranges for a bone marrow aspirate / trephine
which shows increased marrow cellularity with tri-lineage dysplasia. Iron stains show
normal iron stores but a small number of ring sideroblasts are present. The marrow has
been sent for cytogenetic testing.

The haematologist advises Lorraine that she has myelodysplastic syndrome and that
unfortunately this is a very serious incurable condition which will become progressively
worse. It is likely that the first intervention she will require will blood transfusions in the not too distant future.

## Questions (Trigger 4)

-  **What are the various forms of myelodysplasia?**
-  <a href="https://i.imgur.com/Nyrxp78.png">
  <img src="https://imgur.com/Nyrxp78l.png" />
</a>
-  [reading classification](https://academic.oup.com/view-large/35527240)
-  **What proportion of these are secondary to other causes? What are these?**
-  About 80% of people with MDS have primary MDS (no apparent risk factors). Secondary has poorer prognosis and risk factors are: previous chemo (alkylating agents/purine analogues), radiotherapy, tobacco smoking, benzene+derivatives, agricultural and industrial workers more likely
-  **At what point do blood transfusions become necessary? Are there complications associated with transfusion dependence?**
-  Chronic RBC transfusions are associated with chronic anaemia, leading to excess morbidity, and they cannot completely correct impaired QoL, may also be associated with iron overload => deleterious on organs (this is a bit disputed [MDS reading](https://academic.oup.com/annonc/article/25/suppl_3/iii57/1740877))
-  **At what level of platelets or neutrophils do complications usually arise?**

## Questions cont. (Trigger 4)

-  **What is the natural history of this disorder?**
-  Highly variable, survival ranges from a few weeks to several years. The median OS is 15-30 months and the risk of progression to AML is 25%-35% at 5 years. Bone marrow failure (infection and haemorrhage) is the leading cause of death, with more patients dying before overt AML has occurred
-  [Revised International Prognostic Scoring System (IPSS-R) for myelodysplastic syndromes](https://academic.oup.com/view-large/35527248) stratifies patients into five risk groups (v low, low, int, high, v high)
-  If she was at [high risk](https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/annonc/25/suppl_3/10.1093_annonc_mdu180/3/mdu18002.png?Expires=1526047443&Signature=jd--FREEBGN0cNyVIGcYxEAWH01IU1W0N1Vx3Mur4v5rL~8BODfiZBoAlMgJAxu5cDYuOWgpT6ZeFkHU2jSy14oTgF7WDYZQPUpjyVUSSzDjtQXJO80-avbLy6YFDZZV2-D1aOusxmmaGXPXlgppRFqKdzPEHIKbXn60m4JJcwYxX-6Sph~B-3ArEekiCIIySUyD5hhRKR3LruN9FoLR9ir3s9WKb-6255hWMOzKaqXy1RrB-fMBB~dn750VLyn~~EuQQd31Ohu8UukEFNsGC5R3payOuEXIIAjr3btMhCw2yABTWmpS8zHqCLe9QAD3C4DmbGLl~kpQDCwUD436YA__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA)

## References

-  [RACGP fatigue](https://www.racgp.org.au/afp/2014/july/fatigue/)
-  [uptodate fatigue in adults](https://www-uptodate-com.ezproxy.library.uq.edu.au/contents/approach-to-the-adult-patient-with-fatigue)
-  [Thrombocytopenia aafp](https://www.aafp.org/afp/2012/0315/p612.html)
-  [Neutropenia by randos](http://www.neutropenianet.org/wp-content/uploads/2013/07/publication-eval-treatment-of-scn.pdf)
-  [Macrocytic anaemia uptodate](https://www-uptodate-com.ezproxy.library.uq.edu.au/contents/macrocytosis-macrocytic-anemia?topicRef=7133&source=see_link)
-  [MDS](https://academic.oup.com/annonc/article/25/suppl_3/iii57/1740877)
-  [Hematologylearning for some images](https://hematologylearning.weebly.com/white-blood-cell-inclusions-and-abnormalities.html)
-  [simpler MDS reading cancer.net](https://www.cancer.net/cancer-types/myelodysplastic-syndromes-mds/subtypes-and-classification)

### 2018-05-15 mh1-mc.rmd
added psych

---
title: Home, Sweet Home! (mh1mc)
date: "16/05/2018"
author: Aaron Khoo
output:
  ioslides_presentation:
    incremental: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vocabulary of mood (source lec 2)

affect mood anhedonia alexithymia euthymic

-  Affect: superficial and immediate emotional state of pt, observable (facial expression and social gestures)
-  Mood: underlying and usually more enduring emotional state of pt, have to ask
-  Anhedonia: inability to experience pleasure
-  Alexithymia: no words for feelings (?inability to identify and describe emotions in the self??)
-  Euthymic: normal mood

## Trigger 1: Presentation

Marlene Truscott is a 44-year-old patient of your suburban group practice, whom you have seen only occasionally. She says "I need some more sleepers, Doctor - you know, the Temazepams." You notice that Marlene looks clearly agitated. For a start, she is quite adamant that she just needs more "sleepers", but as you gently persist in finding out more about her circumstances at home, she relaxes a little and opens up in conversation.

## Questions (trig 1)

-  **What are the important features of Marlene’s presentation? Why?**
    -  Agitation is an acute state of anxiety, heightened emotional arousal, and increased motor activity
    -  Benzodiazepines: augment GABA's inhibitory effect by increasing the frequency of channel openings. This inhibitory effect leads to the anxiolytic, hypnotic, anticonvulsive, and muscle-relaxing properties of benzodiazepines.
-  **List possible causes of these features and explain the underlying mechanisms.**
    -  Primary psychiatric: MDE/MDD, chronic depression, bipolar disorder, psychotic disorders, dementia, substance use (benzodiazepines) disorders, personality disorders, anxiety disorders, schizophrenia, ADHD
    -  Primary medical: insomnia, medications, fatigue/weight loss/insomnia (cancer, and endocrine, cardiovascular or neurological diseases), hypothyroidism. "Presence of medical conditions that may be causally related to the depression itself through pathophysiological mechanisms and in such cases the depression may remit completely with the successful treatment of the underlying medical condition." [ref](https://www-clinicalkey-com-au.ezproxy.library.uq.edu.au/#!/content/book/3-s2.0-B9780323295079000299?scrollTo=%23hl0000204)

## Questions (cont)

-  **Marlene’s agitation raises questions:**
    -  **Is her agitation related to her request for medications? Is she depressed? Is she anxious?**
    -  **Might she be using the benzodiazepines for daytime sedation? Is she dependent on benzodiazepines?**
    -  **Is she contemplating suicide, and collecting the means? Has she been "shopping around", visiting different doctors?**
    -  **Is she actually seeking someone sympathetic to talk to about her life's stresses, and the request for medication is merely an opening gambit?**
-  **How might the possibility of Marlene having suicidal thoughts be approached?**
    -  Ask about the specific nature of the ideation, intent, plans, available means (eg, firearms), and actions
    -  Assess the patient for risk factors for suicide, including prior history of suicide attempts, comorbid psychiatric and general medical illnesses, and family history of suicidal behavior
    -  Develop a safety plan for further evaluation and treatment that depends upon the level of risk and may range from continued primary care follow-up alone to outpatient psychiatric or emergency room psychiatric evaluation

## Questions

-  **What else would you need to find out to consider a diagnosis of depression? (write up 9 things)**
    -  <a href="https://i.imgur.com/SD0uQQ1.png">
  <img src="https://imgur.com/SD0uQQ1l.png" />
</a>
-  **What information do you need to refine your initial hypothesis/differential diagnosis?**

## Trigger 2: History

After 20 minutes of conversation (during which time you actively and empathetically listen, and gently ask her questions that facilitate the flow of her disclosures), Marlene’s story emerges.

She is married, and has two children. Her 16-year-old son, Tom, is in grade 11 and progressing well, but her 14-year-old daughter, Tracey, is "uncontrollable" and is often out all night, coming home intoxicated. Marlene believes she is sexually active but their relationship is so poor they can never have any meaningful discussion with her daughter about the subject. Marlene states that she and her daughter have never been "close".

Marlene's husband of 20 years, Euan, an accountant aged 39, drinks heavily and spends little time with the family. He also often comes home intoxicated and is verbally abusive; he regards the children’s problems as her fault. She and Euan rarely have sex, and that is too often, Marlene says. She uses prescription sedatives for sleep, and often a stiff Scotch as well. She denies drinking during the daytime. Her insomnia is of the early-waking type. She feels churned up most of the time and finds it hard to make decisions, even about minor domestic issues. She says that she can't be bothered taking care of her appearance, and that she avoids going out. Nothing gives her pleasure. She does not know why her life has turned out this way, or what she has done wrong to deserve it, but is sure that somehow, it is her fault. She doesn't think that things will ever get any better, and she wonders if the family would be better off without her.

Her past medical history is unremarkable. Marlene is not inclined to share much about her childhood, except to say that her parents' divorce when she was in her early teens was "a terrible time".

## Questions (trig 2)

-  **How does this information help you refine your differential diagnosis?**
-  **Which ones are now the most likely?**
-  **What information from examining Marlene would help you?**
-  **What are the symptoms and signs of depression?**
    -  Above + aches or pains, headaches, cramps, or digestive problems without a clear physical cause and/or that do not ease even with treatment
-  **Is depression a serious (fatal?) illness?**
    -  mean(Depressed patients) = 77.6% of normal function, with advanced CAD/angina = 65.8%/71.6%， back problems/arthritis/diabetes/HTN = (79%， 88.1%), increased mortality, 21% of recurrent depressive disorder patients attempt suicide, 11th largest burden of disease worldwide by DALYs, comorbidities, influences the course of concomitant medical diseases (increased cardiac mortality, increased diabetic-complications)
-  **What other family factors may a GP need to consider?**

## Risk factors for mood disorders

There are 9:

-  <a href="https://i.imgur.com/QwEP94a.png">
  <img src="https://imgur.com/QwEP94al.png" />
</a>

## Trigger 3: Progress

One week later (during which time Tom and Tracey heard their parents fighting every night), Tom phones the doctor's surgery, very concerned that he has come home from school and found his mother "asleep" in her bed and very difficult to rouse. When shaken vigorously, she murmurs in a very slurred voice, Sorry. "So stupid!" An empty bottle of Marlene's Temazepam was beside her, and a nearly empty bottle of Scotch. The practice nurse calls an ambulance for Tom and instructs him to stay with his mother and ensure she lies on her side until the ambulance arrives.

## Questions (trig 3)

-  **What principles would underlie the acute management of Marlene? Note that acute poisoning will be covered in more detail within clinical science 4.**
    -  Resus: ABCD(isability via GCS, blood glucose), Hx (may be unreliable but try to establish what was taken, vomiting, PMHx/meds/allergies, suicide note?, histories from others), general Ex (vitals done above too, temp, rigidity [ecstasy, amphetamines], skin, breath for ketones), Ix
    -  Obtain more info, decontaminate if appropriate, decrease absorption, increase elimination, supportive, specific antidote
-  **How might Tom be feeling, finding his mother in this state? Frightened? Shaky? Confused?**
-  **How might this event impact his relationship with his mother over time?**
-  **The distress in this family is multifaceted. What thoughts for future intervention will you consider?**

## Trigger 4: Progress

Marlene is taken to hospital and managed with a Guedel airway, administration of intravenous fluids and oxygen via a face mask. Marlene wakes up enough to spit out the airway, and cough. She is very drunk and minimally responds to commands. The emergency department registrar orders a routine blood test for levels of paracetamol, alcohol and benzodiazepines. Paracetamol levels are normal. Marlene is transferred from the emergency department to the ward. Marlene does not experience any medical complications. Once she is medically stabilised and able to engage in a psychiatric assessment, Marlene is referred to the psychiatry registrar who comes to assess Marlene.

## Questions (trig 4)

-  **What might the registrar ask Marlene?**
-  **What are the risk factors for suicide? (8...)**
    -   Male gender, recent losses, drug/alc abuse, MDD, lack of social support, serious physical illness, past history of suicide attempts, family history of completed suicide
-  **What are the preventable factors for suicide in adolescents and adults?**
-  **How common is suicide?**
    -  M>F, [rates by year](https://i.imgur.com/xJrpFRh.png), [rates by age M](https://i.imgur.com/Jj9yfKL.png), most common cause of death in Aus aged 15-44, over 3000 die each year (approx 8 a day), more than 65,000 make an attempt
-  **What public health strategies have been introduced to prevent suicide?**
    -  the thing on cliff edges from lifeline, beyondblue (if theyre funded by gov)
-  **What barriers may prevent patients from discussing suicide and accepting help?**
-  **Could Marlene’s suicide attempt have been prevented?**

## Trigger 5: Progress

Marlene discloses to the psychiatry registrar that she regrets her actions, which were impulsive. She is distressed to think about how her family would be affected if she did suicide. She admits to feeling depressed and is given information about Sertraline, a selective serotonin reuptake inhibitor as treatment for her depression. Marlene is reluctant to take medication because she feels, "I got myself into this, I’ll have to sort it out".

## Questions (trig 5)

-  **How do selective serotonin reuptake inhibitors work?**
-  **Why are they used in the treatment of depression?**
    -  Current first line drugs for MDD (sertraline: zoloft, xydep,.. and fluoxetine: prozac, lovan,...)
    -  Inhibit reuptake into presynaptic cell => more serotonin in synaptic cleft => more available to bind to postsynaptic-R
-  **What other conditions are they used to treat? What are the side effects? How long do they take to work?**
    -  bullimia, depression, generalised anxiety disorder, obsessive-compulsive disorder, panic disorder, PTSD, stroke recovery(?)
    -  Serotonin syndrome (may consist of increased heart rate, shivering, sweating, dilated pupils, myoclonus (intermittent jerking or twitching), as well as overresponsive reflexes), nervousness, anxiety, insomnia, nausea, disarrhoea, dizziness, anorexia, weight loss, fatigue, headache, sedation, rash, pruritis, sexual dysfn (in 30-60%), increased suicide risk <= 18. Genetics plays a factor in side effects (L~A~ 5-HTTPLR allele is associated with decreased adverse effects)
    -  2-3 weeks
-  **What are the other classes of antidepressants?**
    -  MAO inhibitor

## Overview of drugs

<a href="https://i.imgur.com/6c1Muzv.png">
  <img src="https://imgur.com/6c1Muzvl.png" />
</a>

## Questions (cont.)

-  **What other non-pharmacological treatments are available for depression?**
    -  Therapeutic alliance, psychotherapy (CBT, IPT, PST,...), psychoeducation, scheduling of pleasant activities, medications, behavioural activation, combined medications + psychotherapy, ECT, exercise, bright light

## Trigger 6: Closure

Euan is away for a few days and is contacted by Tom with the news of Marlene's suicide attempt. Tracey is horrified by Marlene's overdose and blames herself and her father. She is counselled by the psychiatry registrar, who arranges to see her again later in the day.

During the second conversation, Tracey tearfully discloses to the registrar how much she and Tom hate hearing their parents fight, and how she often goes out to get away from all the strife. The registrar is supportive and validates her need to get away from the tension of her parents' conflict.

Marlene is discharged three days later. She accepts the recommendation from the psychiatry registrar to have sessions with a clinical psychologist at a local private practice on referral from her GP (under a Better Access to Mental Health Scheme). Marlene's psychologist explores her depressed mood and troubled relationships with her daughter and husband. Marlene is offered the option of bringing Tracey to some sessions to work on restoring a more supportive relationship with her daughter. Marlene is also offered a referral to a couples' therapist should she and Euan indicate willingness to work on their distressed relationship.

Tracey visits their family GP and finds relief in talking about her "screwed up" family. She admits to wanting to be more aware of her mother's problems now. Marlene considers leaving Euan but is too financially dependent and cannot see how she could look after the children. Some months after Marlene's hospital admission, Euan (after a number of "secret" visits to his GP) tells Marlene that he is going to get some help for his drinking. She is surprised and thinks to herself, "I'll believe that when I see it."

## Questions (trig 6)

-  **What help is available for patients and their families following a suicide attempt?**
    -  Patients: around 50% of those discharged from hospital after a suicide attempt do not attend any subsequent treatment
    -  Families: [SuicideLine Victoria](https://www.suicideline.org.au/resource/supporting-someone-after-suicide-attempt/) (probably have to live there),
-  **What is the role of psychotherapy in treating patients with mental illness?**
-  **What are the different types of psychotherapy?**
    -  CBT = cognitive behavioural psychotherapy
    -  One of the most effective treatments for conditions where anxiety or depression is the main problem
    -  Most effective psychological treatment for moderate and severe depression
    -  As effective as antidepressants for many types of depression (lecture 4)
-  **What is the evidence for marital therapy as an intervention for depression?**
    -  There probably is some

## Mindfullness meditation

<iframe width="560" height="315" src="https://www.youtube.com/embed/6p_yaNFSYao" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

## References

-  [uptodate psychosis](https://www-uptodate-com.ezproxy.library.uq.edu.au/contents/clinical-manifestations-differential-diagnosis-and-initial-management-of-psychosis-in-adults#H261969392)
-  [racgp illlicit drug overdose](https://www.racgp.org.au/afp/2013/july/illicit-drug-overdose/)
-  Screenshots from lectures
-  [Suicide prevention](https://www.psychology.org.au/inpsych/2016/feb/christensen/)
